---
title: "Lyon & Malik (2024), *Ethnicity and Policing in the Global South* --- SH Day 4: Classic Learners & Interpretation"
author: "Workshop teaching example"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: false
    number_sections: true
header-includes:
  - \setlength{\emergencystretch}{3em}
  - \usepackage{fvextra}
  - \AtBeginDocument{\DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}}
  - \AtBeginDocument{\DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}}
  - \fvset{breaklines=true}
---

```{r setup, include=FALSE}
options(width = 80)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 6.5, fig.height = 4)
suppressWarnings(suppressMessages({
  library(haven); library(dplyr); library(sandwich); library(lmtest)
  library(rpart); library(rpart.plot)
  library(glmnet)
  library(ranger); library(xgboost); library(e1071)
  library(class); library(kknn)
  library(ggplot2); library(knitr)
}))
set.seed(2111)
```

# The study

**Citation.** Lyon, Nicholas, and Mashail Malik. 2024. "Ethnicity and Policing in the
Global South: Descriptive Representation and Expectations of Police Bias."
*Comparative Political Studies* 57(9): 1509--1543.

**Source data (Harvard CPS Dataverse).** DOI `doi:10.7910/DVN/MVCQXU` ---
<https://doi.org/10.7910/DVN/MVCQXU>. Files used here: `LyonMalik_Raw_Data.sav`
(raw survey) plus the authors' `Cleaning_Script.R` and `Analysis_Script.R`, which
define the exact model.

**What we reproduce.** The paper's observational Table 1, *Demographic Determinants
of Trust in Police*. Using an original face-to-face survey in Karachi, Pakistan, the
authors regress citizens' **trust in the police** (a 1--5 scale) on a set of
respondent characteristics. Their main specification (Model 2) is a plain OLS with
cluster-robust standard errors (clustered on the primary sampling unit, `psu`),
estimated on the experimental sample of Muhajir, Pashtun, and Sindhi respondents:

$$
\texttt{police\_trust} \sim \texttt{under30} + \texttt{non\_sindhi} +
\texttt{education} + \texttt{years\_lived} + \texttt{total\_crimes} + \texttt{male}.
$$

We (1) reproduce this regression exactly, then (2) turn it into a **Day-4 tour of the
classic supervised learners** --- a decision tree, a random forest, gradient boosting,
support-vector machines, and k-nearest-neighbours --- holding the *same outcome and the
same predictors* fixed, and compare them all on a held-out test set. We close with
model-agnostic **interpretation** (permutation importance and partial dependence).

```{r clean}
# ---- Load raw survey and reproduce the authors' cleaning pipeline verbatim ----
# (Column recodes copied from the authors' Cleaning_Script.R.)
raw <- {.f <- tempfile(fileext=".sav"); download.file(paste0(
  "https://raw.githubusercontent.com/bdesmarais/intro",
  "-ml-statistical-horizons-4day/main/tutorials/day4_",
  "classic_learners/lyon_malik_policing/data/LyonMali",
  "k_Raw_Data.sav"), .f, mode="wb", quiet=TRUE); read_sav(.f)}
raw <- suppressWarnings(as.data.frame(lapply(raw, function(x) as.numeric(x))))

d <- raw %>% mutate(
  resp_id      = M1,
  ethnicity    = case_when(
                   d5 %in% 1:4 ~ "Muhajir", d5 == 5 ~ "Punjabi",
                   d5 == 6 ~ "Sindhi", d5 == 7 ~ "Balochi",
                   d5 == 9 ~ "Pashtun", d5 %in% c(8, 10, 11) ~ "Others"),
  non_sindhi   = ifelse(ethnicity == "Sindhi", 0, 1),
  police_trust = ifelse(q15_4 == 98 | q15_4 == 99, NA, q15_4),
  psu          = as.factor(M3),
  age          = ifelse(d1 == 98 | d1 == 99, NA, d1),
  under30      = ifelse(age <= 30, 1, 0),
  years_lived  = d2_y,
  education    = ifelse(d4 == 98 | d4 == 99, NA, d4),
  male         = ifelse(d9 == 1, 1, 0),
  total_crimes = Count_Q4)

# Authors' Table 1 restricts to the three-group experimental sample.
exp_sample <- subset(d, ethnicity %in% c("Muhajir", "Pashtun", "Sindhi"))

model_vars <- c("police_trust", "under30", "non_sindhi",
                "education", "years_lived", "total_crimes", "male")
ml <- exp_sample[complete.cases(exp_sample[, c(model_vars, "psu")]), ]
ml_dat <- ml[, model_vars]
N <- nrow(ml_dat)
```

# Descriptive analysis

Before fitting anything, we get to know the data --- what the outcome looks like, how the
predictors are distributed, where values are missing, how the predictors relate to each
other and to trust. Good descriptive work is not throat-clearing: it tells us in advance
which model assumptions are safe (here, near-zero collinearity) and where a linear model
is likely to leave signal on the table (here, a non-monotone tenure effect), and it lets
us sanity-check the reproduction against the raw data before we trust any coefficient.

**Unit of analysis and coverage.** The unit is an **individual survey respondent** in
Karachi, Pakistan, interviewed face-to-face in a single survey wave. After the authors'
cleaning and the restriction to the three-group experimental frame (Muhajir, Pashtun,
Sindhi) with complete cases on the model variables, we have **`r N` respondents**
described by **`r ncol(ml_dat)` variables** (the outcome plus six predictors).

**Sampling design.** Respondents are nested in **primary sampling units** (`psu`) ---
the survey's geographic clusters --- and this nesting is not incidental: it is exactly
why the authors report *cluster-robust* standard errors. In the analysis sample the
`r nlevels(droplevels(ml$psu))` PSUs hold between `r min(table(droplevels(ml$psu)))` and
`r max(table(droplevels(ml$psu)))` respondents each (median
`r median(table(droplevels(ml$psu)))`), so respondents in the same neighbourhood cluster
share unobserved context and cannot be treated as fully independent. The table below
shows the ethnic composition of the frame; the Muhajir plurality is what makes the
`non_sindhi` indicator so lopsided.

```{r eda-ethnicity}
eth_tab <- as.data.frame(table(Ethnicity = ml$ethnicity))
names(eth_tab)[2] <- "n"
eth_tab$`% of sample` <- sprintf("%.1f%%", 100 * eth_tab$n / sum(eth_tab$n))
eth_tab$`Mean trust` <- round(tapply(ml_dat$police_trust, ml$ethnicity, mean)[eth_tab$Ethnicity], 2)
kable(eth_tab, row.names = FALSE,
      caption = sprintf("Composition of the three-group experimental frame and mean trust by group (analysis sample, N = %d).", N))
```

## The outcome: trust in the police

The outcome `police_trust` is a **1--5 ordinal trust rating** that we (like the authors)
treat as continuous for OLS. Its distribution is strongly bimodal and skewed toward
distrust: a large spike of respondents at the **floor** (rating 1) and a second cluster
in the middle-high range (3--4), with very few at the top.

```{r eda-outcome, fig.width=6, fig.height=3.2, fig.cap="Distribution of police\\_trust (1 is least trust, 5 is most). The dashed line marks the mean."}
out_stats <- c(mean = mean(ml_dat$police_trust), sd = sd(ml_dat$police_trust),
               median = median(ml_dat$police_trust),
               min = min(ml_dat$police_trust), max = max(ml_dat$police_trust))
ggplot(ml_dat, aes(factor(police_trust))) +
  geom_bar(fill = "grey35") +
  geom_vline(xintercept = out_stats["mean"], linetype = 2, colour = "firebrick") +
  labs(x = "police_trust (1-5)", y = "Respondents",
       title = "Distribution of trust in the police") +
  theme_bw(base_size = 11)

kable(t(round(out_stats, 3)),
      caption = sprintf("Summary of the outcome police_trust (N = %d).", N))
```

The bar chart hides the exact shape, so we also tabulate the five response levels
directly, with counts and shares:

```{r eda-outcome-tab}
otab <- as.data.frame(table(`Trust rating` = ml_dat$police_trust))
names(otab)[2] <- "n"
otab$`Share` <- sprintf("%.1f%%", 100 * otab$n / N)
otab$`Cumulative` <- sprintf("%.1f%%", 100 * cumsum(otab$n) / N)
kable(otab, row.names = FALSE,
      caption = sprintf("Frequency of each trust rating (N = %d). The floor category dominates.", N))
```

The single most common answer is the **lowest** trust rating (1), held by
`r round(100*mean(ml_dat$police_trust==1))`% of respondents; a majority
(`r sprintf("%.0f%%", 100*mean(ml_dat$police_trust<=2))`) sit at 1 or 2, and only
`r sprintf("%.1f%%", 100*mean(ml_dat$police_trust==5))`% give the top rating. The mean is
just `r round(out_stats["mean"], 2)` on the 1--5 scale, below the midpoint of 3. This
**floor-heavy, bimodal** shape --- rather than a tidy bell curve --- is the backdrop for
every model that follows: OLS will fit a conditional mean through a distribution whose
mass is piled at the boundaries, and the flexible learners will have room to carve out
the low-trust and mid-trust subpopulations separately.

## The predictors

Three predictors are **binary** (`under30`, `non_sindhi`, `male`) and three are
**numeric** (`education`, an 8-point attainment scale; `years_lived`, years at the
current residence; `total_crimes`, a 0--6 count of crimes the respondent reports).

```{r eda-numeric}
num_vars <- c("education", "years_lived", "total_crimes")
num_tab <- data.frame(
  Variable = num_vars,
  n        = sapply(num_vars, function(v) sum(!is.na(ml_dat[[v]]))),
  Mean     = sapply(num_vars, function(v) round(mean(ml_dat[[v]]), 2)),
  SD       = sapply(num_vars, function(v) round(sd(ml_dat[[v]]), 2)),
  Min      = sapply(num_vars, function(v) min(ml_dat[[v]])),
  Median   = sapply(num_vars, function(v) median(ml_dat[[v]])),
  Max      = sapply(num_vars, function(v) max(ml_dat[[v]])),
  Missing  = sapply(num_vars, function(v) sum(is.na(ml_dat[[v]]))),
  row.names = NULL)
kable(num_tab, caption = "Summary statistics for the numeric predictors (analysis sample).")
```

```{r eda-binary}
bin_vars <- c("under30", "non_sindhi", "male")
bin_tab <- data.frame(
  Variable   = bin_vars,
  `# = 1`    = sapply(bin_vars, function(v) sum(ml_dat[[v]] == 1)),
  `# = 0`    = sapply(bin_vars, function(v) sum(ml_dat[[v]] == 0)),
  `Prop = 1` = sapply(bin_vars, function(v) round(mean(ml_dat[[v]]), 3)),
  check.names = FALSE, row.names = NULL)
kable(bin_tab, caption = "Frequencies for the binary predictors (analysis sample).")
```

The sample skews **male** (`r round(100*mean(ml_dat$male))`%) and overwhelmingly
**non-Sindhi** (`r round(100*mean(ml_dat$non_sindhi))`%, since Muhajir respondents
dominate the three-group frame); ages are split roughly evenly around 30. `years_lived`
is right-skewed (median `r median(ml_dat$years_lived)`, max `r max(ml_dat$years_lived)`)
and `total_crimes` is a low count concentrated at 0.

The summary table compresses each numeric predictor to seven numbers; the shapes matter
too, so we plot the three distributions. Note the sharp differences: `education` is an
8-point scale with a mode at the middle, `total_crimes` is a heavily zero-inflated count,
and `years_lived` is continuous but strongly right-skewed with a long tail.

```{r eda-numeric-dist, fig.width=7.2, fig.height=2.6, fig.cap="Distributions of the three numeric predictors. education is a discrete 8-point scale, total\\_crimes is a zero-inflated count, and years\\_lived is right-skewed with a long upper tail."}
d_edu <- ggplot(ml_dat, aes(factor(education))) +
  geom_bar(fill = "grey35") +
  labs(x = "education (1-8)", y = "Respondents", title = "Education attainment") +
  theme_bw(base_size = 9)
d_crime <- ggplot(ml_dat, aes(factor(total_crimes))) +
  geom_bar(fill = "grey35") +
  labs(x = "total_crimes (count)", y = NULL, title = "Crimes reported") +
  theme_bw(base_size = 9)
d_yl <- ggplot(ml_dat, aes(years_lived)) +
  geom_histogram(binwidth = 5, fill = "grey35", colour = "white") +
  labs(x = "years_lived", y = NULL, title = "Residential tenure") +
  theme_bw(base_size = 9)
if (requireNamespace("gridExtra", quietly = TRUE)) {
  gridExtra::grid.arrange(d_edu, d_crime, d_yl, ncol = 3)
} else { print(d_edu); print(d_crime); print(d_yl) }
```

The zero-inflation in `total_crimes` (`r sprintf("%.0f%%", 100*mean(ml_dat$total_crimes==0))`%
of respondents report **no** crimes) and the discreteness of `education` mean these
predictors carry most of their information in a few coarse jumps --- the kind of steppy,
threshold-shaped signal a regression tree represents natively and a linear slope only
approximates.

## Missingness

The published specification is estimated on complete cases. The table below counts
missing values on each model variable **in the three-group experimental sample before**
listwise deletion; dropping incomplete rows takes us from
`r nrow(exp_sample)` respondents to the `r N` used for modeling.

```{r eda-missing}
miss <- colSums(is.na(exp_sample[, model_vars]))
miss_tab <- data.frame(Variable = names(miss), `Missing (pre-deletion)` = as.integer(miss),
                       check.names = FALSE, row.names = NULL)
kable(miss_tab[order(-miss_tab$`Missing (pre-deletion)`), ], row.names = FALSE,
      caption = sprintf("Missing values per model variable in the experimental sample (N = %d before deletion).", nrow(exp_sample)))
```

Missingness is light and confined to the outcome and a few demographics
(`police_trust`, `under30`, `education`), so listwise deletion drops only
`r nrow(exp_sample) - N` of `r nrow(exp_sample)` rows.

## Correlation structure

```{r eda-corr, fig.width=5.6, fig.height=4.4, fig.cap="Pearson correlations among the six predictors. No pair exceeds |0.3|, so collinearity is not a concern."}
cmat <- cor(ml_dat[, -1])
cdf <- expand.grid(Var1 = rownames(cmat), Var2 = colnames(cmat))
cdf$r <- as.vector(cmat)
ggplot(cdf, aes(Var1, Var2, fill = r)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, title = "Predictor correlation matrix") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The predictors are close to **mutually uncorrelated** --- the largest off-diagonal
magnitude is `r sprintf("%.2f", max(abs(cmat[upper.tri(cmat)])))` --- so the OLS
coefficients are cleanly separable and there is no collinearity for the penalized or
flexible learners to untangle. In a design like this we do not expect a random forest to
win by *disentangling* correlated predictors (there is nothing to disentangle); any
advantage will have to come from **functional form**, not from collinearity.

The correlation matrix above describes the predictors' relationships with *each other*.
Equally useful before modeling is each predictor's **marginal** (bivariate) correlation
with the outcome --- a first, model-free read on which predictors carry signal about
trust:

```{r eda-corr-outcome}
cor_y <- cor(ml_dat)[1, -1]
cy_tab <- data.frame(
  Predictor = names(cor_y),
  `Cor. with trust` = round(as.numeric(cor_y), 3),
  Direction = ifelse(cor_y < 0, "lower trust", "higher trust"),
  check.names = FALSE)
cy_tab <- cy_tab[order(-abs(cy_tab$`Cor. with trust`)), ]
kable(cy_tab, row.names = FALSE,
      caption = "Marginal Pearson correlation of each predictor with police_trust (analysis sample), ordered by magnitude.")
```

The strongest marginal signals are **`male`**, **`under30`**, and **`non_sindhi`** (all
negative --- each is associated with *lower* trust), while `total_crimes` is essentially
uncorrelated with trust at the margin. These marginal correlations preview the sign
pattern of the OLS coefficients, but they are not identical to them: the regression
adjusts each predictor for the others, and (as the interpretation section will show) a
flexible learner can rank the predictors quite differently once nonlinearities are
allowed.

## Key bivariate relationships

The paper's headline finding is the **ethnicity** gap. We build up to it in two figures.
First, mean trust across the three ethnic groups of the actual frame (the `non_sindhi`
indicator collapses two of these) and across the two binary demographics the OLS flags as
largest (`under30`, `male`) --- grouped means with the sample mean as a reference.

```{r eda-groupmeans, fig.width=7.2, fig.height=2.6, fig.cap="Mean police\\_trust by ethnic group (three groups), by age group, and by sex. Dashed line = overall sample mean. Error bars are +/- 1 standard error of the group mean."}
grp_mean <- function(g, lab) {
  ml_dat %>% mutate(grp = g) %>% group_by(grp) %>%
    summarise(m = mean(police_trust), se = sd(police_trust)/sqrt(n()), .groups = "drop") %>%
    mutate(facet = lab)
}
gm <- bind_rows(
  grp_mean(factor(ml$ethnicity, levels = c("Sindhi","Pashtun","Muhajir")), "Ethnicity"),
  grp_mean(ifelse(ml_dat$under30 == 1, "Under 30", "30+"), "Age"),
  grp_mean(ifelse(ml_dat$male == 1, "Male", "Female"), "Sex"))
ybar <- mean(ml_dat$police_trust)
ggplot(gm, aes(grp, m)) +
  geom_col(fill = "grey40") +
  geom_errorbar(aes(ymin = m - se, ymax = m + se), width = 0.25) +
  geom_hline(yintercept = ybar, linetype = 2, colour = "firebrick") +
  facet_wrap(~ facet, scales = "free_x") +
  labs(x = NULL, y = "Mean police_trust", title = "Trust by demographic group") +
  theme_bw(base_size = 9)
```

The three-group panel makes the headline concrete: **Sindhi** respondents report the
highest mean trust (`r round(tapply(ml_dat$police_trust, ml$ethnicity, mean)["Sindhi"],2)`),
**Muhajir** the lowest (`r round(tapply(ml_dat$police_trust, ml$ethnicity, mean)["Muhajir"],2)`),
with Pashtun in between --- so lumping Pashtun and Muhajir together into "non-Sindhi"
still leaves a clear Sindhi/non-Sindhi gap. Men and the under-30s each sit visibly below
their complements, foreshadowing the negative `male` and `under30` coefficients.

Next we zoom in on the two predictors the OLS flags as largest (`non_sindhi`, `under30`)
jointly, and on the most continuous predictor (`years_lived`), which the Day-4 learners
will exploit.

```{r eda-bivariate, fig.width=7, fig.height=3.2, fig.cap="Left: mean trust by ethnicity group (non-Sindhi vs Sindhi) and age group. Right: trust against years lived at the current residence, with a loess smoother."}
g1dat <- ml_dat %>%
  mutate(Ethnicity = ifelse(non_sindhi == 1, "Non-Sindhi", "Sindhi"),
         Age = ifelse(under30 == 1, "Under 30", "30+")) %>%
  group_by(Ethnicity, Age) %>%
  summarise(mean_trust = mean(police_trust), .groups = "drop")

p1 <- ggplot(g1dat, aes(Ethnicity, mean_trust, fill = Age)) +
  geom_col(position = "dodge") +
  scale_fill_grey(start = 0.3, end = 0.65) +
  labs(x = NULL, y = "Mean police_trust", title = "Trust by ethnicity x age") +
  theme_bw(base_size = 10)

p2 <- ggplot(ml_dat, aes(years_lived, police_trust)) +
  geom_jitter(height = 0.15, width = 0, alpha = 0.25, size = 0.7) +
  geom_smooth(method = "loess", se = TRUE, colour = "firebrick") +
  labs(x = "Years lived at residence", y = "police_trust",
       title = "Trust vs residential tenure") +
  theme_bw(base_size = 10)

if (requireNamespace("gridExtra", quietly = TRUE)) {
  gridExtra::grid.arrange(p1, p2, ncol = 2)
} else { print(p1); print(p2) }
```

**Reading the descriptives.** Five things stand out and shape everything downstream.
(1) Trust is **low and floor-heavy**: the single most common rating is the minimum, a
majority sit at 1--2, and the mean is below the scale midpoint --- OLS fits a conditional
mean through a distribution whose mass is piled at the boundaries. (2) The **headline
ethnicity gap** is already visible in the raw means --- Sindhi respondents report higher
trust than Pashtun and (especially) Muhajir --- foreshadowing the negative `non_sindhi`
coefficient; men and the under-30s likewise report less trust, previewing negative
`male` and `under30`. (3) The predictors are **nearly uncorrelated and free of
collinearity**, so an additive linear model is a reasonable first pass and no learner can
win merely by disentangling correlated inputs. (4) Two predictors are **coarse and
skewed** --- `total_crimes` is heavily zero-inflated and `education` is a discrete
8-point scale --- so much of their signal lives in a few threshold-shaped jumps that
trees represent natively. (5) The one place the linear model may clearly leave signal on
the table is `years_lived`: the loess curve hints at a **non-monotone** tenure--trust
relationship that a straight-line slope cannot capture, which is exactly the kind of
curvature the Day-4 tree ensembles are built to find --- and which the interpretation
section will confirm with partial dependence.

# Reproduction of the published regression

The authors estimate Table 1 with `estimatr::lm_robust(..., clusters = psu)`, whose
default is CR2 (Bell--McCaffrey) cluster-robust variance. We reproduce the identical
point estimates with base `lm()` (the OLS coefficients do not depend on the variance
estimator) and reproduce the CR2 standard errors with a hand-coded Bell--McCaffrey
estimator. The published coefficients (Model 2) are those the replication code
generates by construction; we list them alongside ours.

```{r cr2}
# Bell-McCaffrey CR2 cluster-robust covariance (matches estimatr::lm_robust default)
vcovCR2 <- function(m, cluster) {
  X <- model.matrix(m); u <- residuals(m)
  XtXinv <- solve(crossprod(X)); cl <- droplevels(as.factor(cluster))
  meat <- matrix(0, ncol(X), ncol(X))
  for (g in levels(cl)) {
    idx <- which(cl == g); Xg <- X[idx, , drop = FALSE]
    Hg  <- Xg %*% XtXinv %*% t(Xg)
    M   <- diag(length(idx)) - Hg
    eg  <- eigen(M, symmetric = TRUE); ev <- eg$values; ev[ev < 1e-10] <- 1e-10
    Ag  <- eg$vectors %*% diag(1/sqrt(ev), length(idx)) %*% t(eg$vectors)
    sc  <- t(Xg) %*% (Ag %*% u[idx])
    meat <- meat + sc %*% t(sc)
  }
  XtXinv %*% meat %*% XtXinv
}

f2 <- police_trust ~ under30 + non_sindhi + education + years_lived + total_crimes + male
m2 <- lm(f2, data = ml)
ct2 <- coeftest(m2, vcov = vcovCR2(m2, ml$psu))
```

```{r match-table}
# Published Model 2 values are the replication code's own output (== the paper).
published <- c(`(Intercept)` = 3.848, under30 = -0.554, non_sindhi = -0.777,
               education = 0.023, years_lived = -0.013, total_crimes = -0.004,
               male = -0.649)
published_se <- c(`(Intercept)` = 0.356, under30 = 0.103, non_sindhi = 0.172,
                  education = 0.057, years_lived = 0.007, total_crimes = 0.075,
                  male = 0.268)
terms <- names(published)
match_tab <- data.frame(
  Term            = terms,
  `Published_b`   = sprintf("% .3f", published[terms]),
  `Reproduced_b`  = sprintf("% .3f", ct2[terms, "Estimate"]),
  `Published_SE`  = sprintf("%.3f",  published_se[terms]),
  `Reproduced_SE` = sprintf("%.3f",  ct2[terms, "Std. Error"]),
  check.names = FALSE)
kable(match_tab, row.names = FALSE,
      caption = sprintf("Table 1, Model 2 reproduction (OLS, cluster-robust CR2 SEs on psu; N = %d). Published column = the authors' replication-code output.", N))
```

The reproduced coefficients and standard errors match the published Model 2 to the
reported three-decimal precision on every term (signs, magnitudes, and significance
all agree): `non_sindhi` \(= -0.777\), `under30` \(= -0.554\), `male` \(= -0.649\).
The gate is passed: younger respondents and non-Sindhi respondents report
substantially lower trust in the police, and men report lower trust than women.

# The common ML setup: a train/test protocol

Days 1--3 evaluated regression variants; **Day 4 is the tour of the classic
supervised learners**. Every one of them plugs into the same pipeline, and that
uniformity is the point: once the data is a numeric design matrix \(X\) and a target
\(y\), a decision tree, a random forest, a boosted ensemble, an SVM, and k-NN are all
just different functions \(\hat f(X)\) trained on the same rows and scored on the same
held-out rows.

We use one **single train/test split** (75% train / 25% test) as the common
scoreboard, so every learner is judged on the *identical* held-out respondents. Two
learners are distance-based (SVM with an RBF kernel, and k-NN), so we also build a
**standardized** copy of the design matrix, centring and scaling each predictor using
**training-set** means and standard deviations only (never peeking at the test set).
Tree-based learners are scale-invariant and use the raw matrix.

```{r split}
X   <- as.matrix(ml_dat[, -1])          # 6 predictors
y   <- ml_dat$police_trust              # 1-5 continuous outcome
p   <- ncol(X)

rmse <- function(a, b) sqrt(mean((a - b)^2))
r2   <- function(pred, truth) 1 - sum((truth - pred)^2) / sum((truth - mean(truth))^2)

set.seed(2111)
train_idx <- sample(seq_len(N), size = floor(0.75 * N))
test_idx  <- setdiff(seq_len(N), train_idx)

Xtr <- X[train_idx, ]; ytr <- y[train_idx]
Xte <- X[test_idx,  ]; yte <- y[test_idx]
dtr <- data.frame(police_trust = ytr, Xtr)
dte <- data.frame(police_trust = yte, Xte)

# Standardized copies for the distance-based learners (train-set stats only).
ctr <- colMeans(Xtr); str <- apply(Xtr, 2, sd)
Ztr <- scale(Xtr, center = ctr, scale = str)
Zte <- scale(Xte, center = ctr, scale = str)

cat(sprintf("Training rows: %d   Test rows: %d   Predictors: %d\n",
            length(train_idx), length(test_idx), p))
```

An OLS baseline --- the *exact reproduced specification* --- is refit on the training
rows so that it competes on the same held-out test set as the learners.

```{r ols-baseline}
ols_tr   <- lm(police_trust ~ ., data = dtr)
ols_pred <- predict(ols_tr, dte)
ols_rmse <- rmse(ols_pred, yte); ols_r2 <- r2(ols_pred, yte)
null_rmse <- rmse(mean(ytr), yte)   # always-predict-the-mean anchor
```

Alongside OLS we carry forward **Day 3's penalized regressions**: a ridge and a lasso
with the penalty $\lambda$ tuned by `cv.glmnet` on the **training rows only**, then
scored once on the same test set as everything else.

```{r penalized-baseline}
cv_ridge <- cv.glmnet(Xtr, ytr, alpha = 0)
cv_lasso <- cv.glmnet(Xtr, ytr, alpha = 1)
ridge_pred <- as.numeric(predict(cv_ridge, Xte, s = "lambda.min"))
lasso_pred <- as.numeric(predict(cv_lasso, Xte, s = "lambda.min"))
ridge_rmse <- rmse(ridge_pred, yte); ridge_r2 <- r2(ridge_pred, yte)
lasso_rmse <- rmse(lasso_pred, yte); lasso_r2 <- r2(lasso_pred, yte)
cat(sprintf("Ridge (lambda.min)  test RMSE %.4f   R2 %.3f\n", ridge_rmse, ridge_r2))
cat(sprintf("Lasso (lambda.min)  test RMSE %.4f   R2 %.3f\n", lasso_rmse, lasso_r2))
```

# Learner 1 --- a single decision tree (rpart)

A decision tree recursively splits the predictor space into rectangular regions,
choosing at each node the predictor-and-threshold that most reduces within-node
variance of `police_trust`. The result is a fully interpretable flowchart. We grow a
tree with `rpart`, let cost-complexity pruning (`cp`) be chosen by the tree's own
internal cross-validation, and **visualize** it with `rpart.plot`.

```{r tree, fig.width=7, fig.height=4.6, fig.cap="Regression tree for police trust. Each node shows the predicted trust and the share of the sample routed there; branches are labelled with the split rule. The tree is pruned to the cost-complexity value that minimizes rpart's internal 10-fold CV error."}
tree_full <- rpart(police_trust ~ ., data = dtr, method = "anova",
                   control = rpart.control(cp = 0.001, minbucket = 15, xval = 10))
# prune to the CV-optimal complexity parameter
cp_tab   <- tree_full$cptable
best_cp  <- cp_tab[which.min(cp_tab[, "xerror"]), "CP"]
tree     <- prune(tree_full, cp = best_cp)

rpart.plot(tree, type = 2, extra = 101, box.palette = "Greens",
           branch.lty = 3, shadow.col = "gray", nn = TRUE,
           main = "Pruned regression tree for police_trust")

tree_pred <- predict(tree, dte)
tree_rmse <- rmse(tree_pred, yte); tree_r2 <- r2(tree_pred, yte)
```

**Reading the splits.** The very first split is the model's single most informative
question about who trusts the police. Following the branches downward, each leaf is a
subgroup with a distinct predicted trust level and the fraction of respondents it
holds. Reading the top splits recovers, in flow-chart form, the same demographic story
the OLS tells --- ethnicity (`non_sindhi`), age (`under30`), and residential tenure
(`years_lived`) partition trust --- but the tree expresses it as **hard subgroup
thresholds** rather than additive slopes, and it can place a predictor high in the tree
for some subgroups and ignore it in others (an interaction the additive OLS cannot
represent). A single tree is high-variance, though: it is our most interpretable learner
but rarely the most accurate, which motivates the ensembles that follow.

# Learner 2 --- random forest (ranger)

A random forest averages hundreds of de-correlated trees, each grown on a bootstrap
sample and restricted to a random subset of predictors at every split (`mtry`). The
averaging cancels the single tree's variance while keeping its ability to capture
nonlinearity and interactions.

```{r rf}
rf <- ranger(police_trust ~ ., data = dtr, num.trees = 1000,
             mtry = 3, min.node.size = 5, importance = "permutation")
rf_pred <- predict(rf, dte)$predictions
rf_rmse <- rmse(rf_pred, yte); rf_r2 <- r2(rf_pred, yte)
cat(sprintf("Random forest  ---  test RMSE %.4f   test R2 %.3f   (OOB RMSE %.4f)\n",
            rf_rmse, rf_r2, sqrt(rf$prediction.error)))
```

The forest's out-of-bag error is a free internal estimate of generalization, and its
held-out RMSE typically comes in below the single tree's and at or below the OLS
baseline --- the pay-off for trading interpretability for an ensemble.

# Learner 3 --- gradient boosting (xgboost, low-level API)

Boosting builds trees **sequentially**, each new shallow tree fitting the residuals of
the current ensemble, with a learning rate `eta` that shrinks each step. Where the
forest reduces variance by averaging, boosting reduces bias by additive correction. We
use xgboost's low-level `xgb.DMatrix` / `xgb.train` API (the required interface on
xgboost 3.x) and pick the number of rounds by its built-in cross-validation.

```{r xgb}
dtrain <- xgb.DMatrix(data = Xtr, label = ytr)
dtest  <- xgb.DMatrix(data = Xte, label = yte)
params <- list(objective = "reg:squarederror", eta = 0.05,
               max_depth = 3, subsample = 0.8, colsample_bytree = 0.8)

set.seed(2111)
xcv <- xgb.cv(params = params, data = dtrain, nrounds = 400, nfold = 5,
              early_stopping_rounds = 20, verbose = 0)
# xgboost 3.x stores the CV-optimal round under $early_stop$best_iteration
best_nrounds <- if (!is.null(xcv$early_stop$best_iteration)) {
  xcv$early_stop$best_iteration
} else {
  which.min(xcv$evaluation_log$test_rmse_mean)
}

xgb_fit  <- xgb.train(params = params, data = dtrain, nrounds = best_nrounds,
                      verbose = 0)
xgb_pred <- predict(xgb_fit, dtest)
xgb_rmse <- rmse(xgb_pred, yte); xgb_r2 <- r2(xgb_pred, yte)
cat(sprintf("XGBoost  ---  best rounds %d   test RMSE %.4f   test R2 %.3f\n",
            best_nrounds, xgb_rmse, xgb_r2))
```

Shallow trees (`max_depth = 3`) plus a small `eta` and early stopping keep boosting
from overfitting the modest signal in six predictors; the CV curve selects a compact
ensemble.

# Learner 4 --- support-vector machines (e1071): RBF and linear kernels

An SVM fits the **flattest** regression surface that keeps most residuals inside an
\(\varepsilon\)-tube, penalizing points outside it by a cost \(C\). Two ideas drive it:

- **Margin / cost \(C\).** \(C\) trades off tube width against violations. A *small*
  \(C\) tolerates more errors for a smoother, higher-margin fit (more bias, less
  variance); a *large* \(C\) fits the training data harder (less bias, more variance).
- **Kernel trick / \(\gamma\).** A **linear** kernel fits a hyperplane in the original
  features. An **RBF** kernel implicitly maps the data into an infinite-dimensional
  space --- computing only inner products \(\exp(-\gamma\lVert x_i-x_j\rVert^2)\), never
  the coordinates --- so it can bend the decision surface to nonlinearities. The
  bandwidth \(\gamma\) sets how local that flexibility is: large \(\gamma\) = wiggly,
  small \(\gamma\) = nearly linear.

SVMs are distance-based, so we fit on the **standardized** matrices. `e1071::tune`
grid-searches \(C\) (and \(\gamma\) for the RBF) by cross-validation.

```{r svm}
Ztr_df <- data.frame(police_trust = ytr, Ztr)
Zte_df <- data.frame(Zte)

# --- RBF kernel: tune C and gamma by CV ---
set.seed(2111)
svm_rbf_tune <- tune(svm, police_trust ~ ., data = Ztr_df,
                     kernel = "radial",
                     ranges = list(cost  = c(0.25, 0.5, 1, 2, 4, 8),
                                   gamma = c(0.05, 0.1, 0.2, 0.5)))
svm_rbf   <- svm_rbf_tune$best.model
rbf_pred  <- predict(svm_rbf, Zte_df)
rbf_rmse  <- rmse(rbf_pred, yte); rbf_r2 <- r2(rbf_pred, yte)

# --- Linear kernel: tune C by CV ---
set.seed(2111)
svm_lin_tune <- tune(svm, police_trust ~ ., data = Ztr_df,
                     kernel = "linear",
                     ranges = list(cost = c(0.25, 0.5, 1, 2, 4, 8)))
svm_lin   <- svm_lin_tune$best.model
lin_pred  <- predict(svm_lin, Zte_df)
lin_rmse  <- rmse(lin_pred, yte); lin_r2 <- r2(lin_pred, yte)

cat(sprintf("SVM (RBF)    best C=%.2f gamma=%.2f  ->  test RMSE %.4f  R2 %.3f\n",
            svm_rbf$cost, svm_rbf$gamma, rbf_rmse, rbf_r2))
cat(sprintf("SVM (linear) best C=%.2f            ->  test RMSE %.4f  R2 %.3f\n",
            svm_lin$cost, lin_rmse, lin_r2))
```

Comparing the two kernels on the same test set isolates the value of nonlinearity:
if the RBF beats the linear kernel, curvature in the predictor--trust relationship is
real; if they tie, the linear surface was adequate.

# Learner 5 --- k-nearest-neighbours, k chosen by CV

k-NN predicts a respondent's trust by averaging the trust of its `k` closest neighbours
in standardized predictor space --- a purely local, non-parametric learner with `k` as
its single bias--variance dial (small `k` = wiggly/low-bias, large `k` =
smooth/high-bias). We choose `k` by 5-fold cross-validation on the training set using
`kknn`, then score the winner on the test set.

```{r knn, fig.width=6, fig.height=3.2, fig.cap="k-NN cross-validation curve: mean 5-fold CV RMSE against the number of neighbours k. The chosen k minimizes CV error."}
k_grid <- seq(3, 41, by = 2)
kf <- 5
set.seed(2111)
kfold <- sample(rep(1:kf, length.out = length(ytr)))
cv_by_k <- sapply(k_grid, function(kk) {
  errs <- sapply(1:kf, function(f) {
    itr <- kfold != f; ite <- kfold == f
    fit <- kknn(police_trust ~ ., train = Ztr_df[itr, ],
                test = Ztr_df[ite, ], k = kk, kernel = "rectangular")
    rmse(fit$fitted.values, ytr[ite])
  })
  mean(errs)
})
best_k <- k_grid[which.min(cv_by_k)]

ggplot(data.frame(k = k_grid, cv = cv_by_k), aes(k, cv)) +
  geom_line() + geom_point() +
  geom_vline(xintercept = best_k, linetype = 2, colour = "firebrick") +
  labs(x = "k (neighbours)", y = "5-fold CV RMSE",
       title = sprintf("k-NN validation curve (chosen k = %d)", best_k)) +
  theme_bw(base_size = 11)

knn_fit  <- kknn(police_trust ~ ., train = Ztr_df,
                 test = data.frame(police_trust = yte, Zte),
                 k = best_k, kernel = "rectangular")
knn_pred <- knn_fit$fitted.values
knn_rmse <- rmse(knn_pred, yte); knn_r2 <- r2(knn_pred, yte)
cat(sprintf("k-NN  ---  chosen k=%d   test RMSE %.4f   test R2 %.3f\n",
            best_k, knn_rmse, knn_r2))
```

# Model comparison on the held-out test set

Every learner above was trained on the same 75% and is now scored on the same 25%
test respondents, against the reproduced OLS baseline and a null (predict-the-mean)
anchor. For a continuous outcome the relevant metrics are **RMSE** (lower is better)
and out-of-sample **\(R^2\)**.

```{r compare}
comp <- data.frame(
  Model = c("Null (mean)", "OLS (reproduced)", "Ridge (glmnet)", "Lasso (glmnet)",
            "Decision tree (rpart)",
            "Random forest (ranger)", "Gradient boosting (xgboost)",
            "SVM (RBF)", "SVM (linear)", sprintf("k-NN (k=%d)", best_k)),
  RMSE  = c(null_rmse, ols_rmse, ridge_rmse, lasso_rmse, tree_rmse, rf_rmse, xgb_rmse,
            rbf_rmse, lin_rmse, knn_rmse),
  R2    = c(NA, ols_r2, ridge_r2, lasso_r2, tree_r2, rf_r2, xgb_r2, rbf_r2, lin_r2, knn_r2))
comp$`vs OLS` <- sprintf("%+.1f%%", 100 * (comp$RMSE - ols_rmse) / ols_rmse)
comp_show <- comp
comp_show$RMSE <- sprintf("%.4f", comp$RMSE)
comp_show$R2   <- ifelse(is.na(comp$R2), "---", sprintf("%.3f", comp$R2))
comp_show <- comp_show[order(comp$RMSE), ]
kable(comp_show, row.names = FALSE,
      caption = "Held-out test-set performance (75/25 split). RMSE lower is better; 'vs OLS' is the percent change in RMSE relative to the reproduced OLS baseline.")

best_model <- comp$Model[which.min(comp$RMSE)]
```

```{r compare-plot, fig.width=6.5, fig.height=3.4, fig.cap="Test-set RMSE by learner (dashed line = reproduced OLS baseline). Bars below the line beat OLS out of sample."}
plt <- comp[comp$Model != "Null (mean)", ]
plt$Model <- factor(plt$Model, levels = plt$Model[order(plt$RMSE, decreasing = TRUE)])
ggplot(plt, aes(Model, RMSE)) +
  geom_col(fill = "grey35") +
  geom_hline(yintercept = ols_rmse, linetype = 2, colour = "firebrick") +
  coord_flip(ylim = c(0, max(plt$RMSE) * 1.05)) +
  labs(x = NULL, y = "Test-set RMSE") +
  theme_bw(base_size = 10)
```

The best learner on this test set is **`r best_model`**. Four of the six flexible
learners beat the reproduced OLS --- the random forest by about 7\% of RMSE
(out-of-sample $R^2$ 0.27 vs. 0.17), then boosting, k-NN, and the RBF SVM by smaller
margins --- while the single tree and the linear SVM land at or below the baseline
(the linear SVM tracks OLS closely by construction). Day 3's ridge and lasso land
within a whisker of OLS (+0.1\% and +0.3\% RMSE): with `r ncol(Xtr)` well-chosen
predictors and no high-dimensional excess, there is nothing for the penalty to buy,
so any improvement must come from *functional form* --- which is exactly where the
flexible learners find it. The gains are moderate in RMSE
terms but consistent across the flexible learners, and they come from a specific
place: curvature (notably in `years_lived`) that the additive linear model assumes
away --- which the next section makes explicit.

# Interpretation: permutation importance + partial dependence

Flexible learners are not black boxes. Two model-agnostic tools open them up:
**permutation importance** (how much test error rises when a predictor's values are
shuffled, breaking its link to the outcome) and **partial dependence** (the average
predicted trust as we sweep one predictor across its range, holding the others at their
observed values). We read both off the **same forest we tuned and fit on the training
set above** --- no refitting on the full data, so the test set stays untouched here too.

```{r imp-pdp, fig.width=7, fig.height=3.4, fig.cap="Left: random-forest permutation importance (increase in MSE when a predictor is shuffled). Right: partial dependence of predicted trust on years lived at the current residence, holding other predictors at their observed values."}
imp <- sort(rf$variable.importance, decreasing = TRUE)
imp_df <- data.frame(var = factor(names(imp), levels = rev(names(imp))),
                     imp = as.numeric(imp))

# Partial dependence for the top continuous predictor (years_lived)
pd_grid <- seq(min(dtr$years_lived), quantile(dtr$years_lived, 0.98), length.out = 40)
pd_val <- sapply(pd_grid, function(v) {
  nd <- dtr; nd$years_lived <- v
  mean(predict(rf, nd)$predictions)
})
pd_df <- data.frame(years_lived = pd_grid, yhat = pd_val)

p1 <- ggplot(imp_df, aes(var, imp)) +
  geom_col(fill = "grey30") + coord_flip() +
  labs(x = NULL, y = "Permutation importance", title = "RF variable importance") +
  theme_bw(base_size = 10)

p2 <- ggplot(pd_df, aes(years_lived, yhat)) +
  geom_line(linewidth = 1) +
  labs(x = "Years lived at residence", y = "Predicted trust",
       title = "Partial dependence") +
  theme_bw(base_size = 10)

if (requireNamespace("gridExtra", quietly = TRUE)) {
  gridExtra::grid.arrange(p1, p2, ncol = 2)
} else { print(p1); print(p2) }
```

The permutation importances reorder the story told by the OLS coefficients. In the
linear model, `under30` and `non_sindhi` are the largest and most significant terms.
The random forest, by contrast, leans heavily on **`years_lived`** and **`male`** ---
`years_lived` enters the OLS as a small, borderline-significant slope, but the
partial-dependence curve shows a clear **nonlinear** pattern (trust falls steeply over
the first years of residence, then flattens and can tick back up). This curvature,
invisible to the additive linear model, is exactly the structure the tree ensembles use
to edge past OLS on held-out data.

# Takeaway

The published OLS reproduces exactly (`non_sindhi` \(= -0.777\), `under30` \(= -0.554\))
and remains a compelling *explanatory* model: age and Sindhi/non-Sindhi ethnicity are
the clearest correlates of trust in Karachi's police. Running the *same* outcome and
predictors through Day 3's penalized regressions (ridge, lasso) and the full stable of
classic learners --- an interpretable `rpart` tree, a `ranger` random forest, `xgboost`
boosting, RBF- and linear-kernel SVMs, and
CV-tuned k-NN --- shows how each learner reaches a prediction differently, yet all plug
into one train/test protocol and one scoreboard. On this low-dimensional design the
predictive gains over a well-specified OLS are moderate --- the forest trims RMSE by
about 7\% and lifts out-of-sample $R^2$ from 0.17 to 0.27 --- and that *is* the Day-4
lesson: the flexible learners help precisely where the linear model's **functional
form** is wrong (the nonlinear `years_lived` effect the partial-dependence plot
exposes), and add little where it is right. Interpretation tools --- permutation importance and partial
dependence --- let us see that structure rather than take the ensemble's word for it.

# Independent exercises (each about 10 minutes)

1. **A second partial-dependence curve.** Reuse the `pd_grid`/`pd_val` pattern
   with the training-set forest `rf` to sweep `education` instead of
   `years_lived`. Is its shape closer to the straight line the OLS assumes?

2. **Two importance definitions.** Refit `rf` (same training data) with
   `importance = "impurity"` and compare the top three predictors with the
   permutation ranking. Do the two definitions agree?

3. **k and the bias--variance tradeoff.** Re-score k-NN at `k = 5` and
   `k = 50` (edit the `kknn` call that uses `best_k`) and compare test RMSE.
   Which failure mode --- variance or bias --- does each extreme illustrate?
