---
title: "Reproduce-then-Extend — State Policy Liberalism (American State Politics) (Day 1)"
author: "Intro to Machine Learning · Bruce A. Desmarais"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: true
geometry: margin=1in
fontsize: 10pt
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
  - \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 3.6, fig.align = "center")
options(width = 80)
library(glmnet)   # lasso / ridge / elastic net
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(e1071)    # skewness / kurtosis
library(patchwork)
theme_set(theme_minimal(base_size = 9))
```

> **Published study.** Caughey, D. & Warshaw, C. (2016). "The Dynamics of State Policy Liberalism,
> 1936--2014." *American Journal of Political Science* 60(4):899--913. Building on the state-representation
> tradition (Erikson, Wright & McIver, *Statehouse Democracy*, 1993), the paper measures each state's
> **policy liberalism** and links it to public opinion and state political conditions -- a **cross-sectional
> OLS** of policy liberalism on state characteristics. This example illustrates **regularization and variable
> selection** in a small-sample (n = 50) design, where predictors are many relative to observations.

# Background

A central question in American state politics is why some states adopt systematically more **liberal
policies** than others. The state-representation literature argues that policy tracks **public opinion**
(the "opinion--policy link"), mediated by **party control** and the organized strength of liberal and
conservative constituencies. Caughey & Warshaw construct a dynamic measure of state policy liberalism and
show it responds to mass preferences. Here we treat the **cross-section of the 50 states** as the unit of
analysis and ask which state characteristics predict policy liberalism -- the setting the state-politics
literature has studied with ordinary regression for decades.

# Data and codebook

**Unit of analysis:** a U.S. **state** (cross-section averaged over 2005--2012); n = 50. Data are drawn from
the **Correlates of State Policy Project** (Jordan & Grossmann 2020), which compiles the standard
state-politics measures.

| Variable | Definition |
|---|---|
| `pollib_median` | state policy liberalism (Caughey--Warshaw), higher = more liberal **(outcome)** |
| `mood` | public policy mood (higher = more liberal preferences) |
| `citi6013` | citizen ideology (Berry et al.), higher = more liberal |
| `ranney4_control` | Ranney index of Democratic party control (higher = more Democratic) |
| `union_density` | percent of workers unionized |
| `evangelical_pop` | evangelical share of the population |
| `gini_coef` | income inequality (Gini coefficient) |
| `gsppcap` | gross state product per capita |

# Descriptive analysis

We first load the data and get to know it thoroughly before any modeling. The
data-loading step is unchanged from the reproduction below; the descriptive
analysis operates on the loaded data frame `d`.

```{r data}
d <- read.csv(paste0(
  "https://raw.githubusercontent.com/bdesmarais/",
  "intro-ml-statistical-horizons-4day/main/",
  "tutorials/day3_regularization_and_selection/11_state_policy_liberalism/data/state_policy_liberalism.csv"))
preds <- setdiff(names(d), c("st", "pollib_median"))
cat(sprintf("%d states x %d predictors (p/n = %.2f)\n",
            nrow(d), length(preds), length(preds) / nrow(d)))
```

## Dataset and provenance

**Unit of analysis.** One row per U.S. **state**, a pure cross-section averaged
over roughly 2005--2012, so `n = 50`. There is no time dimension and no
sampling: this is the *entire population* of states, which is exactly why the
small-`n` problem is structural rather than fixable by collecting more data.
With `r length(preds)` predictors and 50 states, the predictor-to-observation
ratio is `r round(length(preds)/nrow(d), 2)` -- modest in absolute terms, but
large enough that ordinary least squares has few states per coefficient and
begins to over-fit. That tension is the entire point of this Day-3 example.

**Source.** Variables are drawn from the **Correlates of State Policy Project**
(Jordan & Grossmann 2020), which compiles the standard state-politics measures;
the outcome is the Caughey--Warshaw (2016) dynamic policy-liberalism index.

The variable dictionary below lists the outcome and all `r length(preds)`
predictors, their role, type, and coding direction.

```{r dict}
dict <- data.frame(
  Variable = c("pollib_median", "mood", "citi6013", "ranney4_control",
               "union_density", "evangelical_pop", "gini_coef", "gsppcap"),
  Role = c("outcome", rep("predictor", 7)),
  Type = "numeric",
  Description = c(
    "Policy liberalism (Caughey-Warshaw); higher = more liberal",
    "Public policy mood; higher = more liberal preferences",
    "Citizen ideology (Berry et al.); higher = more liberal",
    "Ranney index of Democratic party control; higher = more Dem.",
    "Percent of workers unionized",
    "Evangelical share of the population",
    "Income inequality (Gini coefficient)",
    "Gross state product per capita (US$)"))
kable(dict)
```

All eight variables are continuous. There are no categorical or binary
predictors, which keeps the descriptive work focused on distributions and
correlations.

## The outcome in depth

`pollib_median` is a standardized latent index, so it is centered near zero and
takes both signs (negative = more conservative than average, positive = more
liberal). We examine its full distribution and shape.

```{r out-dist, fig.height=2.6, fig.cap="Distribution of state policy liberalism: histogram with density overlay (left) and boxplot (right)."}
p1 <- ggplot(d, aes(pollib_median)) +
  geom_histogram(aes(y = after_stat(density)), bins = 12,
                 fill = "#a6cee3", colour = "white") +
  geom_density(colour = "#1f78b4", linewidth = 0.7) +
  labs(x = "policy liberalism", y = "density", title = "Outcome distribution")
p2 <- ggplot(d, aes(y = pollib_median)) +
  geom_boxplot(fill = "#a6cee3", width = 0.4) +
  labs(y = "policy liberalism", x = NULL, title = "Boxplot")
p1 + p2
```

```{r out-stats}
stat_row <- function(x) {
  q <- quantile(x, c(.25, .5, .75))
  c(n = length(x), mean = mean(x), sd = sd(x), min = min(x),
    Q1 = q[1], median = q[2], Q3 = q[3], max = max(x),
    IQR = q[3] - q[1], skew = skewness(x), kurt = kurtosis(x))
}
out_tab <- t(stat_row(d$pollib_median))
rownames(out_tab) <- "pollib_median"
kable(round(out_tab, 2))
```

**Shape.** The outcome is close to **symmetric** (skewness
`r round(skewness(d$pollib_median), 2)`) and mildly **platykurtic**
(excess kurtosis `r round(kurtosis(d$pollib_median), 2)`) -- flatter than a
normal, with mass spread fairly evenly from conservative to liberal states
rather than bunched at a mode. There is no zero-inflation, no floor or ceiling,
and no obvious outlier: the boxplot whiskers cover the full range with no points
beyond them. A response transform is **not warranted**; we model `pollib_median`
on its natural scale, as the published OLS does.

## Univariate predictor distributions

With only seven predictors we can profile each one. The faceted histograms show
the marginal shape of every predictor, and the summary table quantifies it.

```{r pred-hist, fig.height=3.4, fig.cap="Marginal distributions of the seven predictors."}
d |>
  select(all_of(preds)) |>
  pivot_longer(everything(), names_to = "var", values_to = "val") |>
  ggplot(aes(val)) +
  geom_histogram(bins = 12, fill = "#b2df8a", colour = "white") +
  facet_wrap(~ var, scales = "free", ncol = 4) +
  labs(x = NULL, y = "count")
```

```{r pred-stats}
pred_tab <- t(sapply(d[preds], function(x) {
  q <- quantile(x, c(.25, .5, .75))
  c(n = length(x), n_miss = sum(is.na(x)), mean = mean(x), sd = sd(x),
    min = min(x), Q1 = q[1], median = q[2], Q3 = q[3], max = max(x),
    skew = skewness(x))
}))
colnames(pred_tab) <- c("n", "n_miss", "mean", "sd", "min",
                        "Q1", "median", "Q3", "max", "skew")
kable(round(pred_tab, 2))
```

The public-opinion and party-control measures (`mood`, `citi6013`,
`ranney4_control`) are roughly symmetric. The socio-economic predictors are
**right-skewed**: `evangelical_pop` (skew
`r round(skewness(d$evangelical_pop), 2)`), `mood`, and `gsppcap` all have a
long upper tail driven by a handful of extreme states (high-evangelical
Southern states; high-income outliers). The predictor scales differ by orders
of magnitude (`gsppcap` in the tens of thousands vs. proportions near zero),
which is exactly why `glmnet` **standardizes predictors internally** before
penalizing -- otherwise the penalty would fall almost entirely on the
large-scale variables.

## Missing data

```{r miss}
miss <- data.frame(
  variable = c("pollib_median", preds),
  n_missing = sapply(d[c("pollib_median", preds)], function(x) sum(is.na(x))))
miss$pct_missing <- round(100 * miss$n_missing / nrow(d), 1)
kable(miss, row.names = FALSE)
cat(sprintf("Complete cases: %d of %d states (%.0f%%).\n",
            sum(complete.cases(d)), nrow(d),
            100 * mean(complete.cases(d))))
```

The data are **fully complete**: every state has every variable, so there is no
co-missingness pattern to display and no listwise deletion in the reproduction.
This is expected for an administrative population-level compilation -- the
Correlates of State Policy Project reports each measure for all 50 states -- and
means the missing-data mechanism is a non-issue here. All 50 states enter every
model.

## Bivariate relationships with the outcome

We next look at how each predictor co-moves with policy liberalism. Because
every predictor is numeric, we use scatterplots with a linear smoother and
annotate each panel with its Pearson correlation `r`.

```{r biv-scatter, fig.height=4, fig.cap="Policy liberalism vs each predictor, with linear fit and Pearson r."}
long <- d |>
  select(pollib_median, all_of(preds)) |>
  pivot_longer(all_of(preds), names_to = "var", values_to = "val")
rlab <- long |>
  group_by(var) |>
  summarise(r = cor(val, pollib_median), .groups = "drop") |>
  mutate(lab = sprintf("r = %.2f", r))
ggplot(long, aes(val, pollib_median)) +
  geom_point(colour = "#1f78b4", alpha = 0.6, size = 1) +
  geom_smooth(method = "lm", se = FALSE, colour = "#e31a1c",
              linewidth = 0.6) +
  geom_text(data = rlab, aes(x = -Inf, y = Inf, label = lab),
            hjust = -0.1, vjust = 1.4, size = 2.6, inherit.aes = FALSE) +
  facet_wrap(~ var, scales = "free_x", ncol = 4) +
  labs(x = NULL, y = "policy liberalism")
```

```{r biv-tab}
marg <- long |>
  group_by(var) |>
  summarise(r = cor(val, pollib_median),
            r2 = r^2, .groups = "drop") |>
  arrange(desc(abs(r)))
kable(marg, digits = 2,
      col.names = c("predictor", "r with outcome", "r-squared"))
```

**Reading the associations.** The opinion--policy link dominates: **public
mood** and **citizen ideology** are the strongest marginal predictors of policy
liberalism (large positive `r`), followed by **Democratic party control** and
**union density**. **Evangelical population** is the strongest *negative*
correlate -- more evangelical states are more conservative. Income inequality
and per-capita product are weaker, more ambiguous marginal predictors. These
raw associations preview the signs the reproduction recovers.

## Multivariate structure and collinearity

Marginal correlations can mislead when predictors are themselves correlated. We
examine the full predictor correlation matrix and the variance-inflation factors
(VIFs) from the reproduction model.

```{r corr-heat, fig.height=3.4, fig.cap="Correlation heatmap of the seven predictors."}
cm <- cor(d[preds])
cm_long <- as.data.frame(as.table(cm))
names(cm_long) <- c("v1", "v2", "r")
ggplot(cm_long, aes(v1, v2, fill = r)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 2.4) +
  scale_fill_gradient2(low = "#e31a1c", mid = "white",
                       high = "#1f78b4", midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r vif}
# VIFs via auxiliary regressions (car is not installed)
vif <- sapply(preds, function(p) {
  aux <- lm(as.formula(paste(p, "~ .")), data = d[preds])
  1 / (1 - summary(aux)$r.squared)
})
kable(data.frame(predictor = preds, VIF = round(vif, 2)),
      row.names = FALSE)
```

Only one predictor pair exceeds the `|r| > 0.7` flag: **citizen ideology and
evangelical population** correlate about **-0.76** (conservative,
high-evangelical states have low citizen-ideology scores). No VIF is alarmingly
high -- the largest is `citi6013` at
`r round(max(vif), 1)`, well under the usual 10 rule of thumb -- so the
collinearity is moderate, not pathological. But combined with only 50 states,
even moderate collinearity inflates OLS coefficient variance, which is precisely
the instability that **regularization** is built to tame in the sections that
follow.

## From description to modeling

The descriptive picture sets up the rest of the tutorial cleanly. The outcome is
symmetric and well-behaved (no transform needed); the predictors carry a clear
opinion--policy signal but differ wildly in scale (so `glmnet` standardizes
them); the data are complete (all 50 states model-able); and with seven
correlated predictors over just 50 states, OLS has little slack. We now
reproduce the published regression, then show that penalized and best-subset
methods predict new states more accurately while naming the compact set of
characteristics that drive policy liberalism.

# Reproduce the published regression

The state-representation model regresses policy liberalism on public opinion and state political conditions.

```{r reproduce}
fit <- lm(pollib_median ~ ., data = d[, c("pollib_median", preds)])
round(summary(fit)$coefficients, 3)
```

**Confirmation against the literature.** In-sample $R^2$ = `r round(summary(fit)$r.squared, 2)`, and the
**opinion--policy link** is recovered: more liberal **public mood** and **citizen ideology**, stronger
**Democratic control**, and higher **union density** all predict more liberal policy, while a larger
**evangelical** population predicts more conservative policy -- the signs the state-politics literature
reports.

# Regularization & variable selection

With only **50 states** and **`r length(preds)` predictors**, ordinary least squares has few observations per
coefficient and **over-fits**: it fits the 50 states well but predicts new states poorly. Penalized
regression -- lasso ($L_1$), ridge ($L_2$), elastic net -- shrinks the coefficients to trade a little bias
for much lower variance.

```{r reg-coef}
X <- as.matrix(d[, preds]); y <- d$pollib_median
set.seed(2026)
lasso <- cv.glmnet(X, y, alpha = 1); ridge <- cv.glmnet(X, y, alpha = 0); enet <- cv.glmnet(X, y, alpha = 0.5)
n_keep <- sum(coef(lasso, s = "lambda.1se")[-1] != 0)
cat(sprintf("At lambda.1se the lasso keeps %d of %d predictors (the rest are set to exactly 0).\n",
            n_keep, length(preds)))
data.frame(
  OLS        = round(coef(fit)[-1], 3),
  Lasso      = round(as.numeric(coef(lasso, s = "lambda.1se"))[-1], 3),
  Ridge      = round(as.numeric(coef(ridge, s = "lambda.1se"))[-1], 3),
  ElasticNet = round(as.numeric(coef(enet,  s = "lambda.1se"))[-1], 3),
  row.names  = preds)
```

**Which fits best out of sample?** With n = 50 a single split is noisy, so we **repeat a 70/30 split 50
times**: each replicate tunes the penalty by cross-validation on the training states and scores once on the
held-out states. Penalized models are scored at `lambda.min` (the loss-minimizing penalty).

```{r reg-perf}
rmse <- function(p, yt) sqrt(mean((yt - p)^2))
REPS <- 50; alphas <- c(Lasso = 1, Ridge = 0, ElasticNet = 0.5)
M <- matrix(NA, REPS, 4, dimnames = list(NULL, c("OLS", names(alphas))))
Xdf <- as.data.frame(X)
for (r in 1:REPS) {
  set.seed(2025 + r)
  te <- sample(nrow(X), round(0.30 * nrow(X))); tr <- setdiff(seq_len(nrow(X)), te)
  ols <- lm(y[tr] ~ ., data = Xdf[tr, ]); M[r, "OLS"] <- rmse(predict(ols, Xdf[te, ]), y[te])
  for (a in names(alphas)) {
    cv <- cv.glmnet(X[tr, ], y[tr], alpha = alphas[a])
    M[r, a] <- rmse(as.numeric(predict(cv, X[te, ], s = "lambda.min")), y[te])
  }
}
round(colMeans(M), 3)                                    # mean held-out RMSE over 50 splits
best <- names(which.min(colMeans(M)[names(alphas)]))
cat(sprintf("\nBest penalization: %s. Held-out RMSE: %s = %.3f vs OLS %.3f (%.0f%% lower).\n",
            best, best, min(colMeans(M)[names(alphas)]), colMeans(M)["OLS"],
            100 * (colMeans(M)["OLS"] - min(colMeans(M)[names(alphas)])) / colMeans(M)["OLS"]))
```

Because there are so few states per predictor, unpenalized OLS over-fits and the penalized models **predict
new states more accurately**, while the lasso simultaneously **selects** the handful of state characteristics
that carry the signal.

```{r reg-path, echo=FALSE, fig.height=3.4, fig.cap="Lasso coefficient paths vs the penalty (log lambda); paths reaching zero are de-selected. The dashed line marks the CV-optimal penalty."}
par(mar = c(4.2, 4.2, 3, 1))
plot(lasso$glmnet.fit, xvar = "lambda", col = "#1f78b455")
abline(v = log(lasso$lambda.min), lty = 2, col = "#e31a1c")
```

# Best-subset selection with ABESS

Lasso reaches a sparse model through shrinkage. **Best-subset selection** instead searches directly for the
subset of predictors that fits best; the **adaptive best-subset (ABESS)** algorithm does so efficiently and
chooses the subset size automatically.

```{r abess}
library(abess)
ab <- abess(X, y)
cat(sprintf("ABESS selects a best subset of %d of %d predictors:\n", ab$best.size, length(preds)))
cf <- as.numeric(coef(ab, support.size = ab$best.size))[-1]
data.frame(predictor = preds, coef = round(cf, 3))[order(-abs(cf)), ]
```

Best subset and the lasso agree on the core drivers of state policy liberalism -- public opinion, party
control, and organized labor -- reached by direct search rather than shrinkage.

# Takeaway

This example illustrates **regularization in a small-sample design**. State-politics regressions have few
observations (50 states) relative to their predictors, so ordinary regression over-fits; lasso/elastic-net
(`glmnet`) and best-subset selection (`abess`) predict new states more accurately **and** name the compact
set of characteristics that drive policy liberalism.

# Recommended exercises

1. Add a couple more state characteristics from the Correlates of State Policy data and see whether the
   held-out gain over OLS grows as the predictor-to-state ratio rises.
2. Compare `lambda.min` and `lambda.1se` predictions: how much accuracy does the sparser 1-SE model give up?
3. Refit predicting a *specific* policy (e.g., minimum wage) instead of the liberalism index and report which
   predictors the lasso keeps.
