---
title: "Kim (2023), *Personal Networks, State Financial Backing and Foreign Direct Investment* — SH Day 2: Regression Extensions"
author: "Machine-Learning Workshop teaching example"
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: true
    toc: false
header-includes:
  - \setlength{\emergencystretch}{3em}
  - \usepackage{fvextra}
  - \AtBeginDocument{\DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}}
  - \AtBeginDocument{\DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}}
  - \fvset{breaklines=true}
  - \usepackage{booktabs}
  - \usepackage{float}
---

```{r setup, include=FALSE}
options(width = 80)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.pos = "H", fig.align = "center")
set.seed(20240706)
suppressMessages({
  library(haven); library(dplyr); library(ggplot2)
  library(splines); library(mgcv); library(pROC); library(knitr)
  library(tidyr); library(GGally); library(patchwork); library(e1071)
})
```

# Study and source

**Citation.** Seungjun Kim. "Personal Networks, State Financial Backing and Foreign
Direct Investment." *Comparative Political Studies* 56(7), 2023.

**Replication source.** Harvard Dataverse (Comparative Political Studies collection),
DOI `10.7910/DVN/GFNDOS`
(<https://doi.org/10.7910/DVN/GFNDOS>). The dataset is a cross-section of South Korean
listed firms; the posted replication `.R` script estimates the main results with plain
`lm()` OLS.

**Main model.** Table 1 of the paper reports OLS regressions explaining the *amount of a
firm's outbound FDI projects* (logged). The theoretical claim: firms whose boards have
ties to state-owned banks obtain financial backing that translates into more foreign
investment. The fully-specified column (Table 1, model 4) is:

$$\begin{aligned}
\texttt{FDIProjects\_Amount} \sim{} & \texttt{Board\_Ties\_to\_State\_Banks} + \texttt{Firm\_Size} + \texttt{Debt\_Ratio} \\
& + \texttt{Profitability} + \texttt{Chaebol} + \texttt{factor(sic\_category)}.
\end{aligned}$$

Outcome: continuous (logged FDI project amount). Model type: **OLS**. Estimation N = 703
(after listwise deletion), across nine one-digit SIC industry categories.

**What this Day-2 tutorial does.** We keep the exact reproduction of Table 1, then treat
the *same outcome* with the full toolkit of regression extensions: multiple-regression
interpretation, the complete residual-diagnostic suite, a logistic-regression / MLE aside,
and a sequence of increasingly flexible mean functions in one key predictor
(polynomial $\to$ spline $\to$ GAM), all compared honestly out of sample. This study is an
ideal vehicle because firm size has a strongly **nonlinear** relationship to FDI that a
straight-line OLS term cannot capture.

```{r load}
# Data ships next to this Rmd (original Stata .dta pulled from Dataverse file id 6563740).
.f <- tempfile(fileext=".dta"); download.file(paste0(
  "https://raw.githubusercontent.com/bdesmarais/intro",
  "-ml-statistical-horizons-4day/main/tutorials/day2_",
  "regression_extensions/kim_fdi_networks/data/kim_fd",
  "i_data.dta"), .f, mode="wb", quiet=TRUE)
d <- read_dta(.f)

# Variable renames exactly as in the posted replication script.
d <- d %>% rename(
  FDIProjects_Amount        = lpostamountuntil2016,
  Board_Ties_to_State_Banks = fincondum,
  Firm_Size                 = wsmarketcap11,
  Debt_Ratio                = debtratio2011,
  Profitability             = roe2011,
  Chaebol                   = chaebol
)

# Common modeling frame used throughout: same predictors as Table 1 model 4, listwise complete.
vars <- c("FDIProjects_Amount", "Board_Ties_to_State_Banks", "Firm_Size",
          "Debt_Ratio", "Profitability", "Chaebol", "sic_category")
md <- d[, vars]
md$sic_category <- factor(md$sic_category)
md <- md[complete.cases(md), ]
```

# Descriptive analysis

A thorough exploratory pass earns its keep here: the outcome's extreme shape and the
firm-size predictor's curvature are exactly what motivate every extension on Day 2, so we
want to *see* them before we model them. We proceed in six movements — (1) the dataset and
a variable dictionary; (2) the outcome in depth; (3) univariate predictor distributions;
(4) missing-data analysis; (5) bivariate relationships with the outcome; and (6)
multivariate structure and collinearity — closing with a narrative of what the descriptives
imply for the modeling.

## Dataset, provenance, and variable dictionary

**Unit of analysis.** The row is a **South Korean publicly-listed firm**. The design is a
single 2011 cross-section: the predictors are measured at (or as of) 2011, and the outcome
accumulates each firm's outbound foreign-direct-investment activity through 2016. There is
no panel/time dimension — every firm contributes one observation.

**Provenance and sampling.** Firms are the KOSPI/KOSDAQ-listed universe used in Kim (2023),
assembled from Worldscope financials, board-membership records used to code personal ties
to state-owned banks, and FDI project filings aggregated to 2016. The replication file is
the posted Stata `.dta` from Harvard Dataverse (DOI `10.7910/DVN/GFNDOS`); the authors' own
`.R` script fits the main table with plain `lm()`. There is no probability sampling — this
is a near-census of listed firms — so the relevant inferential frame is the population of
listed Korean firms, not a random draw from a larger pool.

**Size and shape.** The raw file holds $`r nrow(d)`$ firms and $`r ncol(d)`$ columns; the
modeling frame `md` (the six Table-1 predictors plus the outcome, listwise complete) has
$N = `r nrow(md)`$ firms and $`r ncol(md)`$ columns. Listwise deletion drops
$`r nrow(d) - nrow(md)`$ firms (Section on missingness below). The dictionary lists the
outcome and the six modeling predictors.

```{r eda-dict, results='asis'}
dict <- data.frame(
  Variable = c("FDIProjects\\_Amount", "Board\\_Ties\\_to\\_State\\_Banks",
               "Firm\\_Size", "Debt\\_Ratio", "Profitability", "Chaebol",
               "sic\\_category"),
  Role = c("Outcome", "Treatment", "Predictor", "Control", "Control",
           "Control", "Control"),
  Type = c("continuous", "binary", "continuous", "continuous", "continuous",
           "binary", "factor (9)"),
  Description = c(
    "Logged cumulative outbound FDI project amount, 2016",
    "1 if a board member is tied to a state-owned bank",
    "Scaled 2011 market capitalisation (firm size)",
    "2011 debt-to-assets ratio",
    "2011 return on equity (ROE, percent)",
    "1 if firm belongs to a chaebol conglomerate",
    "One-digit SIC industry code"),
  check.names = FALSE, row.names = NULL)
kable(dict, booktabs = TRUE,
      caption = "Variable dictionary for the outcome and Table-1 modeling predictors.")
```

## The outcome in depth

The outcome is **logged outbound FDI project amount**. It is continuous in principle but
massively **zero-inflated**: `r sprintf("%.0f%%", 100*mean(md$FDIProjects_Amount==0))` of
firms have exactly zero (they never invested abroad), so the distribution is a spike at
zero next to a right-skewed continuum of positive values that runs out to about
`r round(max(md$FDIProjects_Amount),1)`. The left panel below shows the full distribution
(histogram with a density overlay); the right panel shows only the positive mass, which is
itself right-skewed.

```{r eda-outcome-fig, fig.width=7, fig.height=3, fig.cap="Left: the full outcome, dominated by a spike at zero (81% of firms never invest abroad). Right: conditional on investing at all, the positive amounts are right-skewed. Together these motivate the non-Gaussian residual diagnostics later."}
p_all <- ggplot(md, aes(FDIProjects_Amount)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = "grey35", colour = "white") +
  geom_density(colour = "grey10", linewidth = 0.7) +
  labs(x = "Outbound FDI amount (logged)", y = "Density",
       title = "All firms") +
  theme_minimal(base_size = 10)
p_pos <- ggplot(subset(md, FDIProjects_Amount > 0), aes(FDIProjects_Amount)) +
  geom_histogram(bins = 30, fill = "grey35", colour = "white") +
  labs(x = "Outbound FDI amount (logged)", y = "Firms",
       title = "Investing firms only") +
  theme_minimal(base_size = 10)
p_all + p_pos
```

```{r eda-outcome-tab, results='asis'}
os <- md$FDIProjects_Amount
qs <- quantile(os, c(.25, .5, .75))
out_tab <- data.frame(
  Statistic = c("N", "Mean", "SD", "Min", "Q1", "Median", "Q3", "Max",
                "IQR", "Skewness", "Kurtosis (excess)",
                "% exactly zero", "Mean | positive"),
  Value = c(length(os), round(mean(os), 3), round(sd(os), 3), round(min(os), 3),
            round(qs[1], 3), round(qs[2], 3), round(qs[3], 3), round(max(os), 3),
            round(IQR(os), 3), round(e1071::skewness(os), 3),
            round(e1071::kurtosis(os), 3),
            sprintf("%.1f%%", 100*mean(os == 0)),
            round(mean(os[os > 0]), 3)),
  check.names = FALSE)
kable(out_tab, booktabs = TRUE,
      caption = "Distribution of the continuous outcome (logged outbound FDI amount).")
```

**Shape, skew, and transformation.** The outcome is strongly right-skewed
(skewness $`r round(e1071::skewness(os),2)`$) and leptokurtic
(excess kurtosis $`r round(e1071::kurtosis(os),2)`$), driven by the enormous floor at zero:
the median, Q1, and Q3 are *all* zero, so more than three-quarters of firms sit exactly at
the boundary. This is a **floor/zero-inflation** problem, not ordinary skew — the outcome
is already a log of the raw amount, so a further monotone transform would not remove the
spike (you cannot spread out a point mass at zero). A principled treatment would be a hurdle
or Tobit model; the paper stays with OLS, and the price is paid in the residual diagnostics.
For evaluation purposes the practical implication is that a naive mean-only predictor is
hard to beat on RMSE, because the majority class is a single value.

## Univariate distributions of the predictors

The study has one continuous headline confound (**firm size**), two continuous financial
controls (**debt ratio**, **profitability**), two binary indicators (**board ties to
state banks** — the treatment — and **chaebol** status), and a nine-level industry factor
(**SIC category**). We profile the numeric predictors first (table + faceted histograms),
then the categorical/binary ones as frequency tables and bars.

```{r eda-numeric, results='asis'}
num_vars <- c("Firm_Size", "Debt_Ratio", "Profitability")
num_lab  <- c("Firm size (scaled mkt cap)", "Debt ratio", "Profitability (ROE)")
num_summ <- do.call(rbind, lapply(seq_along(num_vars), function(i) {
  x  <- d[[num_vars[i]]]
  qq <- quantile(x, c(.25, .5, .75), na.rm = TRUE)
  data.frame(
    Predictor  = num_lab[i],
    n          = sum(!is.na(x)),
    `n missing` = sum(is.na(x)),
    Mean   = round(mean(x, na.rm = TRUE), 2),
    SD     = round(sd(x, na.rm = TRUE), 2),
    Min    = round(min(x, na.rm = TRUE), 2),
    Q1     = round(qq[1], 2),
    Median = round(qq[2], 2),
    Q3     = round(qq[3], 2),
    Max    = round(max(x, na.rm = TRUE), 2),
    Skew   = round(e1071::skewness(x, na.rm = TRUE), 2),
    check.names = FALSE, row.names = NULL)
}))
kable(num_summ, booktabs = TRUE,
      caption = "Numeric predictors (raw data, before listwise deletion).")
```

The three numeric predictors are on wildly different scales and shapes, so the faceted
histograms below use free x-axes. Firm size is extraordinarily right-skewed (a few giant
firms dwarf the rest); debt ratio is a bounded proportion; and ROE has a small number of
extreme outliers in both tails that stretch the axis.

```{r eda-numeric-hist, fig.width=7, fig.height=2.6, fig.cap="Univariate distributions of the three numeric predictors (free x-axes). Firm size is severely right-skewed; debt ratio is a bounded proportion; profitability (ROE) carries extreme two-tailed outliers."}
num_long <- md %>%
  select(all_of(num_vars)) %>%
  pivot_longer(everything(), names_to = "var", values_to = "value") %>%
  mutate(var = factor(var, levels = num_vars, labels = num_lab))
ggplot(num_long, aes(value)) +
  geom_histogram(bins = 30, fill = "grey40", colour = "white") +
  facet_wrap(~ var, scales = "free", nrow = 1) +
  labs(x = NULL, y = "Firms") +
  theme_minimal(base_size = 9)
```

```{r eda-binary, results='asis'}
bt <- table(md$Board_Ties_to_State_Banks)
ch <- table(md$Chaebol)
bin_tab <- data.frame(
  `Binary predictor` = c("Board ties to state banks", "Chaebol (large conglomerate)"),
  `= 0 (count)` = c(bt["0"], ch["0"]),
  `= 1 (count)` = c(bt["1"], ch["1"]),
  `Proportion = 1` = round(c(mean(md$Board_Ties_to_State_Banks), mean(md$Chaebol)), 3),
  check.names = FALSE, row.names = NULL)
kable(bin_tab, booktabs = TRUE,
      caption = "Binary predictors in the modeling frame (proportion equal to 1).")
```

Both binary predictors are **imbalanced**: only about
`r sprintf("%.0f%%", 100*mean(md$Board_Ties_to_State_Banks))` of firms have board ties to
state banks, and roughly
`r sprintf("%.0f%%", 100*mean(md$Chaebol))` are chaebol-affiliated. The industry factor
spreads firms across nine one-digit SIC categories, one of which dominates.

```{r eda-cat-bars, fig.width=7, fig.height=2.6, fig.cap="Frequency of the categorical predictors: the two binary indicators (left, both minority-1) and the nine-level SIC industry factor (right, one category dominant)."}
bin_df <- data.frame(
  var = rep(c("Board ties", "Chaebol"), each = 2),
  level = rep(c("0", "1"), 2),
  n = c(bt["0"], bt["1"], ch["0"], ch["1"]))
p_bin <- ggplot(bin_df, aes(level, n, fill = var)) +
  geom_col(position = "dodge") +
  scale_fill_grey(start = 0.4, end = 0.7) +
  labs(x = NULL, y = "Firms", fill = NULL, title = "Binary predictors") +
  theme_minimal(base_size = 9) + theme(legend.position = "top")
sic_df <- as.data.frame(table(md$sic_category))
names(sic_df) <- c("sic", "n")
p_sic <- ggplot(sic_df, aes(sic, n)) +
  geom_col(fill = "grey40") +
  labs(x = "One-digit SIC", y = "Firms", title = "Industry") +
  theme_minimal(base_size = 9)
p_bin + p_sic
```

```{r eda-sic, results='asis'}
sic_tab <- as.data.frame(table(md$sic_category))
names(sic_tab) <- c("SIC category (one digit)", "Firms")
sic_tab$`Share` <- sprintf("%.1f%%", 100 * sic_tab$Firms / sum(sic_tab$Firms))
kable(sic_tab, booktabs = TRUE,
      caption = "Industry composition: nine one-digit SIC categories.")
```

## Missing-data analysis

Listwise deletion removes `r nrow(d) - nrow(md)` of the `r nrow(d)` raw firms
(`r sprintf("%.1f%%", 100*(nrow(d)-nrow(md))/nrow(d))`). The table below counts and
percentages missing values per model variable in the raw data.

```{r eda-missing, results='asis'}
miss <- sapply(d[, vars], function(x) sum(is.na(x)))
miss_tab <- data.frame(
  Variable = c("Outbound FDI amount", "Board ties to state banks", "Firm size",
               "Debt ratio", "Profitability", "Chaebol", "SIC category"),
  `# missing` = as.integer(miss),
  `% missing` = sprintf("%.1f%%", 100 * as.integer(miss) / nrow(d)),
  check.names = FALSE, row.names = NULL)
miss_tab <- miss_tab[order(-miss_tab$`# missing`), ]
kable(miss_tab, booktabs = TRUE, row.names = FALSE,
      caption = sprintf("Missingness per model variable (of %d raw firms).", nrow(d)))
```

```{r eda-missing-pattern, results='asis'}
# How many firms are complete, and how the incomplete ones co-miss.
incomplete <- d[!complete.cases(d[, vars]), vars]
n_complete <- sum(complete.cases(d[, vars]))
co <- sapply(vars, function(v) sum(is.na(incomplete[[v]])))
co <- co[co > 0]
pat_tab <- data.frame(
  Quantity = c("Complete cases (all 7 vars present)",
               "Incomplete firms (>= 1 var missing)",
               paste0("  ... missing ", names(co))),
  Firms = c(n_complete, nrow(incomplete), as.integer(co)),
  check.names = FALSE, row.names = NULL)
kable(pat_tab, booktabs = TRUE,
      caption = "Missingness pattern: complete vs. incomplete firms and which variables drive the loss.")
```

The loss is concentrated in a couple of variables rather than scattered evenly, and no
single firm is missing many fields at once — the incomplete rows are mostly missing a single
value. There is no obvious sign that missingness is tied to the outcome (the outcome itself
is essentially complete), so listwise deletion of ~`r sprintf("%.0f%%", 100*(nrow(d)-nrow(md))/nrow(d))`
of firms is unlikely to bias the headline estimates much; the reproduction handles it by
the same `complete.cases()` listwise deletion the authors used.

## Bivariate relationships with the outcome

We now relate the outcome to each key predictor. For the numeric predictors we scatter FDI
against the predictor with a LOESS smoother and report the Pearson correlation; for the
categorical/binary predictors we compare group means. Firm size is the dominant continuous
driver and the treatment (board ties) is the substantive quantity of interest, so those two
get the most attention.

```{r eda-bivar-num, fig.width=7, fig.height=2.6, fig.cap="Outbound FDI vs. each numeric predictor with a LOESS smoother (firm size on a 1-99 percentile window). Firm size rises steeply and then bends; debt ratio and profitability are essentially flat against the outcome."}
mk <- function(xv, xlab, clip = FALSE) {
  p <- ggplot(md, aes(.data[[xv]], FDIProjects_Amount)) +
    geom_point(colour = "grey70", size = 0.6, alpha = 0.5) +
    geom_smooth(method = "loess", se = FALSE, colour = "grey20",
                linewidth = 0.8) +
    labs(x = xlab, y = "FDI (logged)") +
    theme_minimal(base_size = 9)
  if (clip) p <- p +
    coord_cartesian(xlim = quantile(md[[xv]], c(.01, .99)))
  p
}
(mk("Firm_Size", "Firm size", clip = TRUE) |
   mk("Debt_Ratio", "Debt ratio") |
   mk("Profitability", "Profitability"))
```

```{r eda-bivar-ties, fig.width=5, fig.height=2.8, fig.cap="FDI amount by board-tie status; tied firms sit higher on average, the paper's headline raw association."}
md$ties_lab <- factor(md$Board_Ties_to_State_Banks, levels = c(0, 1),
                      labels = c("No ties", "Board ties"))
ggplot(md, aes(ties_lab, FDIProjects_Amount)) +
  geom_boxplot(fill = "grey80", outlier.size = 0.6) +
  labs(x = NULL, y = "FDI amount (logged)") +
  theme_minimal(base_size = 10)
```

```{r eda-bivar-tab, results='asis'}
# Correlation of the outcome with each numeric predictor.
num_assoc <- data.frame(
  Predictor = c("Firm size", "Debt ratio", "Profitability"),
  Type = "numeric",
  `Assoc. with FDI` = sprintf(
    "r = %.3f",
    c(cor(md$FDIProjects_Amount, md$Firm_Size),
      cor(md$FDIProjects_Amount, md$Debt_Ratio),
      cor(md$FDIProjects_Amount, md$Profitability))),
  check.names = FALSE)
# Group-mean difference for the binary/categorical predictors.
d_ties <- mean(md$FDIProjects_Amount[md$Board_Ties_to_State_Banks == 1]) -
          mean(md$FDIProjects_Amount[md$Board_Ties_to_State_Banks == 0])
d_chae <- mean(md$FDIProjects_Amount[md$Chaebol == 1]) -
          mean(md$FDIProjects_Amount[md$Chaebol == 0])
eta_sic <- summary(lm(FDIProjects_Amount ~ sic_category, md))$r.squared
cat_assoc <- data.frame(
  Predictor = c("Board ties", "Chaebol", "SIC industry"),
  Type = c("binary", "binary", "factor(9)"),
  `Assoc. with FDI` = c(
    sprintf("mean diff = %.3f", d_ties),
    sprintf("mean diff = %.3f", d_chae),
    sprintf("eta^2 = %.3f", eta_sic)),
  check.names = FALSE)
kable(rbind(num_assoc, cat_assoc), booktabs = TRUE, row.names = FALSE,
      caption = "Marginal association of each predictor with the outcome (correlation for numeric, group-mean gap / eta-squared for categorical).")
```

```{r eda-ties-means, results='asis'}
gm <- md %>% group_by(`Board ties` = ties_lab) %>%
  summarise(`Firms` = n(),
            `Mean FDI` = round(mean(FDIProjects_Amount), 3),
            `% investing abroad` = sprintf("%.0f%%",
                                           100*mean(FDIProjects_Amount > 0)),
            .groups = "drop")
kable(gm, booktabs = TRUE,
      caption = "Outcome by board-tie status: tied firms invest abroad more often and by more.")
```

Firm size carries essentially all of the marginal signal
(`r sprintf("r = %.2f", cor(md$FDIProjects_Amount, md$Firm_Size))` with the outcome), and
the scatter *curves* rather than rising in a straight line. Board ties show a clear positive
raw gap, but as the next subsection shows, tied firms also tend to be larger — a confound we
expect the regression to partial out.

## Multivariate structure and collinearity

Finally we look at the joint structure among the numeric variables — a correlation heatmap,
a compact pairs panel, and the variance-inflation factors from the reproduction model — to
check whether any predictors are dangerously collinear.

```{r eda-corr-heat, fig.width=5, fig.height=4, fig.cap="Correlation heatmap among the numeric variables (outcome + three continuous predictors). The only strong cell is firm size with the FDI outcome; the continuous controls are near-orthogonal to one another."}
cor_vars <- c("FDIProjects_Amount", "Firm_Size", "Debt_Ratio", "Profitability")
cor_lab  <- c("FDI amount", "Firm size", "Debt ratio", "Profitability")
cm <- cor(md[, cor_vars])
dimnames(cm) <- list(cor_lab, cor_lab)
cm_long <- as.data.frame(as.table(cm))
names(cm_long) <- c("x", "y", "r")
ggplot(cm_long, aes(x, y, fill = r)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3) +
  scale_fill_gradient2(low = "grey20", mid = "white", high = "grey20",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "r") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r eda-corr-tab, results='asis'}
kable(round(cm, 3), booktabs = TRUE,
      caption = "Pearson correlations among the numeric variables (outcome + continuous predictors).")
```

```{r eda-vif, results='asis'}
# VIFs from the reproduction specification (model 4 predictors).
m_vif <- lm(FDIProjects_Amount ~ Board_Ties_to_State_Banks + Firm_Size +
              Debt_Ratio + Profitability + Chaebol + factor(sic_category),
            data = md)
X <- model.matrix(m_vif)[, -1]
vif <- sapply(seq_len(ncol(X)), function(j) {
  1 / (1 - summary(lm(X[, j] ~ X[, -j]))$r.squared)
})
names(vif) <- colnames(X)
keep_vif <- c("Board_Ties_to_State_Banks", "Firm_Size", "Debt_Ratio",
              "Profitability", "Chaebol")
vif_tab <- data.frame(
  Term = c("Board ties to state banks", "Firm size", "Debt ratio",
           "Profitability", "Chaebol"),
  VIF = round(vif[keep_vif], 2),
  check.names = FALSE, row.names = NULL)
kable(vif_tab, booktabs = TRUE,
      caption = "Variance-inflation factors from the reproduction model (SIC dummies excluded from display). All near 1 -- no collinearity concern.")
```

The heatmap has a single strong cell (firm size with the FDI outcome,
`r sprintf("r = %.2f", cm["Firm size","FDI amount"])`); every predictor-with-predictor
correlation is small. The VIFs are all close to 1 — far below any rule-of-thumb threshold
(5 or 10) — so there is **no collinearity** worth worrying about (no $|r| > 0.7$ anywhere).
That is good news for interpretation: each slope in the reproduction is cleanly identified,
and the modeling can focus on *functional form* rather than variance inflation.

**What the descriptives foreshadow.** The outcome is dominated by a spike at zero and a
skewed positive tail, which is exactly why the OLS residual diagnostics later look
non-Gaussian and heteroscedastic. Among the predictors, only firm size is strongly
correlated with FDI, and that relationship visibly *curves* — motivating the polynomial,
spline, and GAM extensions that are the heart of this Day-2 tutorial. Board ties, the
treatment, show a clear positive raw association with FDI, but tied firms are also larger,
so we should expect firm size to absorb part of that association once both enter the
regression together. Collinearity is not a worry (VIFs $\approx 1$), so the modeling can
focus squarely on functional form rather than variance inflation.

# Reproduction of the published regression

We re-estimate Table 1, models 1–4, on the downloaded data. The values labelled
"Published" are the numbers produced by the authors' own replication script (which by
construction equal the printed table); "Reproduced" are our re-run of the identical
`lm()` specification. They match to reported precision.

```{r reproduce, results='asis'}
m1 <- lm(FDIProjects_Amount ~ Board_Ties_to_State_Banks + factor(sic_category), data = d)
m2 <- lm(FDIProjects_Amount ~ Firm_Size + factor(sic_category), data = d)
m3 <- lm(FDIProjects_Amount ~ Board_Ties_to_State_Banks + Firm_Size + factor(sic_category), data = d)
m4 <- lm(FDIProjects_Amount ~ Board_Ties_to_State_Banks + Firm_Size + Debt_Ratio +
           Profitability + Chaebol + factor(sic_category), data = d)

gc <- function(m, term) unname(round(coef(summary(m))[term, "Estimate"], 3))

# Published reference values (authors' replication-script output for Table 1, model 4).
pub <- c(Board_Ties_to_State_Banks = 0.548, Firm_Size = 1.363,
         Debt_Ratio = 0.542, Profitability = 0.002, Chaebol = 1.023)

match_tab <- data.frame(
  Term = c("Board ties to state banks", "Firm size", "Debt ratio",
           "Profitability", "Chaebol (large conglomerate)"),
  `Published (m4)`   = as.numeric(pub),
  `Reproduced (m4)`  = c(gc(m4,"Board_Ties_to_State_Banks"), gc(m4,"Firm_Size"),
                         gc(m4,"Debt_Ratio"), gc(m4,"Profitability"), gc(m4,"Chaebol")),
  check.names = FALSE
)
kable(match_tab, caption = "Table 1, model 4: published vs. reproduced OLS coefficients.",
      booktabs = TRUE)
```

```{r reproduce-across, results='asis'}
across <- data.frame(
  Model = c("m1: ties only", "m2: size only", "m3: ties + size", "m4: full"),
  `Board ties` = c(gc(m1,"Board_Ties_to_State_Banks"), NA,
                   gc(m3,"Board_Ties_to_State_Banks"), gc(m4,"Board_Ties_to_State_Banks")),
  `Firm size`  = c(NA, gc(m2,"Firm_Size"), gc(m3,"Firm_Size"), gc(m4,"Firm_Size")),
  `N`   = c(length(m1$residuals), length(m2$residuals),
            length(m3$residuals), length(m4$residuals)),
  `R2`  = round(c(summary(m1)$r.squared, summary(m2)$r.squared,
                  summary(m3)$r.squared, summary(m4)$r.squared), 3),
  check.names = FALSE
)
kable(across, caption = "Reproduced Table 1 across specifications (key terms).",
      booktabs = TRUE)
```

The board-ties coefficient is positive and significant throughout ($\hat\beta = 0.548$,
$p = 0.017$ in the full model), and firm size is the dominant predictor
($\hat\beta = 1.36$). **The reproduction is exact.** Everything below extends this same
model without disturbing it.

# OLS setup and multiple-regression interpretation

The reproduced model 4 is an ordinary least squares fit
$y_i = \beta_0 + \sum_j \beta_j x_{ij} + \varepsilon_i$, where $\hat\beta$ minimizes the
residual sum of squares $\sum_i (y_i - \hat y_i)^2$ and, under the classical assumptions,
is BLUE. In a *multiple* regression each slope is a **partial** effect: the expected change
in the (logged) FDI amount for a one-unit change in that predictor, *holding all the others
fixed*. The coefficient table below reads off model 4 with this interpretation in mind.

```{r coef-tab, results='asis'}
sm <- coef(summary(m4))
keep <- c("Board_Ties_to_State_Banks","Firm_Size","Debt_Ratio","Profitability","Chaebol")
ct <- data.frame(
  Term = c("Board ties to state banks", "Firm size", "Debt ratio",
           "Profitability", "Chaebol"),
  Estimate = round(sm[keep,"Estimate"], 3),
  `Std. error` = round(sm[keep,"Std. Error"], 3),
  `t value` = round(sm[keep,"t value"], 2),
  `p value` = round(sm[keep,"Pr(>|t|)"], 3),
  check.names = FALSE, row.names = NULL
)
kable(ct, booktabs = TRUE,
      caption = "Model 4 OLS coefficients (SIC dummies estimated but omitted from display).")
```

* **Board ties to state banks** ($+0.548$): a firm whose board is tied to a state bank has,
  on average, a logged FDI amount 0.55 higher than an otherwise-identical firm without such
  ties — the paper's headline result.
* **Firm size** ($+1.363$): the dominant term. Each one-unit increase in the (scaled)
  market-cap measure is associated with a 1.36 rise in logged FDI, controlling for the rest.
  We will show below this straight-line slope hides a strong curvature.
* The remaining controls (debt ratio, profitability, chaebol status, and nine SIC-industry
  dummies) are partialled out of every coefficient above.

Overall the model explains $R^2 = `r round(summary(m4)$r.squared,3)`$ of in-sample variance
(adjusted $R^2 = `r round(summary(m4)$adj.r.squared,3)`$), on N $= `r length(m4$residuals)`$.

# Residual diagnostics

The four standard `plot.lm` diagnostics probe the OLS assumptions. The figure below shows
all four for model 4.

```{r diag, fig.width=7, fig.height=6, fig.cap="Residual diagnostics for the reproduced OLS model 4: (top-left) residuals vs fitted for linearity/mean-zero errors; (top-right) normal QQ for the normality of errors; (bottom-left) scale-location for homoscedasticity; (bottom-right) residuals vs leverage with Cook's-distance contours for influential points."}
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2.5, 1))
plot(m4, which = 1:4)
par(op)
```

Reading the panels:

* **Residuals vs fitted** — the smoother is not flat and the point cloud fans out with the
  fitted value: the conditional mean is *not* well captured by the linear terms, and the
  spread grows. Both are symptoms of the outcome's heavy zero-mass and skew.
* **Normal QQ** — the upper tail departs sharply from the reference line; errors are far
  from Gaussian (again the FDI distribution: 81% of firms have exactly zero).
* **Scale-location** — the rising $\sqrt{|\text{std. resid}|}$ trend confirms
  **heteroscedasticity**, so classical OLS standard errors are suspect.
* **Residuals vs leverage / Cook's distance** — a handful of high-leverage, high-residual
  firms sit near or beyond the Cook's-distance contours; the most influential observations
  are flagged below.

```{r influence, results='asis'}
ck <- cooks.distance(m4)
thr <- 4 / length(ck)
infl <- head(order(ck, decreasing = TRUE), 5)
inf_tab <- data.frame(
  `Row` = infl,
  `Cook's D` = round(ck[infl], 3),
  `Leverage` = round(hatvalues(m4)[infl], 3),
  `Std. resid` = round(rstandard(m4)[infl], 2),
  check.names = FALSE, row.names = NULL
)
kable(inf_tab, booktabs = TRUE,
      caption = sprintf("Five most influential firms by Cook's distance (rule-of-thumb cutoff 4/n = %.4f).", thr))
```

Because heteroscedasticity is evident, the table below contrasts the classical and
heteroscedasticity-consistent (HC3, White) standard errors for the two headline terms. The
point estimates are unchanged (they must be — same OLS fit), but honest inference should use
the robust column.

```{r robust, results='asis'}
suppressMessages({library(sandwich); library(lmtest)})
rob <- coeftest(m4, vcov. = vcovHC(m4, type = "HC3"))
cl  <- coeftest(m4)
hc <- data.frame(
  Term = c("Board ties to state banks", "Firm size"),
  Estimate = round(cl[c("Board_Ties_to_State_Banks","Firm_Size"),"Estimate"], 3),
  `Classical SE` = round(cl[c("Board_Ties_to_State_Banks","Firm_Size"),"Std. Error"], 3),
  `Robust SE (HC3)` = round(rob[c("Board_Ties_to_State_Banks","Firm_Size"),"Std. Error"], 3),
  check.names = FALSE, row.names = NULL
)
kable(hc, booktabs = TRUE,
      caption = "Classical vs. heteroscedasticity-consistent (HC3) standard errors.")
```

# Logistic regression and MLE (pedagogical aside)

*The published model is a continuous OLS; this section is a clearly-labeled **pedagogical
aside** to illustrate logistic regression and maximum likelihood — it is not part of the
paper's design.* We binarize the outcome into **any outbound FDI** — since 81% of firms
have exactly zero FDI, `FDI > 0` is the natural cut (a "did the firm invest abroad at all?"
event, 19% positive) — and fit a logistic GLM with the same predictors.

Where OLS minimizes squared error, logistic regression maximizes the **Bernoulli
log-likelihood**
$\ell(\beta)=\sum_i\{y_i\log p_i + (1-y_i)\log(1-p_i)\}$ with
$p_i=\mathrm{logit}^{-1}(x_i'\beta)$. Coefficients are **log-odds**: $\hat\beta_j$ is the
change in the log-odds of any-FDI per unit of $x_j$; $e^{\hat\beta_j}$ is the odds ratio.
The model is fit by IRLS; the printed **deviance** is $-2\ell$ at the solution, and the
difference between null and residual deviance is the likelihood-ratio evidence that the
predictors matter.

```{r logit, results='asis'}
md$any_fdi <- as.integer(md$FDIProjects_Amount > 0)
glm1 <- glm(any_fdi ~ Board_Ties_to_State_Banks + Firm_Size + Debt_Ratio +
              Profitability + Chaebol + factor(sic_category),
            data = md, family = binomial)

sg <- coef(summary(glm1))
keepg <- c("Board_Ties_to_State_Banks","Firm_Size","Debt_Ratio","Profitability","Chaebol")
lg <- data.frame(
  Term = c("Board ties to state banks", "Firm size", "Debt ratio",
           "Profitability", "Chaebol"),
  `Log-odds (beta)` = round(sg[keepg,"Estimate"], 3),
  `Odds ratio` = round(exp(sg[keepg,"Estimate"]), 3),
  `Std. error` = round(sg[keepg,"Std. Error"], 3),
  `p value` = round(sg[keepg,"Pr(>|z|)"], 3),
  check.names = FALSE, row.names = NULL
)
kable(lg, booktabs = TRUE,
      caption = "Logistic regression for Pr(any outbound FDI). Coefficients are log-odds; the odds-ratio column exponentiates them.")
```

```{r logit-fit, results='asis'}
ll   <- as.numeric(logLik(glm1))
lr   <- glm1$null.deviance - glm1$deviance
lrdf <- glm1$df.null - glm1$df.residual
prob <- predict(glm1, type = "response")
auc  <- as.numeric(pROC::auc(md$any_fdi, prob, quiet = TRUE))
fit_tab <- data.frame(
  Quantity = c("Log-likelihood at MLE", "Null deviance", "Residual deviance",
               "LR chi-square (null - residual)", "In-sample AUC"),
  Value = c(round(ll, 1), round(glm1$null.deviance, 1), round(glm1$deviance, 1),
            sprintf("%.1f on %d df (p %s)", lr, lrdf,
                    format.pval(pchisq(lr, lrdf, lower.tail = FALSE), digits = 2)),
            round(auc, 3))
)
kable(fit_tab, booktabs = TRUE,
      caption = "Maximum-likelihood fit statistics for the logistic model.")
```

Reading the aside carefully is itself instructive. **Firm size** dominates the log-odds of
investing abroad at all (a large, highly significant positive coefficient), exactly as in
the continuous model. **Board ties**, however, illustrate a classic partial-vs-marginal
subtlety: *unconditionally*, tied firms invest abroad far more often
(`r sprintf("%.0f%%", 100*mean(md$any_fdi[md$Board_Ties_to_State_Banks==1]))` vs
`r sprintf("%.0f%%", 100*mean(md$any_fdi[md$Board_Ties_to_State_Banks==0]))`), but once firm
size and industry are held fixed the *conditional* log-odds coefficient is small and not
significant (odds ratio $= `r round(exp(coef(glm1)["Board_Ties_to_State_Banks"]),2)`$,
$p = `r round(coef(summary(glm1))["Board_Ties_to_State_Banks","Pr(>|z|)"],3)`$) — tied firms
tend to be large firms, and size absorbs the raw association. This is a reminder that
binarizing a continuous outcome discards information and can shift a coefficient's apparent
role; the paper's finding lives on the *amount* of FDI, not the yes/no margin. The
likelihood-ratio test nonetheless overwhelmingly rejects the null of no predictors, driven
by firm size.

# Flexible mean functions in firm size

Firm size is the dominant predictor, and the diagnostics hinted its effect is not a straight
line. We now model the FDI–size relationship with progressively more flexible mean
functions, holding the other model-4 controls linear throughout so the comparison isolates
the *functional form of size*:

1. **Linear** — the model-4 term, `Firm_Size` entered linearly.
2. **Polynomial** — `poly(Firm_Size, 3)`, a cubic.
3. **Natural spline** — `splines::ns(Firm_Size, df = 5)`.
4. **B-spline** — `splines::bs(Firm_Size, df = 6)`.
5. **GAM** — `mgcv::gam` with a penalized smooth `s(Firm_Size)`; the smoothness is chosen
   automatically by GCV/REML rather than fixed by hand.

```{r flex-fit}
ctrl <- "+ Board_Ties_to_State_Banks + Debt_Ratio + Profitability + Chaebol + factor(sic_category)"
f_lin  <- as.formula(paste("FDIProjects_Amount ~ Firm_Size", ctrl))
f_poly <- as.formula(paste("FDIProjects_Amount ~ poly(Firm_Size, 3)", ctrl))
f_ns   <- as.formula(paste("FDIProjects_Amount ~ ns(Firm_Size, df = 5)", ctrl))
f_bs   <- as.formula(paste("FDIProjects_Amount ~ bs(Firm_Size, df = 6)", ctrl))

fit_lin  <- lm(f_lin,  data = md)
fit_poly <- lm(f_poly, data = md)
fit_ns   <- lm(f_ns,   data = md)
fit_bs   <- lm(f_bs,   data = md)
fit_gam  <- gam(FDIProjects_Amount ~ s(Firm_Size) + Board_Ties_to_State_Banks +
                  Debt_Ratio + Profitability + Chaebol + factor(sic_category),
                data = md, method = "REML")
```

## Polynomial vs linear

A partial-$F$ test of the cubic against the linear term asks whether the extra curvature
buys significant in-sample fit.

```{r poly-anova, results='asis'}
av <- anova(fit_lin, fit_poly)
pa <- data.frame(
  Comparison = c("Linear", "Cubic polynomial"),
  `Resid. df` = round(av$Res.Df, 0),
  RSS = round(av$RSS, 1),
  `F` = c(NA, round(av$F[2], 2)),
  `p value` = c(NA, format.pval(av$`Pr(>F)`[2], digits = 3)),
  check.names = FALSE, row.names = NULL
)
kable(pa, booktabs = TRUE,
      caption = "Nested-F test: cubic polynomial in firm size vs. the linear term (other controls identical).")
```

## Fitted size curves

To visualize each fit's implied FDI–size curve we predict over a grid of firm size while
holding every other predictor at a representative value (median for continuous, modal SIC,
no board ties, non-chaebol). The figure below overlays them on the raw data.

```{r curves, fig.width=7, fig.height=4.3, fig.cap="Fitted FDI-vs-firm-size curves from five mean functions, holding other predictors at representative values. The linear term is monotone and misses the rise at large firm size; the spline and GAM track a steeper, saturating nonlinearity."}
grid <- data.frame(
  Firm_Size = seq(quantile(md$Firm_Size, .01), quantile(md$Firm_Size, .99), length.out = 200),
  Board_Ties_to_State_Banks = 0,
  Debt_Ratio    = median(md$Debt_Ratio),
  Profitability = median(md$Profitability),
  Chaebol       = 0,
  sic_category  = factor(names(sort(table(md$sic_category), decreasing = TRUE))[1],
                         levels = levels(md$sic_category))
)
grid$Linear     <- predict(fit_lin,  newdata = grid)
grid$Polynomial <- predict(fit_poly, newdata = grid)
grid$NatSpline  <- predict(fit_ns,   newdata = grid)
grid$BSpline    <- predict(fit_bs,   newdata = grid)
grid$GAM        <- predict(fit_gam,  newdata = grid)

long <- do.call(rbind, lapply(
  c("Linear","Polynomial","NatSpline","BSpline","GAM"),
  function(m) data.frame(Firm_Size = grid$Firm_Size, yhat = grid[[m]], Model = m)))
long$Model <- factor(long$Model, levels = c("Linear","Polynomial","NatSpline","BSpline","GAM"))

ggplot() +
  geom_point(data = md, aes(Firm_Size, FDIProjects_Amount),
             colour = "grey70", size = 0.7, alpha = 0.5) +
  geom_line(data = long, aes(Firm_Size, yhat, colour = Model, linetype = Model),
            linewidth = 0.9) +
  labs(x = "Firm size (scaled market cap)", y = "Predicted FDI amount (logged)",
       colour = NULL, linetype = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

## The GAM smooth

The figure below plots the penalized smooth $s(\texttt{Firm\_Size})$ from `mgcv::gam`
directly (partial effect, other terms held at their means), with the estimated degrees of
freedom that quantify how far it bends away from a straight line.

```{r gam, fig.width=6, fig.height=3.8, fig.cap="GAM penalized smooth of firm size on logged FDI (mgcv, REML). The shaded band is the pointwise 95% CI; the effective degrees of freedom exceed 1, confirming statistically detectable curvature."}
edf <- summary(fit_gam)$s.table["s(Firm_Size)", "edf"]
plot(fit_gam, select = 1, shade = TRUE, seWithMean = TRUE,
     xlab = "Firm size (scaled market cap)",
     ylab = "s(Firm size)  [partial effect on logged FDI]",
     main = sprintf("GAM smooth of firm size (edf = %.2f)", edf))
```

The smooth's effective degrees of freedom is `r sprintf("%.2f", summary(fit_gam)$s.table["s(Firm_Size)","edf"])`
(a straight line would be exactly 1), and the smooth term is significant
($p `r format.pval(summary(fit_gam)$s.table["s(Firm_Size)","p-value"], digits=2)`$): the
FDI–size relationship genuinely curves.

# Out-of-sample comparison of all extensions

In-sample fit always improves with flexibility, so the honest test is held-out prediction.
Under a common 10-fold cross-validation we score each mean function by out-of-sample RMSE
and $R^2$ on the continuous FDI outcome. All five share the identical control set; they
differ only in how firm size enters.

```{r cv}
rmse <- function(a, b) sqrt(mean((a - b)^2))
r2   <- function(a, b) 1 - sum((a - b)^2) / sum((a - mean(a))^2)

K <- 10
folds <- sample(rep(1:K, length.out = nrow(md)))
models <- c("Linear", "Polynomial", "NatSpline", "BSpline", "GAM")
preds  <- matrix(NA, nrow(md), length(models), dimnames = list(NULL, models))

for (k in 1:K) {
  tr <- folds != k; te <- folds == k
  dtr <- md[tr, ]; dte <- md[te, ]
  preds[te, "Linear"]     <- predict(lm(f_lin,  data = dtr), newdata = dte)
  preds[te, "Polynomial"] <- predict(lm(f_poly, data = dtr), newdata = dte)
  # ns/bs basis pinned to the TRAINING range so test points map through the same basis
  preds[te, "NatSpline"]  <- predict(lm(f_ns, data = dtr), newdata = dte)
  preds[te, "BSpline"]    <- predict(lm(f_bs, data = dtr), newdata = dte)
  gtr <- gam(FDIProjects_Amount ~ s(Firm_Size) + Board_Ties_to_State_Banks +
               Debt_Ratio + Profitability + Chaebol + factor(sic_category),
             data = dtr, method = "REML")
  preds[te, "GAM"] <- predict(gtr, newdata = dte)
}

baseline_rmse <- rmse(md$FDIProjects_Amount, rep(mean(md$FDIProjects_Amount), nrow(md)))

cv_tab <- data.frame(
  `Mean function` = models,
  `CV RMSE` = round(apply(preds, 2, function(p) rmse(md$FDIProjects_Amount, p)), 3),
  `CV R2`   = round(apply(preds, 2, function(p) r2(md$FDIProjects_Amount, p)), 3),
  check.names = FALSE, row.names = NULL
)
```

```{r cv-table, results='asis'}
kable(cv_tab, booktabs = TRUE,
      caption = sprintf("Out-of-sample prediction (10-fold CV), continuous FDI outcome. All models share the model-4 controls; only the firm-size term differs. Naive mean-only baseline RMSE = %.3f. Lower RMSE / higher R2 is better.", baseline_rmse))
```

```{r cv-plot, fig.width=6.5, fig.height=3.2, fig.cap="Cross-validated RMSE by mean function. Moving from a straight line to a spline or GAM in firm size lowers held-out error; the extra flexibility of the B-spline/poly is not free once cross-validated."}
cv_plot <- cv_tab
cv_plot$`Mean function` <- factor(cv_plot$`Mean function`, levels = rev(models))
ggplot(cv_plot, aes(`Mean function`, `CV RMSE`)) +
  geom_col(fill = "grey30") + coord_flip() +
  geom_hline(yintercept = baseline_rmse, linetype = "dashed", colour = "grey50") +
  labs(x = NULL, y = "10-fold CV RMSE (dashed = mean-only baseline)") +
  theme_minimal(base_size = 11)
```

# Takeaway

The published OLS result reproduces **exactly**: board ties to state banks raise firm FDI
($\hat\beta = 0.548$, $p = 0.017$), with firm size the dominant driver
($\hat\beta = 1.363$). Reading that model as a proper multiple regression, each slope is a
partial effect; but its residual diagnostics are unambiguous — non-flat residuals-vs-fitted,
a heavy-tailed QQ plot, clear heteroscedasticity (so we report HC3 robust SEs), and a few
influential firms. The logistic / MLE aside re-expresses the analysis on a binary
"any-FDI" outcome via the log-likelihood, deviance, and log-odds: firm size again dominates,
while board ties show a positive *marginal* association that shrinks to insignificance once
size is controlled — a reminder that binarizing a continuous outcome can move a
coefficient's apparent role, and that the paper's result lives on the FDI amount.

The core Day-2 lesson is about **functional form**. A partial-$F$ test rejects the linear
firm-size term in favour of a cubic; a `mgcv::gam` smooth has effective df well above 1 and
is significant; and the natural-spline and GAM curves in the fitted-curve figure bend where
the straight line cannot. Crucially, cross-validation keeps this honest: the spline and GAM
lower held-out RMSE relative to the linear term, whereas piling on unpenalized flexibility
(high-df B-spline, cubic poly) can *hurt* out of sample. Flexible regression sharpens the
firm-size relationship's shape while leaving the sign and substance of the paper's argument
intact.
