---
title: "Gerber, Green & Larimer (2008), *Social Pressure and Voter Turnout: Evidence from a Large-Scale Field Experiment* --- SH Day 1: Model Evaluation"
author: "Machine-Learning Workshop teaching example"
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: true
    toc: false
header-includes:
  - \setlength{\emergencystretch}{3em}
  - \usepackage{fvextra}
  - \AtBeginDocument{\DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}}
  - \AtBeginDocument{\DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}}
  - \fvset{breaklines=true}
---

```{r setup, include=FALSE}
options(width = 80)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 4)
suppressPackageStartupMessages({
  library(ggplot2)
  library(sandwich)
  library(lmtest)
  library(pROC)
})
set.seed(42)
```

# The study

**Citation.** Alan S. Gerber, Donald P. Green, and Christopher W. Larimer (2008).
"Social Pressure and Voter Turnout: Evidence from a Large-Scale Field Experiment."
*American Political Science Review* 102(1): 33--48.

**The design.** This is one of the most famous field experiments in modern social
science. Ahead of the **August 2006 primary election in Michigan**, the authors
randomly assigned **344,084 registered voters in 180,002 households** --- with
randomization at the *household* level --- to a no-mail **Control** condition or to
one of four mailings sent shortly before the election, each turning up the dial of
social pressure:

1. **Civic Duty.** A mailing that simply appeals to the recipient's sense of civic
   duty ("Remember your rights and responsibilities as a citizen. Remember to
   vote."). This is the baseline mobilization message.
2. **Hawthorne.** Adds mild surveillance: recipients are told they are part of a
   study and that "You are being studied!" --- their turnout will be examined by
   researchers via public records. No information is disclosed to anyone else.
3. **Self.** Adds *self*-monitoring: the mailing lists the recipient's own
   household's past turnout (who voted in 2004, who did not) and promises an
   updated mailing after the election showing whether each member voted.
4. **Neighbors.** The maximal-pressure treatment: the mailing lists the recent
   turnout of the recipient's household *and of their neighbors*, and promises a
   follow-up mailing telling the whole neighborhood who voted.

Because voting in the United States is a matter of public record, the threat to
publicize turnout is credible, and the design lets the authors measure --- with
experimental precision --- how much *social pressure*, over and above a civic-duty
appeal, moves people to the polls.

**Why it is a landmark.** The result was enormous by the standards of the
get-out-the-vote literature: the Neighbors mailing raised turnout by **8.1
percentage points** off a control base of 29.7% --- an effect comparable to
face-to-face canvassing, achieved with a single piece of mail costing pennies. The
paper has shaped both the science of voter mobilization (spawning a large
literature on social-pressure interventions) and its practice (variants of the
Neighbors letter have been used, sometimes controversially, in real campaigns).
It is also a model of design-based inference: with random assignment, a simple
difference in means is an unbiased causal estimate.

**The published quantities we reproduce.** Two headline results:

- **Table 2**: the percentage voting in the August 2006 primary within each of the
  five experimental groups (Control 29.7%; Civic Duty 31.5%; Hawthorne 32.2%;
  Self 34.5%; Neighbors 37.8%).
- **Table 3, model (a)**: OLS of the turnout indicator on the four treatment
  dummies (Control as reference), with robust standard errors **clustered on
  household** --- Civic Duty .018, Hawthorne .026, Self .049, Neighbors .081, all
  with clustered SEs of .003 and $p < .001$; $N$ = 344,084 individuals.

```{r load-data, cache=TRUE}
ggl <- read.csv(paste0(
  "https://raw.githubusercontent.com/bdesmarais/",
  "intro-ml-statistical-horizons-4day/main/",
  "tutorials/day1_intro_and_evaluation/",
  "gerber_green_larimer_turnout/data/",
  "ggl_individual.csv"))
dim(ggl)
```

# Data and codebook

The unit of analysis is the **individual registered voter**, nested in households.
The file has `r format(nrow(ggl), big.mark=",")` rows and `r ncol(ggl)` columns.

```{r codebook}
codebook <- data.frame(
  Variable = c("sex", "yob", "g2000, g2002, g2004",
               "p2000, p2002, p2004", "treatment", "cluster",
               "voted", "hh_id", "hh_size", "numberofnames",
               "p2004_mean", "g2004_mean", "X", "origorder"),
  Description = c(
    "male / female",
    "year of birth",
    "voted in the 2000/2002/2004 GENERAL election (yes/no)",
    "voted in the 2000/2002/2004 PRIMARY election (yes/no)",
    "experimental condition (Control + 4 mailings)",
    "grouping identifier from the assignment procedure",
    "OUTCOME: voted in the August 2006 primary (0/1)",
    "household identifier (randomization unit)",
    "number of registered voters in the household",
    "number of names shown on the Neighbors mailing",
    "household mean of p2004 (pre-treatment)",
    "household mean of g2004 (pre-treatment)",
    "empty unnamed column in the source file (dropped)",
    "original row order in the replication file"))
knitr::kable(codebook, caption = "Codebook for ggl_individual.csv.")
```

**Cleaning.** Three quirks of the source file need attention before any analysis.
First, the `treatment` strings carry a **leading space** (e.g. `" Control"`).
Second, the vote-history columns mix capitalization and can carry stray whitespace
(`p2004` is recorded as `"Yes"`/`"No"` while the others are lowercase
`"yes"`/`"no"`); we inspect the raw values, then `trimws()` and lowercase
everything and convert to 0/1 indicators. Third, there is an empty unnamed column
(read in as `X`), which we drop.

```{r clean}
# Inspect the raw coding quirks before touching anything.
table(ggl$treatment)          # note the leading space in every level
table(ggl$p2004)              # 'Yes'/'No' capitalization differs from the rest

ggl$X <- NULL                                     # drop the empty column
ggl$treatment <- trimws(ggl$treatment)            # strip the leading space
hist_vars <- c("g2000", "g2002", "g2004", "p2000", "p2002", "p2004")
for (v in hist_vars)                              # 'yes'/'Yes ' etc. -> 0/1
  ggl[[v]] <- as.integer(trimws(tolower(ggl[[v]])) == "yes")
ggl$treatment <- relevel(factor(ggl$treatment), ref = "Control")
ggl$age    <- 2006 - ggl$yob                      # age at the 2006 primary
ggl$female <- as.integer(ggl$sex == "female")
stopifnot(!anyNA(ggl[c("voted", "treatment", "age", hist_vars)]))
```

One vote-history variable deserves a flag: **`g2004` equals 1 for every single
row** (`r format(sum(ggl$g2004), big.mark=",")` of
`r format(nrow(ggl), big.mark=",")`). That is by design --- the authors restricted
the experimental universe to registered voters who had voted in the 2004 general
election. A constant carries no information, so `g2004` is documented here but
excluded from every model below; the five informative history indicators are
`g2000`, `g2002`, `p2000`, `p2002`, `p2004`.

# Descriptive analysis

Before reproducing anything we get to know the data thoroughly. With an
experiment the stakes of EDA are slightly different than with observational data:
beyond documenting distributions, the key descriptive product is the **balance
check** --- randomization should make the five groups look alike on everything
determined *before* treatment, and we verify that it does.

## Experimental group sizes

```{r group-sizes}
grp_n <- as.data.frame(table(Condition = ggl$treatment),
                       responseName = "Individuals")
grp_hh <- tapply(ggl$hh_id, ggl$treatment,
                 function(x) length(unique(x)))
grp_n$Households <- as.integer(grp_hh[as.character(grp_n$Condition)])
grp_n$Share <- round(100 * grp_n$Individuals / nrow(ggl), 1)
knitr::kable(grp_n, caption =
  "Experimental group sizes: individuals, households, and percent of sample.")
```

The design is unbalanced on purpose: about 56% of the sample is Control, and each
mailing goes to roughly 38,000 individuals in about 20,000 households. In total
there are `r format(length(unique(ggl$hh_id)), big.mark=",")` households ---
matching the paper's 180,002.

## Turnout by experimental condition

The single most important descriptive picture: the share voting in the August
2006 primary in each condition, with 95% confidence intervals. (With group sizes
in the tens of thousands the CIs are so tight they are barely visible --- one of
the luxuries of a 344,084-person experiment.)

```{r turnout-by-condition, fig.height=3.4, fig.cap="Turnout in the August 2006 primary by experimental condition, with 95 percent confidence intervals (barely visible at these sample sizes). Turnout climbs monotonically with the social-pressure dosage."}
tb <- aggregate(voted ~ treatment, ggl, mean)
tb$n  <- as.integer(table(ggl$treatment)[as.character(tb$treatment)])
tb$se <- sqrt(tb$voted * (1 - tb$voted) / tb$n)
tb$lo <- tb$voted - 1.96 * tb$se
tb$hi <- tb$voted + 1.96 * tb$se
tb$treatment <- factor(tb$treatment, levels = c(
  "Control", "Civic Duty", "Hawthorne", "Self", "Neighbors"))
ggplot(tb, aes(x = treatment, y = voted)) +
  geom_col(fill = "steelblue", width = 0.65) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.15) +
  geom_text(aes(label = sprintf("%.1f%%", 100 * voted)),
            vjust = -1.1, size = 3.2) +
  scale_y_continuous(labels = function(x) sprintf("%.0f%%", 100 * x),
                     limits = c(0, 0.45)) +
  labs(x = NULL, y = "Voted in Aug 2006 primary",
       title = "Turnout rises with the social-pressure dosage") +
  theme_minimal(base_size = 11)
```

The bar chart *is* the experiment's result: turnout climbs monotonically from
Control (29.7%) through Civic Duty, Hawthorne, and Self to Neighbors (37.8%),
exactly the ordering of social-pressure intensity.

## Vote history

The five informative history indicators record turnout in the 2000 and 2002
general elections and the 2000, 2002, and 2004 primaries.

```{r vote-history}
hist5 <- c("g2000", "g2002", "p2000", "p2002", "p2004")
vh <- data.frame(
  Election = c("General 2000", "General 2002", "Primary 2000",
               "Primary 2002", "Primary 2004"),
  Voted_pct = round(100 * sapply(ggl[hist5], mean), 1))
knitr::kable(vh, row.names = FALSE, caption =
  "Share voting in each prior election (g2004 = 100% by sample construction).")
```

```{r vote-history-plot, fig.height=3.0, fig.cap="Prior turnout by election. Generals draw far more of this sample than primaries; the 2004 general is 100 percent by construction and omitted."}
vh$Election <- factor(vh$Election, levels = vh$Election)
ggplot(vh, aes(x = Election, y = Voted_pct)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_text(aes(label = paste0(Voted_pct, "%")), vjust = -0.5, size = 3.2) +
  ylim(0, 100) +
  labs(x = NULL, y = "% voted", title = "Prior turnout by election") +
  theme_minimal(base_size = 11)
```

The familiar American pattern appears: general elections mobilize far more voters
than primaries (~70--85% vs. ~25--40% in this sample of proven 2004-general
voters). Primary turnout --- the behavior the experiment targets --- is the rare
event.

## Age

```{r age-dist, fig.height=3.2, fig.cap="Age distribution at the time of the 2006 primary. The registration-list sample skews middle-aged and older."}
age_counts <- as.data.frame(table(age = ggl$age), stringsAsFactors = FALSE)
age_counts$age <- as.integer(age_counts$age)
ggplot(age_counts, aes(x = age, y = Freq)) +
  geom_col(fill = "steelblue", width = 1) +
  labs(x = "Age in 2006", y = "Registered voters",
       title = "Age distribution") +
  theme_minimal(base_size = 11)
summary(ggl$age)
```

```{r age-turnout, fig.height=3.2, fig.cap="Turnout in the 2006 primary by year of age (ages with at least 200 voters). The relationship is strong and curved -- rising steeply from the 20s to the 60s, then flattening and turning down at advanced ages. A straight line in age would miss this shape; polynomials will pick it up in Section 5."}
age_to <- aggregate(voted ~ age, ggl, mean)
age_to$n <- as.integer(table(ggl$age)[as.character(age_to$age)])
age_to <- age_to[age_to$n >= 200, ]
ggplot(age_to, aes(x = age, y = voted)) +
  geom_point(colour = "steelblue", size = 1) +
  geom_smooth(method = "loess", se = FALSE, colour = "grey30",
              linewidth = 0.7) +
  scale_y_continuous(labels = function(x) sprintf("%.0f%%", 100 * x)) +
  labs(x = "Age in 2006", y = "Turnout, Aug 2006 primary",
       title = "Turnout by age is strongly nonlinear") +
  theme_minimal(base_size = 11)
```

Turnout rises steeply with age from the 20s into the 60s--70s and then bends back
down --- a strong, visibly **nonlinear** relationship. Keep this plot in mind: in
Section 5 the degree of a polynomial in age becomes our bias--variance knob.

## Turnout by past-vote count

Summing the five informative history indicators gives each voter a **past-vote
count** from 0 to 5 --- a one-number habit measure.

```{r past-count}
ggl$past_votes <- rowSums(ggl[hist5])
pc <- aggregate(voted ~ past_votes, ggl, mean)
pc$n <- as.integer(table(ggl$past_votes))
knitr::kable(data.frame(Past_votes = pc$past_votes,
                        N = format(pc$n, big.mark = ","),
                        Turnout_2006_pct = round(100 * pc$voted, 1)),
             caption = "2006 primary turnout by past-vote count (0-5 prior elections).")
```

```{r past-count-plot, fig.height=3.0, fig.cap="Turnout by past-vote count. Voting is habitual: those who voted in all five prior elections turn out at several times the rate of those who voted in none."}
ggplot(pc, aes(x = past_votes, y = voted)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_text(aes(label = sprintf("%.0f%%", 100 * voted)),
            vjust = -0.5, size = 3.2) +
  scale_y_continuous(labels = function(x) sprintf("%.0f%%", 100 * x),
                     limits = c(0, 0.75)) +
  scale_x_continuous(breaks = 0:5) +
  labs(x = "Prior elections voted in (of 5)", y = "Turnout, Aug 2006",
       title = "Voting is a habit") +
  theme_minimal(base_size = 11)
```

The gradient is enormous: turnout runs from `r round(100*pc$voted[1])`% among
those with no prior votes to `r round(100*pc$voted[6])`% among those who voted in
all five prior elections. **Past behavior dwarfs any treatment effect as a
predictor** --- the Neighbors mailing moves turnout 8 points; vote history spans
more than 40. This asymmetry is the engine of the predictive modeling in
Section 5.

## Household size

```{r hh-size}
hh_tab <- as.data.frame(table(hh_size = ggl$hh_size),
                        responseName = "Individuals")
hh_tab$Pct <- round(100 * hh_tab$Individuals / nrow(ggl), 2)
knitr::kable(hh_tab, caption =
  "Individuals by household size (number of registered voters at the address).")
```

Almost everyone lives in a one- or two-voter household --- a consequence of the
authors' screening of the registration list. Household structure matters for two
reasons: treatment was assigned at the household level (everyone in a household
gets the same condition), and household members' turnout is correlated. Both
facts will force us to **cluster** --- standard errors on households in the
reproduction, and train/test splits on households in the extension.

## Balance check: covariates across conditions

The whole point of randomization is that pre-treatment covariates should be
(near-)identical across conditions. We check age, sex, household size, and all
five history indicators.

```{r balance}
bal_vars <- c("age", "female", "hh_size", hist5)
bal <- sapply(split(ggl[bal_vars], ggl$treatment), colMeans)
bal <- bal[, c("Control", "Civic Duty", "Hawthorne", "Self", "Neighbors")]
knitr::kable(round(bal, 3), caption =
  "Balance check: pre-treatment covariate means by experimental condition.")
```

Balance holds to the third decimal place essentially everywhere: mean age varies
by a few tenths of a year across groups, and the history indicators by a few
tenths of a percentage point. This is what a well-executed randomization looks
like, and it is why the raw turnout differences in the bar chart above can be
read directly as causal effects.

## Co-occurrence of the vote-history indicators

Finally, how do the five history indicators relate to each other and to the
outcome? For binary variables the Pearson correlation is the $\phi$ coefficient.

```{r history-cor, fig.height=3.4, fig.cap="Correlation (phi) matrix of the five vote-history indicators and 2006 turnout. Same-type elections cohere most strongly; every history indicator is positively related to the 2006 outcome."}
cm <- cor(ggl[c(hist5, "voted")])
cl <- as.data.frame(as.table(cm))
names(cl) <- c("V1", "V2", "r")
ggplot(cl, aes(V1, V2, fill = r)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 2.8) +
  scale_fill_gradient2(low = "#b2182b", mid = "white", high = "#2166ac",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL,
       title = "Vote-history co-occurrence (phi coefficients)") +
  theme_minimal(base_size = 10)
```

All correlations are positive --- habitual voters vote everywhere --- but they are
moderate (`r sprintf("%.2f", min(cm[hist5, hist5][upper.tri(cm[hist5, hist5])]))`
to
`r sprintf("%.2f", max(cm[hist5, hist5][upper.tri(cm[hist5, hist5])]))`
among the history indicators), with primaries correlating more strongly with other
primaries than with generals. No pair is close to collinear, so all five can enter
a model together. The primary indicators (especially `p2004`,
$\phi = `r sprintf("%.2f", cm["p2004","voted"])`$ with the outcome) are the
strongest single predictors of 2006 primary turnout --- like predicts like.

## Reading the descriptives as a whole

Pulling the EDA together: the outcome is a **moderately rare binary event** (base
rate ~32%), the five conditions are cleanly balanced on everything pre-treatment
(so raw group differences are causal), and turnout climbs monotonically with
social-pressure dosage. On the predictive side, the strongest signals in the data
are **not** the experimental treatments: past turnout (a 40+-point gradient) and
age (a strong, curved profile) dominate the 2--8-point treatment contrasts. And
the household structure --- shared treatment and correlated behavior within
addresses --- warns us that both inference and evaluation must respect the
household clustering. All three observations script what follows.

# Reproduction of the published results

## Table 2: percentage voting by experimental group

The paper's Table 2 reports the share voting in the August 2006 primary in each
group. We compute it and set it beside the published values.

```{r table2}
pub_grp <- c("Control", "Civic Duty", "Hawthorne", "Self", "Neighbors")
pub_pct <- c(29.7, 31.5, 32.2, 34.5, 37.8)
pub_N   <- c(191243, 38218, 38204, 38218, 38201)

comp_pct <- round(100 * tapply(ggl$voted, ggl$treatment, mean)[pub_grp], 1)
comp_N   <- as.integer(table(ggl$treatment)[pub_grp])

t2 <- data.frame(Group = pub_grp,
                 Published_pct = pub_pct, Computed_pct = as.numeric(comp_pct),
                 Published_N = pub_N, Computed_N = comp_N)
knitr::kable(t2, row.names = FALSE, caption =
  "Table 2 reproduction: percentage voting and N by experimental group, published vs. computed.")

# --- Reproduction gate 1: Table 2 must match exactly at printed precision ---
stopifnot(max(abs(as.numeric(comp_pct) - pub_pct)) < 1e-9,
          identical(comp_N, as.integer(pub_N)))
```

Every percentage matches the published value exactly at the paper's printed
precision, and every group N matches exactly. The Neighbors--Control contrast is
`r sprintf("%.1f", comp_pct["Neighbors"] - comp_pct["Control"])` percentage
points --- the paper's headline 8.1-point effect, straight from the raw means.

## Table 3, model (a): OLS with household-clustered standard errors

Table 3(a) regresses the turnout indicator on the four treatment dummies (Control
is the reference). Because treatment was randomized by household and household
members' outcomes are correlated, the paper reports robust standard errors
**clustered on household**; we do the same via `sandwich::vcovCL()`.

```{r table3}
m_t3 <- lm(voted ~ treatment, data = ggl)
ct   <- coeftest(m_t3, vcov = vcovCL(m_t3, cluster = ~ hh_id))
ct

rows    <- paste0("treatment", c("Civic Duty", "Hawthorne", "Self", "Neighbors"))
pub_b   <- c(0.018, 0.026, 0.049, 0.081)
pub_se  <- c(0.003, 0.003, 0.003, 0.003)
comp_b  <- ct[rows, "Estimate"]
comp_se <- ct[rows, "Std. Error"]

t3 <- data.frame(
  Term = c("Civic Duty", "Hawthorne", "Self", "Neighbors"),
  Published_b  = pub_b,  Computed_b  = round(comp_b, 4),
  Published_SE = pub_se, Computed_SE = round(comp_se, 4),
  p = signif(ct[rows, "Pr(>|t|)"], 2))
knitr::kable(t3, row.names = FALSE, caption =
  "Table 3, model (a) reproduction: OLS treatment effects with household-clustered SEs, published vs. computed. N = 344,084 individuals; all p < .001.")

# --- Reproduction gate 2: coefficients and clustered SEs must round to the
# --- published three-decimal values.
stopifnot(max(abs(round(comp_b, 3)  - pub_b))  < 1e-9,
          max(abs(round(comp_se, 3) - pub_se)) < 1e-9,
          nobs(m_t3) == 344084)
```

The fresh fit gives `r sprintf("%.4f", comp_b[1])`, `r sprintf("%.4f", comp_b[2])`,
`r sprintf("%.4f", comp_b[3])`, and `r sprintf("%.4f", comp_b[4])` with clustered
SEs between `r sprintf("%.4f", min(comp_se))` and
`r sprintf("%.4f", max(comp_se))` --- a **digit-perfect match at the paper's
printed precision** (.018/.026/.049/.081, all SEs .003), on the full
`r format(nobs(m_t3), big.mark=",")` individuals, all $p < .001$. Both
`stopifnot()` gates pass: **the reproduction gate is passed**, and this fitted
experiment is the anchor for everything below.

# Day 1: Model evaluation --- from average effects to individual predictions

The experiment answers a *causal* question superbly: sending the Neighbors mailing
raises turnout by 8.1 points **on average**. Day 1 asks the *predictive* question:
given what we know about a registered voter --- their vote history, age, sex,
household, and which mailing (if any) they received --- **can we predict whether
that particular person will vote?** The two questions are different, and the EDA
told us why: the treatments shift turnout by 2--8 points, but vote history spans
40+ points. A good *predictor* of turnout may lean mostly on variables the
*experiment* never manipulated. We build the standard Day-1 evaluation toolkit ---
train/test splits, cross-validation, the bias--variance tradeoff, ROC and
precision--recall, calibration --- on top of the reproduced experiment.

## The model ladder and the metrics

We compare four logistic-regression models of `voted`, each adding information:

1. **Baseline** --- intercept only: predict the base rate for everyone.
2. **Treatment only** --- the experiment's information set: the four dummies.
3. **+ Vote history** --- adds the five informative history indicators.
4. **Full** --- adds age (with a quadratic), sex, and household size.

Because the outcome is binary we score models with **log-loss** (average negative
log-likelihood of the observed outcome under the predicted probability --- lower
is better, and it punishes confident wrong predictions hard), **AUC** (the
probability a random voter is ranked above a random non-voter --- 0.5 is
coin-flipping), and later the **Brier score** for calibration.

```{r day1-setup}
logloss <- function(y, p) {
  p <- pmin(pmax(p, 1e-12), 1 - 1e-12)
  -mean(y * log(p) + (1 - y) * log(1 - p))
}
auc_of <- function(y, p) {
  if (length(unique(p)) < 2) return(0.5)   # constant predictor: chance
  as.numeric(pROC::auc(y, p, quiet = TRUE, direction = "<"))
}
f_treat <- voted ~ treatment
f_hist  <- voted ~ treatment + g2000 + g2002 + p2000 + p2002 + p2004
f_full  <- update(f_hist, . ~ . + poly(age, 2) + female + hh_size)
```

## An honest train/test split --- by household

We hold out 30% of the data as a test set. But we do **not** sample individuals:
we sample **households**. Treatment is assigned by household and turnout is
correlated within households (spouses tend to vote --- or not --- together). If we
split individuals, one member of a household could land in training and the other
in test; the model would be graded partly on households it has effectively already
seen, and the test score would be **optimistically biased** by this leakage.
Splitting on the randomization/clustering unit keeps the test set genuinely
unseen. This is the general rule: *split on the unit whose observations are
dependent.*

```{r split}
set.seed(2006)
hh_ids   <- unique(ggl$hh_id)
tr_hh    <- sample(hh_ids, round(0.70 * length(hh_ids)))
in_train <- ggl$hh_id %in% tr_hh
train <- ggl[in_train, ]
test  <- ggl[!in_train, ]
c(train_ind = nrow(train), test_ind = nrow(test),
  train_hh = length(unique(train$hh_id)),
  test_hh  = length(unique(test$hh_id)))
```

```{r split-fit}
p_base  <- rep(mean(train$voted), nrow(test))       # model 1: base rate
m_tr    <- glm(f_treat, train, family = binomial)   # model 2
m_hi    <- glm(f_hist,  train, family = binomial)   # model 3
m_fu    <- glm(f_full,  train, family = binomial)   # model 4
p_tr <- predict(m_tr, test, type = "response")
p_hi <- predict(m_hi, test, type = "response")
p_fu <- predict(m_fu, test, type = "response")

split_tab <- data.frame(
  Model = c("1. Baseline (intercept)", "2. Treatment only",
            "3. + Vote history", "4. Full (+ age, sex, hh size)"),
  LogLoss = round(c(logloss(test$voted, p_base), logloss(test$voted, p_tr),
                    logloss(test$voted, p_hi),   logloss(test$voted, p_fu)), 4),
  AUC     = round(c(auc_of(test$voted, p_base), auc_of(test$voted, p_tr),
                    auc_of(test$voted, p_hi),   auc_of(test$voted, p_fu)), 3))
knitr::kable(split_tab, caption =
  "Held-out (30% of households) test performance of the model ladder.")
```

The ordering is instructive. The treatment dummies --- the entire causal content
of the experiment --- move the AUC from 0.500 to only about
`r sprintf("%.2f", split_tab$AUC[2])`: knowing which mailing someone received
barely helps you guess *whether they will vote*, even though it shifts the
*probability* causally. Vote history is where the predictive action is (AUC jumps
to about `r sprintf("%.2f", split_tab$AUC[3])`), and demographics add a further
increment. Big average effects and big predictive power are simply different
things.

## 10-fold cross-validation of the ladder

A single split is one random draw; cross-validation uses every household in both
roles. We assign each **household** (again: the clustering unit, not the
individual) to one of $K = 10$ folds; each fold takes a turn as the test set. We
pool the out-of-fold predictions so every individual is predicted exactly once,
by a model that never saw their household.

```{r cv-ladder, cache=TRUE}
K <- 10
set.seed(2008)
fold_of_hh <- sample(rep(1:K, length.out = length(hh_ids)))
ggl$fold <- fold_of_hh[match(ggl$hh_id, hh_ids)]
knitr::kable(t(as.matrix(table(Fold = ggl$fold))), caption =
  "Individuals per household-level fold (folds are equal in households, so individual counts vary slightly).")

n <- nrow(ggl)
cv_p <- matrix(NA_real_, n, 4,
               dimnames = list(NULL, c("base", "treat", "hist", "full")))
for (k in 1:K) {
  tri <- ggl$fold != k; tei <- !tri
  cv_p[tei, "base"]  <- mean(ggl$voted[tri])
  cv_p[tei, "treat"] <- predict(glm(f_treat, ggl[tri, ], family = binomial),
                                ggl[tei, ], type = "response")
  cv_p[tei, "hist"]  <- predict(glm(f_hist,  ggl[tri, ], family = binomial),
                                ggl[tei, ], type = "response")
  cv_p[tei, "full"]  <- predict(glm(f_full,  ggl[tri, ], family = binomial),
                                ggl[tei, ], type = "response")
}
cv_tab <- data.frame(
  Model   = split_tab$Model,
  CV_LogLoss = round(apply(cv_p, 2, logloss, y = ggl$voted), 4),
  CV_AUC     = round(apply(cv_p, 2, auc_of,  y = ggl$voted), 3),
  row.names = NULL)
knitr::kable(cv_tab, caption =
  "10-fold cross-validated (household folds, pooled out-of-fold) log-loss and AUC for the model ladder.")
```

The CV numbers agree with the single split to the third decimal --- with 344,084
observations, evaluation noise is tiny --- and the story is identical: each rung
of the ladder helps, history helps most, and the full model wins on both metrics.

## Bias--variance: a validation curve in the age polynomial

The turnout--age profile in the EDA was strongly curved, so *some* flexibility in
age must help --- but how much? We take the full model and sweep the **degree of
the age polynomial** from 1 (a straight line in the log-odds; most rigid, most
*bias*) to 8 (a wiggly curve that can chase sample noise; most *variance*),
recording the **training** log-loss (fit and scored on all rows) and the
**cross-validated** log-loss (household folds as above) at each degree.

```{r validation-curve, cache=TRUE, fig.height=3.6, fig.cap="Validation curve for the age-polynomial degree in the full logit. Training log-loss falls monotonically with flexibility; CV log-loss drops sharply from degree 1 to 2-3 and then flattens, with the minimum marked by the dashed line. At n = 344,084 even a degree-8 polynomial barely overfits, so the right-hand rise is shallow -- the classic U is compressed by sample size."}
degs <- 1:8
form_deg <- function(p)
  update(f_hist, as.formula(paste(
    ". ~ . + poly(age,", p, ") + female + hh_size")))
vc <- data.frame(degree = degs, Train = NA_real_, CV = NA_real_)
for (i in seq_along(degs)) {
  p <- degs[i]
  fit_all <- glm(form_deg(p), ggl, family = binomial)
  vc$Train[i] <- logloss(ggl$voted, fitted(fit_all))
  oof <- numeric(n)
  for (k in 1:K) {
    tri <- ggl$fold != k; tei <- !tri
    fit <- glm(form_deg(p), ggl[tri, ], family = binomial)
    oof[tei] <- predict(fit, ggl[tei, ], type = "response")
  }
  vc$CV[i] <- logloss(ggl$voted, oof)
}
best_deg <- vc$degree[which.min(vc$CV)]
vl <- rbind(data.frame(degree = vc$degree, LogLoss = vc$Train,
                       Curve = "Training (in-sample)"),
            data.frame(degree = vc$degree, LogLoss = vc$CV,
                       Curve = "Cross-validated (held-out)"))
ggplot(vl, aes(degree, LogLoss, colour = Curve)) +
  geom_line(linewidth = 0.8) + geom_point() +
  geom_vline(xintercept = best_deg, linetype = "dashed",
             colour = "grey40") +
  scale_x_continuous(breaks = degs) +
  labs(x = "age polynomial degree --- more flexible to the right",
       y = "log-loss", colour = NULL,
       title = paste0("Bias-variance: CV log-loss minimized at degree = ",
                      best_deg)) +
  theme_minimal(base_size = 11) + theme(legend.position = "top")
```

```{r validation-table}
knitr::kable(data.frame(degree = vc$degree,
                        Train_LogLoss = round(vc$Train, 5),
                        CV_LogLoss    = round(vc$CV, 5)),
             caption = "Validation-curve values: training vs. cross-validated log-loss by age-polynomial degree.")
```

Two lessons. First, the familiar mechanics: training loss falls monotonically with
degree (flexibility always fits the data it sees at least as well), while CV loss
drops sharply as the polynomial captures the real curvature in the age profile
and is minimized at degree `r best_deg`. Second, a lesson about **sample size and
variance**: with 344,084 observations even a degree-8 polynomial has little room
to chase noise, so the right-hand side of the U barely rises. Overfitting is a
*ratio* of flexibility to data --- run this same curve on a 2,000-row subsample
(Exercise 3) and the U reappears in force.

## ROC and precision--recall for the best model

We now evaluate the CV-best model (full logit, age degree `r best_deg`) on the
held-out household test set from the single split, starting with the ROC curve:
the tradeoff between the true-positive rate and the false-positive rate as the
classification threshold sweeps from 1 to 0.

```{r roc, fig.height=3.8, fig.cap="ROC curves on the held-out test set. The full model dominates the treatment-only model everywhere; the treatment-only curve hugs the diagonal, showing that experimental assignment alone barely ranks voters above non-voters."}
m_best  <- glm(form_deg(best_deg), train, family = binomial)
p_best  <- predict(m_best, test, type = "response")
roc_best <- pROC::roc(test$voted, p_best,  quiet = TRUE, direction = "<")
roc_tr   <- pROC::roc(test$voted, p_tr,    quiet = TRUE, direction = "<")
plot(roc_best, col = "steelblue", lwd = 2, legacy.axes = TRUE,
     main = sprintf("ROC, held-out test set (full AUC = %.3f)",
                    as.numeric(pROC::auc(roc_best))))
plot(roc_tr, col = "firebrick", lwd = 2, add = TRUE)
legend("bottomright", lwd = 2, col = c("steelblue", "firebrick"),
       legend = c(sprintf("Full model (AUC = %.3f)",
                          as.numeric(pROC::auc(roc_best))),
                  sprintf("Treatment only (AUC = %.3f)",
                          as.numeric(pROC::auc(roc_tr)))), bty = "n")
```

With a base rate of `r sprintf("%.1f", 100*mean(test$voted))`% --- roughly one
voter for every two non-voters --- the ROC curve can flatter a model, because the
false-positive *rate* has a large denominator of non-voters. The
**precision--recall curve** asks the question a campaign would ask: *of the people
the model flags as likely voters, what share actually vote (precision), and what
share of all actual voters do we capture (recall)?*

```{r pr-curve, fig.height=3.6, fig.cap="Precision-recall on the held-out test set. The dashed line is the base rate (~32 percent), the precision of flagging everyone. The full model reaches roughly 80 percent precision on its most confident picks and degrades gracefully toward the base rate as recall approaches 1."}
ord  <- order(p_best, decreasing = TRUE)
yy   <- test$voted[ord]
tp   <- cumsum(yy)
prec <- tp / seq_along(yy)
rec  <- tp / sum(yy)
thin <- seq(1, length(yy), by = 200)     # thin for plotting only
pr <- data.frame(recall = rec[thin], precision = prec[thin])
base_rate <- mean(test$voted)
ggplot(pr, aes(recall, precision)) +
  geom_line(colour = "steelblue", linewidth = 0.9) +
  geom_hline(yintercept = base_rate, linetype = "dashed",
             colour = "grey40") +
  annotate("text", x = 0.15, y = base_rate - 0.035, size = 3.2,
           label = sprintf("base rate = %.3f", base_rate)) +
  ylim(0, 1) +
  labs(x = "Recall (share of actual voters captured)",
       y = "Precision (share of flagged who vote)",
       title = "Precision-recall, full model, held-out test set") +
  theme_minimal(base_size = 11)
```

Among the model's top-ranked prospects, precision is high --- the top 5% of the
test set votes at `r sprintf("%.0f", 100*prec[round(0.05*length(yy))])`% ---
more than double the base rate. Any threshold is a business decision: a campaign
with money for one mailing per ten households wants the high-precision end; a
canvass that must reach every plausible voter accepts lower precision for high
recall.

## Calibration and the Brier score

AUC only checks *ranking*. **Calibration** checks the probabilities themselves:
among people to whom the model assigns, say, a 40% chance of voting, do about 40%
actually vote? We bin the test set into deciles of predicted probability and plot
the mean prediction against the observed turnout in each bin. Experimental data
makes this exercise especially clean: the predictions embed causal treatment
contrasts we *know* are right on average, so a well-specified model should sit on
the 45-degree line.

```{r calibration, fig.height=3.8, fig.cap="Reliability curve on the held-out test set: mean predicted probability vs. observed turnout within deciles of the prediction. Points sit on the 45-degree line across the whole range -- the logit's probabilities mean what they say."}
dec <- cut(p_best, quantile(p_best, 0:10 / 10), include.lowest = TRUE)
cal <- data.frame(
  mean_pred = tapply(p_best, dec, mean),
  obs_rate  = tapply(test$voted, dec, mean),
  n         = as.integer(table(dec)))
ggplot(cal, aes(mean_pred, obs_rate)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey50") +
  geom_point(colour = "steelblue", size = 2.4) +
  geom_line(colour = "steelblue", linewidth = 0.6) +
  coord_equal(xlim = c(0, 0.8), ylim = c(0, 0.8)) +
  labs(x = "Mean predicted P(vote), decile of prediction",
       y = "Observed turnout in decile",
       title = "Calibration: predicted vs. observed by decile") +
  theme_minimal(base_size = 11)
```

```{r calibration-table}
knitr::kable(data.frame(Decile = 1:10,
                        Mean_pred = round(cal$mean_pred, 3),
                        Observed  = round(cal$obs_rate, 3),
                        N = format(cal$n, big.mark = ",")),
             row.names = FALSE, caption =
  "Decile reliability table for the full model on the held-out test set.")
brier <- data.frame(
  Model = c("Baseline (base rate)", "Treatment only",
            paste0("Full (age degree ", best_deg, ")")),
  Brier = round(c(mean((p_base - test$voted)^2),
                  mean((p_tr   - test$voted)^2),
                  mean((p_best - test$voted)^2)), 4))
knitr::kable(brier, caption =
  "Brier scores (mean squared error of the predicted probability) on the held-out test set; lower is better.")
```

The reliability curve tracks the diagonal from the ~10%-probability decile up to
the ~70% decile: when this model says 60%, about 60% vote. The Brier score
summarizes the same thing in one number, and its ladder mirrors the log-loss
ladder --- most of the improvement over the base-rate benchmark comes from vote
history and age, not from treatment.

## What ML adds --- and what it does not

The closing contrast of Day 1. The **experiment** delivers something no
predictive model can: an unbiased estimate of what *would happen* if we mailed
the Neighbors letter to anyone --- +8.1 points, full stop, no modeling
assumptions. The **predictive models** deliver something the experiment alone
does not: individual-level probabilities --- *this* voter is a 9% prospect, *that*
one is 71% --- driven overwhelmingly by vote history and age, variables the
experiment never manipulated and which no one can manipulate. The two outputs
answer different questions and are useful together: a modern GOTV campaign uses
the *causal* estimate to decide whether the mailing is worth sending and the
*predictive* model to decide whom to send it to (don't waste postage on
sure-thing voters or hopeless abstainers). Keeping "predicts well" and "identifies
a causal effect" distinct --- and knowing which one your model is delivering ---
is the most important habit this course teaches.

# Exercises

1. **Logit reproduction.** Refit the Table 3(a) specification as a logistic
   regression (`glm(voted ~ treatment, family = binomial)`) with
   household-clustered SEs. Convert the Neighbors coefficient into an average
   marginal effect (average the difference in predicted probabilities with
   everyone assigned to Neighbors vs. Control). How close is it to the OLS
   coefficient of .081, and why does the linear probability model work so well
   here?

2. **Individual vs. household splits.** Rerun the train/test comparison in
   Section 5.2, but sample *individuals* rather than households into the test
   set. Compare the test log-loss and AUC of the full model under the two
   protocols. In which direction does the individual-level split bias the
   evaluation, and how large is the leakage effect?

3. **Bias--variance and sample size.** Draw a random subsample of 2,000
   individuals and recompute the validation curve of Section 5.4 (age degree 1
   through 8, using simple 10-fold CV within the subsample). Compare the shape of
   the CV curve to the full-data version: where is the minimum now, and how sharp
   is the overfitting penalty at degree 8?

4. **Household context and heterogeneous targeting.** Add the pre-treatment
   household means `p2004_mean` and `g2004_mean`, plus an interaction between
   `treatment` and `p2004`, to the full model. Does household context improve
   held-out log-loss? Then use the interaction model to ask a targeting question:
   among which voters (habitual primary voters vs. non-voters) does the model
   predict the Neighbors mailing to matter most, and does that match the
   subgroup means you can compute directly from the experiment?
