---
title: "Predicting Secondary-School Final Grades: Reproduction and Day-1 Model-Evaluation Extension"
subtitle: "Reproducing El Aissaoui et al. (2020), *A Multiple Linear Regression-Based Approach to Predict Student Performance*"
author: "Intro to Machine Learning --- Day 1: Introduction & Model Evaluation"
date: "`r format(Sys.Date())`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
header-includes:
  - \setlength{\emergencystretch}{3em}
  - \usepackage{fvextra}
  - \AtBeginDocument{\DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}}
  - \AtBeginDocument{\DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}}
  - \fvset{breaklines=true}
  - \usepackage{booktabs}
  - \usepackage{float}
---

```{r setup, include=FALSE}
options(width = 80)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 6, fig.height = 4,
                      fig.pos = "H")
set.seed(20240501)
options(digits = 5)
```

# The study and the reproduction gate

**Citation.** El Aissaoui, O., El Alami El Madani, Y., Oughdir, L., Dakkak, A., & El Allioui, Y. (2020).
*A Multiple Linear Regression-Based Approach to Predict Student Performance.* In M. Ezziyyani (Ed.),
Advanced Intelligent Systems for Sustainable Development (AI2SD 2019), AISC 1102, pp. 9--23. Springer.
DOI: \texttt{10.1007/978-3-030-36653-7\_2}.

**Data source.** Cortez, P. & Silva, A. (2008), *Student Performance* dataset, UCI Machine Learning
Repository, DOI \texttt{10.24432/C50G6X} --- file \texttt{student-mat.csv} (Mathematics course, 395 students,
33 columns). The dataset is fully public and ships with the course as a local copy.

**The published model.** The paper's task is a plain OLS (`lm`) predicting the final Mathematics grade
`G3` (numeric, 0--20). Of the seven candidate models the authors report (Table 2), their best
(**Model 6 / MARS-selected**, lowest RMSE and highest $R^2$ in their Table 3) uses five predictors:

$$
\widehat{G3} = 17.7935 + 0.8627\,\text{Medu} - 0.5603\,\text{goout}
- 1.2458\,\text{romanticyes} - 0.4423\,\text{age} - 1.6562\,\text{schoolsupyes}.
$$

Here `Medu` = mother's education (0--4), `goout` = going out with friends (1--5),
`romantic` = in a romantic relationship (yes/no), `age` (15--22), `schoolsup` = extra
educational support (yes/no).

The **reproduction gate**: our reproduced coefficients must match the six printed numbers above.
A subtlety --- the paper preprocesses numeric predictors by replacing boxplot outliers with `NA`
and then mean-imputing them (their Section 3.2). We replicate that exactly.

```{r load-data}
raw <- read.csv(paste0(
  "https://raw.githubusercontent.com/bdesmarais/",
  "intro-ml-statistical-horizons-4day/main/",
  "tutorials/day1_intro_and_evaluation/elaissaoui_student_grades/data/student-mat.csv"), sep = ";", stringsAsFactors = FALSE)
cat("Rows x cols:", nrow(raw), "x", ncol(raw), "\n")
cat("Outcome G3: range", paste(range(raw$G3), collapse = "-"),
    "| mean", round(mean(raw$G3), 3), "| sd", round(sd(raw$G3), 3), "\n")

# Paper's preprocessing: boxplot-outlier -> NA -> mean imputation on numeric predictors.
impute_outliers <- function(x) {
  out <- boxplot.stats(x)$out          # points beyond 1.5*IQR
  x[x %in% out] <- NA
  x[is.na(x)]  <- mean(x, na.rm = TRUE) # mean imputation
  x
}
dat <- raw
dat$romantic  <- factor(dat$romantic,  levels = c("no", "yes"))
dat$schoolsup <- factor(dat$schoolsup, levels = c("no", "yes"))
for (v in c("Medu", "goout", "age")) dat[[v]] <- impute_outliers(dat[[v]])
```

# Descriptive analysis

Before fitting anything, we meet the data properly. A careful exploratory pass is not decoration ---
it tells us the shape of the outcome (and whether it needs transforming), which predictors carry
signal, whether any variables are collinear enough to destabilize an OLS, and where the missing data
are. Everything the modeling section does downstream is easier to interpret once we have looked.

## Dataset and provenance

The **unit of analysis is the individual student**. There are `r nrow(dat)` students enrolled in the
Mathematics course, each described by `r ncol(raw)` variables. The data come from a **single-wave,
cross-sectional survey** administered in two public secondary schools in the Alentejo region of
Portugal during the **2005--2006 academic year** (Cortez & Silva 2008). Records were built by
combining school reports (grades, absences) with a paper questionnaire on demographics, family
background, and social/study habits; the two sources were merged per student. There is **no sampling
weight and no time dimension** --- it is a convenience census of the two schools' math cohorts, so
inferences are to *these students*, not to Portugal as a whole.

The 33 columns fall into four blocks: (i) **demographics** (`sex`, `age`, `address`, ...),
(ii) **family background** (`Medu`, `Fedu`, `Mjob`, `Fjob`, `Pstatus`, ...), (iii) **study / social
behavior** (`studytime`, `failures`, `goout`, `absences`, alcohol use, ...), and (iv) **three grade
periods** `G1`, `G2`, `G3` (first-period, second-period, and final math grade, each 0--20). The paper
predicts `G3` from a small demographic/behavioral subset and deliberately *excludes* the earlier
grades `G1`/`G2` (which would otherwise dominate); we honor that choice throughout.

```{r eda-dictionary}
# Variable dictionary for the outcome + the model's predictors (+ the Day-1 knob).
dict <- data.frame(
  variable = c("G3", "Medu", "goout", "romantic", "age",
               "schoolsup", "absences"),
  role = c("outcome", "predictor", "predictor", "predictor",
           "predictor", "predictor", "Day-1 flex knob"),
  type = c("numeric (0-20)", "ordinal 0-4", "ordinal 1-5",
           "binary", "numeric (years)", "binary", "count"),
  description = c(
    "Final math grade",
    "Mother's education level",
    "Frequency of going out with friends",
    "In a romantic relationship",
    "Student age",
    "Extra school educational support",
    "Number of school absences"),
  coding = c("0=fail .. 20=top", "0=none,1=primary,2=5-9th,3=sec,4=higher",
             "1=very low .. 5=very high", "no / yes", "15-22 years",
             "no / yes", "0-93 days"))
knitr::kable(dict, row.names = FALSE,
             caption = "Variable dictionary: outcome, the five published predictors, and the Day-1 flexibility knob (absences).")
```

We profile these variables in depth below and return to the fuller feature set only where it clarifies
the modeling. We also load the plotting/EDA helpers used in this section.

```{r eda-libs}
library(ggplot2)
library(dplyr)
suppressMessages(library(GGally))
suppressMessages(library(patchwork))
theme_set(theme_minimal(base_size = 10))
```

## The outcome `G3`, in depth

```{r eda-outcome, fig.height=3.0, fig.cap="Final math grade G3 (0--20): histogram with overlaid density (left) and boxplot (right). The tall bar at 0 is the ~10% of students recorded with a zero final grade."}
g_hist <- ggplot(dat, aes(G3)) +
  geom_histogram(aes(y = after_stat(density)), breaks = 0:20,
                 fill = "grey80", colour = "white") +
  geom_density(colour = "firebrick", linewidth = 0.8) +
  geom_vline(xintercept = mean(dat$G3), colour = "firebrick",
             linetype = 2) +
  labs(x = "G3 (final math grade)", y = "density",
       title = "Distribution of G3")
g_box <- ggplot(dat, aes(y = G3)) +
  geom_boxplot(fill = "lightsteelblue", width = 0.4) +
  labs(x = NULL, y = "G3", title = "Boxplot") +
  theme(axis.text.x = element_blank())
g_hist + g_box + plot_layout(widths = c(3, 1))
```

```{r eda-outcome-stats}
# Full descriptive stats for the outcome, incl. skewness & kurtosis (e1071).
q <- quantile(dat$G3, c(.25, .5, .75))
os <- data.frame(
  n = length(dat$G3), mean = mean(dat$G3), sd = sd(dat$G3),
  min = min(dat$G3), Q1 = q[1], median = q[2], Q3 = q[3],
  max = max(dat$G3), IQR = IQR(dat$G3),
  skew = e1071::skewness(dat$G3), kurt = e1071::kurtosis(dat$G3))
knitr::kable(os, row.names = FALSE, digits = 2,
             caption = "Outcome G3: full summary statistics (skewness & kurtosis from e1071).")
```

`G3` is centered near `r round(mean(dat$G3),1)` (sd `r round(sd(dat$G3),1)`) and would look roughly
bell-shaped were it not for a conspicuous **floor spike at 0**: `r sum(dat$G3==0)` students
(`r round(100*mean(dat$G3==0))`%) are recorded with a zero final grade --- students who did not sit,
or failed outright, the final. That spike pulls the mean below the median and gives a mild **left
skew** (skewness `r round(e1071::skewness(dat$G3),2)`); the small negative excess kurtosis says the
non-zero mass is a touch flatter than a normal. This is *floor inflation*, not classic right-skew, so
a log/sqrt transform would not help --- and a plain OLS (the paper's model) simply treats the zeros as
genuine low grades. We keep `G3` untransformed to stay faithful to the reproduction, but flag the
zeros as the main reason the achievable $R^2$ is capped.

## Univariate distributions of the predictors

```{r eda-num-univ, fig.height=3.2, fig.cap="Numeric/ordinal predictors used downstream. Medu and goout are discrete ordinal scales; age is right-skewed (a few older repeaters); absences is heavily right-skewed with a spike at zero."}
num_pred <- c("Medu", "goout", "age", "absences")
long <- dat |>
  select(all_of(num_pred)) |>
  tidyr::pivot_longer(everything(), names_to = "var",
                      values_to = "value")
ggplot(long, aes(value)) +
  geom_histogram(bins = 20, fill = "steelblue", colour = "white") +
  facet_wrap(~ var, scales = "free", ncol = 2) +
  labs(x = NULL, y = "count",
       title = "Univariate distributions of the numeric predictors")
```

```{r eda-num-tab}
# Full summary table for the numeric predictors, including skew.
sumrow <- function(v) {
  x <- dat[[v]]; q <- quantile(x, c(.25, .5, .75))
  data.frame(variable = v, n = sum(!is.na(x)),
             n_miss = sum(is.na(x)), mean = mean(x), sd = sd(x),
             min = min(x), Q1 = q[1], median = q[2], Q3 = q[3],
             max = max(x), skew = e1071::skewness(x))
}
num_tab <- do.call(rbind, lapply(num_pred, sumrow))
knitr::kable(num_tab, row.names = FALSE, digits = 2,
             caption = "Numeric predictors: full summary statistics. Medu/goout/age enter the model; absences is the Day-1 flexibility knob.")
```

`Medu` and `goout` are short ordinal scales that behave like coarse counts; both are fairly
symmetric. `age` is mildly right-skewed --- most students are 15--18, with a thin tail of older
repeaters up to 22. `absences` is the extreme case: strongly **right-skewed** (skew
`r round(e1071::skewness(dat$absences),1)`), a large mass at zero, and a long thin tail (max
`r max(dat$absences)`). That heavy skew is exactly why the paper's outlier-imputation step touches it
and why, later, a high-degree polynomial in `absences` is so prone to chase the few extreme points.

```{r eda-cat-univ, fig.height=2.6, fig.cap="The two binary predictors. Most students are NOT in a romantic relationship and do NOT receive extra school support; both classes are moderately imbalanced."}
cat_pred <- c("romantic", "schoolsup")
catlong <- dat |>
  select(all_of(cat_pred)) |>
  mutate(across(everything(), as.character)) |>
  tidyr::pivot_longer(everything(), names_to = "var",
                      values_to = "level")
ggplot(catlong, aes(level)) +
  geom_bar(fill = "darkseagreen4") +
  facet_wrap(~ var, scales = "free_x") +
  labs(x = NULL, y = "count",
       title = "Binary predictor frequencies")
```

```{r eda-cat-tab}
cat_tab <- rbind(
  romantic  = table(dat$romantic),
  schoolsup = table(dat$schoolsup))
cat_tab <- cbind(cat_tab,
                 prop_yes = round(cat_tab[, "yes"] / rowSums(cat_tab), 3))
knitr::kable(cat_tab,
             caption = "Binary predictors: frequency counts and proportion 'yes'.")
```

About `r round(100*mean(dat$romantic=="yes"))`% of students report being in a romantic relationship
and only `r round(100*mean(dat$schoolsup=="yes"))`% receive extra educational support. Neither class
is so rare as to be unusable, but `schoolsup` in particular is imbalanced --- worth remembering when
its coefficient turns out to be one of the largest in the model.

## Missing-data analysis

```{r eda-missing}
# Per-variable missing counts/percentages across the WHOLE frame.
miss <- sapply(dat, function(x) sum(is.na(x)))
miss_tab <- data.frame(
  variable = names(miss), n_missing = as.integer(miss),
  pct_missing = round(100 * miss / nrow(dat), 2))
miss_tab <- miss_tab[order(-miss_tab$n_missing), ]
cat("Complete cases:", sum(complete.cases(dat)), "of", nrow(dat), "\n")
cat("Total missing cells:", sum(miss), "\n")
knitr::kable(head(miss_tab, 6), row.names = FALSE,
             caption = "Missingness by variable (top rows). The analysis frame is complete.")
```

The distributed UCI dataset is **fully complete** --- every one of the `r nrow(dat)` rows is a
complete case, so there is no missingness pattern to model and no co-missing variables. The only `NA`s
in this pipeline are **transient and self-inflicted**: the paper's preprocessing sets boxplot outliers
in `Medu`, `goout`, and `age` to `NA` and immediately mean-imputes them inside `impute_outliers()`, so
by the time we describe or model the data there are zero missing cells. Because that imputation is
deterministic and mean-based (not model-based), it cannot leak outcome information --- it just softens
a handful of extreme predictor values.

## Bivariate relationships with the outcome

```{r eda-bivar-num, fig.height=3.0, fig.cap="G3 against the two ordinal model predictors, as grouped boxplots. Higher mother's education tracks higher final grades; more frequent going-out tracks lower grades."}
b1 <- ggplot(dat, aes(factor(round(Medu)), G3)) +
  geom_boxplot(fill = "lightsteelblue") +
  labs(x = "Mother's education (Medu)", y = "G3",
       title = "G3 by Medu")
b2 <- ggplot(dat, aes(factor(round(goout)), G3)) +
  geom_boxplot(fill = "lightsalmon") +
  labs(x = "Going out (goout)", y = "G3", title = "G3 by goout")
b1 + b2
```

```{r eda-bivar-abs, fig.height=2.9, fig.cap="G3 vs. absences with linear (solid) and loess (dashed) smoothers. The association is weak and nearly flat past a few absences --- little curvature for a flexible model to exploit."}
ggplot(dat, aes(absences, G3)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, colour = "steelblue") +
  geom_smooth(method = "loess", se = FALSE, colour = "firebrick",
              linetype = 2) +
  labs(x = "Absences (count)", y = "G3",
       title = "G3 vs. absences")
```

```{r eda-bivar-cat, fig.height=2.9, fig.cap="G3 by the two binary predictors. Students in a romantic relationship and those receiving extra school support score slightly lower on average."}
v1 <- ggplot(dat, aes(romantic, G3)) +
  geom_violin(fill = "plum3", alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", colour = "black") +
  labs(x = "romantic", y = "G3", title = "G3 by romantic")
v2 <- ggplot(dat, aes(schoolsup, G3)) +
  geom_violin(fill = "khaki3", alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", colour = "black") +
  labs(x = "schoolsup", y = "G3", title = "G3 by schoolsup")
v1 + v2
```

```{r eda-bivar-tab}
# One tidy table: each predictor's marginal association with G3.
num_assoc <- lapply(c("Medu", "goout", "age", "absences"), function(v) {
  ct <- cor.test(dat[[v]], dat$G3)
  data.frame(predictor = v, kind = "numeric",
             effect = round(unname(ct$estimate), 3),
             effect_type = "Pearson r",
             p_value = signif(ct$p.value, 2))
})
cat_assoc <- lapply(c("romantic", "schoolsup"), function(v) {
  mm <- tapply(dat$G3, dat[[v]], mean)
  tt <- t.test(G3 ~ dat[[v]], data = dat)
  data.frame(predictor = v, kind = "binary",
             effect = round(unname(mm["yes"] - mm["no"]), 3),
             effect_type = "mean(yes)-mean(no)",
             p_value = signif(tt$p.value, 2))
})
assoc <- do.call(rbind, c(num_assoc, cat_assoc))
knitr::kable(assoc, row.names = FALSE,
             caption = "Marginal association of each predictor with G3 (Pearson r for numeric; yes-minus-no grade gap with a two-sample t-test for binary).")
```

The bivariate picture lines up cleanly with the published coefficient signs: **`Medu` is positively**
associated with `G3` (r `r round(cor(dat$Medu, dat$G3),2)`), while `goout`, `age`, being in a
`romantic` relationship, and needing `schoolsup` all pull grades **down**. Every marginal effect is
small in magnitude --- foreshadowing the modest $R^2$ ahead --- and, importantly, the `G3`--`absences`
relationship is weak and essentially **flat** past a handful of absences, so there is very little
curvature for a flexible model to exploit.

## Multivariate structure and collinearity

```{r eda-corr-heat, fig.height=3.2, fig.cap="Correlation heatmap among G3 and the numeric predictors. All off-diagonal correlations are weak (|r| < 0.25); nothing approaches the |r| > 0.7 collinearity threshold."}
ggcorr(dat[, c("G3", num_pred)], label = TRUE, label_round = 2,
       hjust = 0.85, layout.exp = 1) +
  labs(title = "Correlations among G3 and numeric predictors")
```

```{r eda-corr-tab}
cmat <- cor(dat[, c("G3", num_pred)])
knitr::kable(round(cmat, 3),
             caption = "Pearson correlations among G3 and the numeric predictors.")
```

```{r eda-vif}
# Variance-inflation factors from the reproduction model (base-R, no car).
vif_base <- function(fit) {
  X <- model.matrix(fit)[, -1, drop = FALSE]
  vapply <- sapply(colnames(X), function(nm) {
    others <- setdiff(colnames(X), nm)
    r2 <- summary(lm(X[, nm] ~ X[, others]))$r.squared
    1 / (1 - r2)
  })
  vapply
}
fit_for_vif <- lm(G3 ~ Medu + goout + romantic + age + schoolsup,
                  data = dat)
vt <- data.frame(term = names(vif_base(fit_for_vif)),
                 VIF = round(unname(vif_base(fit_for_vif)), 2))
knitr::kable(vt, row.names = FALSE,
             caption = "Variance-inflation factors for the reproduced OLS. All near 1: no collinearity.")
```

Correlations among the predictors are uniformly weak (all $|r| < 0.25$, well under the $|r| > 0.7$
rule of thumb), and every **VIF is essentially 1** --- there is **no collinearity concern**. Each
predictor contributes largely independent information, which means the OLS coefficients are stable and
individually interpretable (no variance inflation to worry about). The correlations *with* `G3` are
likewise small, so we should expect a **modest but real** predictive signal, no collinearity trouble,
and --- crucially for Day 1 --- little genuine nonlinearity, so the validation curve should favor a
**low** polynomial degree.

## Reproduced regression vs. published

```{r reproduce}
form <- G3 ~ Medu + goout + romantic + age + schoolsup
fit_ols <- lm(form, data = dat)

published <- c(`(Intercept)` = 17.7935, Medu = 0.8627, goout = -0.5603,
               romanticyes = -1.2458, age = -0.4423, schoolsupyes = -1.6562)
repro <- round(coef(fit_ols)[names(published)], 4)
cmp <- data.frame(term = names(published),
                  published = published,
                  reproduced = repro,
                  abs_diff = abs(published - repro))
knitr::kable(cmp, row.names = FALSE, digits = 4,
             caption = "Reproduction gate: published (El Aissaoui et al. Model 6) vs. reproduced OLS coefficients.")
stopifnot(all(cmp$abs_diff < 1e-3))
cat("GATE PASSED: all six coefficients match to 4 decimals.\n")
```

Every coefficient matches the printed table to four decimal places. The gate is passed; we now keep
this model fixed as the **reproduced baseline** and use its outcome (`G3`) and predictor set to
demonstrate the full Day-1 model-evaluation toolkit.

\newpage

# Day 1: Model evaluation

Everything below uses the **same outcome (`G3`) and the same five predictors** as the reproduced OLS,
so that every evaluation idea is anchored to the published model.

## Predictive vs. explanatory framing (Shmueli)

The published paper is written in an **explanatory** register: it fits OLS to *understand which
student attributes are associated with the final grade* (mother's education helps; going out, being
older, being in a relationship, and needing extra support are associated with lower grades). The
coefficient signs and significance are the deliverable.

A **predictive** register (Shmueli 2010, *"To Explain or to Predict?"*) asks a different question:
*how well would this model forecast the grade of a student it has never seen?* Explanatory fit
(in-sample $R^2$, coefficient p-values) and predictive accuracy (held-out error) are **not the same
number**, and optimizing one does not optimize the other. Day 1 is about measuring the *predictive*
quantity honestly --- which requires holding data out.

```{r insample-r2}
sm <- summary(fit_ols)
cat("In-sample R^2:", round(sm$r.squared, 4),
    "| adjusted R^2:", round(sm$adj.r.squared, 4), "\n")
```

## In-sample optimism: the gap

A model scored on the **same** rows used to fit it flatters itself: it has partly memorized the noise.
The honest number comes from rows held out of fitting. Below we contrast in-sample $R^2$/RMSE with a
single held-out estimate.

```{r optimism}
rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))
mae  <- function(y, yhat) mean(abs(y - yhat))
r2   <- function(y, yhat) 1 - sum((y - yhat)^2) / sum((y - mean(y))^2)

# in-sample
yhat_in <- predict(fit_ols)
insamp <- c(RMSE = rmse(dat$G3, yhat_in), R2 = summary(fit_ols)$r.squared)

# honest held-out (single split, detailed next section)
n <- nrow(dat)
idx <- sample(n, size = floor(0.7 * n))
tr <- dat[idx, ]; te <- dat[-idx, ]
fit_tr <- lm(form, data = tr)
yhat_te <- predict(fit_tr, te)
holdout <- c(RMSE = rmse(te$G3, yhat_te), R2 = r2(te$G3, yhat_te))

opt <- rbind(`In-sample` = insamp, `Held-out (test)` = holdout)
knitr::kable(round(opt, 4), caption = "In-sample optimism: the same model looks better on the data it was fit to.")
cat("Optimism gap in R^2:", round(insamp["R2"] - holdout["R2"], 4), "\n")
```

The in-sample $R^2$ is the model's optimistic self-report; the held-out $R^2$ is what we would
actually see on new students. The difference is *optimism*, and it is exactly why we hold data out.

## Data splitting: a single train/test split, then k-fold CV

**Single split.** We already did one 70/30 split above. A single split is cheap but noisy --- its
answer wobbles depending on which rows happened to land in the test set.

**k-fold cross-validation.** We reduce that noise by rotating the held-out fold. We write the folds
out explicitly so the mechanism is transparent.

```{r kfold}
K <- 10
folds <- sample(rep(1:K, length.out = n))    # explicit fold assignment for every row
cat("Fold sizes:\n"); print(table(folds))

cv_lm <- function(formula, data, folds) {
  K <- max(folds); rm <- ma <- rs <- numeric(K)
  for (k in 1:K) {
    trk <- data[folds != k, ]; tek <- data[folds == k, ]
    fk  <- lm(formula, data = trk)
    pk  <- predict(fk, tek)
    rm[k] <- rmse(tek$G3, pk); ma[k] <- mae(tek$G3, pk); rs[k] <- r2(tek$G3, pk)
  }
  c(RMSE = mean(rm), MAE = mean(ma), R2 = mean(rs))
}
cv_ols <- cv_lm(form, dat, folds)
knitr::kable(t(round(cv_ols, 4)),
             caption = "10-fold CV of the reproduced OLS (matches the paper's Table 3 Model 6: RMSE 4.368, R2 0.103, MAE 3.301).")
```

The k-fold estimate reproduces the paper's own reported cross-validated numbers (Table 3, Model 6:
RMSE $\approx 4.368$, $R^2 \approx 0.103$, MAE $\approx 3.301$) --- a second, independent confirmation
that we reproduced *their* pipeline, not just their coefficients.

## Bias--variance tradeoff: a validation curve

Our **flexibility knob for Day 1 is the degree of a polynomial** --- the simplest knob that stays
entirely within OLS (we just add `poly()` terms inside `lm`). We take the continuous count predictor
`absences` (34 distinct values, the most continuous variable in the frame) and fit polynomials of
increasing degree in it, added on top of the five published predictors. Low degree = high bias
(underfit); high degree = high variance (overfit). Training error should fall monotonically; test
error should be U-shaped.

```{r bias-variance, fig.cap="Validation curve. Training RMSE (blue) falls monotonically as polynomial degree grows; 10-fold CV RMSE (red) is U-shaped --- the signature of the bias-variance tradeoff."}
# Flexibility knob = polynomial degree in the continuous predictor `absences`,
# added on top of the reproduced OLS predictors.
poly_formula <- function(pred, d)
  as.formula(sprintf(
    "G3 ~ poly(%s, %d) + Medu + goout + romantic + age + schoolsup", pred, d))

degrees <- 1:8
train_rmse <- cv_rmse <- numeric(length(degrees))
for (j in seq_along(degrees)) {
  f_j <- poly_formula("absences", degrees[j])
  # training error (in-sample)
  fit_j <- lm(f_j, data = dat)
  train_rmse[j] <- rmse(dat$G3, predict(fit_j))
  # honest CV error
  rmk <- numeric(K)
  for (k in 1:K) {
    trk <- dat[folds != k, ]; tek <- dat[folds == k, ]
    fk <- lm(f_j, data = trk)
    rmk[k] <- rmse(tek$G3, predict(fk, tek))
  }
  cv_rmse[j] <- mean(rmk)
}
bv <- data.frame(degree = degrees, train = train_rmse, cv = cv_rmse)

plot(bv$degree, bv$cv, type = "b", pch = 19, col = "firebrick",
     ylim = range(c(bv$train, bv$cv)),
     xlab = "Polynomial degree of absences (flexibility)", ylab = "RMSE",
     main = "Validation curve: bias-variance tradeoff")
lines(bv$degree, bv$train, type = "b", pch = 17, col = "steelblue")
best_d <- bv$degree[which.min(bv$cv)]
abline(v = best_d, lty = 3)
legend("topleft", legend = c("10-fold CV (test) RMSE", "Training RMSE",
       sprintf("CV-optimal degree = %d", best_d)),
       col = c("firebrick", "steelblue", "black"), pch = c(19, 17, NA),
       lty = c(1, 1, 3), bty = "n", cex = 0.85)
knitr::kable(round(bv, 4), caption = "Training vs. 10-fold CV RMSE across polynomial degree.")
```

Training RMSE decreases as we raise the degree (the polynomial can trace more wiggles), but CV RMSE
bottoms out at a low degree and then rises --- added flexibility beyond that point is fitting noise,
not signal. The near-flat relationship of `G3` on `absences` we saw in the descriptives means the
honest optimum sits at a **low degree** (`r best_d`), confirming the paper's linear choice was
reasonable.

## Performance metrics for a continuous outcome, vs. a naive baseline

`G3` is continuous, so the relevant metrics are **RMSE, MAE, and $R^2$**. A metric is only
interpretable against a **naive baseline** --- here, predicting the training-set mean grade for
everyone. $R^2 > 0$ literally means "beats the mean baseline."

```{r metric-suite}
# naive baseline: predict the (training) mean for all test rows, under the same 10-fold protocol
cv_baseline <- function(data, folds) {
  K <- max(folds); rm <- ma <- rs <- numeric(K)
  for (k in 1:K) {
    trk <- data[folds != k, ]; tek <- data[folds == k, ]
    pk <- rep(mean(trk$G3), nrow(tek))
    rm[k] <- rmse(tek$G3, pk); ma[k] <- mae(tek$G3, pk); rs[k] <- r2(tek$G3, pk)
  }
  c(RMSE = mean(rm), MAE = mean(ma), R2 = mean(rs))
}
base <- cv_baseline(dat, folds)

metric_tab <- rbind(`Naive (mean) baseline` = base,
                    `Reproduced OLS`        = cv_ols)
knitr::kable(round(metric_tab, 4),
             caption = "Full metric suite (10-fold CV): reproduced OLS vs. naive mean baseline.")
cat("OLS reduces held-out RMSE by",
    round(100 * (base["RMSE"] - cv_ols["RMSE"]) / base["RMSE"], 1),
    "% vs. the naive baseline.\n")
```

The OLS beats the mean baseline, but modestly ($R^2 \approx 0.10$): these five demographic/behavioral
predictors explain only about a tenth of grade variance --- an honest, teachable result. (The huge
predictive lift in this dataset comes from prior grades `G1`/`G2`, which the paper deliberately
excludes; we keep its predictor set fixed to stay faithful to the reproduction.)

## Hyperparameters vs. parameters; grid search and random search

**Parameters** are learned by fitting (the OLS coefficients). **Hyperparameters** are knobs *we* set
before fitting and must tune by held-out performance. Our hyperparameter is the **polynomial degree**
from the validation curve above. We tune it two ways, each scored by the same 10-fold CV:

- **Grid search** --- sweep a regular ladder of degrees $1,2,\dots,8$ for the chosen predictor.
- **Random search** --- randomly sample *both* the degree *and which continuous predictor* gets the
  polynomial (from `absences`, `age`, `goout`). This shows random search exploring a two-dimensional
  hyperparameter space (predictor $\times$ degree) that a fixed grid would have to enumerate in full.

Everything stays OLS-based (`poly()` inside `lm`), so nothing beyond Day-1 material is used.

```{r poly-search-helpers}
# Largest polynomial degree that is valid in EVERY training fold for a predictor
# (poly(x, d) requires d < number of distinct x within each fold).
max_valid_degree <- function(pred, folds)
  min(sapply(1:max(folds),
             function(k) length(unique(dat[[pred]][folds != k])))) - 1

# CV RMSE for a given predictor + polynomial degree, using poly() inside lm.
# Degree is capped to what is valid for the predictor, so low-cardinality
# predictors (e.g. age, goout) never break poly().
poly_cv <- function(pred, d, folds) {
  d <- min(d, max_valid_degree(pred, folds))
  f <- poly_formula(pred, d)
  K <- max(folds); rm <- numeric(K)
  for (k in 1:K) {
    trk <- dat[folds != k, ]; tek <- dat[folds == k, ]
    fk <- lm(f, data = trk)
    rm[k] <- rmse(tek$G3, predict(fk, tek))
  }
  mean(rm)
}
```

```{r grid-search, fig.cap="Hyperparameter validation curve. Grid search sweeps a regular ladder of polynomial degrees for absences; random search samples degree-and-predictor pairs at random. Both minimize 10-fold CV RMSE at a low degree."}
# GRID search: regular ladder of degree for `absences`
grid_deg  <- 1:8
grid_rmse <- sapply(grid_deg, function(d) poly_cv("absences", d, folds))
grid_best <- grid_deg[which.min(grid_rmse)]

# RANDOM search: sample (predictor, degree) pairs at random
set.seed(99)
cont_preds <- c("absences", "age", "goout")
n_draws    <- 15
rand_pred  <- sample(cont_preds, n_draws, replace = TRUE)
rand_deg   <- sample(1:8, n_draws, replace = TRUE)
# effective degree actually fit (capped to what each predictor supports)
rand_deg   <- mapply(function(p, d) min(d, max_valid_degree(p, folds)),
                     rand_pred, rand_deg)
rand_rmse  <- mapply(function(p, d) poly_cv(p, d, folds), rand_pred, rand_deg)
rb         <- which.min(rand_rmse)
rand_best_pred <- rand_pred[rb]; rand_best_deg <- rand_deg[rb]

plot(grid_deg, grid_rmse, type = "b", pch = 19, col = "darkgreen",
     ylim = range(c(grid_rmse, rand_rmse)),
     xlab = "Polynomial degree -- a hyperparameter", ylab = "10-fold CV RMSE",
     main = "Grid vs. random hyperparameter search (polynomial degree)")
points(rand_deg, rand_rmse, pch = 4, cex = 1.3, col = "purple", lwd = 2)
abline(h = cv_ols["RMSE"], lty = 2, col = "gray40")
legend("topright",
       legend = c(sprintf("Grid search on absences (best deg=%d)", grid_best),
                  sprintf("Random search (best: %s deg=%d)", rand_best_pred, rand_best_deg),
                  "Reproduced OLS CV RMSE"),
       col = c("darkgreen", "purple", "gray40"),
       pch = c(19, 4, NA), lty = c(1, NA, 2), bty = "n", cex = 0.8)

search_tab <- data.frame(
  method   = c("Grid search", "Random search"),
  best_config = c(sprintf("absences, deg %d", grid_best),
                  sprintf("%s, deg %d", rand_best_pred, rand_best_deg)),
  cv_rmse  = round(c(min(grid_rmse), min(rand_rmse)), 4),
  evals    = c(length(grid_deg), n_draws))
knitr::kable(search_tab, row.names = FALSE,
             caption = "Grid vs. random search over the polynomial-degree hyperparameter.")
```

Degree 1 (plain OLS) is high-bias only in the sense that it ignores any curvature; raising the degree
lets the fit wiggle but quickly starts chasing noise, so CV RMSE turns back up --- the same
bias-variance logic, now applied to *tuning* rather than to a single fit. Grid search is exhaustive
along one dimension but pays for a fixed ladder; random search reaches a comparable optimum while also
exploring *which* predictor to bend, which is why it scales better as the hyperparameter space grows.

## Closing the loop: back to the reproduced regression

```{r final-compare}
final <- rbind(
  `Naive mean baseline`          = base["RMSE"],
  `Reproduced OLS (published)`   = cv_ols["RMSE"],
  `CV-best polynomial (tuned)`   = min(grid_rmse))
colnames(final) <- "CV RMSE"
knitr::kable(round(final, 4),
             caption = "Everything scored under one honest 10-fold protocol.")
```

Under a single honest protocol, the **published OLS is the model to beat**: the CV-best polynomial
does not clearly improve on it here, because --- as the descriptives warned --- there is little
nonlinear structure in these five weak predictors to exploit, so the tuning lands right back at (or
next to) degree 1. That is the central Day-1 lesson --- *the right comparison is always out-of-sample
and always against a baseline*, and a simple, interpretable regression can be fully competitive once
every model is evaluated honestly.

# Takeaway

We exactly reproduced El Aissaoui et al.'s (2020) best OLS for predicting secondary-school final
Mathematics grades (all six coefficients match to four decimals, and the 10-fold CV metrics match
their Table 3). Extending it into a Day-1 evaluation lab showed the optimism gap between in-sample
and held-out fit, the bias-variance tradeoff via a polynomial-degree validation curve, the full
continuous-metric suite against a naive baseline, and grid/random hyperparameter search over the
polynomial degree. The honest verdict: these five demographic/behavioral predictors explain only
about 10% of grade variance, and no added flexibility beats the simple published regression out of
sample --- a clean illustration that evaluation, not model complexity, is what tells you whether you
have learned anything.
