---
title: "Reproduce-then-Extend — Voter Turnout & Term Limits (Political Science) (Day 1)"
author: "Intro to Machine Learning · Bruce A. Desmarais"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: true
geometry: margin=1in
fontsize: 10pt
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
  - \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 3.6, fig.align = "center")
options(width = 82)
```

> **Published study.** Kuhlmann, R. & Lewis, D. C. (2017). "Legislative Term Limits and Voter Turnout."
> *State Politics & Policy Quarterly* 17(4):372–392. Data: UNC Dataverse doi:10.15139/S3/POGNLC. The original
> analysis is an **OLS** regression of state lower-chamber turnout on institutional predictors (Table 2,
> Col. 1). This example illustrates **out-of-sample prediction with a regression**.

# Background

Kuhlmann and Lewis ask whether **legislative term limits shape voter turnout** in state legislative elections. Term limits reshape the electoral environment -- forcing open-seat contests and changing candidate experience and the information available to voters -- with theoretically ambiguous effects on participation. They use **OLS** to estimate the association between the strength of a state's term limits and its lower-chamber turnout, holding constant the major institutional and demographic determinants of turnout (registration rules, professionalism, party competition, election timing, income, diversity). In the reproduced Table 2 model the **term-limitedness coefficient is positive and statistically significant** -- more term-limited legislatures see modestly higher turnout -- though the largest effects are presidential-election timing and ease of registration.

# Data and codebook

**Unit of analysis:** a U.S. state lower-chamber legislative election (a **state-year**); n = 486.

| Variable | Definition |
|---|---|
| `turnout` | lower-chamber election turnout rate, % of the voting-age population **(outcome)** |
| `tlness` | term-limitedness: strength/bindingness of legislative term limits |
| `lcr2` | log legislator-to-citizen ratio (per 100,000 residents) |
| `mmds` | share of legislative seats elected from multimember districts |
| `regdifficulty` | voter-registration closing date (days before the election) |
| `straight` | 1 if the state offers straight-party-ticket voting |
| `iuse2` | log number of ballot initiatives on the ballot |
| `squire` | Squire index of legislative professionalism |
| `ranney` | Ranney index of interparty electoral competition |
| `income2` | per-capita income (thousands of dollars) |
| `diversity3` | Herfindahl index of racial/ethnic diversity |
| `preselec` | 1 in presidential-election years |
| `redistrict` | 1 if districts were newly redrawn (post-redistricting) |
| `year` | election year |

# Descriptive analysis

Before modeling, we profile the data thoroughly: the sample, the outcome, the
predictors one at a time, missingness, each predictor's bivariate link to
turnout, and the collinearity structure. This EDA motivates every choice in the
reproduction that follows.

```{r data}
d <- read.delim(paste0(
  "https://raw.githubusercontent.com/bdesmarais/intro",
  "-ml-statistical-horizons-4day/main/tutorials/day1_",
  "intro_and_evaluation/01_turnout/data/turnout_state",
  ".tab"))
mvars <- c("turnout", "tlness", "lcr2", "mmds", "regdifficulty", "straight", "iuse2",
           "squire", "ranney", "income2", "diversity3", "preselec", "redistrict", "year")
d <- as.data.frame(na.omit(d[, mvars]))
preds <- setdiff(mvars, "turnout")
```

```{r eda-setup}
library(ggplot2)
library(dplyr)
library(tidyr)
library(e1071)
theme_set(theme_minimal(base_size = 9))

binvars <- c("straight", "preselec", "redistrict")   # 0/1 indicators
numpred  <- setdiff(preds, binvars)                  # continuous predictors
```

## Dataset and provenance

The **unit of analysis** is a U.S. state lower-chamber legislative election (a
state-year). After restricting to the modeling variables and dropping incomplete
rows we have **n = `r nrow(d)`** state-years and **`r length(mvars)`** variables:
one continuous outcome (`turnout`), `r length(numpred)` continuous predictors,
and `r length(binvars)` binary indicators. The data are Kuhlmann & Lewis's
replication file (UNC Dataverse `doi:10.15139/S3/POGNLC`), assembling state
turnout with institutional and demographic covariates across recent election
cycles.

```{r dict, echo=FALSE}
dict <- data.frame(
  variable = mvars,
  role = c("outcome", rep("predictor", length(preds))),
  type = c("continuous", "continuous", "continuous", "continuous", "continuous",
           "binary", "continuous", "continuous", "continuous", "continuous",
           "continuous", "binary", "binary", "continuous"),
  description = c(
    "lower-chamber turnout, % of voting-age population",
    "term-limitedness (strength/bindingness of limits)",
    "log legislator-to-citizen ratio (per 100k)",
    "share of seats from multimember districts",
    "registration closing date (days before election)",
    "1 = straight-party-ticket voting offered",
    "log number of ballot initiatives",
    "Squire index of legislative professionalism",
    "Ranney index of interparty competition",
    "per-capita income ($1000s)",
    "Herfindahl index of racial/ethnic diversity",
    "1 = presidential-election year",
    "1 = newly redrawn (post-redistricting)",
    "election year"))
knitr::kable(dict, caption = "Variable dictionary: role, type, and coding.")
```

## The outcome in depth

```{r outcome-stats, echo=FALSE}
y <- d$turnout
ostat <- data.frame(
  n = length(y), mean = mean(y), sd = sd(y), min = min(y),
  Q1 = quantile(y, .25), median = median(y), Q3 = quantile(y, .75),
  max = max(y), IQR = IQR(y), skew = skewness(y), kurt = kurtosis(y))
knitr::kable(round(ostat, 2), row.names = FALSE,
             caption = "Turnout (%): summary statistics.")
```

```{r outcome-plot, echo=FALSE, fig.height=2.8, fig.cap="Distribution of state lower-chamber turnout, with density overlay and mean (dashed)."}
ggplot(d, aes(turnout)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "#a6cee3", colour = "white") +
  geom_density(colour = "#1f78b4", linewidth = 0.8) +
  geom_vline(xintercept = mean(y), linetype = 2, colour = "#e31a1c") +
  labs(x = "turnout (% of voting-age population)", y = "density")
```

Turnout is close to **symmetric** (skewness `r round(skewness(y), 2)`) and
mildly **platykurtic** (kurtosis `r round(kurtosis(y), 2)`, i.e. thinner tails
than a normal), ranging from about `r round(min(y))`% to `r round(max(y))`% with
no floor/ceiling pile-up and no extreme outliers. Because the distribution is
already well-behaved, **no transformation of the outcome is warranted** — OLS on
`turnout` in its natural units is appropriate.

## Univariate predictor distributions

```{r pred-stats, echo=FALSE}
ptab <- t(sapply(d[numpred], function(x) c(
  n = sum(!is.na(x)), n_miss = sum(is.na(x)), mean = mean(x), sd = sd(x),
  min = min(x), Q1 = quantile(x, .25), median = median(x),
  Q3 = quantile(x, .75), max = max(x), skew = skewness(x))))
colnames(ptab) <- c("n", "n_miss", "mean", "sd", "min", "Q1",
                    "median", "Q3", "max", "skew")
knitr::kable(round(ptab, 2),
             caption = "Continuous predictors: summary statistics.")
```

```{r pred-hist, echo=FALSE, fig.height=3.6, fig.cap="Distributions of the continuous predictors (faceted, free x-axes)."}
d %>% select(all_of(numpred)) %>%
  pivot_longer(everything(), names_to = "var", values_to = "val") %>%
  ggplot(aes(val)) +
  geom_histogram(bins = 22, fill = "#b2df8a", colour = "white") +
  facet_wrap(~ var, scales = "free", ncol = 4) +
  labs(x = NULL, y = "count")
```

Most predictors are moderately behaved; `squire` (professionalism) and `mmds`
(multimember-district share) are right-skewed, reflecting a few highly
professional legislatures and states leaning on multimember districts. The three
binary indicators split as follows:

```{r bin-freq, echo=FALSE}
btab <- do.call(rbind, lapply(binvars, function(v) {
  t <- table(d[[v]])
  data.frame(variable = v, n_0 = t["0"], n_1 = t["1"],
             pct_1 = round(100 * mean(d[[v]]), 1))
}))
knitr::kable(btab, row.names = FALSE,
             caption = "Binary predictors: frequencies and share coded 1.")
```

```{r bin-bar, echo=FALSE, fig.height=2.2, fig.cap="Class balance of the three binary predictors."}
d %>% select(all_of(binvars)) %>%
  pivot_longer(everything(), names_to = "var", values_to = "val") %>%
  count(var, val) %>%
  ggplot(aes(factor(val), n)) +
  geom_col(fill = "#fb9a99") +
  facet_wrap(~ var) +
  labs(x = "value (0/1)", y = "count")
```

## Missing-data analysis

```{r miss, echo=FALSE}
miss <- data.frame(
  variable = mvars,
  n_missing = colSums(is.na(d[mvars])),
  pct_missing = round(100 * colMeans(is.na(d[mvars])), 2))
knitr::kable(miss, row.names = FALSE,
             caption = "Missing values per variable in the modeling frame.")
```

The modeling frame is built by `na.omit()` on the selected columns, so all
`r nrow(d)` retained state-years are **complete cases** — every variable shows
0% missing above. Any incompleteness in the raw file was list-wise deleted up
front; the reproduction therefore fits on a fully observed sample and no
imputation is needed. (Because deletion happens before analysis, any bias would
come from *which* state-years were dropped, not from missingness within the
analysis set.)

## Bivariate relationships with the outcome

For continuous predictors we correlate each with turnout; for the binary
indicators we compare group means with a two-sample t-test.

```{r marg-num, echo=FALSE}
mnum <- data.frame(
  predictor = numpred,
  r_with_turnout = sapply(d[numpred], function(x) cor(x, d$turnout)))
mnum <- mnum[order(-abs(mnum$r_with_turnout)), ]
knitr::kable(mnum, row.names = FALSE, digits = 2,
             caption = "Continuous predictors: Pearson r with turnout.")
```

```{r marg-bin, echo=FALSE}
mbin <- do.call(rbind, lapply(binvars, function(v) {
  g <- split(d$turnout, d[[v]])
  tt <- t.test(g[["1"]], g[["0"]])
  data.frame(predictor = v, mean_1 = mean(g[["1"]]), mean_0 = mean(g[["0"]]),
             diff = mean(g[["1"]]) - mean(g[["0"]]), p_value = tt$p.value)
}))
knitr::kable(mbin, row.names = FALSE, digits = 3,
             caption = "Binary predictors: turnout means by group (diff = 1 minus 0).")
```

```{r biv-plot, echo=FALSE, fig.height=3.4, fig.cap="Turnout vs the two strongest continuous predictors (with linear fit), and by presidential-year timing."}
p1 <- ggplot(d, aes(regdifficulty, turnout)) +
  geom_point(alpha = 0.35, colour = "#1f78b4") +
  geom_smooth(method = "lm", se = FALSE, colour = "#e31a1c") +
  labs(x = "registration closing (days)", y = "turnout")
p2 <- ggplot(d, aes(squire, turnout)) +
  geom_point(alpha = 0.35, colour = "#1f78b4") +
  geom_smooth(method = "lm", se = FALSE, colour = "#e31a1c") +
  labs(x = "professionalism (Squire)", y = NULL)
p3 <- ggplot(d, aes(factor(preselec), turnout)) +
  geom_boxplot(fill = "#a6cee3") +
  labs(x = "presidential year (0/1)", y = NULL)
gridExtra::grid.arrange(p1, p2, p3, nrow = 1)
```

The dominant marginal signal is **election timing**: presidential-year contests
run far higher (see the `preselec` group difference), and easier registration
(smaller `regdifficulty`) tracks higher turnout. Professionalism (`squire`) and
competition (`ranney`) are positively associated, consistent with the paper's
story. `tlness` (term-limitedness) has only a modest marginal correlation — its
role emerges once the institutional controls are held constant in the regression.

## Multivariate structure and collinearity

```{r corr-heat, echo=FALSE, fig.height=4.2, fig.cap="Correlation heatmap among continuous predictors."}
cm <- cor(d[numpred])
cmdf <- as.data.frame(as.table(cm))
names(cmdf) <- c("v1", "v2", "r")
ggplot(cmdf, aes(v1, v2, fill = r)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = round(r, 2)), size = 2.4) +
  scale_fill_gradient2(low = "#e31a1c", mid = "white", high = "#1f78b4",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r vif, echo=FALSE}
ols0 <- lm(turnout ~ ., data = d)
X <- model.matrix(ols0)[, -1]
vif <- sapply(colnames(X), function(v) {
  r2 <- summary(lm(X[, v] ~ X[, setdiff(colnames(X), v)]))$r.squared
  1 / (1 - r2)
})
vtab <- data.frame(predictor = names(vif), VIF = as.numeric(vif))
vtab <- vtab[order(-vtab$VIF), ]
knitr::kable(vtab, row.names = FALSE, digits = 2,
             caption = "Variance inflation factors from the reproduction model.")
```

One correlation exceeds the **|r| > 0.7** flag: `income2` and `year`
(r `r round(cor(d$income2, d$year), 2)`) — per-capita income rises secularly over
the sample window — and these two carry the largest VIFs (~`r round(max(vif), 1)`
and ~`r round(sort(vif, decreasing = TRUE)[2], 1)`). All other VIFs sit below
2.5. The collinearity is mild and confined to the income/time pair, so OLS
coefficients on the substantive institutional predictors remain interpretable;
it is worth noting only because it can inflate the standard errors on `income2`
and `year` specifically. This motivates the regularized comparisons introduced
in later tutorials, where correlated predictors are handled by shrinkage.

# Reproduce the published regression

We fit their exact Table 2, Column 1 specification.

```{r reproduce}
ols <- lm(turnout ~ tlness + lcr2 + mmds + regdifficulty + straight + iuse2 + squire +
            ranney + income2 + diversity3 + preselec + redistrict + year, data = d)
summary(ols)
```

**Confirmation against the published study.** Fitting their specification recovers the published model:
`r nrow(d)` state-years and in-sample $R^2$ = `r round(summary(ols)$r.squared, 3)`. Presidential-election years
raise turnout by ~12.8 points (`preselec`), stricter registration deadlines lower it (`regdifficulty`), and
professionalism (`squire`) and competition (`ranney`) enter positively — reproducing the reported signs and
significance.

# Out-of-sample prediction with the regression

The same OLS that *explains* turnout can also *predict* it. We hold out a random 30% test set, fit on the
other 70%, and evaluate on the unseen states — the core train/test workflow.

```{r oos}
set.seed(2026)
train <- sample(nrow(d), floor(0.7 * nrow(d)))
test  <- setdiff(seq_len(nrow(d)), train)

fit    <- lm(turnout ~ ., data = d[train, ])
pred   <- predict(fit, d[test, ])
actual <- d$turnout[test]

# out-of-sample error and R^2 (relative to the training-set mean)
rmse <- sqrt(mean((actual - pred)^2))
r2   <- 1 - sum((actual - pred)^2) / sum((actual - mean(d$turnout[train]))^2)
cat(sprintf("Train n=%d, test n=%d;  out-of-sample RMSE=%.2f points,  out-of-sample R^2=%.3f\n",
            length(train), length(test), rmse, r2))
```

```{r oos-plot, echo=FALSE, fig.height=3.4, fig.cap="Predicted vs actual turnout on the held-out test set; the red line marks perfect prediction."}
par(mar = c(4.2, 4.2, 2.5, 1))
plot(pred, actual, pch = 16, col = "#1f78b455", xlab = "predicted (held-out test set)",
     ylab = "actual turnout", main = "Out-of-sample prediction")
abline(0, 1, col = "#e31a1c", lwd = 2)
```

The held-out $R^2 \approx$ `r round(r2, 2)` is the **honest** estimate of predictive accuracy — scored on data
the model never saw, rather than the in-sample $R^2 = 0.72$ that always flatters the fit. The points cluster
around the 45° line, so the regression predicts turnout in unseen states reasonably well.

# Takeaway

The published OLS is the right tool for the paper's *explanatory* question, and the *same model*, scored on a
held-out test set, gives a calibrated estimate of how well turnout can be **predicted** out of sample.
Explanation and prediction are different goals, and the train/test split is how we measure the second. (Later
examples extend this baseline with polynomial terms, regularization, and tree-based learners.)


# Recommended exercises

1. Re-run the train/test split under several random seeds (or switch to 5-fold cross-validation) and report the spread in out-of-sample RMSE. How stable is the estimate?
2. Add a quadratic term in `income2` (or an interaction between `preselec` and `regdifficulty`) and check whether held-out RMSE improves.
3. Fit a cross-validated lasso on the same predictors and compare its out-of-sample RMSE and its selected variables to OLS.
