---
title: "Reproduce-then-Extend — Ames House Prices (Real Estate / Hedonic Economics) (Day 1)"
author: "Intro to Machine Learning · Bruce A. Desmarais"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: true
geometry: margin=1in
fontsize: 10pt
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
  - \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 3.6, fig.align = "center")
options(width = 80)
library(glmnet)     # lasso / ridge / elastic net
library(ggplot2)    # EDA plots
library(dplyr)      # data wrangling
library(tidyr)      # reshaping for faceted plots
library(knitr)      # kable tables
library(patchwork)  # compose ggplots
library(e1071)      # skewness / kurtosis
theme_set(theme_minimal(base_size = 9))
```

> **Published study.** De Cock, D. (2011). "Ames, Iowa: Alternative to the Boston Housing Data as an End of
> Semester Regression Project." *Journal of Statistics Education* 19(3). The analysis is an **OLS hedonic
> price regression** of sale price on property features (n = 2930; data at the JSE site). This example
> illustrates **regularization and variable selection** in a wide model.

# Background

De Cock introduces the **Ames housing dataset** as a modern, richer alternative to the classic Boston housing data for teaching regression. The intended use of the data is a **hedonic price regression** -- an OLS model that decomposes a home's **sale price** into the contributions of its individual characteristics (size, quality, age, amenities). The **primary conclusion** is methodological: a well-specified hedonic OLS explains the large majority of the variation in sale price ($R^2$ around 0.9), and the data support a full end-of-semester regression project from specification through diagnostics.

# Data and codebook

**Unit of analysis:** a single-family residential **home sale**; n = 2,930. The printed hedonic regression
uses ten interpretable features; the regularization section uses De Cock's full ~80-feature set.

| Variable | Definition |
|---|---|
| `SalePrice` | sale price in USD **(outcome)** |
| `Overall.Qual` | overall material-and-finish quality (1-10) |
| `Gr.Liv.Area` | above-grade living area (sq ft) |
| `Total.Bsmt.SF` | total basement area (sq ft) |
| `Garage.Cars` | garage capacity (number of cars) |
| `Year.Built` | year of original construction |
| `Year.Remod.Add` | year of the last remodel/addition |
| `Full.Bath` | full bathrooms above grade |
| `TotRms.AbvGrd` | total rooms above grade (excludes bathrooms) |
| `Lot.Area` | lot size (sq ft) |
| `Fireplaces` | number of fireplaces |

# Descriptive analysis

We load the data exactly as the reproduction does, then profile it thoroughly before
fitting anything. The read, the median-imputation of the ten curated features, and the
complete-case filter below are the *published* pipeline and are left untouched; the EDA
that follows is computed on the resulting data frame `d`.

```{r data}
d <- read.delim(paste0(
  "https://raw.githubusercontent.com/bdesmarais/intro",
  "-ml-statistical-horizons-4day/main/tutorials/day3_",
  "regularization_and_selection/08_ames_housing/data/",
  "AmesHousing.txt"),
                stringsAsFactors = FALSE, check.names = TRUE)
feats <- c("Overall.Qual", "Gr.Liv.Area", "Total.Bsmt.SF", "Garage.Cars", "Year.Built",
           "Year.Remod.Add", "Full.Bath", "TotRms.AbvGrd", "Lot.Area", "Fireplaces")
for (j in feats) d[[j]][is.na(d[[j]])] <- median(d[[j]], na.rm = TRUE)   # a few missing values
d <- d[complete.cases(d[, c("SalePrice", feats)]), ]
```

## Dataset and provenance

The **unit of analysis** is a single residential **home sale** in Ames, Iowa recorded by the
City Assessor's Office over **2006-2010**; it is a near-census of transactions in that window,
not a random sample. After the published complete-case filter on the ten curated features we
retain **`r nrow(d)` sales** described by **`r ncol(d)` raw columns** (`r sum(sapply(d, is.numeric))`
numeric, `r sum(sapply(d, is.character))` character). Two columns (`Order`, `PID`) are record
identifiers and are dropped before modeling.

De Cock's intended use is a **hedonic price regression**: decompose `SalePrice` into the marginal
contributions of a home's physical characteristics. The reproduction below uses ten interpretable
features; the regularization section expands to the full ~80-feature set (one-hot encoded to ~250
columns), which is why this is a *wide*-model illustration.

The **variable dictionary** for the outcome and the ten curated predictors:

```{r dict}
dict <- data.frame(
  Variable = c("SalePrice", feats),
  Role = c("outcome", rep("predictor", length(feats))),
  Type = c("numeric ($)", "ordinal 1-10", "numeric (sq ft)", "numeric (sq ft)",
           "count", "year", "year", "count", "count", "numeric (sq ft)", "count"),
  Description = c(
    "sale price in USD",
    "overall material-and-finish quality (1 = poor, 10 = excellent)",
    "above-grade living area (sq ft)",
    "total basement area (sq ft)",
    "garage capacity (number of cars)",
    "year of original construction",
    "year of last remodel/addition",
    "full bathrooms above grade",
    "total rooms above grade (excludes bathrooms)",
    "lot size (sq ft)",
    "number of fireplaces"))
kable(dict, caption = "Variable dictionary: outcome and curated hedonic predictors.")
```

## The outcome: SalePrice

```{r y-stats}
y_summary <- function(x) {
  q <- quantile(x, c(.25, .5, .75))
  data.frame(n = length(x), mean = mean(x), sd = sd(x), min = min(x),
             Q1 = q[1], median = q[2], Q3 = q[3], max = max(x),
             IQR = q[3] - q[1], skew = e1071::skewness(x),
             kurt = e1071::kurtosis(x))
}
kable(round(y_summary(d$SalePrice)), row.names = FALSE,
      caption = "SalePrice distribution summary (USD).")
```

```{r y-dist, fig.height=2.7, fig.cap="SalePrice on the raw dollar scale (left, right-skewed) and after a log10 transform (right, near-symmetric)."}
p1 <- ggplot(d, aes(SalePrice)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = "#a6cee3", colour = "white") +
  geom_density(colour = "#1f78b4", linewidth = .6) +
  scale_x_continuous(labels = function(z) z / 1000) +
  labs(x = "sale price ($000)", y = "density", title = "Raw scale")
p2 <- ggplot(d, aes(log10(SalePrice))) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = "#b2df8a", colour = "white") +
  geom_density(colour = "#33a02c", linewidth = .6) +
  labs(x = "log10(sale price)", y = "density", title = "Log scale")
p1 + p2
```

`SalePrice` ranges from about \$13k to \$755k with a mean near \$180k above the median (~\$160k),
the classic signature of a **right-skewed** price variable (skewness `r round(e1071::skewness(d$SalePrice), 2)`,
excess kurtosis `r round(e1071::kurtosis(d$SalePrice), 2)` — a heavy upper tail of expensive homes).
A **log transform** removes almost all of the skew (right panel), which is why hedonic practice
routinely models `log(price)`. The reproduction below follows De Cock and keeps the **raw dollar
scale** so coefficients read as dollars-per-unit; the skew is worth flagging because it inflates
RMSE through the high-price tail and rewards models that fit the bulk well.

## Univariate predictor distributions

```{r x-stats}
skew_tab <- function(x) {
  q <- quantile(x, c(.25, .5, .75))
  c(n = length(x), n_miss = sum(is.na(x)), mean = mean(x), sd = sd(x),
    min = min(x), Q1 = q[1], median = q[2], Q3 = q[3], max = max(x),
    skew = e1071::skewness(x))
}
xtab <- t(sapply(d[feats], skew_tab))
kable(round(xtab, 1), caption = "Summary statistics for the ten curated predictors.")
```

```{r x-dist, fig.height=3.6, fig.cap="Distributions of the ten curated numeric predictors (post-imputation)."}
d %>%
  select(all_of(feats)) %>%
  pivot_longer(everything(), names_to = "var", values_to = "val") %>%
  ggplot(aes(val)) +
  geom_histogram(bins = 25, fill = "#1f78b4", colour = "white") +
  facet_wrap(~ var, scales = "free", ncol = 5) +
  labs(x = NULL, y = "count") +
  theme(axis.text = element_text(size = 6))
```

Several predictors carry the same right skew as the outcome — `Lot.Area`, `Gr.Liv.Area`, and
`Total.Bsmt.SF` all have long upper tails (a few very large properties), while `Overall.Qual`
is roughly bell-shaped around 5-6 and the counts (`Full.Bath`, `Garage.Cars`, `Fireplaces`,
`TotRms.AbvGrd`) are discrete and multi-modal. `Year.Built`/`Year.Remod.Add` show the vintage
mix of the housing stock. These skews reinforce that linear coefficients describe average
marginal effects across a heterogeneous stock.

## Missing data

The curated ten are essentially complete, but the **full** dataset is not — and its missingness
is the well-known Ames pattern:

```{r missing}
draw <- read.delim(paste0(
  "https://raw.githubusercontent.com/bdesmarais/intro",
  "-ml-statistical-horizons-4day/main/tutorials/day3_",
  "regularization_and_selection/08_ames_housing/data/",
  "AmesHousing.txt"),
                   stringsAsFactors = FALSE, check.names = TRUE)
na_ct <- sapply(draw, function(x) sum(is.na(x)))
na_ct <- sort(na_ct[na_ct > 0], decreasing = TRUE)
miss <- data.frame(variable = names(na_ct), n_missing = as.integer(na_ct),
                   pct_missing = round(100 * na_ct / nrow(draw), 1))
kable(head(miss, 12), row.names = FALSE,
      caption = "Twelve most-missing columns in the full dataset (of 25 with any NA).")
```

The heavily "missing" columns are **structurally absent features, not lost data**: `Pool.QC` is
NA for the `r sum(is.na(draw$Pool.QC))` homes with no pool, `Alley` for homes with no alley
access, `Fireplace.Qu` where there is no fireplace, and the `Garage.*` block where there is no
garage. The mechanism is therefore **missing-not-at-random by construction** — the NA *is* the
information ("none"). Only `Lot.Frontage` (`r round(100*mean(is.na(draw$Lot.Frontage)),1)`%) is
plausibly missing-at-random. The reproduction's pipeline handles this by **median-imputing** the
few NAs in the numeric curated features, and the regularization section recodes character NAs to
an explicit `"NA"` level (so "no garage" becomes its own category) before one-hot encoding.

## Bivariate relationships with SalePrice

```{r biv-tab}
assoc <- sapply(d[feats], function(x) cor(x, d$SalePrice))
bt <- data.frame(predictor = feats,
                 r_with_price = round(as.numeric(assoc), 3),
                 abs_r = round(abs(as.numeric(assoc)), 3))
bt <- bt[order(-bt$abs_r), ]
kable(bt, row.names = FALSE,
      caption = "Pearson correlation of each predictor with SalePrice, sorted.")
```

```{r biv-plot, fig.height=3.6, fig.cap="SalePrice against the four predictors most correlated with it, with linear smoothers and r."}
top4 <- bt$predictor[1:4]
d %>%
  select(SalePrice, all_of(top4)) %>%
  pivot_longer(-SalePrice, names_to = "var", values_to = "val") %>%
  ggplot(aes(val, SalePrice / 1000)) +
  geom_point(alpha = .15, colour = "#1f78b4", size = .5) +
  geom_smooth(method = "lm", se = FALSE, colour = "#e31a1c", linewidth = .7) +
  facet_wrap(~ var, scales = "free_x", ncol = 2) +
  labs(x = NULL, y = "sale price ($000)")
```

**Overall quality** and **above-grade living area** are the dominant correlates of price
(r > 0.7 each), followed by garage capacity, basement size, and construction year — exactly the
features a buyer would price. The relationships are close to linear on the dollar scale over the
bulk of the range, curving up only in the expensive tail, which supports the OLS specification
while foreshadowing why the high-price homes drive the residual variance.

## Multivariate structure and collinearity

```{r corr-heat, fig.height=3.8, fig.cap="Correlation heatmap among SalePrice and the ten curated predictors."}
cm <- cor(d[c("SalePrice", feats)])
cm[lower.tri(cm)] <- NA
as.data.frame(cm) %>%
  mutate(row = rownames(cm)) %>%
  pivot_longer(-row, names_to = "col", values_to = "r") %>%
  filter(!is.na(r)) %>%
  mutate(row = factor(row, levels = rownames(cm)),
         col = factor(col, levels = rownames(cm))) %>%
  ggplot(aes(col, row, fill = r)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 2.1) +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
        axis.text.y = element_text(size = 6))
```

```{r vif}
# VIF for each curated predictor = 1 / (1 - R^2) from regressing it on the others.
vif_calc <- function(vars, data) {
  sapply(vars, function(v) {
    r2 <- summary(lm(reformulate(setdiff(vars, v), v), data = data))$r.squared
    1 / (1 - r2)
  })
}
vifs <- vif_calc(feats, d)
kable(data.frame(predictor = names(vifs), VIF = round(as.numeric(vifs), 2)),
      row.names = FALSE, caption = "Variance inflation factors (curated hedonic model).")
```

No pair of curated predictors is alarmingly collinear (all |r| well below 0.7 except the
size-related features, and every **VIF is under ~3**), so the ten-feature OLS is stable and its
coefficients interpretable. Collinearity becomes a real problem only in the **full ~250-column
design** — many size, quality, and neighborhood indicators are mutually correlated — which is
precisely the regime where unpenalized OLS over-fits and **regularization** earns its keep. The
descriptives thus motivate the whole tutorial: a clean, interpretable ten-feature hedonic model
reproduces the published fit, while the wide design demands a penalty.

# Reproduce the published regression

A standard hedonic OLS on interpretable features (De Cock's regression-project design).

```{r reproduce}
hedonic <- lm(SalePrice ~ ., data = d[, c("SalePrice", feats)])
summary(hedonic)
```

**Confirmation against the published study.** In-sample $R^2$ =
`r round(summary(hedonic)$r.squared, 3)`, with **overall quality** and **above-grade living area** the dominant
positive drivers of price. (Adding the full ~80-feature set with neighborhood indicators raises the OLS to
$R^2 \approx 0.92$, matching De Cock's larger model.)

# Regularization & variable selection

Regularization's natural home is a **wide** model. We expand to the full feature set — every numeric and
one-hot-encoded categorical predictor (~250 columns) — where ordinary least squares over-fits.

```{r build-full}
# build the full design matrix: numeric features as-is, categoricals one-hot encoded
full <- d[, !names(d) %in% c("Order", "PID")]
y    <- full$SalePrice; full$SalePrice <- NULL
for (j in names(full)) {
  if (is.character(full[[j]])) { full[[j]][is.na(full[[j]])] <- "NA"; full[[j]] <- factor(full[[j]]) }
  else full[[j]][is.na(full[[j]])] <- median(full[[j]], na.rm = TRUE)
}
X <- model.matrix(~ . - 1, data = full)
X <- X[, apply(X, 2, function(z) length(unique(z)) > 1)]   # drop constant columns
cat(sprintf("Full design matrix: %d homes x %d features\n", nrow(X), ncol(X)))
```

```{r fit-reg}
set.seed(2026)
lasso <- cv.glmnet(X, y, alpha = 1)     # L1
ridge <- cv.glmnet(X, y, alpha = 0)     # L2
enet  <- cv.glmnet(X, y, alpha = 0.5)   # L1 + L2
n_keep <- sum(coef(lasso, s = "lambda.1se")[-1] != 0)   # nonzero features at the 1-SE penalty
cat(sprintf("At lambda.1se the lasso keeps %d of %d features (the rest are set to 0); ridge keeps all but shrinks them.\n",
            n_keep, ncol(X)))
```

**Which fits best out of sample?** We hold out a random 30% test set, fit each model on the training 70%
(penalty tuned by CV inside the training data), and compare held-out **RMSE** (dollars) on the untouched test
set; we **repeat the split 10 times and average**. Penalized models are scored at `lambda.min`, the
loss-minimizing penalty (the right choice for *prediction*; the sparser `lambda.1se` used for the coefficient
display trades a little accuracy for parsimony). On ~250 features unpenalized OLS over-fits badly.

```{r reg-perf}
rmse   <- function(p, yt) sqrt(mean((yt - p)^2))
REPS   <- 10
alphas <- c(Lasso = 1, Ridge = 0, ElasticNet = 0.5)
Xdf <- as.data.frame(X)
M <- matrix(NA, REPS, 4, dimnames = list(NULL, c("OLS", names(alphas))))
for (r in 1:REPS) {
  set.seed(2025 + r)
  te <- sample(nrow(X), round(0.30 * nrow(X))); tr <- setdiff(seq_len(nrow(X)), te)
  ols <- lm(y[tr] ~ ., data = Xdf[tr, ])                                       # baseline OLS
  M[r, "OLS"] <- rmse(predict(ols, Xdf[te, ]), y[te])
  for (a in names(alphas)) {
    cv <- cv.glmnet(X[tr, ], y[tr], alpha = alphas[a])
    M[r, a] <- rmse(as.numeric(predict(cv, X[te, ], s = "lambda.min")), y[te])
  }
}
round(colMeans(M))                                            # mean held-out RMSE ($) over 10 splits
best <- names(which.min(colMeans(M)[names(alphas)]))
cat(sprintf("\nBest penalization (10 splits): %s (mean test RMSE = $%s vs OLS $%s).\n",
            best, format(round(min(colMeans(M)[names(alphas)])), big.mark = ","),
            format(round(colMeans(M)["OLS"]), big.mark = ",")))
```

On ~250 features OLS over-fits, but lasso/ridge/elastic-net keep only a few dozen effective predictors and
**cut held-out RMSE well below OLS** — regularization buying both parsimony and accuracy. This is the setting
(wide, collinear design) where a penalty pays off in prediction, in contrast to compact, curated models with
large samples where a penalty mainly buys parsimony rather than accuracy.

```{r reg-path, echo=FALSE, fig.height=3.4, fig.cap="Lasso coefficient paths vs the penalty (log lambda) on the full Ames feature set; the dashed line marks the CV-optimal penalty."}
par(mar = c(4.2, 4.2, 3, 1))
plot(lasso$glmnet.fit, xvar = "lambda", col = "#1f78b433")
abline(v = log(lasso$lambda.min), lty = 2, col = "#e31a1c")
```

To see the shrinkage concretely, we refit the three penalties on the **ten interpretable features** and place
the penalized coefficients next to the original OLS (dollars per unit):

```{r reg-coef}
Xi <- as.matrix(d[, feats])
li <- cv.glmnet(Xi, y, alpha = 1); ri <- cv.glmnet(Xi, y, alpha = 0); ei <- cv.glmnet(Xi, y, alpha = 0.5)
coefs_at_1se <- function(cv) as.numeric(coef(cv, s = "lambda.1se"))[-1]   # 1-SE penalty
data.frame(
  OLS        = round(coef(hedonic)[-1]),
  Lasso      = round(coefs_at_1se(li)),
  Ridge      = round(coefs_at_1se(ri)),
  ElasticNet = round(coefs_at_1se(ei)),
  row.names = feats)
```

Each penalty pulls the OLS dollar-coefficients toward zero (ridge most uniformly); at the 1-SE penalty the
lasso drops the weakest of the ten to exactly zero, and on the full ~250-column model above it zeroes the
large majority of columns.

# Best-subset selection with ABESS

Lasso reaches a sparse model *indirectly* through shrinkage. **Best-subset selection** instead searches
directly for the subset of features that fits best, and — via the **adaptive best-subset (ABESS)** algorithm —
does so in polynomial time even on this wide design, choosing the subset size automatically.

```{r abess}
library(abess)
ab <- abess(X, y)                                  # X = full ~250-feature design matrix
cat(sprintf("ABESS selects a best subset of %d of %d features. Top few by |coefficient|:\n",
            ab$best.size, ncol(X)))
cf  <- as.numeric(coef(ab, support.size = ab$best.size))[-1]
top <- order(abs(cf), decreasing = TRUE)[1:8]
data.frame(feature = colnames(X)[top], coef = round(cf[top]))
```

Best subset and the $L_1$ path agree on the price drivers — overall quality, living area, and the strongest
neighborhood and condition indicators — reached by direct search rather than shrinkage.

# Takeaway

This example illustrates **regularization and variable selection** in the setting it was built for: a wide
hedonic model where unpenalized OLS over-fits. Lasso/elastic-net (`glmnet`) and best-subset selection (`abess`)
each pick a compact, accurate model from ~250 candidate features — the principled way to handle many
correlated predictors.


# Recommended exercises

1. Compare lasso, ridge, and elastic net out-of-sample RMSE on identical folds; which wins?
2. Add interaction or polynomial terms for a couple of strong predictors and see whether penalized RMSE improves.
3. Plot the lasso coefficient path and note the order in which variables enter the model.
