---
title: "SH Day 3: Regularization & Variable Selection — Voter Turnout & Term Limits (Political Science)"
author: "Intro to Machine Learning · Bruce A. Desmarais"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: true
geometry: margin=1in
fontsize: 10pt
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
  - \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 3.6, fig.align = "center")
options(width = 82)
```

> **Published study.** Kuhlmann, R. & Lewis, D. C. (2017). "Legislative Term Limits and Voter Turnout."
> *State Politics & Policy Quarterly* 17(4):372–392. Data: UNC Dataverse doi:10.15139/S3/POGNLC. The original
> analysis is an **OLS** regression of state lower-chamber turnout on institutional predictors (Table 2,
> Col. 1). We reproduce that model, then extend it with **regularization and variable selection** —
> ridge, lasso, elastic net (`glmnet`) and best-subset selection (`abess`) — on the same outcome and
> the same 13 predictors.

# Background

Kuhlmann and Lewis ask whether **legislative term limits shape voter turnout** in state legislative elections. Term limits reshape the electoral environment -- forcing open-seat contests and changing candidate experience and the information available to voters -- with theoretically ambiguous effects on participation. They use **OLS** to estimate the association between the strength of a state's term limits and its lower-chamber turnout, holding constant the major institutional and demographic determinants of turnout (registration rules, professionalism, party competition, election timing, income, diversity). In the reproduced Table 2 model the **term-limitedness coefficient is positive and statistically significant** -- more term-limited legislatures see modestly higher turnout -- though the largest effects are presidential-election timing and ease of registration.

# Data and codebook

**Unit of analysis:** a U.S. state lower-chamber legislative election (a **state-year**); n = 486.

| Variable | Definition |
|---|---|
| `turnout` | lower-chamber election turnout rate, % of the voting-age population **(outcome)** |
| `tlness` | term-limitedness: strength/bindingness of legislative term limits |
| `lcr2` | log legislator-to-citizen ratio (per 100,000 residents) |
| `mmds` | share of legislative seats elected from multimember districts |
| `regdifficulty` | voter-registration closing date (days before the election) |
| `straight` | 1 if the state offers straight-party-ticket voting |
| `iuse2` | log number of ballot initiatives on the ballot |
| `squire` | Squire index of legislative professionalism |
| `ranney` | Ranney index of interparty electoral competition |
| `income2` | per-capita income (thousands of dollars) |
| `diversity3` | Herfindahl index of racial/ethnic diversity |
| `preselec` | 1 in presidential-election years |
| `redistrict` | 1 if districts were newly redrawn (post-redistricting) |
| `year` | election year |

# Descriptive analysis

Before modeling, we profile the data thoroughly: the sample, the outcome, the
predictors one at a time, missingness, each predictor's bivariate link to
turnout, and the collinearity structure. This EDA motivates every choice in the
reproduction that follows.

```{r data}
d <- read.delim(paste0(
  "https://raw.githubusercontent.com/bdesmarais/intro",
  "-ml-statistical-horizons-4day/main/tutorials/day3_",
  "regularization_and_selection/01_turnout/data/",
  "turnout_state.tab"))
mvars <- c("turnout", "tlness", "lcr2", "mmds", "regdifficulty", "straight", "iuse2",
           "squire", "ranney", "income2", "diversity3", "preselec", "redistrict", "year")
d <- as.data.frame(na.omit(d[, mvars]))
preds <- setdiff(mvars, "turnout")
```

```{r eda-setup}
library(ggplot2)
library(dplyr)
library(tidyr)
library(e1071)
theme_set(theme_minimal(base_size = 9))

binvars <- c("straight", "preselec", "redistrict")   # 0/1 indicators
numpred  <- setdiff(preds, binvars)                  # continuous predictors
```

## Dataset and provenance

The **unit of analysis** is a U.S. state lower-chamber legislative election (a
state-year). After restricting to the modeling variables and dropping incomplete
rows we have **n = `r nrow(d)`** state-years and **`r length(mvars)`** variables:
one continuous outcome (`turnout`), `r length(numpred)` continuous predictors,
and `r length(binvars)` binary indicators. The data are Kuhlmann & Lewis's
replication file (UNC Dataverse `doi:10.15139/S3/POGNLC`), assembling state
turnout with institutional and demographic covariates across recent election
cycles.

```{r dict, echo=FALSE}
dict <- data.frame(
  variable = mvars,
  role = c("outcome", rep("predictor", length(preds))),
  type = c("continuous", "continuous", "continuous", "continuous", "continuous",
           "binary", "continuous", "continuous", "continuous", "continuous",
           "continuous", "binary", "binary", "continuous"),
  description = c(
    "lower-chamber turnout, % of voting-age population",
    "term-limitedness (strength/bindingness of limits)",
    "log legislator-to-citizen ratio (per 100k)",
    "share of seats from multimember districts",
    "registration closing date (days before election)",
    "1 = straight-party-ticket voting offered",
    "log number of ballot initiatives",
    "Squire index of legislative professionalism",
    "Ranney index of interparty competition",
    "per-capita income ($1000s)",
    "Herfindahl index of racial/ethnic diversity",
    "1 = presidential-election year",
    "1 = newly redrawn (post-redistricting)",
    "election year"))
knitr::kable(dict, caption = "Variable dictionary: role, type, and coding.")
```

## The outcome in depth

```{r outcome-stats, echo=FALSE}
y <- d$turnout
ostat <- data.frame(
  n = length(y), mean = mean(y), sd = sd(y), min = min(y),
  Q1 = quantile(y, .25), median = median(y), Q3 = quantile(y, .75),
  max = max(y), IQR = IQR(y), skew = skewness(y), kurt = kurtosis(y))
knitr::kable(round(ostat, 2), row.names = FALSE,
             caption = "Turnout (%): summary statistics.")
```

```{r outcome-plot, echo=FALSE, fig.height=2.8, fig.cap="Distribution of state lower-chamber turnout, with density overlay and mean (dashed)."}
ggplot(d, aes(turnout)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "#a6cee3", colour = "white") +
  geom_density(colour = "#1f78b4", linewidth = 0.8) +
  geom_vline(xintercept = mean(y), linetype = 2, colour = "#e31a1c") +
  labs(x = "turnout (% of voting-age population)", y = "density")
```

Turnout is close to **symmetric** (skewness `r round(skewness(y), 2)`) and
mildly **platykurtic** (kurtosis `r round(kurtosis(y), 2)`, i.e. thinner tails
than a normal), ranging from about `r round(min(y))`% to `r round(max(y))`% with
no floor/ceiling pile-up and no extreme outliers. Because the distribution is
already well-behaved, **no transformation of the outcome is warranted** — OLS on
`turnout` in its natural units is appropriate.

## Univariate predictor distributions

```{r pred-stats, echo=FALSE}
ptab <- t(sapply(d[numpred], function(x) c(
  n = sum(!is.na(x)), n_miss = sum(is.na(x)), mean = mean(x), sd = sd(x),
  min = min(x), Q1 = quantile(x, .25), median = median(x),
  Q3 = quantile(x, .75), max = max(x), skew = skewness(x))))
colnames(ptab) <- c("n", "n_miss", "mean", "sd", "min", "Q1",
                    "median", "Q3", "max", "skew")
knitr::kable(round(ptab, 2),
             caption = "Continuous predictors: summary statistics.")
```

```{r pred-hist, echo=FALSE, fig.height=3.6, fig.cap="Distributions of the continuous predictors (faceted, free x-axes)."}
d %>% select(all_of(numpred)) %>%
  pivot_longer(everything(), names_to = "var", values_to = "val") %>%
  ggplot(aes(val)) +
  geom_histogram(bins = 22, fill = "#b2df8a", colour = "white") +
  facet_wrap(~ var, scales = "free", ncol = 4) +
  labs(x = NULL, y = "count")
```

Most predictors are moderately behaved; `squire` (professionalism) and `mmds`
(multimember-district share) are right-skewed, reflecting a few highly
professional legislatures and states leaning on multimember districts. The three
binary indicators split as follows:

```{r bin-freq, echo=FALSE}
btab <- do.call(rbind, lapply(binvars, function(v) {
  t <- table(d[[v]])
  data.frame(variable = v, n_0 = t["0"], n_1 = t["1"],
             pct_1 = round(100 * mean(d[[v]]), 1))
}))
knitr::kable(btab, row.names = FALSE,
             caption = "Binary predictors: frequencies and share coded 1.")
```

```{r bin-bar, echo=FALSE, fig.height=2.2, fig.cap="Class balance of the three binary predictors."}
d %>% select(all_of(binvars)) %>%
  pivot_longer(everything(), names_to = "var", values_to = "val") %>%
  count(var, val) %>%
  ggplot(aes(factor(val), n)) +
  geom_col(fill = "#fb9a99") +
  facet_wrap(~ var) +
  labs(x = "value (0/1)", y = "count")
```

## Missing-data analysis

```{r miss, echo=FALSE}
miss <- data.frame(
  variable = mvars,
  n_missing = colSums(is.na(d[mvars])),
  pct_missing = round(100 * colMeans(is.na(d[mvars])), 2))
knitr::kable(miss, row.names = FALSE,
             caption = "Missing values per variable in the modeling frame.")
```

The modeling frame is built by `na.omit()` on the selected columns, so all
`r nrow(d)` retained state-years are **complete cases** — every variable shows
0% missing above. Any incompleteness in the raw file was list-wise deleted up
front; the reproduction therefore fits on a fully observed sample and no
imputation is needed. (Because deletion happens before analysis, any bias would
come from *which* state-years were dropped, not from missingness within the
analysis set.)

## Bivariate relationships with the outcome

For continuous predictors we correlate each with turnout; for the binary
indicators we compare group means with a two-sample t-test.

```{r marg-num, echo=FALSE}
mnum <- data.frame(
  predictor = numpred,
  r_with_turnout = sapply(d[numpred], function(x) cor(x, d$turnout)))
mnum <- mnum[order(-abs(mnum$r_with_turnout)), ]
knitr::kable(mnum, row.names = FALSE, digits = 2,
             caption = "Continuous predictors: Pearson r with turnout.")
```

```{r marg-bin, echo=FALSE}
mbin <- do.call(rbind, lapply(binvars, function(v) {
  g <- split(d$turnout, d[[v]])
  tt <- t.test(g[["1"]], g[["0"]])
  data.frame(predictor = v, mean_1 = mean(g[["1"]]), mean_0 = mean(g[["0"]]),
             diff = mean(g[["1"]]) - mean(g[["0"]]), p_value = tt$p.value)
}))
knitr::kable(mbin, row.names = FALSE, digits = 3,
             caption = "Binary predictors: turnout means by group (diff = 1 minus 0).")
```

```{r biv-plot, echo=FALSE, fig.height=3.4, fig.cap="Turnout vs the two strongest continuous predictors (with linear fit), and by presidential-year timing."}
p1 <- ggplot(d, aes(regdifficulty, turnout)) +
  geom_point(alpha = 0.35, colour = "#1f78b4") +
  geom_smooth(method = "lm", se = FALSE, colour = "#e31a1c") +
  labs(x = "registration closing (days)", y = "turnout")
p2 <- ggplot(d, aes(squire, turnout)) +
  geom_point(alpha = 0.35, colour = "#1f78b4") +
  geom_smooth(method = "lm", se = FALSE, colour = "#e31a1c") +
  labs(x = "professionalism (Squire)", y = NULL)
p3 <- ggplot(d, aes(factor(preselec), turnout)) +
  geom_boxplot(fill = "#a6cee3") +
  labs(x = "presidential year (0/1)", y = NULL)
gridExtra::grid.arrange(p1, p2, p3, nrow = 1)
```

The dominant marginal signal is **election timing**: presidential-year contests
run far higher (see the `preselec` group difference), and easier registration
(smaller `regdifficulty`) tracks higher turnout. Professionalism (`squire`) and
competition (`ranney`) are positively associated, consistent with the paper's
story. `tlness` (term-limitedness) has only a modest marginal correlation — its
role emerges once the institutional controls are held constant in the regression.

## Multivariate structure and collinearity

```{r corr-heat, echo=FALSE, fig.height=4.2, fig.cap="Correlation heatmap among continuous predictors."}
cm <- cor(d[numpred])
cmdf <- as.data.frame(as.table(cm))
names(cmdf) <- c("v1", "v2", "r")
ggplot(cmdf, aes(v1, v2, fill = r)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = round(r, 2)), size = 2.4) +
  scale_fill_gradient2(low = "#e31a1c", mid = "white", high = "#1f78b4",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r vif, echo=FALSE}
ols0 <- lm(turnout ~ ., data = d)
X <- model.matrix(ols0)[, -1]
vif <- sapply(colnames(X), function(v) {
  r2 <- summary(lm(X[, v] ~ X[, setdiff(colnames(X), v)]))$r.squared
  1 / (1 - r2)
})
vtab <- data.frame(predictor = names(vif), VIF = as.numeric(vif))
vtab <- vtab[order(-vtab$VIF), ]
knitr::kable(vtab, row.names = FALSE, digits = 2,
             caption = "Variance inflation factors from the reproduction model.")
```

One correlation exceeds the **|r| > 0.7** flag: `income2` and `year`
(r `r round(cor(d$income2, d$year), 2)`) — per-capita income rises secularly over
the sample window — and these two carry the largest VIFs (~`r round(max(vif), 1)`
and ~`r round(sort(vif, decreasing = TRUE)[2], 1)`). All other VIFs sit below
2.5. The collinearity is mild and confined to the income/time pair, so OLS
coefficients on the substantive institutional predictors remain interpretable;
it is worth noting only because it can inflate the standard errors on `income2`
and `year` specifically. This motivates the regularized comparisons introduced
in later tutorials, where correlated predictors are handled by shrinkage.

# Reproduce the published regression

We fit their exact Table 2, Column 1 specification.

```{r reproduce}
ols <- lm(turnout ~ tlness + lcr2 + mmds + regdifficulty + straight + iuse2 + squire +
            ranney + income2 + diversity3 + preselec + redistrict + year, data = d)
summary(ols)
```

**Confirmation against the published study.** Fitting their specification recovers the published model:
`r nrow(d)` state-years and in-sample $R^2$ = `r round(summary(ols)$r.squared, 3)`. Presidential-election years
raise turnout by ~12.8 points (`preselec`), stricter registration deadlines lower it (`regdifficulty`), and
professionalism (`squire`) and competition (`ranney`) enter positively — reproducing the reported signs and
significance.

# Day 3: Regularization & Variable Selection

The published OLS *explains* turnout with all 13 predictors entered simultaneously. Day 3 asks a different
question: **which predictors do we actually need, and can shrinking their coefficients buy better
out-of-sample prediction?** OLS is unbiased but high-variance — with 13 correlated institutional predictors
(recall the `income2`/`year` collinearity) it can overfit sample noise. **Regularization** trades a little
bias for a large drop in variance by penalizing coefficient size; **variable selection** goes further and
drops predictors entirely. We treat the same continuous outcome (`turnout`) and the same 13 predictors, and
we compare four regularized fits against the OLS baseline.

## Motivation and standardization

The penalized estimators minimize the residual sum of squares **plus** a penalty on the coefficients:

$$\hat\beta \;=\; \arg\min_\beta \;\tfrac{1}{2n}\sum_i (y_i - x_i^\top\beta)^2 \;+\; \lambda\Big[\underbrace{\tfrac{1-\alpha}{2}\lVert\beta\rVert_2^2}_{\text{ridge}} + \underbrace{\alpha\lVert\beta\rVert_1}_{\text{lasso}}\Big].$$

- **Ridge** ($\alpha=0$): an $\ell_2$ penalty shrinks all coefficients toward zero but keeps every predictor;
  ideal when predictors are correlated and we want to stabilize, not eliminate.
- **Lasso** ($\alpha=1$): an $\ell_1$ penalty can shrink coefficients *exactly to zero*, performing
  selection.
- **Elastic net** ($0<\alpha<1$): blends the two — selection with grouped shrinkage of correlated predictors.

Because the penalty acts on the **magnitude** of $\beta$, predictors must be on a common scale — otherwise a
variable measured in small units (e.g. `income2` in \$1000s) would be penalized differently from one in large
units. `glmnet` **standardizes internally** (`standardize = TRUE`, the default) and returns coefficients on
the original scale. For interpreting *importance* we standardize by hand so the coefficients are directly
comparable.

```{r std-setup}
library(glmnet)
library(abess)

y <- d$turnout
X <- as.matrix(d[, preds])              # 13 predictors, original units
Xs <- scale(X)                          # standardized copy (mean 0, sd 1) for importance
```

## Ridge, lasso, and elastic net: coefficient paths

Each estimator traces a **coefficient path** as the penalty $\lambda$ ranges from large (everything shrunk to
zero) to small (approaching OLS). Reading right-to-left shows the order in which predictors "enter."

```{r paths, fig.height=3.2, fig.cap="Coefficient paths (log-lambda on the x-axis) for ridge, lasso, and elastic net. Ridge shrinks all 13 predictors smoothly; lasso and EN zero predictors out as the penalty grows (right)."}
fit_ridge <- glmnet(X, y, alpha = 0)
fit_lasso <- glmnet(X, y, alpha = 1)
fit_en    <- glmnet(X, y, alpha = 0.5)

par(mfrow = c(1, 3), mar = c(4, 4, 2.5, 1))
plot(fit_ridge, xvar = "lambda"); title("ridge (alpha=0)", line = 1.4)
plot(fit_lasso, xvar = "lambda"); title("lasso (alpha=1)", line = 1.4)
plot(fit_en,    xvar = "lambda"); title("elastic net (alpha=0.5)", line = 1.4)
```

The **ridge** panel keeps all 13 paths alive, squeezing them toward zero together. The **lasso** panel
collapses paths onto the zero line one by one as $\lambda$ rises — the top axis counts how many predictors
survive. **Elastic net** looks in between.

## Cross-validation to choose lambda

We pick $\lambda$ by **k-fold cross-validation**, which estimates out-of-sample error for each candidate
penalty. Two conventional choices: `lambda.min` (minimizes CV error) and `lambda.1se` (the largest $\lambda$
whose CV error is within one standard error of the minimum — a simpler, more strongly regularized model).

```{r cv, fig.height=3.0, fig.cap="Cross-validation curves. Dotted verticals mark lambda.min (left) and lambda.1se (right). Top axis = number of nonzero predictors."}
set.seed(2026)
cv_ridge <- cv.glmnet(X, y, alpha = 0)
cv_lasso <- cv.glmnet(X, y, alpha = 1)
cv_en    <- cv.glmnet(X, y, alpha = 0.5)

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
plot(cv_ridge); title("ridge", line = 2.2)
plot(cv_lasso); title("lasso", line = 2.2)
plot(cv_en);    title("elastic net", line = 2.2)
```

```{r cv-lambdas, echo=FALSE}
lam <- data.frame(
  model      = c("ridge", "lasso", "elastic net"),
  lambda_min = c(cv_ridge$lambda.min, cv_lasso$lambda.min, cv_en$lambda.min),
  lambda_1se = c(cv_ridge$lambda.1se, cv_lasso$lambda.1se, cv_en$lambda.1se),
  nonzero_1se = c(
    cv_ridge$nzero[cv_ridge$index["1se", ]],
    cv_lasso$nzero[cv_lasso$index["1se", ]],
    cv_en$nzero[cv_en$index["1se", ]]))
knitr::kable(lam, digits = 3, row.names = FALSE,
             caption = "CV-selected penalties and model size at lambda.1se.")
```

Ridge keeps all 13 predictors at either $\lambda$. At **lambda.1se**, lasso and elastic net prune the model
down to a compact set of institutional drivers — the strongly-signalled predictors (election timing,
registration, professionalism, competition) survive while weak ones are zeroed out. **lambda.min** keeps more
predictors than **lambda.1se** because it chases the CV-minimizing fit rather than a parsimonious one.

## Interpreting shrinkage: coefficients side by side

```{r coef-table, echo=FALSE}
get_coef <- function(cvfit, s) as.numeric(coef(cvfit, s = s))[-1]  # drop intercept
ctab <- data.frame(
  predictor  = preds,
  OLS        = coef(ols)[preds],
  ridge_1se  = get_coef(cv_ridge, "lambda.1se"),
  lasso_min  = get_coef(cv_lasso, "lambda.min"),
  lasso_1se  = get_coef(cv_lasso, "lambda.1se"),
  en_1se     = get_coef(cv_en,    "lambda.1se"))
knitr::kable(ctab, digits = 2, row.names = FALSE,
             caption = "OLS vs regularized coefficients (original units). Zeros show where lasso/EN dropped a predictor.")
```

Two lessons are visible in the table. First, **ridge shrinks every coefficient toward zero** but never to
zero — the large OLS effects (`preselec`, `ranney`, `squire`) are pulled in but retained. Second,
**lasso/EN set several coefficients to exactly zero**, delivering a sparser model; the survivors are the same
predictors the descriptive section flagged as the strongest marginal signals.

## Feature importance from standardized coefficients

On the standardized scale, $|\hat\beta_j|$ is a direct **importance** measure: how many points of turnout move
per one-SD change in predictor $j$, holding the rest fixed.

```{r importance, fig.height=3.4, fig.cap="Feature importance = |standardized coefficient| from a CV-tuned lasso (lambda.1se). Larger bars = more predictive leverage per standard deviation."}
cv_lasso_s <- cv.glmnet(Xs, y, alpha = 1)
imp <- data.frame(
  predictor = preds,
  abs_beta  = abs(as.numeric(coef(cv_lasso_s, s = "lambda.1se"))[-1]))
imp <- imp[imp$abs_beta > 0, ]
imp <- imp[order(imp$abs_beta), ]
imp$predictor <- factor(imp$predictor, levels = imp$predictor)

ggplot(imp, aes(abs_beta, predictor)) +
  geom_col(fill = "#1f78b4") +
  labs(x = "|standardized coefficient| (points of turnout per 1 SD)", y = NULL)
```

Presidential-election timing (`preselec`) and registration difficulty dominate, with professionalism and
competition close behind — matching the paper's substantive story that **when** and **how easily** people can
vote swamps the finer institutional details.

## Best-subset selection with abess

Lasso selects by *continuous shrinkage*; **best-subset selection** instead searches directly for the
best-fitting model of each size. The `abess` package makes this tractable via an adaptive
best-subset algorithm and picks the support size by an information criterion.

```{r abess}
ab <- abess(X, y)                       # searches support sizes; picks best by (default) GIC
ab_size  <- extract(ab)$support.size
ab_coef  <- coef(ab, support.size = ab_size)[-1, 1]   # drop intercept
ab_keep  <- names(ab_coef)[ab_coef != 0]
cat("abess selected support size:", ab_size, "\n")
cat("abess keeps:", paste(ab_keep, collapse = ", "), "\n")
```

```{r abess-plot, fig.height=2.8, fig.cap="abess information criterion vs support size; the dashed line marks the selected model."}
gic <- data.frame(size = ab$support.size, gic = ab$tune.value)
ggplot(gic, aes(size, gic)) +
  geom_line(colour = "#1f78b4") + geom_point(colour = "#1f78b4") +
  geom_vline(xintercept = ab_size, linetype = 2, colour = "#e31a1c") +
  scale_x_continuous(breaks = gic$size) +
  labs(x = "support size (number of predictors)", y = "abess tuning criterion")
```

```{r selection-contrast, echo=FALSE}
lasso_keep <- preds[get_coef(cv_lasso, "lambda.1se") != 0]
en_keep    <- preds[get_coef(cv_en,    "lambda.1se") != 0]
sel <- data.frame(
  predictor    = preds,
  lasso_1se    = ifelse(preds %in% lasso_keep, "keep", "."),
  elastic_net  = ifelse(preds %in% en_keep,    "keep", "."),
  abess        = ifelse(preds %in% ab_keep,    "keep", "."))
knitr::kable(sel, row.names = FALSE,
             caption = "Selected support: which predictors each method keeps ('keep') vs drops ('.').")
```

`abess` selects a subset by *direct* best-subset search rather than shrinkage, so its support can differ from
the lasso's — it tends to include the strongest institutional predictors while dropping the weakest, and
because it does not shrink the retained coefficients it can keep a predictor the (more aggressive) 1se-lasso
had zeroed. Comparing the columns shows **where the two selection philosophies agree and disagree**.

## Hyperparameter tuning: lambda and alpha

Elastic net has **two** tuning parameters: $\lambda$ (penalty strength) and $\alpha$ (the ridge/lasso mix). We
grid-search $\alpha$, running a full CV over $\lambda$ at each value, and keep the $(\alpha,\lambda)$ pair with
the lowest cross-validated error.

```{r tune-alpha, fig.height=3.0, fig.cap="Minimum CV error (lambda.min) across the alpha grid. Lowest point = best ridge/lasso mixture."}
set.seed(2026)
alphas <- seq(0, 1, by = 0.1)
foldid <- sample(rep(seq_len(10), length.out = nrow(X)))   # common folds for a fair comparison
tune <- t(sapply(alphas, function(a) {
  cvf <- cv.glmnet(X, y, alpha = a, foldid = foldid)
  c(alpha = a, cvm = min(cvf$cvm), lambda = cvf$lambda.min)
}))
tune <- as.data.frame(tune)
best <- tune[which.min(tune$cvm), ]

ggplot(tune, aes(alpha, cvm)) +
  geom_line(colour = "#1f78b4") + geom_point(colour = "#1f78b4") +
  geom_point(data = best, colour = "#e31a1c", size = 3) +
  labs(x = expression(alpha ~ "(0 = ridge, 1 = lasso)"),
       y = "CV error (MSE) at lambda.min")
cat(sprintf("Best alpha = %.1f  (lambda = %.3f,  CV MSE = %.2f)\n",
            best$alpha, best$lambda, best$cvm))
```

The tuning curve shows how predictive error changes as we move from pure ridge to pure lasso; the red point is
the CV-optimal blend, which we carry into the out-of-sample comparison.

## Out-of-sample comparison: OLS vs regularized

The decisive test is **held-out prediction**. We split 70/30, fit every method on the training set (choosing
each method's $\lambda$ by CV *inside* the training set only), and score RMSE and $R^2$ on the untouched test
set.

```{r oos}
set.seed(2026)
train <- sample(nrow(d), floor(0.7 * nrow(d)))
test  <- setdiff(seq_len(nrow(d)), train)
Xtr <- X[train, ]; Xte <- X[test, ]
ytr <- y[train];   yte <- y[test]

score <- function(pred) {
  rmse <- sqrt(mean((yte - pred)^2))
  r2   <- 1 - sum((yte - pred)^2) / sum((yte - mean(ytr))^2)
  c(RMSE = rmse, R2 = r2)
}

# OLS baseline
ols_tr  <- lm(turnout ~ ., data = d[train, ])
p_ols   <- predict(ols_tr, d[test, ])

# glmnet family: CV lambda.min within the training set
best_alpha <- best$alpha
cvf <- function(a) cv.glmnet(Xtr, ytr, alpha = a)
cv_r <- cvf(0); cv_l <- cvf(1); cv_e <- cvf(best_alpha)
p_ridge <- as.numeric(predict(cv_r, Xte, s = "lambda.min"))
p_lasso <- as.numeric(predict(cv_l, Xte, s = "lambda.min"))
p_en    <- as.numeric(predict(cv_e, Xte, s = "lambda.min"))

# abess on the training set
ab_tr  <- abess(Xtr, ytr)
p_abess <- as.numeric(predict(ab_tr, newx = Xte,
                              support.size = extract(ab_tr)$support.size))

res <- rbind(OLS = score(p_ols), ridge = score(p_ridge),
             lasso = score(p_lasso),
             `elastic net` = score(p_en), abess = score(p_abess))
knitr::kable(round(res, 3),
             caption = "Held-out performance (test n = 30% of states): RMSE (points) and R^2.")
```

```{r oos-plot, echo=FALSE, fig.height=3.2, fig.cap="Test-set RMSE by method (lower is better). Regularization/selection is measured against the OLS baseline."}
rdf <- data.frame(method = rownames(res), RMSE = res[, "RMSE"])
rdf$method <- factor(rdf$method, levels = rdf$method[order(rdf$RMSE)])
ggplot(rdf, aes(method, RMSE)) +
  geom_col(fill = "#1f78b4") +
  geom_hline(yintercept = res["OLS", "RMSE"], linetype = 2, colour = "#e31a1c") +
  labs(x = NULL, y = "held-out RMSE (turnout points)",
       caption = "Dashed line = OLS baseline")
```

With only 13 predictors and $n \approx 486$, this is a **low-dimensional, high-signal** problem, so OLS is already
hard to beat — the regularized models land within a fraction of a point of it. That is itself the lesson:
**regularization protects against variance without sacrificing much when OLS is well-behaved**, and it pays
off far more when predictors are many relative to n or heavily collinear. The selection methods (lasso,
elastic net, `abess`) deliver a **simpler** model — fewer predictors, easier to communicate — at essentially
the same predictive cost, which is often the whole point.

# Takeaway

- **Ridge** stabilizes correlated predictors by shrinking without dropping; **lasso** and **elastic net** add
  automatic variable selection; **abess** does direct best-subset selection.
- Tuning is by **cross-validation** over $\lambda$ (and $\alpha$ for elastic net); `lambda.1se` buys a simpler
  model than `lambda.min` for a small CV-error cost.
- On this well-behaved, low-dimensional turnout problem the regularized fits roughly **match** OLS
  out-of-sample while yielding a **sparser, more interpretable** model — and the machinery pays off much more
  as dimensionality and collinearity grow.

# Independent exercise (about 10 minutes)

1. In the out-of-sample block, re-score the lasso at `s = "lambda.1se"` instead
   of `"lambda.min"`. How much held-out RMSE do you give up, and how many
   predictors does the simpler model drop?
2. Add five pure-noise columns (`rnorm(nrow(X))`) to `X`, re-split, and refit
   the lasso. Does it exclude the noise, and does regularization now look
   better relative to OLS?
