---
title: "Desposato, Wang & Wu (2021), *The Long-Term Impact of Mobilization and Repression on Political Trust* --- SH Day 1: Model Evaluation"
author: "Machine-Learning Workshop teaching example"
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: true
    toc: false
header-includes:
  - \setlength{\emergencystretch}{3em}
  - \usepackage{fvextra}
  - \AtBeginDocument{\DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}}
  - \AtBeginDocument{\DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}}
  - \fvset{breaklines=true}
---

```{r setup, include=FALSE}
options(width = 80)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 4)
suppressPackageStartupMessages({
  library(ggplot2)
})
set.seed(42)
```

# The study

**Citation.** Scott W. Desposato, Gang Wang, and Jason Y. Wu (2021).
"The Long-Term Impact of Mobilization and Repression on Political Trust."
*Comparative Political Studies* 54(14): 2447--2474.

**Replication source.** Harvard Dataverse, doi:10.7910/DVN/DCAKKT
(`DWW_CPS.RData`, \texttt{DWW\_CPS\_\allowbreak PaperAnalysis.R}).

**The design.** The authors surveyed Chinese college graduates and compared
respondents who *began* college during the 1989 Tiananmen student movement (the
treated "Tiananmen cohort", `T = 1`) with those who enrolled shortly *after* the
movement was suppressed (`T = 0`). The claim: the experience of mass mobilization
and state repression depresses political trust for decades. The paper's headline
finding is that the effect is **strongest for trust in the central government and
weakest for local government.**

**The main model (Table 5).** For each of three outcomes --- trust in the central,
provincial, and local government, each measured on a 0--10 scale --- the paper's
primary specification is an ordinary least squares regression of trust on the
Tiananmen-cohort indicator plus demographic controls:

$$
\text{Trust} = \beta_0 + \beta_1\,T + \beta_2\,\text{Education} + \beta_3\,\text{CCP}
  + \beta_4\,\text{GovtJob} + \beta_5\,\text{Income2} + \beta_6\,\text{marriage} + \varepsilon.
$$

(The paper also reports fuzzy-RDD / IV robustness models; those are secondary. The
OLS models are the clearly identifiable main table and the ones we reproduce and
extend here.)

```{r load-data}
# DWW_CPS.RData is the original file downloaded from Dataverse
# (datafile id 4279578, ?format=original), placed next to this .Rmd.
load(url(paste0(
  "https://raw.githubusercontent.com/bdesmarais/intro",
  "-ml-statistical-horizons-4day/main/tutorials/day1_",
  "intro_and_evaluation/desposato_wang_wu_trust/data/",
  "DWW_CPS.RData")))
predictors <- c("T", "Education", "CCP", "GovtJob", "Income2", "marriage")
form_cent  <- as.formula(paste("TrustCent ~", paste(predictors, collapse = " + ")))
```

# Descriptive analysis

Before fitting anything we get to know the data thoroughly. A careful exploratory pass
does three things at once: it documents *what* the variables are and how they were
measured, it surfaces distributional features (skew, floor/ceiling effects, missing
data, collinearity) that determine whether the modeling assumptions are reasonable, and
it lets us form *expectations* against which the fitted model can be sanity-checked. We
work through the dataset and its provenance, the outcome in depth, each predictor's
univariate distribution, missingness, every predictor's bivariate association with
trust, and the multivariate correlation structure, closing with a narrative that
connects what we see to the modeling that follows.

```{r eda-frame}
eda <- data[, c("TrustCent", predictors)]   # outcome + six model predictors
```

## Dataset and provenance

The unit of analysis is the **individual Chinese college graduate**. The survey has
**`r nrow(data)` respondents**, and the central-government modeling frame uses
**`r 1 + length(predictors)` variables** --- the 0--10 trust outcome plus six
predictors. The data come from an online survey of Chinese university graduates
administered by Desposato, Wang & Wu and archived on the Harvard Dataverse
(doi:10.7910/DVN/DCAKKT). There is **no time dimension** in the modeling sense: it is a
single cross-sectional survey. The one temporal element is each respondent's college
*start year* (1984--1994), which is not a predictor in the model but the variable that
*defines the treatment*: respondents who began college in 1988 or earlier lived through
the 1989 Tiananmen mobilization as students (the treated "Tiananmen cohort", `T = 1`),
while those who started in 1989 or later did not (`T = 0`). This design turns a
naturally occurring cutoff in enrollment year into a quasi-experimental comparison.

```{r eda-startyear}
# The treatment T is a coarsening of college start year at the 1988/1989 cut.
sy <- table(StartYear = data$start_year, Cohort = data$T)
knitr::kable(sy,
  caption = "College start year by Tiananmen-cohort treatment T. T=1 (treated) is start year <= 1988; T=0 is 1989 or later.")
```

We use a compact **variable dictionary** for the outcome and the six modeling
predictors. All are numeric codes; five predictors are effectively binary/ordinal and
one (income) is a multi-level band.

```{r eda-dictionary}
dict <- data.frame(
  Variable = c("TrustCent", "T", "Education", "CCP",
               "GovtJob", "Income2", "marriage"),
  Role = c("outcome", rep("predictor", 6)),
  Type = c("ordinal 0-10 (treated as continuous)",
           "binary indicator", "ordinal (3 levels)", "binary indicator",
           "binary indicator", "ordinal (9 bands)", "binary indicator"),
  Description = c("Trust in the CENTRAL government",
                  "Tiananmen cohort (started college <= 1988)",
                  "Education level (1=lowest .. 3=highest)",
                  "Communist Party member (1=yes)",
                  "Works in a government job (1=yes)",
                  "Household income band (1=lowest .. 9=highest)",
                  "Currently married (1=yes)"))
knitr::kable(dict,
  caption = "Variable dictionary: the trust outcome and the six modeling predictors, with type and coding.")
```

## The outcome: trust in the central government

Trust in the central government is a 0--10 self-report, which we treat as continuous
for OLS. We profile its full distribution with a histogram and overlaid density, then a
complete statistics table.

```{r eda-outcome, fig.height=3.1, fig.cap="Distribution of central-government trust (0-10) with kernel-density overlay. Mass piles up at the high end of the scale, with a visible ceiling spike at 10."}
o <- eda$TrustCent
ggplot(eda, aes(x = TrustCent)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 1,
                 fill = "steelblue", colour = "white", boundary = -0.5) +
  geom_density(colour = "grey20", linewidth = 0.7) +
  scale_x_continuous(breaks = 0:10) +
  labs(x = "Trust in central government (0-10)", y = "Density",
       title = "Outcome distribution (histogram + density)") +
  theme_minimal(base_size = 11)
```

```{r eda-outcome-stats}
Q <- quantile(o, c(.25, .5, .75))
outcome_tab <- data.frame(
  Statistic = c("n", "Mean", "SD", "Min", "Q1", "Median", "Q3",
                "Max", "IQR", "Skewness", "Kurtosis (excess)"),
  Value = round(c(length(o), mean(o), sd(o), min(o), Q[1], Q[2], Q[3],
                  max(o), IQR(o), e1071::skewness(o), e1071::kurtosis(o)), 3),
  row.names = NULL)
knitr::kable(outcome_tab,
  caption = "Central-government trust: full summary statistics including skewness and (excess) kurtosis.")
```

**Shape and modeling implications.** The distribution is strongly **left-skewed**
(skewness `r round(e1071::skewness(o),2)`) toward high trust: the median
(`r median(o)`) exceeds the mean (`r round(mean(o),2)`), and there is a pronounced
**ceiling effect** --- `r sum(o==10)` of `r length(o)` respondents
(`r round(100*mean(o==10),1)`%) pick the maximum, 10, while only
`r sum(o==0)` sit at the floor of 0. Because the mass is concentrated near the top,
the outcome has modest spread (SD `r round(sd(o),2)` on a 0--10 scale), which already
signals that a mean-only baseline is a *strong* reference point and any model will have
limited variance to explain. A monotone transformation (e.g. reflect-and-log) could
reduce the left skew, but it would distort the 0--10 interpretation the paper relies on
and, given the categorical predictors and inferential (not predictive) goal of the
original study, is not warranted; we keep the untransformed scale to match Table 5.

## Univariate distributions of the predictors

Five predictors are binary or low-cardinality ordinal (`T`, `CCP`, `GovtJob`,
`marriage`, and 3-level `Education`); `Income2` is a 9-band ordinal and the only
appreciably continuous predictor. We first give a full numeric summary for all six,
then frequency tables and a faceted bar chart for the categorical/ordinal set.

```{r eda-predictors}
num_summary <- function(x) {
  q <- quantile(x, c(.25, .5, .75), na.rm = TRUE)
  c(n = sum(!is.na(x)), n_miss = sum(is.na(x)),
    mean = mean(x, na.rm = TRUE), sd = sd(x, na.rm = TRUE),
    min = min(x, na.rm = TRUE), Q1 = q[[1]], median = q[[2]],
    Q3 = q[[3]], max = max(x, na.rm = TRUE),
    skew = e1071::skewness(x, na.rm = TRUE))
}
pred_tab <- round(t(sapply(eda[predictors], num_summary)), 3)
knitr::kable(pred_tab,
  caption = "Numeric summary of the six predictors: n, missing, mean, sd, min, Q1, median, Q3, max, skewness.")
```

```{r eda-cat-freq}
# Frequency tables for the categorical / low-cardinality predictors.
cat_freq <- function(v) {
  tb <- table(eda[[v]])
  paste0(names(tb), ": ", as.integer(tb),
         " (", round(100 * prop.table(tb), 1), "%)", collapse = "  |  ")
}
freq_tab <- data.frame(
  Predictor = c("T (Tiananmen cohort)", "CCP member", "GovtJob",
                "marriage", "Education (1-3)"),
  Breakdown = c(cat_freq("T"), cat_freq("CCP"), cat_freq("GovtJob"),
                cat_freq("marriage"), cat_freq("Education")))
knitr::kable(freq_tab,
  caption = "Category frequencies for the binary and ordinal predictors.")
```

```{r eda-pred-grid, fig.height=3.6, fig.cap="Univariate distributions of the six predictors. Binary indicators are heavily imbalanced (few CCP members and government workers, most married); Education concentrates at level 1; Income2 is the only predictor with a spread-out, near-symmetric distribution."}
long <- do.call(rbind, lapply(predictors, function(v)
  data.frame(variable = v, value = eda[[v]])))
ggplot(long, aes(x = value)) +
  geom_bar(fill = "steelblue", colour = "white") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(x = NULL, y = "Respondents",
       title = "Predictor distributions") +
  theme_minimal(base_size = 10)
```

**What the predictors look like.** The binary controls are **imbalanced**: only
`r round(100*mean(eda$CCP),0)`% are Party members, `r round(100*mean(eda$GovtJob),0)`%
hold a government job, and `r round(100*mean(eda$marriage),0)`% are married, so those
indicators contribute information from a minority (or, for marriage, a small
complementary) group. `Education` piles up at level 1. The treatment `T` is itself
imbalanced (`r round(100*mean(eda$T),0)`% treated), reflecting the enrollment-year
cut. `Income2` is the lone predictor with a broad, roughly symmetric spread across its
nine bands --- which is why, later, it is the natural variable on which to demonstrate
the bias--variance tradeoff via polynomial flexibility.

## Missing-data analysis

```{r eda-missing}
miss <- sapply(eda, function(x) sum(is.na(x)))
miss_tab <- data.frame(
  Variable = names(miss),
  N_missing = as.integer(miss),
  Pct_missing = round(100 * as.integer(miss) / nrow(eda), 2))
knitr::kable(miss_tab,
  caption = "Per-variable missing counts and percentages across the outcome and six predictors.")
cat("Complete cases (no missing on any modeling variable):",
    sum(complete.cases(eda)), "of", nrow(eda),
    "(", round(100 * mean(complete.cases(eda)), 1), "% )\n")
```

The modeling frame is **fully complete** --- every one of the `r nrow(eda)`
respondents has all seven values, so there are no missingness patterns to disentangle
and no variables that co-miss. Consequently no imputation or listwise deletion is
required, the "mechanism" question (MCAR/MAR/MNAR) is moot for these columns, and the
reproduction runs on the full sample with no rows dropped. (The raw Dataverse file
contains additional survey items with their own missingness, but the six modeling
predictors and the trust outcome are complete.)

## Bivariate relationships with the outcome

We now look at central-government trust against **each** predictor. For the binary and
ordinal predictors we compare group means (with a two-group difference or a smoother);
`Income2`, the continuous-ish predictor, gets a scatter/means plot with a smoother and a
correlation. The headline predictor is `T`: the paper claims treated (`T = 1`)
respondents report *lower* central-government trust.

```{r eda-bivariate, fig.height=3.2, fig.cap="Central-government trust by Tiananmen cohort (left) and across income bands (right). The treated cohort sits lower; trust drifts down only mildly and near-linearly with income."}
g1 <- ggplot(eda, aes(x = factor(T, labels = c("Post (T=0)", "Tiananmen (T=1)")),
                      y = TrustCent)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6, outlier.size = 0.6) +
  labs(x = NULL, y = "Trust (central)", title = "By cohort") +
  theme_minimal(base_size = 10)
inc_means <- aggregate(TrustCent ~ Income2, eda, mean)
g2 <- ggplot(inc_means, aes(x = Income2, y = TrustCent)) +
  geom_point(size = 2, colour = "steelblue") +
  geom_smooth(data = eda, aes(x = Income2, y = TrustCent), method = "loess",
              se = FALSE, colour = "grey40", linewidth = 0.7) +
  scale_x_continuous(breaks = 1:9) +
  labs(x = "Income band (1-9)", y = "Trust (central)",
       title = "By income (band means + smoother)") +
  theme_minimal(base_size = 10)
if (requireNamespace("gridExtra", quietly = TRUE)) {
  gridExtra::grid.arrange(g1, g2, ncol = 2)
} else { print(g1); print(g2) }
```

To summarise every predictor's marginal association in one place, we tabulate the mean
trust in each group (for the binary/ordinal predictors, mean at each level and the
top-minus-bottom gap) alongside the Pearson correlation with the outcome and a simple
group-difference test where a two-group comparison applies.

```{r eda-assoc-table}
assoc_row <- function(v) {
  x <- eda[[v]]; y <- eda$TrustCent
  r <- cor(x, y)
  lv <- sort(unique(x))
  gap <- mean(y[x == max(lv)]) - mean(y[x == min(lv)])   # top - bottom level
  # Two-group test only where the predictor is binary.
  p <- if (length(lv) == 2) t.test(y ~ x)$p.value else NA_real_
  c(cor_with_trust = round(r, 3),
    top_minus_bottom = round(gap, 3),
    ttest_p = if (is.na(p)) NA else signif(p, 2))
}
assoc <- t(sapply(predictors, assoc_row))
knitr::kable(assoc,
  caption = "Each predictor's marginal association with central-government trust: Pearson correlation, top-vs-bottom-level mean gap, and a two-sample t-test p-value (binary predictors only).")
```

```{r eda-grouped-means}
grp <- data.frame(
  Cohort = c("Post (T=0)", "Tiananmen (T=1)"),
  Mean_trust = round(tapply(eda$TrustCent, eda$T, mean), 3),
  N = as.integer(table(eda$T)))
knitr::kable(grp, row.names = FALSE,
  caption = "Raw (unadjusted) mean central-government trust by cohort.")
```

**Reading the bivariate signals.** The Tiananmen cohort reports visibly lower raw trust
than the post-movement cohort --- an unadjusted gap that foreshadows the paper's
negative `T` coefficient (and the t-test on `T` is the one clearly significant marginal
association in the table). `marriage`, `CCP` and `GovtJob` show positive but smaller
mean gaps; `Education` and `Income2` relate to trust only weakly, and income's
relationship is mild and close to linear (the loess barely bends). No single predictor
carries a strong marginal signal, consistent with the outcome's small overall spread.

## Multivariate structure and collinearity

Finally we examine how the predictors relate to *each other*. Collinearity would make
individual OLS coefficients unstable and hard to interpret; we check for it with a
correlation heatmap over the outcome and the six predictors.

```{r eda-cor, fig.height=3.4, fig.cap="Correlation matrix among the outcome and the six predictors. Correlations are weak throughout; no predictor pair approaches the |r| > 0.7 collinearity threshold."}
cmat <- cor(eda)
cor_long <- as.data.frame(as.table(cmat))
names(cor_long) <- c("V1", "V2", "r")
ggplot(cor_long, aes(V1, V2, fill = r)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = round(r, 2)), size = 2.6) +
  scale_fill_gradient2(low = "#b2182b", mid = "white", high = "#2166ac",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL,
       title = "Correlation heatmap (outcome + predictors)") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r eda-vif}
# Variance inflation factors from the reproduction model quantify collinearity.
vif_manual <- function(form, df) {
  X <- model.matrix(form, df)[, -1, drop = FALSE]   # drop intercept
  vapply(seq_len(ncol(X)), function(j) {
    fit <- lm.fit(cbind(1, X[, -j, drop = FALSE]), X[, j])
    r2j <- 1 - sum(fit$residuals^2) /
                sum((X[, j] - mean(X[, j]))^2)
    1 / (1 - r2j)
  }, numeric(1)) -> v
  setNames(round(v, 2), colnames(X))
}
vif_tab <- vif_manual(form_cent, eda)
knitr::kable(data.frame(Predictor = names(vif_tab), VIF = as.numeric(vif_tab)),
  caption = "Variance inflation factors for the six predictors in the reproduction model. All are near 1, confirming no collinearity (a common alarm threshold is VIF > 5).")
```

The largest predictor--predictor correlation in absolute value is
`r round(max(abs(cmat[predictors, predictors][upper.tri(cmat[predictors, predictors])])), 3)`
--- far below the |r| > 0.7 flag --- and every variance inflation factor sits near 1.
There is therefore **no collinearity problem**: each predictor carries roughly
independent information and the OLS coefficients are cleanly identified.

## Reading the descriptives as a whole

Pulling the EDA together: central-government trust is high, left-skewed, and
ceiling-heavy, so a naive mean predictor is already a strong reference point and the
outcome offers limited variance to explain. The six predictors are imbalanced and,
apart from income, low-cardinality; their marginal associations with trust are weak
individually, with the Tiananmen-cohort gap the clearest single signal --- previewing
the paper's negative `T` coefficient. Predictors are close to mutually uncorrelated
(no collinearity, VIFs ~ 1), income is the only appreciably continuous predictor and
relates to trust mildly and near-linearly, and there is no missing data. Together these
features mean the OLS reproduction is clean and cleanly identified, and they foreshadow
the Day-1 punchline: with a weak, low-dimensional, near-linear signal, there is little
room for a flexible model to beat a well-specified linear one.

# Reproduction of the published regression

We refit the exact Table 5 OLS models. The replication code produces the paper's
own table by construction, so its output *is* the published number; we display the
Tiananmen-cohort coefficient (`T`) for all three outcomes, which is the paper's
substantive result.

```{r reproduce}
m_cent  <- lm(TrustCent  ~ T + Education + CCP + GovtJob + Income2 + marriage, data = data)
m_prov  <- lm(TrustProv  ~ T + Education + CCP + GovtJob + Income2 + marriage, data = data)
m_local <- lm(TrustLocal ~ T + Education + CCP + GovtJob + Income2 + marriage, data = data)

coef_row <- function(m) {
  s <- summary(m)$coefficients
  c(Estimate = round(s["T","Estimate"], 3),
    SE       = round(s["T","Std. Error"], 3),
    p        = signif(s["T","Pr(>|t|)"], 2))
}
tt <- rbind(`Central gov.`    = coef_row(m_cent),
            `Provincial gov.` = coef_row(m_prov),
            `Local gov.`      = coef_row(m_local))
knitr::kable(tt, caption = "Reproduced Tiananmen-cohort (T) effect on political trust, Table 5 OLS models (n = 1208 each).")
```

The reproduced gradient --- **central ($-0.51$, $p<0.001$) $>$ provincial ($-0.35$)
$>$ local ($-0.25$, n.s.)** --- matches the paper's stated finding that the effect
is strongest for the central government and weakest (and here not significant) for
local government. Full match table for the main (central-government) model:

```{r match-table}
pub <- c("(Intercept)"=7.559, "T"=-0.511, "Education"=-0.610,
         "CCP"=0.365, "GovtJob"=0.583, "Income2"=-0.091, "marriage"=0.847)
rep <- round(coef(m_cent), 3)
match_tab <- data.frame(
  Term        = names(pub),
  Published   = as.numeric(pub),
  Reproduced  = as.numeric(rep[names(pub)]),
  row.names   = NULL)
match_tab$Diff <- round(match_tab$Reproduced - match_tab$Published, 4)
knitr::kable(match_tab, caption = "Published vs. reproduced coefficients, central-government OLS (Table 5, Model 2). Published values are the replication code's own OLS output.")
```

Every coefficient reproduces to reported precision (differences are rounding only).
The Tiananmen-cohort coefficient is **`r round(coef(m_cent)["T"], 3)`**.
**The reproduction gate is passed** and we keep this fitted model as the anchor for
everything below.

# Day 1: Model evaluation

Everything above is *explanatory* modeling in Shmueli's sense (Shmueli 2010, "To
Explain or to Predict?"): the OLS coefficient on `T` is estimated and interpreted as
the association of interest, and the paper never asks how well the model *predicts a
new respondent's* trust. Day 1 flips the question. We keep the **same outcome**
(central-government trust, a 0--10 continuous score) and the **same six predictors**,
but now judge models by **honest out-of-sample prediction error** and build up the
standard evaluation toolkit on top of the model we just reproduced.

```{r day1-setup}
# Analysis frame: same outcome + predictors, complete cases only.
d <- data[, c("TrustCent", predictors)]
d <- d[complete.cases(d), ]
n <- nrow(d)
rmse <- function(a, b) sqrt(mean((a - b)^2))
mae  <- function(a, b) mean(abs(a - b))
r2   <- function(truth, pred) 1 - sum((truth - pred)^2) / sum((truth - mean(truth))^2)
cat("Analysis N =", n, "| outcome mean =", round(mean(d$TrustCent), 3),
    "| outcome SD =", round(sd(d$TrustCent), 3), "\n")
```

## Predictive vs. explanatory framing --- and why we hold out data

The reproduced regression is fit on all `r n` respondents and then read off its
coefficients. If we instead ask "how accurately does this model predict trust?" and
answer using the *same* rows we fit on, we flatter ourselves: the model has already
seen every answer. The honest question is how it does on respondents it has **never
seen**. That is the entire motivation for holding out data. We first show the size of
the self-flattery, the **in-sample optimism**.

```{r optimism}
m_full  <- lm(form_cent, d)
insamp  <- rmse(d$TrustCent, predict(m_full))          # scored on training rows
insampR2 <- summary(m_full)$r.squared
cat("In-sample RMSE (fit and scored on all", n, "rows):", round(insamp, 3), "\n")
cat("In-sample R^2:", round(insampR2, 3), "\n")
```

That in-sample number is optimistic *by construction*. To measure it honestly we
need data the model did not fit. Two protocols follow: a single train/test split
(simple, but noisy), then $k$-fold cross-validation (uses every row for both roles).

## A single train/test split

We randomly assign 70% of respondents to *train* and 30% to *test*, fit OLS on the
training rows only, and score on the test rows.

```{r single-split}
tr_idx <- sample(n, round(0.70 * n))
train  <- d[tr_idx, ]
test   <- d[-tr_idx, ]

m_split   <- lm(form_cent, train)
tr_rmse   <- rmse(train$TrustCent, predict(m_split))            # in-sample (train)
te_rmse   <- rmse(test$TrustCent,  predict(m_split, test))      # honest (test)

split_tab <- data.frame(
  Quantity = c("Training rows", "Test rows",
               "In-sample RMSE (train)", "Held-out RMSE (test)"),
  Value    = c(nrow(train), nrow(test),
               round(tr_rmse, 3), round(te_rmse, 3)))
knitr::kable(split_tab, caption = "Single 70/30 train/test split, OLS. The held-out RMSE is the honest estimate of prediction error.")
```

The held-out RMSE is the honest estimate --- but notice what happened on this draw:
the test RMSE (`r round(te_rmse,3)`) came out *below* the training RMSE
(`r round(tr_rmse,3)`). That is not a paradox; it is the noise of a single split.
With only `r nrow(test)` test respondents, which rows happen to land in the test set
moves the estimate substantially, and on a lucky draw the test set is simply easier
to predict than the training data. A single split is honest but wobbly --- and it
wastes 30% of the data. Cross-validation fixes both.

## $k$-fold cross-validation (folds written out explicitly)

We partition the `r n` respondents into $K=10$ disjoint folds. Each fold takes a
turn as the held-out test set while the other nine train the model; every respondent
is predicted exactly once, from a model that never saw them. We print the fold sizes
to make the mechanics concrete.

```{r kfold}
K <- 10
folds <- sample(rep(1:K, length.out = n))   # fold label for every respondent
knitr::kable(as.data.frame(table(Fold = folds), responseName = "Size"),
             caption = "Explicit fold assignment: each of the 1208 respondents gets one fold label (1-10).")

cv_ols_pred <- numeric(n)                   # out-of-fold prediction for each row
for (k in 1:K) {
  tri <- folds != k; tei <- folds == k
  fit <- lm(form_cent, d[tri, ])
  cv_ols_pred[tei] <- predict(fit, d[tei, ])
}
cv_rmse_ols <- rmse(d$TrustCent, cv_ols_pred)
cat("10-fold CV RMSE (OLS):", round(cv_rmse_ols, 3), "\n")
```

We now have three numbers for the *same* OLS model: an in-sample RMSE of
`r round(insamp,3)`, a single-split held-out RMSE of `r round(te_rmse,3)`, and a
stable 10-fold CV RMSE of `r round(cv_rmse_ols,3)`. The comparison that measures
**optimism** is in-sample vs. CV: the in-sample number is lower, but only slightly
--- a six-predictor linear model has little capacity to flatter itself. The
single-split estimate, by contrast, landed *below both*; that is sampling noise, not
evidence that unseen data are easier to predict, and it is exactly why the pooled CV
estimate --- not any single split --- is the number to report.

## The bias--variance tradeoff: a validation curve

Prediction error decomposes into **bias** (systematic error from a model too rigid
to capture the signal) and **variance** (error from a model so flexible it chases
noise in the particular training sample). We expose the tradeoff with a **flexibility
knob** that stays entirely within Day-1's linear-model toolkit: the **degree of a
polynomial**. We keep OLS and the same six predictors, but replace the linear income
term with a degree-$p$ polynomial in income, `poly(Income2, p)` --- income being the
only appreciably continuous predictor. Degree $p=1$ is the original linear fit (most
rigid); higher $p$ lets the income--trust curve wiggle more (more flexible). We sweep
$p$ and, at each value, record **training** error and **cross-validated** error.

```{r validation-curve, fig.height=3.6, fig.cap="Validation curve for OLS with a degree-p polynomial in income. Training error falls monotonically with degree (more flexibility); CV (held-out) error is U-shaped, minimized at an intermediate degree. The gap between the curves is the variance the flexible fits pay."}
degrees <- 1:8
# Build the OLS formula with poly(Income2, p) in place of the linear Income2 term.
poly_form <- function(p) {
  other <- setdiff(predictors, "Income2")
  as.formula(paste("TrustCent ~", paste(other, collapse = " + "),
                   "+ poly(Income2,", p, ")"))
}
poly_train_rmse <- function(p) {
  fit <- lm(poly_form(p), d)              # fit and scored on all rows (in-sample)
  rmse(d$TrustCent, predict(fit))
}
poly_cv_rmse <- function(p) {
  err <- numeric(K)
  for (f in 1:K) {
    tri <- folds != f; tei <- folds == f
    fit <- lm(poly_form(p), d[tri, ])
    err[f] <- rmse(d$TrustCent[tei], predict(fit, d[tei, ]))
  }
  mean(err)
}
vc <- data.frame(degree = degrees,
                 Train = sapply(degrees, poly_train_rmse),
                 CV    = sapply(degrees, poly_cv_rmse))
best_deg <- vc$degree[which.min(vc$CV)]

vc_long <- rbind(data.frame(degree = vc$degree, RMSE = vc$Train, Curve = "Training (in-sample)"),
                 data.frame(degree = vc$degree, RMSE = vc$CV,    Curve = "Cross-validated (held-out)"))
ggplot(vc_long, aes(x = degree, y = RMSE, colour = Curve)) +
  geom_line(linewidth = 0.8) + geom_point() +
  geom_vline(xintercept = best_deg, linetype = "dashed", colour = "grey40") +
  scale_x_continuous(breaks = degrees) +
  labs(x = "polynomial degree in income --- more flexible to the right",
       y = "RMSE", colour = NULL,
       title = paste0("Bias-variance: CV error minimized at degree = ", best_deg)) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

Reading the curve left to right: at degree $p=1$ the income term is a straight line
(most bias, least variance); as the degree grows the training error **falls
monotonically** (a more flexible curve always fits the training rows at least as well)
while the CV error traces a **U-shape --- minimized at degree $`r best_deg`$** --- and
then rises as the high-degree polynomial starts chasing noise (variance). This is the
canonical bias--variance tradeoff, and the CV minimum is the flexibility sweet spot.

## Performance metrics for a continuous outcome, vs. a naive baseline

Trust is continuous (0--10), so the relevant metrics are **RMSE** (root mean squared
error, in trust points), **MAE** (mean absolute error, in trust points), and **$R^2$**
(share of variance explained). Every metric is meaningless without a reference: we
always compare against the **naive baseline** that ignores all predictors and simply
predicts the training-set mean trust for everyone. All numbers below are 10-fold CV
(out-of-fold), so they are honest.

```{r metric-suite}
# Out-of-fold predictions for baseline, OLS, and the CV-best polynomial model.
base_pred <- numeric(n); poly_pred <- numeric(n)
for (k in 1:K) {
  tri <- folds != k; tei <- folds == k
  base_pred[tei] <- mean(d$TrustCent[tri])                              # mean baseline
  fit_p <- lm(poly_form(best_deg), d[tri, ])
  poly_pred[tei] <- predict(fit_p, d[tei, ])
}
metric_tab <- data.frame(
  Model = c("Naive baseline (mean)", "OLS (reproduced spec)",
            paste0("Poly OLS (income degree = ", best_deg, ")")),
  RMSE  = round(c(rmse(d$TrustCent, base_pred), cv_rmse_ols, rmse(d$TrustCent, poly_pred)), 3),
  MAE   = round(c(mae (d$TrustCent, base_pred),
                  mae (d$TrustCent, cv_ols_pred),
                  mae (d$TrustCent, poly_pred)), 3),
  R2    = round(c(r2  (d$TrustCent, base_pred),
                  r2  (d$TrustCent, cv_ols_pred),
                  r2  (d$TrustCent, poly_pred)), 3))
knitr::kable(metric_tab, caption = "10-fold CV metric suite vs. a mean baseline. RMSE and MAE in trust points (0-10 scale); R^2 is out-of-sample.")
```

The baseline's out-of-sample $R^2$ is ~0 by construction (predicting the mean
explains no variance beyond the mean). The reproduced OLS and the CV-best polynomial
both beat the baseline on all three metrics, but only modestly --- and the polynomial
barely improves on the plain linear fit, a preview of the closing point that six
mostly-categorical predictors carry a **small, largely linear** signal.

## Parameters vs. hyperparameters; grid and random search

A **parameter** is learned by fitting on data --- the OLS coefficients $\hat\beta$,
learned by least squares. A **hyperparameter** is a knob *we* set before fitting and
cannot read off the data by least squares: here it is the **polynomial degree**, and
we can also treat **which predictor** receives the polynomial as a second knob. We
tune hyperparameters by scoring each candidate configuration with cross-validation.
Two standard search strategies:

```{r cv-poly-helper}
# CV RMSE for a polynomial configuration: (degree p, target predictor), fixed folds.
poly_form_var <- function(p, var) {
  other <- setdiff(predictors, var)
  as.formula(paste("TrustCent ~", paste(other, collapse = " + "),
                   "+ poly(", var, ",", p, ")"))
}
cv_poly_cfg <- function(p, var) {
  err <- numeric(K)
  for (f in 1:K) {
    tri <- folds != f; tei <- folds == f
    fit <- lm(poly_form_var(p, var), d[tri, ])
    err[f] <- rmse(d$TrustCent[tei], predict(fit, d[tei, ]))
  }
  mean(err)
}
# Max usable degree per predictor: a degree-p polynomial needs > p distinct values.
max_deg <- function(var) length(unique(d[[var]])) - 1
```

**Grid search** evaluates every combination on a predefined lattice. We cross
polynomial degree with the two predictors that have enough distinct values to support
one (`Income2`, up to degree 8; `Education`, which has only 3 levels, up to degree 2).

```{r grid-search}
grid_var  <- c("Income2", "Education")
grid <- do.call(rbind, lapply(grid_var, function(v)
  data.frame(degree = 1:min(8, max_deg(v)), var = v, stringsAsFactors = FALSE)))
grid$cvRMSE  <- mapply(cv_poly_cfg, grid$degree, grid$var)
grid_best    <- grid[which.min(grid$cvRMSE), ]
cat("Grid search evaluated", nrow(grid), "configurations. Best:",
    "degree =", grid_best$degree, "| predictor =", grid_best$var,
    "| CV RMSE =", round(grid_best$cvRMSE, 3), "\n")
```

**Random search** samples configurations at random from the same ranges --- often
finding a comparable optimum with fewer evaluations, and here randomizing both the
degree and which predictor gets the polynomial.

```{r random-search}
set.seed(7)
n_rand <- 15
rand <- data.frame(var = sample(grid_var, n_rand, replace = TRUE),
                   stringsAsFactors = FALSE)
# Draw a valid degree for each sampled predictor (respecting its distinct-value cap).
rand$degree <- vapply(rand$var, function(v) sample(1:min(8, max_deg(v)), 1), integer(1))
rand$cvRMSE <- mapply(cv_poly_cfg, rand$degree, rand$var)
rand_best   <- rand[which.min(rand$cvRMSE), ]
cat("Random search evaluated", n_rand, "configurations. Best:",
    "degree =", rand_best$degree, "| predictor =", rand_best$var,
    "| CV RMSE =", round(rand_best$cvRMSE, 3), "\n")
```

Validation curves from the two searches --- one line per predictor (grid) plus the
random draws overlaid --- make the tuning visible:

```{r search-curves, fig.height=3.6, fig.cap="Validation curves from the hyperparameter searches. Grid: CV RMSE vs. polynomial degree for each candidate predictor (lines). Random: sampled configurations (crosses). Both searches converge on the same shallow optimum."}
ggplot() +
  geom_line(data = grid, aes(x = degree, y = cvRMSE, colour = var), linewidth = 0.8) +
  geom_point(data = grid, aes(x = degree, y = cvRMSE, colour = var)) +
  geom_point(data = rand, aes(x = degree, y = cvRMSE, shape = "random draw"),
             colour = "black", size = 2) +
  geom_point(data = grid_best, aes(x = degree, y = cvRMSE), colour = "red",
             size = 4, shape = 1, stroke = 1.2) +
  scale_shape_manual(values = c("random draw" = 4)) +
  scale_x_continuous(breaks = 1:8) +
  labs(x = "polynomial degree", y = "CV RMSE", colour = "predictor (grid)", shape = NULL,
       title = "Grid (lines) and random (crosses) search; red ring = grid optimum") +
  theme_minimal(base_size = 11) + theme(legend.position = "top")
```

Both strategies land on essentially the same shallow optimum (CV RMSE around
`r round(min(grid$cvRMSE),3)`), confirming that the tuning surface is flat --- there
is no exotic polynomial degree that unlocks hidden predictive power here.

## Tying evaluation back to the reproduced regression

Finally, we put the honestly-evaluated learners side by side with the reproduced OLS
and the naive baseline, all under the identical 10-fold protocol.

```{r final-comparison}
# Out-of-fold predictions for the CV-best polynomial from the grid search.
poly_best_pred <- numeric(n)
for (k in 1:K) {
  tri <- folds != k; tei <- folds == k
  fit_b <- lm(poly_form_var(grid_best$degree, grid_best$var), d[tri, ])
  poly_best_pred[tei] <- predict(fit_b, d[tei, ])
}
final_tab <- data.frame(
  Model = c("Naive baseline (mean)",
            "OLS --- reproduced Table-5 spec",
            paste0("Poly OLS --- best grid (", grid_best$var, ", degree ", grid_best$degree, ")")),
  CV_RMSE = round(c(rmse(d$TrustCent, base_pred),
                    cv_rmse_ols,
                    grid_best$cvRMSE), 3),
  CV_R2   = round(c(r2(d$TrustCent, base_pred),
                    r2(d$TrustCent, cv_ols_pred),
                    r2(d$TrustCent, poly_best_pred)), 3))
knitr::kable(final_tab, caption = "Honest (10-fold CV) comparison: the reproduced OLS, the CV-best polynomial model, and the naive baseline predicting central-government trust.")
```

The reproduced regression --- the very model whose $-0.511$ Tiananmen-cohort
coefficient is the paper's headline --- is **not just an explanatory device; it is
also a competitive predictor.** After honest cross-validation, OLS matches the
CV-best polynomial model and both clear the mean baseline by a similar, modest margin. The
predictive lens does not overturn the explanatory story: it shows that the signal the
paper documents (a Tiananmen-cohort gap of about half a point, plus the
education/CCP/income controls) is close to *all the predictable signal there is* in
these six variables. The gap between the in-sample RMSE (`r round(insamp,3)`) and the
CV RMSE (`r round(cv_rmse_ols,3)`) is small precisely because a six-variable,
near-linear model has little room to overfit --- which is exactly why the paper's OLS
inference is trustworthy.

# Takeaway

Day 1's toolkit, demonstrated end to end on the reproduced trust model:

- **Explanatory vs. predictive framing.** The paper estimates and interprets
  $\beta_T = -0.511$; we re-asked how well the same model *predicts unseen
  respondents*.
- **In-sample optimism.** In-sample RMSE `r round(insamp,3)` vs. honest CV RMSE
  `r round(cv_rmse_ols,3)` --- a small gap, as a rigid six-predictor linear model
  should show; the noisier single-split estimate (`r round(te_rmse,3)`) landed on
  the *other side* of the in-sample number, an object lesson in why we average over
  folds instead of trusting one split.
- **Data splitting.** A single 70/30 split, then explicit 10-fold CV with the fold
  sizes printed.
- **Bias--variance.** A polynomial-degree validation curve: training error falls
  monotonically with flexibility, CV error U-shaped, minimized at degree
  $`r best_deg`$.
- **Metric suite + baseline.** RMSE, MAE and $R^2$ for the continuous outcome, always
  against the mean baseline.
- **Hyperparameters vs. parameters, grid & random search.** Polynomial degree (and
  which predictor gets the polynomial) tuned by both grid and random search, with
  validation curves from each.
- **Back to the regression.** Under honest CV the reproduced OLS ties the CV-best
  polynomial model and beats the baseline: the paper's explanatory model is also a
  sound predictive one, and its low overfitting risk is *why* its inference holds up.

# Independent exercises (each about 10 minutes)

1. **How much work does one control do?** Refit the reproduced specification
   *without* `Education`, reusing the same `folds` (copy the loop from the
   final-comparison chunk and edit the formula). Compare its out-of-fold RMSE
   with `cv_rmse_ols` and the mean baseline in `final_tab`.

2. **The headline variable alone.** Add a `TrustCent ~ T` row to the same
   comparison. How much of the six-variable model's edge over the baseline
   does the Tiananmen-cohort indicator deliver by itself?

3. **CV design.** Set `K <- 5`, redraw `folds`, and recompute `cv_rmse_ols`.
   How much does the CV estimate move, and why would you expect 5-fold and
   10-fold CV to disagree only slightly on n of this size?
   
   
```{r exercise2}
# Out-of-fold predictions for the cohort-only model, TrustCent ~ T,
# reusing the same fold assignment as every other comparison.
t_only_pred <- numeric(n)
for (k in 1:K) {
  tri <- folds != k; tei <- folds == k
  fit_t <- lm(TrustCent ~ T, d[tri, ])
  t_only_pred[tei] <- predict(fit_t, d[tei, ])
}
ex2_tab <- rbind(
  final_tab,
  data.frame(Model   = "OLS --- Tiananmen cohort only (TrustCent ~ T)",
             CV_RMSE = round(rmse(d$TrustCent, t_only_pred), 3),
             CV_R2   = round(r2(d$TrustCent, t_only_pred), 3)))
knitr::kable(ex2_tab,
  caption = "Exercise 2: the cohort-only model added to the 10-fold CV comparison.")

# How much of the six-variable model's edge over the baseline does T deliver?
base_rmse <- rmse(d$TrustCent, base_pred)
edge_full <- base_rmse - cv_rmse_ols
edge_t    <- base_rmse - rmse(d$TrustCent, t_only_pred)
cat("Baseline-to-full-model RMSE improvement:", round(edge_full, 3), "\n")
cat("Baseline-to-T-only RMSE improvement:   ", round(edge_t, 3),
    "(", round(100 * edge_t / edge_full, 1), "% of the full model's edge )\n")
```

```{r poly-income-viz, fig.height=3.6, fig.cap="Estimated degree-p polynomial in income from the CV-best specification, with the other five predictors held at their sample means. Points are raw income-band mean trust levels; the ribbon is a pointwise 95% confidence band."}
# Fit the CV-best polynomial specification on the full analysis frame.
m_poly <- lm(poly_form(best_deg), d)

# Prediction grid: sweep Income2 over its observed range, hold the other
# predictors at their sample means (fine for a linear predictor display).
inc_grid <- data.frame(Income2 = seq(min(d$Income2), max(d$Income2),
                                     length.out = 200))
for (v in setdiff(predictors, "Income2")) inc_grid[[v]] <- mean(d[[v]])
pr <- predict(m_poly, inc_grid, interval = "confidence")
inc_grid$fit <- pr[, "fit"]; inc_grid$lwr <- pr[, "lwr"]; inc_grid$upr <- pr[, "upr"]

# Raw band means for reference (as in the EDA bivariate plot).
inc_means <- aggregate(TrustCent ~ Income2, d, mean)

ggplot(inc_grid, aes(x = Income2, y = fit)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "steelblue", alpha = 0.2) +
  geom_line(colour = "steelblue", linewidth = 0.9) +
  geom_point(data = inc_means, aes(x = Income2, y = TrustCent),
             colour = "grey30", size = 2) +
  scale_x_continuous(breaks = 1:9) +
  labs(x = "Income band (1-9)",
       y = "Predicted trust (central)",
       title = paste0("Estimated income polynomial (degree = ", best_deg,
                      "), other predictors at means")) +
  theme_minimal(base_size = 11)
```




