---
title: "Reproduce-then-Extend — Pinyon Jay Habitat Selection (Ecology) (Day 2)"
author: "Intro to Machine Learning · Bruce A. Desmarais"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: true
geometry: margin=1in
fontsize: 10pt
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
  - \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines,breakanywhere}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 3.6, fig.align = "center")
options(width = 82)
library(glmnet)   # elastic net / lasso / ridge
library(ranger)   # random forests
library(xgboost)  # gradient boosting
library(PRROC)    # precision-recall / ROC curves
```

> **Published study.** Daw, S. K. et al. (2022). "Behavior-specific occurrence patterns of Pinyon Jays."
> *PLoS ONE* (doi:10.1371/journal.pone.0237621; S1 Dataset, open). A **logistic resource-selection
> function (RSF)** — habitat *use* points (1) vs *available* points (0) on environmental predictors. Use is
> the minority class, so out-of-sample skill is scored by **AUC-PR** (area under the precision-recall curve).

# Background

Pinyon Jays are a **declining bird** of the southwestern United States, and managers need to know **which habitat they use** in order to conserve it. Daw and colleagues fit a **resource-selection function** -- a logistic regression comparing locations the jays *used* to *available* background locations -- to identify the environmental conditions associated with jay occurrence. The **primary conclusion** is that jays select particular pinyon-juniper conditions -- more open stands, sites near habitat edges, and areas with downed woody debris -- which the regression translates into a picture of suitable habitat for management.

# Data and codebook

**Unit of analysis:** a sampled ground location -- an observed jay-**use** point vs an **available**
(background) point; n = 466 (124 use).

| Variable | Definition |
|---|---|
| `use` | 1 = observed Pinyon Jay use location, 0 = available location **(outcome)** |
| `Elevation..feet.` | elevation (feet) |
| `Slope..percent.` | ground slope (%) |
| `Stand_Age..years.` | forest stand age (years) |
| `Stand_Density_Index` | stand density index (tree density / competition) |
| `Canopy_Cover..percent.` | overstory canopy cover (%) |
| `Tree_Cover..percent.` | tree cover (%) |
| `Shrub_Cover..percent.` | shrub cover (%) |
| `Forb_Cover..percent.` | forb (herbaceous) cover (%) |
| `Grass_Cover..percent.` | grass cover (%) |
| `Woody_Debris..category.1.`-`.3.` | counts of downed woody debris by size class (1-3) |
| `Distance.to.Road..km.` | distance to the nearest road (km) |
| `DisturbanceYes` | 1 if any recent disturbance was recorded at the site |
| `Distance.to.Edge..m.` | distance to the nearest habitat edge (m) |

# Descriptive results

```{r data}
d <- read.csv("https://raw.githubusercontent.com/bdesmarais/intro-ml-statistical-horizons-4day/main/tutorials/day4_classic_learners/06_pinyonjay_rsf/data/jays_clean.csv")

# Predictors: keep the interpretable environmental variables and drop the sparse one-hot
# habitat/disturbance-type dummies (unidentified in the logit).
preds <- setdiff(names(d), "use")
preds <- preds[sapply(d[preds], function(x) length(unique(x)) > 1)]
preds <- preds[!grepl("Habitat_Type|Disturbance_Type", preds)]

# Descriptive statistics for the outcome and the first several predictors
stats <- t(sapply(d[c("use", head(preds, 8))],
                  function(x) c(mean = mean(x), sd = sd(x), min = min(x), max = max(x))))
round(stats, 2)
```

```{r desc-plot, fig.height=3, fig.cap="Outcome balance (left) and the environmental variables that differ most between used and available points (right)."}
clean <- function(v) gsub("[._]+", " ", v)   # tidy the mangled column names for labels

par(mfrow = c(1, 2))
# (left) class balance of the used-vs-available outcome
par(mar = c(4, 4, 2.5, 1))
barplot(table(d$use), names.arg = c("available", "used"),
        col = c("grey75", "#1f78b4"), main = "Outcome: use", ylab = "count")

# (right) standardized mean difference (used - available) for each predictor
smd <- sapply(d[preds], function(x) (mean(x[d$use == 1]) - mean(x[d$use == 0])) / sd(x))
smd <- tail(smd[order(abs(smd))], 10)             # the 10 largest-magnitude differences
par(mar = c(4, 8.5, 2.5, 1))
barplot(smd, horiz = TRUE, las = 1, names.arg = clean(names(smd)), cex.names = 0.75,
        col = ifelse(smd > 0, "#1f78b4", "#e31a1c"), main = "Used vs available",
        xlab = "std. mean difference")
abline(v = 0, col = "grey50")
```

# Reproduce the published RSF regression

The published model is a logistic regression of use vs availability on the environmental predictors.

```{r reproduce}
rsf <- glm(reformulate(preds, "use"), data = d, family = binomial)
round(summary(rsf)$coefficients, 3)
```

**Confirmation against the published study.** The signed coefficients recover the reported habitat-selection
pattern — jays select more **open** stands (negative on `Stand_Density_Index` and `Slope..percent.`) and sites
**closer to edges** (negative on `Distance.to.Edge..m.`). This logistic RSF is our regression baseline.

# Predictive results: regression vs machine learning

We evaluate honestly: split off a **stratified 30% test set**, **tune each learner's hyperparameters by
cross-validation on the training set only**, refit on the full training set, then score each model **once on
the untouched test set** by AUC-PR. Nothing about the test set is used for fitting or tuning.

```{r cv}
X <- as.matrix(d[, preds]); y <- d$use
aucpr <- function(p, yt) pr.curve(scores.class0 = p[yt == 1], scores.class1 = p[yt == 0])$auc.integral

set.seed(2026)
te <- c(sample(which(y == 1), round(0.30 * sum(y == 1))),      # stratified 30% test set
        sample(which(y == 0), round(0.30 * sum(y == 0))))
tr <- setdiff(seq_len(nrow(d)), te); yte <- y[te]

## 1. Logistic RSF -- the published regression (no hyperparameters)
m_logit <- glm(use ~ ., data = d[tr, c("use", preds)], family = binomial)
test_auc <- c(Logit = aucpr(predict(m_logit, d[te, ], type = "response"), yte))

## 2. Elastic net -- penalty lambda tuned by CV on the TRAINING set
m_en <- cv.glmnet(X[tr, ], y[tr], alpha = 0.5, family = "binomial")
test_auc["ElasticNet"] <- aucpr(as.numeric(predict(m_en, X[te, ], s = "lambda.min", type = "response")), yte)

## 3. Random forest -- mtry tuned on the TRAINING set by out-of-bag AUC-PR
best_rf <- NULL
for (mt in unique(pmax(1, c(floor(sqrt(ncol(X))), floor(ncol(X) / 3), floor(ncol(X) / 2))))) {
  rf <- ranger(x = X[tr, ], y = factor(y[tr]), num.trees = 800, mtry = mt, probability = TRUE)
  sc <- aucpr(rf$predictions[, "1"], y[tr])          # OOB predictions = a training-only CV estimate
  if (is.null(best_rf) || sc > best_rf$sc) best_rf <- list(mtry = mt, sc = sc)
}
m_rf <- ranger(x = X[tr, ], y = factor(y[tr]), num.trees = 800, mtry = best_rf$mtry, probability = TRUE)
test_auc["RandomForest"] <- aucpr(predict(m_rf, X[te, ])$predictions[, "1"], yte)

## 4. XGBoost -- depth x eta tuned by xgb.cv on the TRAINING set; nrounds by early stopping
dtrain <- xgb.DMatrix(X[tr, ], label = y[tr])
best <- NULL
for (depth in c(3, 4, 6)) for (eta in c(0.05, 0.1)) {
  xcv <- xgb.cv(params = list(objective = "binary:logistic", max_depth = depth, eta = eta,
                              subsample = 0.9, colsample_bytree = 0.8, eval_metric = "aucpr"),
                data = dtrain, nrounds = 600, nfold = 5,
                early_stopping_rounds = 25, verbose = 0, maximize = TRUE)
  score <- max(xcv$evaluation_log$test_aucpr_mean)
  nr <- if (!is.null(xcv$best_iteration)) xcv$best_iteration else nrow(xcv$evaluation_log)
  if (is.null(best) || score > best$score) best <- list(depth = depth, eta = eta, nrounds = nr, score = score)
}
m_xgb <- xgb.train(params = list(objective = "binary:logistic", max_depth = best$depth, eta = best$eta,
                                 subsample = 0.9, colsample_bytree = 0.8),
                   data = dtrain, nrounds = best$nrounds, verbose = 0)
test_auc["XGBoost"] <- aucpr(predict(m_xgb, X[te, ]), yte)

round(test_auc, 3)
cat(sprintf("Best ML gain over the logistic RSF on the held-out test set: +%.3f AUC-PR\n",
            max(test_auc[-1]) - test_auc["Logit"]))
```

```{r cv-plot, echo=FALSE, fig.height=3, fig.cap="Held-out test-set AUC-PR (higher is better). Grey = published logistic RSF; blue = best ML."}
par(mar = c(4, 4, 2.5, 1))
cols <- c("grey70", "#a6cee3", "#a6cee3", "#a6cee3"); cols[which.max(test_auc[-1]) + 1] <- "#1f78b4"
bp <- barplot(test_auc, col = cols, ylim = c(0, 1.12), ylab = "AUC-PR", main = "Pinyon Jay: held-out test AUC-PR by model")
text(bp, test_auc, sprintf("%.3f", test_auc), pos = 3, cex = 0.85)
```

On the held-out test set, random forest and XGBoost lift AUC-PR over the logistic RSF — the nonlinear,
interaction-rich structure of habitat selection that a linear-in-the-logit model cannot capture.

# Interpreting the model: importance, SHAP, and partial dependence

We open the fitted learners three ways: **permutation importance**, **SHAP values** (exact TreeSHAP — the
signed contribution of each variable to each prediction), and **partial dependence** (the shape of an effect).

```{r interp-fit}
# Random forest (permutation importance + partial dependence) on all the data
rf <- ranger(x = X, y = factor(y), num.trees = 800, probability = TRUE, importance = "permutation")

# XGBoost fit on all the data, then exact TreeSHAP contributions per observation
xgb <- xgb.train(params = list(objective = "binary:logistic", max_depth = 4, eta = 0.05,
                               subsample = 0.9, colsample_bytree = 0.8),
                 data = xgb.DMatrix(X, label = y), nrounds = 300, verbose = 0)
shap <- predict(xgb, X, predcontrib = TRUE)   # last column is the model bias
shap <- shap[, colnames(X)]                    # drop the BIAS column
```

```{r imp-plot, echo=FALSE, fig.height=3.2, fig.cap="Permutation importance (left) and mean-absolute-SHAP importance (right). The two rankings agree."}
par(mfrow = c(1, 2), mar = c(4, 8.5, 2.5, 1))
imp_rf <- tail(sort(rf$variable.importance), 8)
barplot(imp_rf, horiz = TRUE, las = 1, names.arg = clean(names(imp_rf)), cex.names = 0.75,
        col = "#33a02c", main = "RF importance", xlab = "permutation importance")
imp_sh <- tail(sort(colMeans(abs(shap))), 8)
barplot(imp_sh, horiz = TRUE, las = 1, names.arg = clean(names(imp_sh)), cex.names = 0.75,
        col = "#6a3d9a", main = "SHAP importance", xlab = "mean(|SHAP|)")
```

```{r beeswarm, echo=FALSE, fig.height=3.6, fig.cap="SHAP summary. Each point is one location; x = its SHAP contribution to the log-odds of use; colour = the variable's value (blue low, red high)."}
# beeswarm: for each top feature, plot each observation's SHAP value, jittered
# vertically and coloured by the (normalized) feature value
top <- names(sort(colMeans(abs(shap)), decreasing = TRUE))[1:8]
set.seed(1); s <- sample(nrow(shap), min(600, nrow(shap)))   # subsample points to keep the PDF small
par(mar = c(4, 8.5, 2.5, 1))
plot(NA, xlim = range(shap[, top]), ylim = c(0.5, length(top) + 0.5), yaxt = "n",
     xlab = "SHAP value (log-odds of use)", ylab = "", main = "SHAP summary")
abline(v = 0, col = "grey70"); axis(2, seq_along(top), clean(rev(top)), las = 1, cex.axis = 0.7)
for (i in seq_along(top)) {
  v <- rev(top)[i]
  val <- (X[s, v] - min(X[, v])) / (max(X[, v]) - min(X[, v]))   # 0 = low, 1 = high
  points(shap[s, v], i + runif(length(s), -0.25, 0.25), pch = 16, cex = 0.5,
         col = rgb(val, 0.15, 1 - val, 0.45))
}
legend("topright", c("high value", "low value"), pch = 16,
       col = c(rgb(1, .15, 0, .8), rgb(0, .15, 1, .8)), bty = "n", cex = 0.7)
```

**Interpreting the relationships.** The SHAP summary reads directly as habitat selection: use rises as
**stand-density index falls** (open blue points sit at positive SHAP, dense red at negative) and **closer to
edges**, with **woody debris** present. In plain terms, pinyon jays select **open pinyon–juniper woodland
near edges with downed wood** — and avoid dense, closed stands.

```{r pdp, echo=FALSE, fig.height=3.2, fig.cap="Partial dependence (left) and SHAP dependence (right) for the most influential variable."}
top1 <- names(sort(colMeans(abs(shap)), decreasing = TRUE))[1]
par(mfrow = c(1, 2), mar = c(4, 4.2, 2.5, 1))

# partial dependence: vary top1 over its range, hold every other predictor at its median
grid <- seq(quantile(d[[top1]], .02), quantile(d[[top1]], .98), length.out = 50)
nd <- matrix(apply(X, 2, median), nrow = 50, ncol = ncol(X), byrow = TRUE,
             dimnames = list(NULL, colnames(X)))
nd[, top1] <- grid
pd <- predict(m_rf, nd)$predictions[, "1"]
plot(grid, pd, type = "l", lwd = 2.5, col = "#1f78b4", xlab = clean(top1),
     ylab = "P(use)", main = "Partial dependence")

# SHAP dependence: how top1's SHAP contribution varies with its value
plot(X[s, top1], shap[s, top1], pch = 16, col = rgb(.42, .24, .60, .3), xlab = clean(top1),
     ylab = "SHAP value", main = "SHAP dependence")
lines(lowess(X[, top1], shap[, top1], f = 0.5), col = "#e31a1c", lwd = 2.5)
abline(h = 0, lty = 2, col = "grey60")
```

Both curves show a strongly **nonlinear** effect — the contribution turns over across the range rather than
following one logit slope, which (with the interactions SHAP captures locally) is why the learners beat the
linear RSF.

# Takeaway

Birds select habitat at nonlinear optima and via interactions — where the linear RSF struggles and trees
excel — and SHAP lets us read those relationships (open stands, edges, woody debris) directly off the fitted
model. The first of two ecology cases showing ML both predicts better *and* interprets cleanly.


# Recommended exercises

1. Vary the random-forest hyperparameters (number of trees, features per split) and report the effect on AUC-PR.
2. Generate a partial-dependence plot for a second habitat variable and describe the nonlinearity.
3. Compare RF vs XGBoost AUC-PR on the same folds; which wins, and by how much?
