---
title: "Reproduce-then-Extend — Pinyon Jay Habitat Selection (Ecology) (Day 2)"
author: "Intro to Machine Learning · Bruce A. Desmarais"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: true
geometry: margin=1in
fontsize: 10pt
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
  - \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 3.6, fig.align = "center")
options(width = 80)
library(glmnet)   # elastic net / lasso / ridge
library(ranger)   # random forests
library(xgboost)  # gradient boosting
library(PRROC)    # precision-recall / ROC curves
library(knitr)    # tables
library(e1071)    # skewness / kurtosis
library(ggplot2)  # correlation heatmap
```

> **Published study.** Daw, S. K. et al. (2022). "Behavior-specific occurrence patterns of Pinyon Jays."
> *PLoS ONE* (doi:10.1371/journal.pone.0237621; S1 Dataset, open). A **logistic resource-selection
> function (RSF)** — habitat *use* points (1) vs *available* points (0) on environmental predictors. Use is
> the minority class, so out-of-sample skill is scored by **AUC-PR** (area under the precision-recall curve).

# Background

Pinyon Jays are a **declining bird** of the southwestern United States, and managers need to know **which habitat they use** in order to conserve it. Daw and colleagues fit a **resource-selection function** -- a logistic regression comparing locations the jays *used* to *available* background locations -- to identify the environmental conditions associated with jay occurrence. The **primary conclusion** is that jays select particular pinyon-juniper conditions -- more open stands, sites near habitat edges, and areas with downed woody debris -- which the regression translates into a picture of suitable habitat for management.

# Data and codebook

**Unit of analysis:** a sampled ground location -- an observed jay-**use** point vs an **available**
(background) point; n = 466 (124 use).

| Variable | Definition |
|---|---|
| `use` | 1 = observed Pinyon Jay use location, 0 = available location **(outcome)** |
| `Elevation..feet.` | elevation (feet) |
| `Slope..percent.` | ground slope (%) |
| `Stand_Age..years.` | forest stand age (years) |
| `Stand_Density_Index` | stand density index (tree density / competition) |
| `Canopy_Cover..percent.` | overstory canopy cover (%) |
| `Tree_Cover..percent.` | tree cover (%) |
| `Shrub_Cover..percent.` | shrub cover (%) |
| `Forb_Cover..percent.` | forb (herbaceous) cover (%) |
| `Grass_Cover..percent.` | grass cover (%) |
| `Woody_Debris..category.1.`-`.3.` | counts of downed woody debris by size class (1-3) |
| `Distance.to.Road..km.` | distance to the nearest road (km) |
| `DisturbanceYes` | 1 if any recent disturbance was recorded at the site |
| `Distance.to.Edge..m.` | distance to the nearest habitat edge (m) |

# Descriptive analysis

```{r data}
d <- read.csv(paste0(
  "https://raw.githubusercontent.com/bdesmarais/intro",
  "-ml-statistical-horizons-4day/main/tutorials/day4_",
  "classic_learners/06_pinyonjay_rsf/data/jays_clean.",
  "csv"))

# Predictors: keep the interpretable environmental variables and drop the sparse
# one-hot habitat/disturbance-type dummies (unidentified in the logit).
preds <- setdiff(names(d), "use")
preds <- preds[sapply(d[preds], function(x) length(unique(x)) > 1)]
preds <- preds[!grepl("Habitat_Type|Disturbance_Type", preds)]

clean <- function(v) gsub("[._]+", " ", v)   # tidy mangled names for labels
short <- function(v) {                        # compact labels for tables/axes
  v <- gsub("[._]+", " ", v); v <- gsub("percent", "%", v)
  v <- gsub("category", "cat", v); trimws(v)
}
num <- setdiff(preds, "DisturbanceYes")       # 14 numeric environmental vars
```

## Dataset and provenance

The unit of analysis is a **sampled ground location** in pinyon–juniper woodland of
the southwestern United States. Each row is either a location where Pinyon Jays were
**observed** (a *use* point) or a randomly drawn **available** (background) point that
characterizes what habitat was on offer. This use-vs-available design is the standard
setup for a **resource-selection function**: the model learns which environmental
conditions distinguish used from available locations. The data are the open **S1
Dataset** from Daw et al. (2022, *PLoS ONE*).

There are **`r nrow(d)` locations** (`r sum(d$use)` use, `r sum(1 - d$use)`
available) and **`r length(preds)` predictors** retained for modeling — `r length(num)`
continuous environmental measurements plus one binary disturbance indicator. (The raw
file also carries sparse one-hot `Habitat_Type*` and `Disturbance_Type*` dummies; these
are dropped above because they are near-collinear and unidentified in the logistic RSF.)

```{r dict, echo=FALSE}
dict <- data.frame(
  Variable = c("use", "Elevation..feet.", "Slope..percent.",
    "Stand_Age..years.", "Stand_Density_Index", "Canopy_Cover..percent.",
    "Tree_Cover..percent.", "Shrub_Cover..percent.", "Forb_Cover..percent.",
    "Grass_Cover..percent.", "Woody_Debris..category.1.",
    "Woody_Debris..category.2.", "Woody_Debris..category.3.",
    "Distance.to.Road..km.", "DisturbanceYes", "Distance.to.Edge..m."),
  Role = c("outcome", rep("predictor", 15)),
  Type = c("binary", "numeric", "numeric", "numeric", "numeric", "numeric",
    "numeric", "numeric", "numeric", "numeric", "count", "count", "count",
    "numeric", "binary", "numeric"),
  Description = c("1 = jay use location, 0 = available",
    "elevation (feet)", "ground slope (%)", "forest stand age (years)",
    "stand density index (competition)", "overstory canopy cover (%)",
    "tree cover (%)", "shrub cover (%)", "forb (herb) cover (%)",
    "grass cover (%)", "downed wood, small size class",
    "downed wood, medium size class", "downed wood, large size class",
    "distance to nearest road (km)", "1 if recent disturbance recorded",
    "distance to nearest habitat edge (m)"))
kable(dict, caption = "Variable dictionary: outcome and retained predictors.")
```

## The outcome: used vs available

The outcome `use` is **binary and imbalanced**: only
**`r round(100 * mean(d$use), 1)`%** of locations are use points, so *available* is the
majority class at a **`r round(mean(d$use == 0) / mean(d$use == 1), 1)`:1** ratio. This
base rate is the single most important fact for evaluation. A trivial model that always
predicts "available" would already be `r round(100 * mean(d$use == 0), 1)`% accurate, so
**raw accuracy is misleading** here — a good classifier must find the minority *use*
class. That is exactly why the tutorial scores held-out skill with **AUC-PR** (area under
the precision–recall curve), whose no-skill baseline equals the base rate
(`r round(mean(d$use), 3)`), not 0.5.

```{r outcome-bar, echo=FALSE, fig.height=2.8, fig.cap="Class balance of the use-vs-available outcome. Use is the minority class."}
par(mar = c(4, 4, 2.5, 1))
bp <- barplot(table(d$use), names.arg = c("available (0)", "used (1)"),
              col = c("grey75", "#1f78b4"), ylab = "count",
              main = "Outcome balance", ylim = c(0, 380))
text(bp, table(d$use), table(d$use), pos = 3, cex = 0.9)
```

## Univariate predictor distributions

The `r length(num)` continuous predictors span very different scales (elevation in
thousands of feet, cover in percent, debris as small counts), so we summarize them in a
single table with location, spread, quartiles and **skewness**, then show the shapes as
faceted histograms.

```{r pred-table, echo=FALSE}
ptab <- t(sapply(d[num], function(x) c(
  n = length(x), n_miss = sum(is.na(x)), mean = mean(x), sd = sd(x),
  min = min(x), Q1 = quantile(x, .25), med = median(x),
  Q3 = quantile(x, .75), max = max(x), skew = skewness(x))))
rownames(ptab) <- short(rownames(ptab))
colnames(ptab) <- c("n", "miss", "mean", "sd", "min", "Q1", "med",
                    "Q3", "max", "skew")
kable(round(ptab, 1),
      caption = "Univariate summary of the numeric predictors.")
```

Several patterns stand out. The **cover** and **woody-debris** variables are strongly
**right-skewed** (skew for forb, grass and the debris counts all exceed 2): most
locations have little cover or downed wood, with a long tail of rich sites. **Elevation**
and **canopy cover** are close to symmetric. `Distance.to.Edge..m.` reaches a large
maximum (`r round(max(d[["Distance.to.Edge..m."]]))` m) with a long right tail, and it
contains a few **negative values** (min = `r round(min(d[["Distance.to.Edge..m."]]))`),
an artifact of how signed distance-to-edge was computed in the source GIS layer — worth
noting but harmless for tree-based learners, which respond only to rank order.

```{r pred-hist, echo=FALSE, fig.height=4.6, fig.cap="Distributions of the numeric predictors. Cover and woody-debris variables are right-skewed; elevation and canopy are roughly symmetric."}
op <- par(mfrow = c(4, 4), mar = c(3.2, 2.4, 2, 1), mgp = c(2, 0.6, 0))
for (v in num) hist(d[[v]], col = "#a6cee3", border = "white",
                    main = short(v), xlab = "", cex.main = 0.85)
par(op)
```

The one binary predictor, **DisturbanceYes**, is recorded at
`r round(100 * mean(d$DisturbanceYes), 1)`% of locations
(`r sum(d$DisturbanceYes)` of `r nrow(d)`) — a minority flag for sites with any recorded
recent disturbance.

## Missing data

```{r miss, echo=FALSE}
miss <- sapply(d[c("use", preds)], function(x) sum(is.na(x)))
cat(sprintf("Total missing cells across outcome + predictors: %d\n", sum(miss)))
cat(sprintf("Complete cases: %d of %d (%.0f%%)\n",
            sum(complete.cases(d[c("use", preds)])), nrow(d),
            100 * mean(complete.cases(d[c("use", preds)]))))
```

The analysis matrix is **fully observed** — every outcome and predictor cell is present,
so there is no missingness to model and no co-missingness pattern to inspect. The
published RSF therefore uses all `r nrow(d)` locations with no imputation or listwise
deletion, and the machine-learning comparison inherits the same complete design.

## Bivariate association with the outcome

We next ask which conditions actually separate used from available locations. For each
numeric predictor we report the **point-biserial correlation** with `use`, the
**standardized mean difference** (used minus available, in SDs), and a two-sample *t*-test
*p*-value; the table is sorted by the magnitude of the standardized difference.

```{r biv-table, echo=FALSE}
u <- d$use
biv <- t(sapply(d[num], function(x) c(
  r = cor(x, u),
  smd = (mean(x[u == 1]) - mean(x[u == 0])) / sd(x),
  p = t.test(x ~ u)$p.value)))
rownames(biv) <- short(rownames(biv))
biv <- biv[order(-abs(biv[, "smd"])), ]
biv <- data.frame(round(biv[, c("r", "smd")], 3),
                  p = format.pval(biv[, "p"], digits = 2, eps = 1e-4))
kable(biv, caption = "Marginal association of each numeric predictor with use, sorted by |std. mean difference|.")
```

```{r biv-plot, echo=FALSE, fig.height=3.2, fig.cap="Standardized mean difference (used - available) for each predictor (left) and the strongest gradient, distance-to-edge, by outcome (right)."}
par(mfrow = c(1, 2))
smd <- sapply(d[preds], function(x)
  (mean(x[u == 1]) - mean(x[u == 0])) / sd(x))
smd <- tail(smd[order(abs(smd))], 10)
par(mar = c(4, 8.2, 2.5, 1))
barplot(smd, horiz = TRUE, las = 1, names.arg = short(names(smd)),
        cex.names = 0.7, col = ifelse(smd > 0, "#1f78b4", "#e31a1c"),
        main = "Used vs available", xlab = "std. mean diff.")
abline(v = 0, col = "grey50")

par(mar = c(4, 4, 2.5, 1))
boxplot(d[["Distance.to.Edge..m."]] ~ u, names = c("available", "used"),
        col = c("grey75", "#1f78b4"), ylab = "distance to edge (m)",
        xlab = "", main = "Distance to edge")
```

The signal is clear and ecologically sensible. Use rises with **downed woody debris**
(all three size classes positive), and falls sharply with **distance to edge**,
**distance to road**, **slope**, **stand density** and **elevation** — jays select
**open, low-density stands near edges with downed wood**, and avoid steep, dense,
high-elevation interior sites. Tree and shrub cover show essentially **no marginal
association** (|r| < 0.03), a hint that their role, if any, is nonlinear or conditional
rather than a simple linear gradient. For the binary flag:

```{r dist-tab, echo=FALSE}
tab <- table(Disturbance = d$DisturbanceYes, use = d$use)
cat("Use rate by disturbance flag:\n")
print(round(prop.table(tab, 1)[, "1"], 3))
cat(sprintf("chi-square test of independence: p = %s\n",
            format.pval(chisq.test(tab)$p.value, digits = 2, eps = 1e-4)))
```

Disturbed sites are used far more often
(`r round(100 * prop.table(tab, 1)["1", "1"])`% vs
`r round(100 * prop.table(tab, 1)["0", "1"])`%), a strong and highly significant
association.

## Multivariate structure and collinearity

Finally we check whether the predictors are redundant with one another. The heatmap shows
pairwise correlations among the numeric predictors, and we compute **variance-inflation
factors** from the reproduction model (via auxiliary regressions, since a linear RSF
shares the design matrix).

```{r corr-heat, echo=FALSE, fig.height=4.2, fig.cap="Correlation heatmap of the numeric predictors. No pair exceeds |r| = 0.7; the strongest is canopy vs stand-density index."}
cm <- cor(d[num]); colnames(cm) <- rownames(cm) <- short(num)
cmd <- data.frame(as.table(cm))
names(cmd) <- c("v1", "v2", "r")
ggplot(cmd, aes(v1, v2, fill = r)) + geom_tile() +
  scale_fill_gradient2(low = "#b2182b", mid = "white",
                       high = "#2166ac", midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "r") +
  theme_minimal(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())
```

```{r vif, echo=FALSE}
Xp <- d[preds]
vif <- sapply(preds, function(v) {
  r2 <- summary(lm(Xp[[v]] ~ ., data = Xp[setdiff(preds, v)]))$r.squared
  1 / (1 - r2)
})
vif <- sort(vif, decreasing = TRUE)
cat("Largest off-diagonal |r| among numeric predictors:",
    round(max(abs(cm[upper.tri(cm)])), 3), "\n")
cat("Variance-inflation factors (top 5):\n")
print(round(head(vif, 5), 2))
```

Collinearity is **mild**: the strongest correlation is only
`r round(max(abs(cm[upper.tri(cm)])), 2)` (canopy cover vs stand-density index — both
overstory-structure measures), **no pair exceeds |r| = 0.7**, and the **maximum VIF is
`r round(max(vif), 2)`**, far below the usual threshold of 5–10. The predictors therefore
carry largely distinct information; the coefficients in the reproduction are individually
interpretable, and the modest correlations that do exist (structure, debris size classes)
are the kind of interaction-rich texture the tree-based learners can exploit later.

# Reproduce the published RSF regression

The published model is a logistic regression of use vs availability on the environmental predictors.

```{r reproduce}
rsf <- glm(reformulate(preds, "use"), data = d, family = binomial)
round(summary(rsf)$coefficients, 3)
```

**Confirmation against the published study.** The signed coefficients recover the reported habitat-selection
pattern — jays select more **open** stands (negative on `Stand_Density_Index` and `Slope..percent.`) and sites
**closer to edges** (negative on `Distance.to.Edge..m.`). This logistic RSF is our regression baseline.

# Predictive results: regression vs machine learning

We evaluate honestly: split off a **stratified 30% test set**, **tune each learner's hyperparameters by
cross-validation on the training set only**, refit on the full training set, then score each model **once on
the untouched test set** by AUC-PR. Nothing about the test set is used for fitting or tuning.

```{r cv}
X <- as.matrix(d[, preds]); y <- d$use
aucpr <- function(p, yt) pr.curve(scores.class0 = p[yt == 1], scores.class1 = p[yt == 0])$auc.integral

set.seed(2026)
te <- c(sample(which(y == 1), round(0.30 * sum(y == 1))),      # stratified 30% test set
        sample(which(y == 0), round(0.30 * sum(y == 0))))
tr <- setdiff(seq_len(nrow(d)), te); yte <- y[te]

## 1. Logistic RSF -- the published regression (no hyperparameters)
m_logit <- glm(use ~ ., data = d[tr, c("use", preds)], family = binomial)
test_auc <- c(Logit = aucpr(predict(m_logit, d[te, ], type = "response"), yte))

## 2. Elastic net -- penalty lambda tuned by CV on the TRAINING set
m_en <- cv.glmnet(X[tr, ], y[tr], alpha = 0.5, family = "binomial")
test_auc["ElasticNet"] <- aucpr(as.numeric(predict(m_en, X[te, ], s = "lambda.min", type = "response")), yte)

## 3. Random forest -- mtry tuned on the TRAINING set by out-of-bag AUC-PR
best_rf <- NULL
for (mt in unique(pmax(1, c(floor(sqrt(ncol(X))), floor(ncol(X) / 3), floor(ncol(X) / 2))))) {
  rf <- ranger(x = X[tr, ], y = factor(y[tr]), num.trees = 800, mtry = mt, probability = TRUE)
  sc <- aucpr(rf$predictions[, "1"], y[tr])          # OOB predictions = a training-only CV estimate
  if (is.null(best_rf) || sc > best_rf$sc) best_rf <- list(mtry = mt, sc = sc)
}
m_rf <- ranger(x = X[tr, ], y = factor(y[tr]), num.trees = 800, mtry = best_rf$mtry, probability = TRUE)
test_auc["RandomForest"] <- aucpr(predict(m_rf, X[te, ])$predictions[, "1"], yte)

## 4. XGBoost -- depth x eta tuned by xgb.cv on the TRAINING set; nrounds by early stopping
dtrain <- xgb.DMatrix(X[tr, ], label = y[tr])
best <- NULL
for (depth in c(3, 4, 6)) for (eta in c(0.05, 0.1)) {
  xcv <- xgb.cv(params = list(objective = "binary:logistic", max_depth = depth, eta = eta,
                              subsample = 0.9, colsample_bytree = 0.8, eval_metric = "aucpr"),
                data = dtrain, nrounds = 600, nfold = 5,
                early_stopping_rounds = 25, verbose = 0, maximize = TRUE)
  score <- max(xcv$evaluation_log$test_aucpr_mean)
  nr <- if (!is.null(xcv$best_iteration)) xcv$best_iteration else nrow(xcv$evaluation_log)
  if (is.null(best) || score > best$score) best <- list(depth = depth, eta = eta, nrounds = nr, score = score)
}
m_xgb <- xgb.train(params = list(objective = "binary:logistic", max_depth = best$depth, eta = best$eta,
                                 subsample = 0.9, colsample_bytree = 0.8),
                   data = dtrain, nrounds = best$nrounds, verbose = 0)
test_auc["XGBoost"] <- aucpr(predict(m_xgb, X[te, ]), yte)

round(test_auc, 3)
cat(sprintf("Best ML gain over the logistic RSF on the held-out test set: +%.3f AUC-PR\n",
            max(test_auc[-1]) - test_auc["Logit"]))
```

```{r cv-plot, echo=FALSE, fig.height=3, fig.cap="Held-out test-set AUC-PR (higher is better). Grey = published logistic RSF; blue = best ML."}
par(mar = c(4, 4, 2.5, 1))
cols <- c("grey70", "#a6cee3", "#a6cee3", "#a6cee3"); cols[which.max(test_auc[-1]) + 1] <- "#1f78b4"
bp <- barplot(test_auc, col = cols, ylim = c(0, 1.12), ylab = "AUC-PR", main = "Pinyon Jay: held-out test AUC-PR by model")
text(bp, test_auc, sprintf("%.3f", test_auc), pos = 3, cex = 0.85)
```

On the held-out test set, random forest and XGBoost lift AUC-PR over the logistic RSF — the nonlinear,
interaction-rich structure of habitat selection that a linear-in-the-logit model cannot capture.

# Interpreting the model: importance, SHAP, and partial dependence

We open the fitted learners three ways: **permutation importance**, **SHAP values** (exact TreeSHAP — the
signed contribution of each variable to each prediction), and **partial dependence** (the shape of an effect).

```{r interp-fit}
# Random forest (permutation importance + partial dependence) on all the data
rf <- ranger(x = X, y = factor(y), num.trees = 800, probability = TRUE, importance = "permutation")

# XGBoost fit on all the data, then exact TreeSHAP contributions per observation
xgb <- xgb.train(params = list(objective = "binary:logistic", max_depth = 4, eta = 0.05,
                               subsample = 0.9, colsample_bytree = 0.8),
                 data = xgb.DMatrix(X, label = y), nrounds = 300, verbose = 0)
shap <- predict(xgb, X, predcontrib = TRUE)   # last column is the model bias
shap <- shap[, colnames(X)]                    # drop the BIAS column
```

```{r imp-plot, echo=FALSE, fig.height=3.2, fig.cap="Permutation importance (left) and mean-absolute-SHAP importance (right). The two rankings agree."}
par(mfrow = c(1, 2), mar = c(4, 8.5, 2.5, 1))
imp_rf <- tail(sort(rf$variable.importance), 8)
barplot(imp_rf, horiz = TRUE, las = 1, names.arg = clean(names(imp_rf)), cex.names = 0.75,
        col = "#33a02c", main = "RF importance", xlab = "permutation importance")
imp_sh <- tail(sort(colMeans(abs(shap))), 8)
barplot(imp_sh, horiz = TRUE, las = 1, names.arg = clean(names(imp_sh)), cex.names = 0.75,
        col = "#6a3d9a", main = "SHAP importance", xlab = "mean(|SHAP|)")
```

```{r beeswarm, echo=FALSE, fig.height=3.6, fig.cap="SHAP summary. Each point is one location; x = its SHAP contribution to the log-odds of use; colour = the variable's value (blue low, red high)."}
# beeswarm: for each top feature, plot each observation's SHAP value, jittered
# vertically and coloured by the (normalized) feature value
top <- names(sort(colMeans(abs(shap)), decreasing = TRUE))[1:8]
set.seed(1); s <- sample(nrow(shap), min(600, nrow(shap)))   # subsample points to keep the PDF small
par(mar = c(4, 8.5, 2.5, 1))
plot(NA, xlim = range(shap[, top]), ylim = c(0.5, length(top) + 0.5), yaxt = "n",
     xlab = "SHAP value (log-odds of use)", ylab = "", main = "SHAP summary")
abline(v = 0, col = "grey70"); axis(2, seq_along(top), clean(rev(top)), las = 1, cex.axis = 0.7)
for (i in seq_along(top)) {
  v <- rev(top)[i]
  val <- (X[s, v] - min(X[, v])) / (max(X[, v]) - min(X[, v]))   # 0 = low, 1 = high
  points(shap[s, v], i + runif(length(s), -0.25, 0.25), pch = 16, cex = 0.5,
         col = rgb(val, 0.15, 1 - val, 0.45))
}
legend("topright", c("high value", "low value"), pch = 16,
       col = c(rgb(1, .15, 0, .8), rgb(0, .15, 1, .8)), bty = "n", cex = 0.7)
```

**Interpreting the relationships.** The SHAP summary reads directly as habitat selection: use rises as
**stand-density index falls** (open blue points sit at positive SHAP, dense red at negative) and **closer to
edges**, with **woody debris** present. In plain terms, pinyon jays select **open pinyon–juniper woodland
near edges with downed wood** — and avoid dense, closed stands.

```{r pdp, echo=FALSE, fig.height=3.2, fig.cap="Partial dependence (left) and SHAP dependence (right) for the most influential variable."}
top1 <- names(sort(colMeans(abs(shap)), decreasing = TRUE))[1]
par(mfrow = c(1, 2), mar = c(4, 4.2, 2.5, 1))

# partial dependence: vary top1 over its range, hold every other predictor at its median
grid <- seq(quantile(d[[top1]], .02), quantile(d[[top1]], .98), length.out = 50)
nd <- matrix(apply(X, 2, median), nrow = 50, ncol = ncol(X), byrow = TRUE,
             dimnames = list(NULL, colnames(X)))
nd[, top1] <- grid
pd <- predict(m_rf, nd)$predictions[, "1"]
plot(grid, pd, type = "l", lwd = 2.5, col = "#1f78b4", xlab = clean(top1),
     ylab = "P(use)", main = "Partial dependence")

# SHAP dependence: how top1's SHAP contribution varies with its value
plot(X[s, top1], shap[s, top1], pch = 16, col = rgb(.42, .24, .60, .3), xlab = clean(top1),
     ylab = "SHAP value", main = "SHAP dependence")
lines(lowess(X[, top1], shap[, top1], f = 0.5), col = "#e31a1c", lwd = 2.5)
abline(h = 0, lty = 2, col = "grey60")
```

Both curves show a strongly **nonlinear** effect — the contribution turns over across the range rather than
following one logit slope, which (with the interactions SHAP captures locally) is why the learners beat the
linear RSF.

# Takeaway

Birds select habitat at nonlinear optima and via interactions — where the linear RSF struggles and trees
excel — and SHAP lets us read those relationships (open stands, edges, woody debris) directly off the fitted
model. The first of two ecology cases showing ML both predicts better *and* interprets cleanly.


# Recommended exercises

1. Vary the random-forest hyperparameters (number of trees, features per split) and report the effect on AUC-PR.
2. Generate a partial-dependence plot for a second habitat variable and describe the nonlinearity.
3. Compare RF vs XGBoost AUC-PR on the same folds; which wins, and by how much?
