---
title: "Reproduce-then-Extend — Building Energy Efficiency (Engineering) (Day 1)"
author: "Intro to Machine Learning · Bruce A. Desmarais"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: true
geometry: margin=1in
fontsize: 10pt
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
  - \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 3.6, fig.align = "center")
options(width = 80)
library(readxl)    # read the .xlsx data file
library(ggplot2)   # EDA plots
library(dplyr)     # EDA data wrangling
library(tidyr)     # reshaping for faceted / tile plots
library(knitr)     # kable tables
library(patchwork) # compose multi-panel figures
library(e1071)     # skewness() and kurtosis()
theme_set(theme_minimal(base_size = 9))
```

> **Published study.** Tsanas, A. & Xifara, A. (2012). "Accurate quantitative estimation of energy performance
> of residential buildings using statistical machine learning tools." *Energy and Buildings* 49:560–567. The
> paper's baseline is a **linear regression** predicting a building's heating load from 8 geometry inputs
> (n = 768; UCI id 242). This example illustrates the first regression extension: **polynomial terms**.

# Background

Tsanas and Xifara study how a building's **energy performance** depends on its **design geometry**, to help architects design more efficient buildings. Using 768 simulated building shapes, they estimate the **heating (and cooling) load** from eight geometry inputs, using a **linear regression** as the interpretable baseline alongside nonlinear machine-learning tools. The **primary conclusion** is that energy load is accurately predictable from geometry -- overall height, relative compactness, and glazing are the strongest drivers -- and that flexible learners (random forests) substantially outperform the linear model, motivating their use for this strongly nonlinear relationship.

# Data and codebook

**Unit of analysis:** a simulated residential **building design** (12 shapes and variants, via Ecotect); n = 768.

| Variable | Definition |
|---|---|
| `Y1` | heating load (kWh per m^2 of floor area) **(outcome)** |
| `X1` | relative compactness (volume-to-surface ratio, normalized) |
| `X2` | surface area (m^2) |
| `X3` | wall area (m^2) |
| `X4` | roof area (m^2) |
| `X5` | overall height (m) |
| `X6` | orientation (2/3/4/5 = N/E/S/W) |
| `X7` | glazing area (fraction of floor area) |
| `X8` | glazing-area distribution (0-5 scheme) |

# Descriptive analysis

Before fitting anything, we profile the data thoroughly. The goal is to
understand the outcome's shape, each geometry input's distribution, how the
inputs relate to heating load, and — critically for a regression — how the
inputs relate to *each other*.

```{r data}
tmp_xlsx <- tempfile(fileext = ".xlsx")
download.file(paste0(
  "https://raw.githubusercontent.com/bdesmarais/",
  "intro-ml-statistical-horizons-4day/main/",
  "tutorials/day2_regression_extensions/09_energy_efficiency/data/ENB2012_data.xlsx"), tmp_xlsx, mode = "wb", quiet = TRUE)
d <- as.data.frame(read_excel(tmp_xlsx))   # data ships in the tutorial folder
d <- d[, colSums(!is.na(d)) > 0]; d <- d[complete.cases(d), ]
preds <- paste0("X", 1:8)
```

## Dataset and provenance

The **unit of analysis** is a *simulated residential building design*. Angeliki
Xifara generated 12 base building shapes in the Ecotect simulator and varied
their geometry (glazing, orientation, etc.), producing **n = `r nrow(d)`**
distinct designs. For each design a physics simulation reports the annual
**heating load** (`Y1`) and cooling load (`Y2`); this tutorial models heating
load. The data are deterministic simulation output, not a survey sample, so
there is **no sampling error and no missingness** — every input combination is
fully observed. The predictors are `r length(preds)` geometry variables. The
data come from the UCI Machine Learning Repository (id 242), released with
Tsanas & Xifara (2012).

```{r dict}
dict <- data.frame(
  Variable = c("Y1", paste0("X", 1:8)),
  Role = c("outcome", rep("predictor", 8)),
  Type = c("numeric", "numeric", "numeric", "numeric", "numeric",
           "binary (2 levels)", "categorical (4 orient.)",
           "ordinal (4 levels)", "ordinal (6 levels)"),
  Description = c(
    "heating load (kWh per m^2 floor)",
    "relative compactness (volume/surface, normalized)",
    "surface area (m^2)", "wall area (m^2)", "roof area (m^2)",
    "overall height (3.5 or 7 m)", "orientation (2/3/4/5 = N/E/S/W)",
    "glazing area (0, .1, .25, .4 of floor)",
    "glazing distribution (0-5 scheme)"))
kable(dict, caption = "Variable dictionary: outcome and predictors.")
```

Although `X1`-`X5` are stored as continuous numbers, the simulation grid means
several take only a handful of distinct values (`X5` is effectively binary:
3.5 m or 7 m; `X6`-`X8` are categorical design choices). We treat them
numerically to match the published baseline, but keep their discreteness in
mind when reading the plots.

## The outcome in depth

```{r y-stats}
ystat <- function(x) c(
  n = length(x), mean = mean(x), sd = sd(x), min = min(x),
  Q1 = quantile(x, .25), median = median(x), Q3 = quantile(x, .75),
  max = max(x), IQR = IQR(x),
  skew = e1071::skewness(x), kurtosis = e1071::kurtosis(x))
kable(t(round(ystat(d$Y1), 2)),
      caption = "Distribution of heating load (Y1).")
```

```{r y-dist, fig.height=2.6, fig.cap="Heating load: histogram with density overlay (left) and boxplot (right). The distribution is broad and bimodal rather than a single bell."}
p_hist <- ggplot(d, aes(Y1)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "#a6cee3", colour = "white") +
  geom_density(colour = "#1f78b4", linewidth = 0.7) +
  labs(x = "heating load (Y1)", y = "density", title = "Distribution")
p_box <- ggplot(d, aes(y = Y1)) +
  geom_boxplot(fill = "#a6cee3", width = 0.4) +
  labs(y = "heating load (Y1)", x = NULL, title = "Boxplot") +
  theme(axis.text.x = element_blank())
p_hist + p_box + plot_layout(widths = c(2.4, 1))
```

Heating load ranges from about `r round(min(d$Y1))` to
`r round(max(d$Y1))` kWh/m^2 (mean `r round(mean(d$Y1), 1)`, sd
`r round(sd(d$Y1), 1)`). Skewness is only
`r round(e1071::skewness(d$Y1), 2)` — nearly symmetric — but the negative
excess kurtosis (`r round(e1071::kurtosis(d$Y1), 2)`) and the histogram reveal
a **bimodal, plateau-shaped** distribution, not a bell. This is a fingerprint
of the design: buildings cluster into low-load and high-load regimes driven
mostly by overall height. There is no zero-inflation, floor/ceiling piling, or
long tail, so **no outcome transformation is warranted** — the spread is
structural, and the regression's job is to explain it with the geometry inputs.

## Univariate predictor distributions

```{r pred-summary}
psum <- function(x) c(
  n = length(x), n_miss = sum(is.na(x)), mean = mean(x), sd = sd(x),
  min = min(x), Q1 = quantile(x, .25), median = median(x),
  Q3 = quantile(x, .75), max = max(x), skew = e1071::skewness(x),
  n_unique = length(unique(x)))
ptab <- t(sapply(d[preds], psum))
kable(round(ptab, 2),
      caption = "Univariate summary of the 8 geometry predictors.")
```

```{r pred-hist, fig.height=3.4, fig.cap="Distribution of each geometry input. Several inputs take only a few discrete values (the simulation grid), so they appear as spikes rather than smooth curves."}
d %>%
  select(all_of(preds)) %>%
  pivot_longer(everything(), names_to = "var", values_to = "value") %>%
  ggplot(aes(value)) +
  geom_histogram(bins = 20, fill = "#b2df8a", colour = "white") +
  facet_wrap(~ var, scales = "free", ncol = 4) +
  labs(x = NULL, y = "count")
```

The summary table exposes the grid structure: `X5` (height) takes exactly two
values, `X6`-`X8` a handful, and even the "continuous" `X1`-`X4` take 4-12
distinct levels. All predictors are close to symmetric (|skew| < 0.6), so no
predictor transformations are indicated. The near-uniform spikes reflect a
**balanced factorial design** — each geometry level appears in roughly equal
numbers of buildings, which is ideal for isolating individual effects (were it
not for the collinearity documented below).

## Missing-data analysis

```{r missing}
miss <- data.frame(
  variable = c("Y1", preds),
  n_missing = sapply(d[c("Y1", preds)], function(x) sum(is.na(x))),
  pct_missing = round(100 * sapply(d[c("Y1", preds)],
                                   function(x) mean(is.na(x))), 1))
kable(miss, row.names = FALSE,
      caption = "Missing values per variable.")
```

There is **no missing data**: all `r nrow(d)` rows are complete across all
`r length(preds) + 1` variables. Because these are exhaustive simulation
outputs over a designed grid, missingness is impossible by construction — the
`complete.cases()` filter in the loading chunk is a safeguard that removes
nothing here. The reproduction therefore uses the full n = `r nrow(d)` with no
imputation or listwise deletion.

## Bivariate relationships with the outcome

```{r biv-tab}
marg <- data.frame(
  predictor = preds,
  r_with_Y1 = round(sapply(d[preds], function(x) cor(x, d$Y1)), 3))
marg <- marg[order(-abs(marg$r_with_Y1)), ]
kable(marg, row.names = FALSE,
      caption = "Marginal (Pearson) correlation of each input with Y1.")
```

```{r biv-plot, fig.height=3.6, fig.cap="Heating load vs each geometry input, with a linear smoother. Height (X5) and compactness (X1) drive load up; surface and roof area (X2, X4) drive it down."}
d %>%
  select(all_of(preds), Y1) %>%
  pivot_longer(all_of(preds), names_to = "var", values_to = "value") %>%
  ggplot(aes(value, Y1)) +
  geom_point(alpha = 0.15, colour = "#333333", size = 0.6) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              colour = "#e31a1c", linewidth = 0.6) +
  facet_wrap(~ var, scales = "free_x", ncol = 4) +
  labs(x = NULL, y = "heating load (Y1)")
```

Four inputs carry almost all of the marginal signal. **Overall height** (`X5`,
r = `r round(cor(d$X5, d$Y1), 2)`) and **relative compactness** (`X1`,
r = `r round(cor(d$X1, d$Y1), 2)`) are the strongest positive drivers, while
**roof area** (`X4`, r = `r round(cor(d$X4, d$Y1), 2)`) and **surface area**
(`X2`, r = `r round(cor(d$X2, d$Y1), 2)`) pull load down. Orientation (`X6`)
and glazing distribution (`X8`) are essentially uncorrelated with load —
consistent with the physics that *how much* glazing and *how tall* a building
is matter far more than which way it faces. Several scatterplots (notably `X1`)
show visible curvature around the linear smoother, foreshadowing the
polynomial extension below.

## Multivariate structure and collinearity

```{r corr-heat, fig.height=3.4, fig.cap="Correlation heatmap of the 8 geometry inputs. Deep red / blue cells mark |r| > 0.7 — the geometry variables are heavily interdependent."}
cmat <- cor(d[preds])
cmat[lower.tri(cmat)] <- NA
as.data.frame(cmat) %>%
  mutate(v1 = rownames(cmat)) %>%
  pivot_longer(-v1, names_to = "v2", values_to = "r") %>%
  filter(!is.na(r)) %>%
  ggplot(aes(v2, v1, fill = r)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 2.4) +
  scale_fill_gradient2(low = "#e31a1c", mid = "white",
                       high = "#1f78b4", midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL) +
  theme(panel.grid = element_blank())
```

```{r vif}
X <- model.matrix(~ X1 + X2 + X3 + X4 + X5 + X6 + X7 + X8, d)[, -1]
vif <- sapply(colnames(X), function(v) {
  r2 <- summary(lm(X[, v] ~ X[, setdiff(colnames(X), v)]))$r.squared
  1 / (1 - r2)
})
kable(t(round(vif, 1)),
      caption = "Variance inflation factors from the reproduction model.")
```

This is the most important diagnostic in the analysis. The geometry inputs are
**severely collinear**: `X1` and `X2` correlate
`r round(cor(d$X1, d$X2), 2)`, and roof area `X4` vs height `X5` correlate
`r round(cor(d$X4, d$X5), 2)`. Six predictor pairs exceed |r| = 0.7. Worse, the
predictor matrix has **rank `r qr(as.matrix(d[preds]))$rank` of
`r length(preds)`** — an *exact* linear dependency, because surface, wall, and
roof areas are geometrically derived from the same base shapes. That is why the
VIFs for `X2`, `X3`, and `X4` blow up to infinity: those coefficients are not
separately identified. The individual `lm` coefficients on the correlated
geometry terms are therefore **unstable and hard to interpret in isolation**,
even though the model as a whole predicts load extremely well
(R^2 ~ 0.92). This tension — excellent joint prediction but unidentifiable
individual effects — is exactly why the reproduction below focuses on overall
fit and why the extension turns to nonlinear (polynomial) structure in a single
well-identified driver (`X1`) rather than trying to disentangle the collinear
area variables.

# Reproduce the published regression

```{r reproduce}
lin <- lm(Y1 ~ X1 + X2 + X3 + X4 + X5 + X6 + X7 + X8, data = d)
summary(lin)
```

**Confirmation against the published study.** The linear model attains in-sample $R^2$ =
`r round(summary(lin)$r.squared, 3)`, matching the ~0.92 reported by Tsanas & Xifara for their linear baseline.
Overall height (`X5`) and relative compactness (`X1`) dominate — but the fit leaves clear structure, which is
the motivation for going beyond a straight line.

# Polynomial regression extension

Heating load responds **nonlinearly** to geometry. The classic regression remedy is to add **polynomial
terms**. We compare the linear model to polynomials of increasing degree in relative compactness (`X1`),
keeping the other inputs linear. We pick the degree by cross-validation on a training set and report final RMSE on a held-out test set.

```{r poly}
set.seed(2026)
others <- paste0("X", 2:8)
te <- sample(nrow(d), round(0.30 * nrow(d)))            # hold out a 30% test set up front
tr <- setdiff(seq_len(nrow(d)), te)

# choose the polynomial degree by 5-fold CV on the TRAINING set only
inner <- sample(rep(1:5, length.out = length(tr)))
cv_rmse <- sapply(1:4, function(deg) {
  r <- numeric(5)
  for (k in 1:5) {
    itr <- tr[inner != k]; ite <- tr[inner == k]
    form <- reformulate(c(sprintf("poly(X1, %d, raw = TRUE)", deg), others), "Y1")
    r[k] <- sqrt(mean((d$Y1[ite] - predict(lm(form, data = d[itr, ]), d[ite, ]))^2))
  }
  mean(r)
})
names(cv_rmse) <- paste0("degree ", 1:4)
best_deg <- which.min(cv_rmse)
round(cv_rmse, 3)                                    # CV RMSE per degree (training set)

# refit at the CV-chosen degree on the full training set; score ONCE on the untouched test set
form_best <- reformulate(c(sprintf("poly(X1, %d, raw = TRUE)", best_deg), others), "Y1")
test_poly <- sqrt(mean((d$Y1[te] - predict(lm(form_best, data = d[tr, ]), d[te, ]))^2))
test_lin  <- sqrt(mean((d$Y1[te] - predict(lm(reformulate(c("X1", others), "Y1"), data = d[tr, ]), d[te, ]))^2))
cat(sprintf("CV-selected degree: %d.  Held-out test RMSE: polynomial = %.3f vs linear = %.3f
",
            best_deg, test_poly, test_lin))
```

```{r poly-plot, echo=FALSE, fig.height=3.4, fig.cap="Degree-3 polynomial fit of heating load on relative compactness, holding the other inputs at their means."}
fit3 <- lm(reformulate(c("poly(X1, 3, raw = TRUE)", others), "Y1"), data = d)
grid <- seq(quantile(d$X1, .02), quantile(d$X1, .98), length.out = 100)
nd   <- as.data.frame(lapply(d[preds], function(x) rep(mean(x), 100))); nd$X1 <- grid
yhat <- predict(fit3, nd)
par(mar = c(4.2, 4.2, 2.5, 1))
plot(d$X1, d$Y1, pch = 16, col = "#33333322", xlab = "X1 (relative compactness)", ylab = "Y1 (heating load)",
     main = "Degree-3 polynomial fit", ylim = range(c(d$Y1, yhat)))
lines(grid, yhat, col = "#e31a1c", lwd = 3)
```

On the held-out test set the CV-selected polynomial improves on the straight-line fit; the curve **bends** with compactness rather than running straight, capturing nonlinearity the
linear model misses while remaining a fully interpretable regression.

# Takeaway

This example illustrates **polynomial regression** as the first, most interpretable extension of a linear
model: adding low-order polynomial terms lowers cross-validated error by bending the fit to the data's
curvature. (Tree-based learners push this further at the cost of a closed-form equation; Tsanas & Xifara
report ML reaching ~0.999 $R^2$ on these data.)


# Recommended exercises

1. Try polynomial degrees 1-5 for the strongest predictor and use cross-validation to choose the degree.
2. Predict the second response (cooling load) and compare which predictors matter most.
3. Add an interaction between two predictors and test whether cross-validated error improves.
