---
title: "Reproduce-then-Extend — Wine Quality (Food Science / Chemometrics) (Day 1)"
author: "Intro to Machine Learning · Bruce A. Desmarais"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: true
geometry: margin=1in
fontsize: 10pt
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
  - \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 3.6, fig.align = "center")
options(width = 80)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(patchwork)
library(e1071)
```

> **Published study.** Cortez, P. et al. (2009). "Modeling wine preferences by data mining from physicochemical
> properties." *Decision Support Systems* 47(4):547–553. The baseline is a **multiple (linear) regression** of
> sensory quality on 11 physicochemical measurements. Data: UCI (Vinho Verde red wine, n = 1599). This example
> illustrates **polynomial regression** — and using cross-validation to judge whether it is worth it.

# Background

Cortez and colleagues ask whether **objective physicochemical measurements** of a wine can predict its **subjective sensory quality** (the median score of blind tasters), which would support quality certification and production decisions. They model quality from 11 physicochemical properties using several methods, with **multiple linear regression** as the interpretable baseline. The **primary conclusion** is that physicochemical properties do carry real signal -- **alcohol** (positive) and **volatile acidity** (negative) matter most -- but the relationship is modest and nonlinear, so flexible models (support vector machines in their paper) predict quality better than the linear regression.

# Data and codebook

**Unit of analysis:** an individual red 'Vinho Verde' **wine sample**; n = 1,599.

| Variable | Definition |
|---|---|
| `quality` | sensory quality, median of >=3 blind tasters (0-10) **(outcome)** |
| `fixed.acidity` | fixed (tartaric) acidity (g/L) |
| `volatile.acidity` | volatile (acetic) acidity (g/L) |
| `citric.acid` | citric acid (g/L) |
| `residual.sugar` | residual sugar (g/L) |
| `chlorides` | chlorides / salt (g/L) |
| `free.sulfur.dioxide` | free SO2 (mg/L) |
| `total.sulfur.dioxide` | total SO2 (mg/L) |
| `density` | density (g/mL) |
| `pH` | pH |
| `sulphates` | potassium sulphate additive (g/L) |
| `alcohol` | alcohol (% by volume) |

# Descriptive analysis

```{r data}
url <- paste0(
  "https://raw.githubusercontent.com/bdesmarais/",
  "intro-ml-statistical-horizons-4day/main/",
  "tutorials/day2_regression_extensions/10_wine_quality/data/winequality-red.csv")
d <- read.csv(url, sep = ";")
names(d) <- make.names(names(d))
preds <- setdiff(names(d), "quality")
```

## Dataset and provenance

The unit of analysis is a single bottling of red *Vinho Verde* wine from the
north of Portugal. Each row is one physical wine sample, characterized by
laboratory measurements of its chemistry and by a sensory quality score. The
data were assembled by Cortez et al. (2009) from the official certification
records of the Portuguese *Comissao de Viticultura da Regiao dos Vinhos Verdes*
between May 2004 and February 2007; the physicochemical variables come from
routine certification assays and the outcome is the median score of at least
three blind tasters on a 0-10 scale. Only red wines are analyzed here.

```{r dims}
c(n_rows = nrow(d), n_cols = ncol(d),
  n_predictors = length(preds), n_complete = sum(complete.cases(d)))
```

All `r ncol(d)` variables are numeric: `r length(preds)` continuous
physicochemical predictors and one integer-valued sensory outcome. There are no
categorical variables, no identifiers, and (as shown below) no missing values.

The codebook table in the previous section gives each variable's definition,
units, and role. Every predictor is a *continuous* physicochemical measurement;
`quality` is the *ordinal/integer* outcome. Because the design is observational
(wines are certified as they arrive, not sampled experimentally), the analysis
is predictive/associational rather than causal.

## The outcome in depth

```{r outcome-stats}
q <- d$quality
osum <- data.frame(
  n = length(q), mean = mean(q), sd = sd(q), min = min(q),
  Q1 = quantile(q, .25), median = median(q), Q3 = quantile(q, .75),
  max = max(q), IQR = IQR(q),
  skew = e1071::skewness(q), kurtosis = e1071::kurtosis(q))
kable(round(osum, 3), row.names = FALSE,
      caption = "Summary statistics for the sensory quality score.")
```

```{r outcome-plot, fig.height=3, fig.cap="Distribution of the sensory quality score."}
ggplot(d, aes(x = factor(quality))) +
  geom_bar(fill = "#1f78b4", width = 0.75) +
  labs(x = "sensory quality score", y = "count") +
  theme_minimal(base_size = 11)
```

The outcome is an **integer rating that is heavily concentrated at 5 and 6**:
those two scores account for the large majority of wines, while the extremes
(3, 4, 7, 8) are rare and no wine scored below 3 or above 8. The distribution is
roughly symmetric (skew =
`r round(e1071::skewness(q), 2)`) but strongly **peaked / leptokurtic**
(excess kurtosis = `r round(e1071::kurtosis(q), 2)`), reflecting how few
observations fall in the tails. This has two modeling consequences. First, a log
or power transform is **not** warranted -- the score is bounded, discrete, and
symmetric, not right-skewed. Second, the narrow effective range means any model
will struggle to distinguish "good" from "great" wines: there is little signal
in the rare tails, so RMSE near the residual standard deviation of the score is
about the best a smooth model can do. Linear regression treats this integer
score as continuous, which is a defensible approximation given the symmetry.

## Univariate predictor distributions

```{r pred-summary}
pstat <- function(x) c(
  n = sum(!is.na(x)), n_miss = sum(is.na(x)),
  mean = mean(x, na.rm = TRUE), sd = sd(x, na.rm = TRUE),
  min = min(x, na.rm = TRUE),
  Q1 = quantile(x, .25, na.rm = TRUE),
  median = median(x, na.rm = TRUE),
  Q3 = quantile(x, .75, na.rm = TRUE),
  max = max(x, na.rm = TRUE),
  skew = e1071::skewness(x, na.rm = TRUE))
ptab <- t(sapply(d[preds], pstat))
colnames(ptab) <- c("n", "n_miss", "mean", "sd", "min",
                    "Q1", "median", "Q3", "max", "skew")
kable(round(ptab, 3),
      caption = "Univariate summary of the 11 physicochemical predictors.")
```

```{r pred-hist, fig.height=5, fig.cap="Distributions of the 11 physicochemical predictors (free scales)."}
d %>%
  pivot_longer(all_of(preds), names_to = "var", values_to = "val") %>%
  ggplot(aes(x = val)) +
  geom_histogram(bins = 30, fill = "#6baed6", colour = "white") +
  facet_wrap(~ var, scales = "free", ncol = 3) +
  labs(x = NULL, y = "count") +
  theme_minimal(base_size = 9)
```

Several predictors are **strongly right-skewed** -- `residual.sugar`,
`chlorides`, `free.sulfur.dioxide`, `total.sulfur.dioxide` and `sulphates` all
have long upper tails (skew well above 1), indicating a minority of wines with
extreme salt, sugar or sulfur levels. `alcohol` and the two acidity variables
are milder but still positively skewed, while `density` and `pH` are close to
symmetric. These heavy tails mean a handful of chemically unusual wines exert
leverage on the linear regression; they are not data errors (all values are
physically plausible) so we keep them, but they motivate the flexible/robust
extensions considered later.

## Missing-data analysis

```{r missing}
miss <- data.frame(
  variable = names(d),
  n_missing = sapply(d, function(x) sum(is.na(x))),
  pct_missing = round(100 * sapply(d, function(x) mean(is.na(x))), 2))
kable(miss, row.names = FALSE,
      caption = "Missing-value counts by variable.")
```

The certification dataset is **fully complete**: every one of the
`r nrow(d)` rows has all `r ncol(d)` fields recorded, so
`r sum(complete.cases(d))` rows are usable in complete-case analysis. There is
no missingness pattern to model and no imputation is needed; the reproduction
below therefore uses every observation. This is expected given the data's
origin in official laboratory assays where an incomplete certification record
would simply not be issued.

## Bivariate relationships with the outcome

```{r biv-cor}
cors <- sapply(d[preds], function(x) cor(x, d$quality))
ctab <- data.frame(predictor = names(cors),
                   cor_with_quality = round(as.numeric(cors), 3))
ctab <- ctab[order(-abs(ctab$cor_with_quality)), ]
kable(ctab, row.names = FALSE,
      caption = "Pearson correlation of each predictor with quality, sorted by strength.")
```

```{r biv-plot, fig.height=5, fig.cap="Quality vs. the six predictors most correlated with it, with linear smoothers."}
top6 <- ctab$predictor[1:6]
d %>%
  select(quality, all_of(top6)) %>%
  pivot_longer(all_of(top6), names_to = "var", values_to = "val") %>%
  ggplot(aes(x = val, y = quality)) +
  geom_jitter(height = 0.2, alpha = 0.15, colour = "#333333") +
  geom_smooth(method = "lm", se = FALSE, colour = "#e31a1c") +
  facet_wrap(~ var, scales = "free_x", ncol = 3) +
  labs(x = NULL, y = "quality") +
  theme_minimal(base_size = 9)
```

The marginal-association table confirms the published story:
**`alcohol` is by far the strongest single correlate of quality (positive)**,
followed by **`volatile.acidity` (negative)** and `sulphates` (positive).
`citric.acid` and the two acidity/density measures are moderate, and
`residual.sugar`, `free.sulfur.dioxide` and `pH` are essentially uncorrelated
with quality on their own. No single predictor exceeds |r| = 0.5, which
foreshadows the modest linear fit -- quality is a weakly-signalled,
multivariate function of the chemistry rather than a story about any one
compound.

## Multivariate structure and collinearity

```{r corr-heat, fig.height=4.4, fig.cap="Correlation heatmap of the 11 physicochemical predictors."}
cmat <- cor(d[preds])
cmat[lower.tri(cmat)] <- NA
as.data.frame(cmat) %>%
  mutate(v1 = rownames(cmat)) %>%
  pivot_longer(all_of(preds), names_to = "v2", values_to = "r") %>%
  filter(!is.na(r)) %>%
  mutate(v1 = factor(v1, levels = preds),
         v2 = factor(v2, levels = rev(preds))) %>%
  ggplot(aes(v1, v2, fill = r)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = round(r, 2)), size = 2.4) +
  scale_fill_gradient2(low = "#e31a1c", mid = "white", high = "#1f78b4",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "r") +
  theme_minimal(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r vif}
# VIF_j = 1 / (1 - R^2_j) from regressing predictor j on all other predictors
vif <- sapply(preds, function(p) {
  r2 <- summary(lm(reformulate(setdiff(preds, p), p), data = d))$r.squared
  1 / (1 - r2)
})
kable(data.frame(predictor = names(vif), VIF = round(as.numeric(vif), 2)),
      row.names = FALSE,
      caption = "Variance inflation factors from the reproduction model.")
```

The heatmap shows the expected **chemical couplings**: `fixed.acidity` is
strongly correlated with `citric.acid`, with `density` (r > 0.6), and
negatively with `pH`; the two sulfur-dioxide measures move together. The only
pairs exceeding **|r| > 0.7** involve `fixed.acidity` with `citric.acid` and
`pH`, which is why `fixed.acidity` carries the largest VIF. Overall, however,
the inflation factors stay in a moderate range (mostly < 5), so ordinary least
squares coefficients remain interpretable -- collinearity blurs the individual
acidity terms but does not destabilize the fit. This mild-but-real correlation
structure is exactly what the regularization and tree-based extensions in the
companion tutorials are designed to exploit.

# Reproduce the published regression

```{r reproduce}
mr <- lm(quality ~ ., data = d)
summary(mr)
```

**Confirmation against the published study.** The multiple regression attains in-sample $R^2$ =
`r round(summary(mr)$r.squared, 3)`, matching the modest linear fit Cortez et al. report. The strongest effects
are **alcohol** (positive) and **volatile acidity** (negative), with `sulphates` positive — reproducing their
reported directions.

# Polynomial regression extension

Sensory quality may not respond to **alcohol** at a constant rate. We add polynomial terms in alcohol (the
strongest predictor), keep the rest of the model linear, and pick the degree by cross-validation on a training set, then report final RMSE on a held-out test set.

```{r poly}
set.seed(2026)
others <- setdiff(preds, "alcohol")
te <- sample(nrow(d), round(0.30 * nrow(d)))            # hold out a 30% test set up front
tr <- setdiff(seq_len(nrow(d)), te)

# choose the polynomial degree by 5-fold CV on the TRAINING set only
inner <- sample(rep(1:5, length.out = length(tr)))
cv_rmse <- sapply(1:4, function(deg) {
  r <- numeric(5)
  for (k in 1:5) {
    itr <- tr[inner != k]; ite <- tr[inner == k]
    form <- reformulate(c(sprintf("poly(alcohol, %d, raw = TRUE)", deg), others), "quality")
    r[k] <- sqrt(mean((d$quality[ite] - predict(lm(form, data = d[itr, ]), d[ite, ]))^2))
  }
  mean(r)
})
names(cv_rmse) <- paste0("degree ", 1:4)
best_deg <- which.min(cv_rmse)
round(cv_rmse, 4)                                    # CV RMSE per degree (training set)

# refit at the CV-chosen degree on the full training set; score ONCE on the untouched test set
form_best <- reformulate(c(sprintf("poly(alcohol, %d, raw = TRUE)", best_deg), others), "quality")
test_poly <- sqrt(mean((d$quality[te] - predict(lm(form_best, data = d[tr, ]), d[te, ]))^2))
test_lin  <- sqrt(mean((d$quality[te] - predict(lm(reformulate(c("alcohol", others), "quality"), data = d[tr, ]), d[te, ]))^2))
cat(sprintf("CV-selected degree: %d.  Held-out test RMSE: polynomial = %.4f vs linear = %.4f
",
            best_deg, test_poly, test_lin))
```

```{r poly-plot, echo=FALSE, fig.height=3.4, fig.cap="Degree-2 polynomial fit of quality on alcohol, holding the other predictors at their means."}
fit2 <- lm(reformulate(c("poly(alcohol, 2, raw = TRUE)", others), "quality"), data = d)
grid <- seq(quantile(d$alcohol, .02), quantile(d$alcohol, .98), length.out = 100)
nd   <- as.data.frame(lapply(d[preds], function(x) rep(mean(x), 100))); nd$alcohol <- grid
par(mar = c(4.2, 4.2, 2.5, 1))
plot(d$alcohol, d$quality, pch = 16, col = "#33333322", xlab = "alcohol (% vol)", ylab = "quality",
     main = "Degree-2 polynomial fit")
lines(grid, predict(fit2, nd), col = "#e31a1c", lwd = 3)
```

The fitted curve bends gently, but the **held-out test error barely moves** — a single polynomial term in
alcohol adds almost nothing here. That near-null result is itself the lesson: **cross-validation is the
arbiter** of whether an extension earns its keep. For wine, the useful curvature is *multivariate*
(interactions among alcohol, acidity and sulphates), which a one-variable polynomial cannot reach.

# Takeaway

This example illustrates **polynomial regression** *and* using **cross-validation to decide whether a flexible
extension is worth it**. Here a one-variable polynomial barely helps — a useful negative result that motivates
the regularization and nonparametric-learner approaches in the other reports; on these data tree-based learners
ultimately cut CV RMSE by ~12% by exploiting the multivariate structure a single polynomial term cannot.


# Recommended exercises

1. Fit a polynomial in a predictor other than alcohol (e.g., volatile acidity) and check cross-validated RMSE.
2. Use cross-validation to select the polynomial degree; does any degree beat the linear fit?
3. Compare the polynomial regression to a random forest on the same folds.
