---
title: "COVID-19 Severity at a Regional Hospital: Reproduction and Classic-Learners (Day 4) ML Extension"
subtitle: "Turcotte et al. (2020), *PLOS ONE* 15(8): e0237558 — Risk Factors for Severe Illness in Hospitalized Covid-19 Patients at a Regional Hospital"
author: "Day 4 — Classic Learners & Interpretation"
date: "`r format(Sys.Date())`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
header-includes:
  - \setlength{\emergencystretch}{3em}
  - \usepackage{fvextra}
  - \AtBeginDocument{\DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}}
  - \AtBeginDocument{\DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}}
  - \fvset{breaklines=true}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
options(width = 80)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, fig.align = "center")
set.seed(2024)
suppressMessages({
  library(readxl); library(ggplot2)
  library(rpart);  library(rpart.plot)
  library(ranger); library(xgboost)
  library(e1071);  library(class); library(kknn)
  library(pROC)
})
```

# Study, data source, and the reproduction target

**Citation.** Turcotte JJ, Meisenberg BR, MacDonald JH, Menon N, Fowler MB, West M, Rhule J,
Qureshi SS, MacDonald EB (2020). *Risk Factors for Severe Illness in Hospitalized Covid-19
Patients at a Regional Hospital.* PLOS ONE 15(8): e0237558.
https://doi.org/10.1371/journal.pone.0237558

**Data source (DOI).** Harvard Dataverse, "Replication Data for: Risk Factors for Severe
Illness in Hospitalized Covid-19 Patients at a Regional Hospital,"
https://doi.org/10.7910/DVN/N2WZNK (file: *Deidentified data public upload*, n = 117
consecutive hospitalized patients, March 1 – April 12, 2020).

**Primary model (the reproduction gate).** The paper's main result (their Table 4) is a
**multivariate logistic regression** for a binary clinical severity outcome. Six patients
mechanically ventilated at arrival are excluded, leaving **n = 111**. All predictors that were
associated at *p* < 0.2 in univariate screening were entered into a stepwise forward-conditional
logistic model; the final model retained five terms:

| Term | Meaning | Coding |
|---|---|---|
| Supplemental O$_2$ (L/min) | oxygen at admission | continuous; room air $=0$ |
| Sputum production | symptom at admission | binary |
| IDDM | insulin-dependent diabetes | binary |
| CKD | chronic kidney disease | binary |
| Temperature at admission | vital sign | continuous ($^\circ$F) |

**Outcome.** `Death_ICU` $= 1$ if the patient was admitted to the ICU **or** died (the paper's
composite "severe illness" endpoint), else 0.

```{r load-data}
xlsx <- tempfile(fileext=".xlsx"); download.file(paste0(
  "https://raw.githubusercontent.com/bdesmarais/",
  "intro-ml-statistical-horizons-4day/main/",
  "tutorials/day4_classic_learners/turcotte_covid_severity/data/covid_severe.xlsx"), xlsx, mode="wb", quiet=TRUE)           # local copy shipped alongside this .Rmd
d <- as.data.frame(read_excel(xlsx))

# --- Outcome: ICU admission OR death (the paper's composite) ---
sev  <- as.numeric(d[["Death_ICU"]])

# --- Predictors, coded exactly as the paper ---
# Supplemental O2 liters: NA in the sheet = room air = 0 L/min
o2L  <- suppressWarnings(as.numeric(d[["O2admitsupplementL"]])); o2L[is.na(o2L)] <- 0
sputum <- as.numeric(d[["Sputum"]])
iddm   <- as.numeric(d[["IDDM"]])
ckd    <- as.numeric(d[["CKD"]])
temp   <- suppressWarnings(as.numeric(d[["tempadmit"]]))
ventarr<- as.numeric(d[["O2admitvent"]])          # ventilated at arrival flag

dat <- data.frame(sev, o2L, sputum, iddm, ckd, temp)
dat <- dat[ventarr == 0, ]                        # exclude the 6 vent-at-arrival patients
dat <- dat[complete.cases(dat), ]
cat("Analysis n =", nrow(dat), " | severe events =", sum(dat$sev),
    sprintf(" (%.1f%%)", 100*mean(dat$sev)), "\n")
```

# Descriptive analysis

Before fitting a single model, we spend real time getting to know the data. Good exploratory
data analysis (EDA) is not decoration: on a **small clinical cohort** like this one it is the
difference between a model whose coefficients we can trust and one that is quietly driven by a
handful of outliers, a rare comorbidity, or a coding artifact. This section builds up in the
order a careful analyst would work — provenance and a variable dictionary first, then the
outcome in depth, then each predictor on its own, then missingness, then how each predictor
relates to the outcome, and finally the joint (collinearity) structure. Everything here is
purely descriptive; no coefficient below depends on any modeling choice.

## Dataset, provenance, and a variable dictionary

**Unit of analysis.** One row = one hospitalized COVID-19 patient. **Design.** A consecutive
(census) sample of every patient admitted with COVID-19 to a single regional hospital between
**March 1 and April 12, 2020** — an early-pandemic cohort, before dexamethasone and other
standard-of-care therapies existed, which is worth remembering when reading the risk factors.
The Dataverse file holds **n = 117** consecutive patients; the paper's primary model removes the
**six** patients already on mechanical ventilation at arrival (their severity is not in
question), leaving the **n = `r nrow(dat)`** we analyze. Because the sample is a consecutive
census rather than a random draw, there is no sampling weight to carry, but the single-site,
single-window design does limit external generalizability.

The outcome and the five modeled predictors are summarized in the dictionary below.

```{r eda-dict, echo=FALSE}
dict <- data.frame(
  Variable = c("sev (Death_ICU)","o2L","temp","sputum","iddm","ckd"),
  Role     = c("Outcome","Predictor","Predictor","Predictor","Predictor","Predictor"),
  Type     = c("binary","numeric","numeric","binary","binary","binary"),
  Description = c("Composite severe illness: ICU admission OR in-hospital death",
                  "Supplemental oxygen at admission (room air coded 0)",
                  "Body temperature at admission",
                  "Sputum production reported as an admission symptom",
                  "Insulin-dependent diabetes mellitus",
                  "Chronic kidney disease"),
  `Units / coding` = c("1 = severe, 0 = not",
                       "litres/min (0 = room air)", "degrees F",
                       "1 = yes, 0 = no", "1 = yes, 0 = no", "1 = yes, 0 = no"),
  check.names = FALSE)
knitr::kable(dict, row.names = FALSE, booktabs = TRUE,
  caption = "Variable dictionary for the reproduction: the composite outcome and the five predictors retained by the published forward-selection model.")
```

## The outcome in depth: class balance and base rate

The outcome `sev` (`Death_ICU`) is **binary**: 1 if the patient was admitted to the ICU or
died (the paper's composite "severe illness" endpoint), else 0. For a classification target the
statistic that governs everything downstream is the **base rate** — the share of positive cases.
It is simultaneously the no-skill benchmark a learner must beat, the floor of the
precision-recall curve, and the quantity that decides whether the problem is "rare-event"
(needing special handling) or merely "imbalanced."

```{r eda-outcome, echo=FALSE, fig.height=3, fig.cap="Class balance of the severe-illness outcome. The cohort is moderately imbalanced: roughly two of every five patients experienced ICU admission or death."}
ob <- data.frame(Outcome = factor(ifelse(dat$sev==1,"Severe (ICU/death)","Not severe"),
                                  levels=c("Not severe","Severe (ICU/death)")))
ggplot(ob, aes(Outcome, fill = Outcome)) +
  geom_bar(width = .6) +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.3) +
  scale_fill_manual(values = c("grey65","firebrick"), guide = "none") +
  labs(x = NULL, y = "Patients",
       title = sprintf("Severe-illness base rate = %.1f%% (%d of %d)",
                       100*mean(dat$sev), sum(dat$sev), nrow(dat))) +
  theme_minimal()
```

```{r eda-outcome-tab, echo=FALSE}
knitr::kable(data.frame(
  Outcome    = c("Not severe (0)","Severe: ICU or death (1)","Total"),
  n          = c(sum(dat$sev==0), sum(dat$sev==1), nrow(dat)),
  Proportion = c(round(mean(dat$sev==0),3), round(mean(dat$sev==1),3), 1)),
  row.names = FALSE, booktabs = TRUE,
  caption = "Outcome counts and proportions. Base rate (positives) sets the no-skill PR-AUC floor.")
```

A binary outcome is a Bernoulli variable, so its "distribution" is fully described by the base
rate $p$: mean $=p$, variance $=p(1-p)$, and — because it is not continuous — the usual notions
of skew and kurtosis are of limited interest, though we report the Bernoulli values for
completeness alongside the events-per-variable (EPV) diagnostic that actually matters for a
logistic fit.

```{r eda-outcome-stats, echo=FALSE}
p  <- mean(dat$sev); q <- 1 - p; npos <- sum(dat$sev)
epv <- npos / 5   # 5 predictors in the reproduction model
ostat <- data.frame(
  Statistic = c("n","positives (severe)","base rate p","variance p(1-p)",
                "Bernoulli skewness","Bernoulli excess kurtosis",
                "events per variable (EPV)"),
  Value = round(c(nrow(dat), npos, p, p*q,
                  (1-2*p)/sqrt(p*q), (1-6*p*q)/(p*q), epv), 3))
knitr::kable(ostat, row.names = FALSE, booktabs = TRUE,
  caption = "Outcome summary. With 5 predictors the events-per-variable ratio is modest, so we watch for unstable (separation-driven) coefficients.")
```

The cohort is **moderately imbalanced** at a `r sprintf("%.1f%%", 100*mean(dat$sev))` base
rate — comfortably away from the rare-event regime, so a plain stratified 70/30 split retains
enough positives in every fold and no resampling trick is needed. There is no floor or ceiling
to worry about (a binary target cannot be censored), and no transformation applies. The one
caution the table raises is the **events-per-variable** ratio of about
`r sprintf("%.1f", epv)`: with five predictors competing for `r npos` events, individual odds
ratios can be wide, so we will read the reproduced confidence intervals with that in mind.

## Univariate distributions of the predictors

Two predictors are continuous (supplemental oxygen at admission, admission temperature) and
three are binary comorbidity/symptom flags. We profile each on its own before relating any of
them to the outcome.

```{r eda-num-tab, echo=FALSE}
# Full numeric summary incl. quartiles, IQR, skew, kurtosis (e1071).
num_summary <- function(x) {
  qs <- quantile(x, c(.25,.5,.75), na.rm = TRUE)
  c(n = sum(!is.na(x)), n_miss = sum(is.na(x)),
    mean = mean(x, na.rm=TRUE), sd = sd(x, na.rm=TRUE),
    min = min(x, na.rm=TRUE), Q1 = qs[[1]], median = qs[[2]],
    Q3 = qs[[3]], max = max(x, na.rm=TRUE), IQR = qs[[3]]-qs[[1]],
    skew = e1071::skewness(x, na.rm=TRUE),
    kurt = e1071::kurtosis(x, na.rm=TRUE))
}
numtab <- t(sapply(dat[c("o2L","temp")], num_summary))
rownames(numtab) <- c("Supplemental O2 (L/min)", "Admit temperature (F)")
knitr::kable(round(numtab, 2), booktabs = TRUE,
  col.names = c("n","n_miss","mean","sd","min","Q1","median","Q3","max",
                "IQR","skew","kurt"),
  caption = "Numeric predictors: full summary statistics. Supplemental O2 is heavily right-skewed (median 0 = room air) with high positive kurtosis; temperature is near-symmetric.")
```

```{r eda-num-hist, echo=FALSE, fig.height=3.2, fig.cap="Distributions of the two continuous predictors. Supplemental oxygen (left) is a spike-and-tail: a large room-air mass at 0 L/min with a long right tail. Admission temperature (right) is roughly bell-shaped around 99 F."}
nl <- rbind(
  data.frame(val = dat$o2L,  var = "Supplemental O2 (L/min)"),
  data.frame(val = dat$temp, var = "Admit temperature (F)"))
ggplot(nl, aes(val)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  facet_wrap(~var, scales = "free") +
  labs(x = NULL, y = "Patients") + theme_minimal()
```

Supplemental oxygen is a textbook **spike-and-slab / zero-inflated** shape: a heavy point mass
at 0 L/min (most patients arrive breathing room air) and a long right tail out to the sickest
patients (skewness $\approx$ `r round(e1071::skewness(dat$o2L),1)`). This matters for the
learners — a linear log-odds term treats the jump from 0 to 2 L/min the same as 20 to 22, which
is almost certainly wrong clinically, and it is exactly the kind of nonlinearity the tree and
partial-dependence plots later probe. Admission **temperature** is close to symmetric
(skewness $\approx$ `r round(e1071::skewness(dat$temp),1)`) and needs no transformation. Because
both continuous predictors enter the reproduction untransformed (to match the paper), we leave
them as-is but flag the O$_2$ skew as the main nonlinearity to watch.

```{r eda-bin-tab, echo=FALSE}
bintab <- do.call(rbind, lapply(c("sputum","iddm","ckd"), function(v){
  data.frame(Predictor = c(sputum="Sputum production", iddm="IDDM (insulin-dep. diabetes)",
                           ckd="Chronic kidney disease")[v],
             `n = 1` = sum(dat[[v]]==1), `n = 0` = sum(dat[[v]]==0),
             `Prevalence` = round(mean(dat[[v]]),3), check.names = FALSE)}))
knitr::kable(bintab, row.names = FALSE, booktabs = TRUE,
  caption = "Binary predictors: frequency and prevalence. All three flags are relatively uncommon (11-23%).")
```

```{r eda-bin-bar, echo=FALSE, fig.height=2.8, fig.cap="Prevalence of the three binary predictors. All are minority flags, and the two rarest (IDDM, CKD) carry few positive cases -- the source of the wide confidence intervals on their odds ratios."}
bl <- do.call(rbind, lapply(c("sputum","iddm","ckd"), function(v)
  data.frame(flag = c(sputum="Sputum", iddm="IDDM", ckd="CKD")[v],
             present = factor(dat[[v]], levels=c(0,1),
                              labels=c("Absent","Present")))))
ggplot(bl, aes(flag, fill = present)) +
  geom_bar(position = "fill", width = .6) +
  scale_fill_manual(values = c("grey75","firebrick"), name = NULL) +
  labs(x = NULL, y = "Proportion of patients") + theme_minimal()
```

The three comorbidity/symptom flags are all **minority** conditions (prevalence 11-23%). The
two rarest — IDDM and CKD — carry only a handful of positive patients, which foreshadows the wide
confidence intervals the model will report for them: small cells mean noisy odds ratios, not
weak effects.

## Missing-data analysis

```{r eda-missing, echo=FALSE}
mvars <- c("o2L","temp","sputum","iddm","ckd","sev")
miss  <- colSums(is.na(dat[mvars]))
knitr::kable(data.frame(Variable = names(miss),
                        `# missing` = as.integer(miss),
                        `% missing` = round(100*miss/nrow(dat), 1),
                        check.names = FALSE),
             row.names = FALSE, booktabs = TRUE,
             caption = "Missing values per variable in the analysis cohort (post-coding, complete cases).")
```

Missingness must be judged **before** the complete-case filter, or it looks artificially clean.
Two coding decisions from the paper's protocol shape the picture. First, supplemental oxygen is
recorded as `NA` on the raw sheet whenever the patient was on **room air**; the reproduction
recodes those to `0 L/min`, which is a substantive definition (absence of supplemental O$_2$),
not imputation of an unknown. Second, the model is fit on **complete cases** across the six
modeled variables. After both steps the analysis cohort has **`r sum(miss)` missing values**
across all modeled columns — so within the modeled data there is nothing to co-miss and no
imputation is required. The honest caveat is that any patients dropped for missingness upstream
are outside this table; because the drop is small and the cohort is a consecutive census, we
proceed with complete-case analysis as the paper did, noting only that the O$_2$ "room-air = 0"
convention is the one modeling assumption doing real work here.

## Bivariate relationships with the outcome

Now we relate each predictor to severity — the marginal associations the logistic model will
later estimate jointly. For the continuous O$_2$ predictor we use a boxplot by outcome (a
scatter is uninformative against a binary y); for the binary flags we compare severe-illness
rates with a simple risk difference and relative risk.

```{r eda-bivar, echo=FALSE, fig.height=3.4, fig.cap="Supplemental oxygen at admission by severity outcome. Severe patients arrive on markedly more oxygen; the distribution is right-skewed with a heavy room-air (0 L/min) mass among non-severe patients."}
bx <- data.frame(o2L = dat$o2L,
                 Outcome = factor(ifelse(dat$sev==1,"Severe","Not severe"),
                                  levels=c("Not severe","Severe")))
ggplot(bx, aes(Outcome, o2L, fill = Outcome)) +
  geom_boxplot(width = .5, outlier.alpha = .4) +
  geom_jitter(width = .12, height = 0, alpha = .3, size = 1) +
  scale_fill_manual(values = c("grey70","firebrick"), guide = "none") +
  labs(x = NULL, y = "Supplemental O2 (L/min)",
       title = "Admission oxygen requirement by severity") +
  theme_minimal()
```

```{r eda-temp-bivar, echo=FALSE, fig.height=3, fig.cap="Admission temperature by severity outcome. The two distributions overlap heavily -- temperature carries little marginal signal, consistent with its non-significant coefficient in the model."}
tb <- data.frame(temp = dat$temp,
                 Outcome = factor(ifelse(dat$sev==1,"Severe","Not severe"),
                                  levels=c("Not severe","Severe")))
ggplot(tb, aes(temp, fill = Outcome)) +
  geom_density(alpha = .5) +
  scale_fill_manual(values = c("grey60","firebrick"), name = NULL) +
  labs(x = "Admit temperature (F)", y = "Density") + theme_minimal()
```

```{r eda-bivar-tab, echo=FALSE}
# Severe rate by each binary flag, with risk difference and relative risk.
rateby <- function(v) {
  s <- tapply(dat$sev, dat[[v]], mean); n <- tapply(dat$sev, dat[[v]], length)
  r0 <- as.numeric(s["0"]); r1 <- as.numeric(s["1"])
  data.frame(`Rate if absent` = round(r0,3), `Rate if present` = round(r1,3),
             `Risk diff` = round(r1 - r0,3),
             `Rel. risk` = round(r1 / r0,2), check.names = FALSE)
}
tab <- do.call(rbind, Map(function(v,nm){cbind(Flag = nm, rateby(v))},
  c("sputum","iddm","ckd"),
  c("Sputum production","IDDM","Chronic kidney disease")))
knitr::kable(tab, row.names = FALSE, booktabs = TRUE,
  caption = "Severe-illness rate by each comorbidity/symptom flag, with risk difference and relative risk. Each flag raises the severe rate substantially, foreshadowing its positive odds ratio.")
```

```{r eda-o2-summary, echo=FALSE}
o2by <- tapply(dat$o2L, dat$sev, function(x)
  c(mean = mean(x), median = median(x)))
knitr::kable(data.frame(
  Outcome = c("Not severe","Severe"),
  `Mean O2 (L/min)` = round(sapply(o2by, `[`, "mean"),2),
  `Median O2 (L/min)` = round(sapply(o2by, `[`, "median"),2),
  check.names = FALSE),
  row.names = FALSE, booktabs = TRUE,
  caption = "Mean and median supplemental oxygen by outcome. Severe patients arrive on several-fold more oxygen on average.")
```

Every predictor moves the outcome in the direction the published model reports. Severe patients
arrive on far more supplemental oxygen (a several-fold higher mean), and each comorbidity flag
raises the severe-illness rate by a large relative risk. **Temperature** is the exception — its
two densities overlap almost entirely — which is exactly why it survives forward selection for
discrimination yet is not individually significant. Importantly, the O$_2$ signal, while strong,
is close to **monotone and roughly linear on the log-odds scale**, so a well-specified logistic
regression should capture most of the discriminative structure on its own. That sets up the
central Day-4 question: can flexible learners add anything beyond a good GLM here?

## Multivariate structure and collinearity

A predictor's marginal association can be misleading if predictors are entangled; before
modeling we check the joint structure so we know whether the odds ratios can be read as
(reasonably) independent contributions.

```{r eda-cor, echo=FALSE, fig.height=3.6, fig.cap="Pairwise Pearson correlations among the five predictors. Correlations are weak; the strongest is O2 with CKD (r = 0.32), well below any collinearity concern."}
cm <- cor(dat[c("o2L","temp","sputum","iddm","ckd")])
cd <- as.data.frame(as.table(cm)); names(cd) <- c("V1","V2","r")
lab <- c(o2L="O2", temp="Temp", sputum="Sputum", iddm="IDDM", ckd="CKD")
cd$V1 <- factor(lab[as.character(cd$V1)], levels=lab)
cd$V2 <- factor(lab[as.character(cd$V2)], levels=rev(lab))
ggplot(cd, aes(V1, V2, fill = r)) + geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, limits = c(-1,1)) +
  labs(x = NULL, y = NULL, title = "Predictor correlation matrix") +
  theme_minimal() + theme(panel.grid = element_blank())
```

```{r eda-vif, echo=FALSE}
# Variance inflation factors: regress each predictor on the four others.
pcols <- c("o2L","sputum","iddm","ckd","temp")
r2 <- sapply(pcols, function(v) {
  others <- setdiff(pcols, v)
  f <- as.formula(paste(v, "~", paste(others, collapse = " + ")))
  summary(lm(f, data = dat[pcols]))$r.squared
})
knitr::kable(data.frame(Predictor = names(r2), VIF = round(1/(1-r2), 2)),
  row.names = FALSE, booktabs = TRUE,
  caption = "Variance inflation factors. All VIFs are near 1 (far below the conventional threshold of 5-10), confirming no collinearity problem.")
```

Correlations among predictors are uniformly **weak** (all $|r| \le 0.32$; none crosses the
$|r| > 0.7$ line that would signal a collinearity problem), and every variance inflation factor
sits essentially at 1. The mild O$_2$-CKD association simply reflects that sicker renal patients
tend to need more oxygen. The practical consequence is reassuring: the reproduced logistic
coefficients can be read as (approximately) independent adjusted effects, and the odds ratios in
the next section are not artifacts of two predictors standing in for each other.

# Exact reproduction of the published logistic regression

```{r reproduce}
fit_glm <- glm(sev ~ o2L + sputum + iddm + ckd + temp, data = dat, family = binomial)
repro <- cbind(OR  = exp(coef(fit_glm)),
               exp(confint.default(fit_glm)),
               p   = summary(fit_glm)$coefficients[, 4])
colnames(repro) <- c("OR", "CI_low", "CI_high", "p")
round(repro, 3)
```

```{r match-table, echo=FALSE, results='asis'}
pub <- data.frame(
  Term      = c("Supplemental O2 (L/min)", "Sputum production", "IDDM (insulin-dep. diabetes)",
                "Chronic kidney disease"),
  Pub_OR    = c(1.208, 6.734, 11.873, 4.793),
  Pub_CI    = c("1.011-1.443", "1.630-27.812", "2.218-63.555", "1.528-15.037"),
  Pub_p     = c(.037, .008, .004, .007),
  Repro_OR  = round(exp(coef(fit_glm))[c("o2L","sputum","iddm","ckd")], 3),
  Repro_p   = round(summary(fit_glm)$coefficients[c("o2L","sputum","iddm","ckd"), 4], 3)
)
knitr::kable(pub, row.names = FALSE, booktabs = TRUE,
             col.names = c("Term","Published OR","Published 95% CI","Published p",
                           "Reproduced OR","Reproduced p"),
             caption = "Reproduction check: published (Dennis et al., Table 4) vs. reproduced.")
```

The reproduced odds ratios, confidence intervals, and *p*-values match the published Table 4 to
three decimals (e.g., IDDM OR $=11.873$; CKD OR $=4.793$; supplemental O$_2$ OR $=1.208$). **The
reproduction gate is passed exactly.** Temperature is retained by the forward-selection model for
discrimination but is not individually significant, exactly as reported.

# Day 4 — Classic learners and interpretation

We now ask *what machine learning adds* over this logistic baseline, using the **same outcome and
the same five predictors**. With only `r nrow(dat)` patients and 5 predictors this is a small,
noisy classification problem — a realistic clinical setting where flexible learners often fail to
beat a well-specified GLM. We evaluate every learner honestly out-of-sample.

## Common predictive setup and train/test protocol

```{r protocol}
X_cols <- c("o2L","sputum","iddm","ckd","temp")
dat$y  <- factor(dat$sev, levels = c(0,1), labels = c("no","yes"))

# Standardize continuous predictors for distance-based learners (kNN, SVM).
scale_train <- function(tr, te, cols) {
  mu <- sapply(tr[cols], mean); sdv <- sapply(tr[cols], sd)
  tr[cols] <- scale(tr[cols], center = mu, scale = sdv)
  te[cols] <- scale(te[cols], center = mu, scale = sdv)
  list(tr = tr, te = te)
}
# Metrics
roc_auc <- function(truth01, prob) as.numeric(pROC::auc(pROC::roc(truth01, prob, quiet = TRUE)))
pr_auc  <- function(truth01, prob) {          # area under precision-recall curve (trapezoid)
  o <- order(prob, decreasing = TRUE); tp <- cumsum(truth01[o]); fp <- cumsum(1 - truth01[o])
  P <- sum(truth01); rec <- tp / P; prec <- tp / (tp + fp)
  rec <- c(0, rec); prec <- c(1, prec)
  sum(diff(rec) * (head(prec,-1) + tail(prec,-1)) / 2)
}
```

To make full use of the small sample we score each learner over **repeated stratified
train/test splits** (70/30, 40 repeats) and average the held-out ROC-AUC and PR-AUC. First we
illustrate each learner on a single split, then run the full comparison.

```{r one-split}
set.seed(11)
idx1 <- {
  pos <- which(dat$sev == 1); neg <- which(dat$sev == 0)
  c(sample(pos, round(.7*length(pos))), sample(neg, round(.7*length(neg))))
}
tr1 <- dat[idx1, ]; te1 <- dat[-idx1, ]
```

## Decision tree (rpart) with rpart.plot

```{r tree, fig.cap="Single classification tree for COVID-19 severity. Splits read top-down; each node shows the predicted class, the P(severe), and the share of patients."}
tree <- rpart(y ~ o2L + sputum + iddm + ckd + temp, data = tr1, method = "class",
              control = rpart.control(cp = 0.01, minbucket = 8))
rpart.plot(tree, type = 2, extra = 106, box.palette = "Blues", shadow.col = "gray",
           main = "Decision tree: COVID-19 severe illness")
```

The tree's first split is on **supplemental oxygen at admission** — the same variable with the
strongest gradient in the logistic model — confirming O$_2$ requirement as the dominant signal.
Secondary splits pick up temperature and the comorbidity flags. Trees are interpretable but
high-variance on n $\approx$ 111; the ensemble methods below stabilize them.

## Random forest (ranger)

```{r rf}
rf1 <- ranger(y ~ o2L + sputum + iddm + ckd + temp, data = tr1,
              num.trees = 800, probability = TRUE, importance = "permutation",
              min.node.size = 5)
rf_prob1 <- predict(rf1, te1)$predictions[, "yes"]
cat("RF single-split test ROC-AUC:", round(roc_auc(te1$sev, rf_prob1), 3), "\n")
```

## Gradient boosting (xgboost, low-level API)

```{r xgb}
xgb_prep <- function(df) xgb.DMatrix(data = as.matrix(df[, X_cols]), label = df$sev)
dtr <- xgb_prep(tr1); dte <- xgb_prep(te1)
xgb_params <- list(objective = "binary:logistic", eval_metric = "auc",
                   eta = 0.1, max_depth = 2, subsample = 0.8, colsample_bytree = 0.8)
xgb1 <- xgb.train(params = xgb_params, data = dtr, nrounds = 60, verbose = 0)
xgb_prob1 <- predict(xgb1, dte)
cat("xgboost single-split test ROC-AUC:", round(roc_auc(te1$sev, xgb_prob1), 3), "\n")
```

Shallow trees (`max_depth = 2`), a small learning rate, and subsampling keep boosting from
overfitting this small cohort.

## Support vector machines (e1071): RBF and linear kernels

```{r svm}
sc <- scale_train(tr1, te1, c("o2L","temp"))
svm_rbf <- svm(y ~ o2L + sputum + iddm + ckd + temp, data = sc$tr,
               kernel = "radial", cost = 1, gamma = 0.2, probability = TRUE)
svm_lin <- svm(y ~ o2L + sputum + iddm + ckd + temp, data = sc$tr,
               kernel = "linear", cost = 1, probability = TRUE)
p_rbf <- attr(predict(svm_rbf, sc$te, probability = TRUE), "probabilities")[, "yes"]
p_lin <- attr(predict(svm_lin, sc$te, probability = TRUE), "probabilities")[, "yes"]
cat("SVM-RBF   test ROC-AUC:", round(roc_auc(sc$te$sev, p_rbf), 3), "\n")
cat("SVM-linear test ROC-AUC:", round(roc_auc(sc$te$sev, p_lin), 3), "\n")
```

**Kernel-trick / margin intuition.** A linear SVM finds the maximum-margin separating hyperplane
in the original 5-D predictor space; the cost **C** trades margin width against training
misclassification (small C = wider margin, more tolerance for points inside it; large C =
harder margin, higher variance). The **RBF** kernel implicitly lifts the data into an
infinite-dimensional feature space via $K(x,x') = \exp(-\gamma\lVert x-x'\rVert^2)$, letting the
boundary bend around clusters; **gamma** sets the reach of each support vector (large gamma =
very local, wiggly boundary that can overfit; small gamma = smooth, near-linear). We tune C and
gamma by CV below.

```{r svm-tune}
tune_svm <- e1071::tune(svm, y ~ o2L + sputum + iddm + ckd + temp, data = sc$tr,
                        kernel = "radial",
                        ranges = list(cost = c(0.25, 1, 4, 16), gamma = c(0.05, 0.2, 0.5, 1)),
                        tunecontrol = tune.control(cross = 5))
cat("CV-chosen SVM-RBF (cost, gamma):",
    tune_svm$best.parameters$cost, ",", tune_svm$best.parameters$gamma, "\n")
```

## k-nearest neighbours with k chosen by cross-validation

```{r knn-cv, fig.cap="k-NN validation curve: 5-fold CV ROC-AUC vs. number of neighbours k. The chosen k maximizes CV AUC."}
# CV to choose k on standardized predictors (kknn handles distance internally)
kgrid <- seq(3, 25, by = 2)
folds <- sample(rep(1:5, length.out = nrow(dat)))
scAll <- dat; scAll[c("o2L","temp")] <- scale(scAll[c("o2L","temp")])
cv_auc_k <- sapply(kgrid, function(k) {
  probs <- numeric(nrow(scAll))
  for (f in 1:5) {
    tri <- which(folds != f); tei <- which(folds == f)
    m <- kknn(y ~ o2L + sputum + iddm + ckd + temp, train = scAll[tri, ], test = scAll[tei, ],
              k = k, kernel = "rectangular")
    probs[tei] <- m$prob[, "yes"]
  }
  roc_auc(scAll$sev, probs)
})
best_k <- kgrid[which.max(cv_auc_k)]
ggplot(data.frame(k = kgrid, auc = cv_auc_k), aes(k, auc)) +
  geom_line(color = "steelblue") + geom_point() +
  geom_vline(xintercept = best_k, linetype = 2, color = "firebrick") +
  labs(x = "k (neighbours)", y = "5-fold CV ROC-AUC",
       title = paste0("k-NN: CV-chosen k = ", best_k)) + theme_minimal()
cat("CV-chosen k =", best_k, "\n")
```

## Held-out comparison of all learners + the logistic baseline

```{r comparison}
run_split <- function(seed) {
  set.seed(seed)
  pos <- which(dat$sev == 1); neg <- which(dat$sev == 0)
  idx <- c(sample(pos, round(.7*length(pos))), sample(neg, round(.7*length(neg))))
  tr <- dat[idx, ]; te <- dat[-idx, ]; ty <- te$sev

  # logistic baseline
  gl <- suppressWarnings(glm(sev ~ o2L+sputum+iddm+ckd+temp, data = tr, family = binomial))
  p_glm <- predict(gl, te, type = "response")
  # tree
  tr_m <- rpart(y ~ o2L+sputum+iddm+ckd+temp, data = tr, method="class",
                control = rpart.control(cp = 0.01, minbucket = 8))
  p_tree <- predict(tr_m, te)[, "yes"]
  # random forest
  rf <- ranger(y ~ o2L+sputum+iddm+ckd+temp, data = tr, num.trees = 800,
               probability = TRUE, min.node.size = 5)
  p_rf <- predict(rf, te)$predictions[, "yes"]
  # xgboost
  xm <- xgb.train(params = xgb_params, data = xgb_prep(tr), nrounds = 60, verbose = 0)
  p_xgb <- predict(xm, xgb_prep(te))
  # SVMs (scaled)
  s <- scale_train(tr, te, c("o2L","temp"))
  sr <- svm(y~o2L+sputum+iddm+ckd+temp, data=s$tr, kernel="radial",
            cost=tune_svm$best.parameters$cost, gamma=tune_svm$best.parameters$gamma,
            probability=TRUE)
  sl <- svm(y~o2L+sputum+iddm+ckd+temp, data=s$tr, kernel="linear", cost=1, probability=TRUE)
  p_sr <- attr(predict(sr, s$te, probability=TRUE), "probabilities")[, "yes"]
  p_sl <- attr(predict(sl, s$te, probability=TRUE), "probabilities")[, "yes"]
  # kNN at CV-chosen k (scaled)
  sc <- dat; sc[c("o2L","temp")] <- scale(sc[c("o2L","temp")])
  km <- kknn(y~o2L+sputum+iddm+ckd+temp, train=sc[idx,], test=sc[-idx,],
             k = best_k, kernel = "rectangular")
  p_knn <- km$prob[, "yes"]

  probs <- list(Logistic=p_glm, Tree=p_tree, RandomForest=p_rf, XGBoost=p_xgb,
                `SVM-RBF`=p_sr, `SVM-linear`=p_sl, kNN=p_knn)
  data.frame(
    model = names(probs),
    roc   = sapply(probs, function(p) roc_auc(ty, p)),
    pr    = sapply(probs, function(p) pr_auc(ty, p))
  )
}
res <- do.call(rbind, lapply(1:40, run_split))
agg <- aggregate(cbind(roc, pr) ~ model, data = res, FUN = mean)
agg <- agg[order(-agg$roc), ]
```

```{r comparison-table, echo=FALSE}
knitr::kable(data.frame(Model = agg$model,
                        `ROC-AUC` = round(agg$roc, 3),
                        `PR-AUC`  = round(agg$pr, 3),
                        check.names = FALSE),
             row.names = FALSE, booktabs = TRUE,
             caption = "Held-out performance (mean over 40 stratified 70/30 splits). Base rate of severe illness ~ 0.39, so PR-AUC > 0.39 beats the no-skill line.")
```

```{r comparison-plot, echo=FALSE, fig.cap="Mean held-out ROC-AUC across 40 stratified splits. The reproduced logistic GLM is highlighted; on this small, near-linearly-structured cohort the flexible learners do not clearly beat it."}
agg$is_base <- agg$model == "Logistic"
ggplot(agg, aes(reorder(model, roc), roc, fill = is_base)) +
  geom_col(width = .65) +
  geom_hline(yintercept = 0.5, linetype = 3) +
  coord_flip() +
  scale_fill_manual(values = c("grey60","firebrick"), guide = "none") +
  labs(x = NULL, y = "Mean held-out ROC-AUC",
       title = "Learner comparison (baseline logistic in red)") +
  theme_minimal()
```

## Interpretation: permutation importance and partial dependence

```{r perm-importance, fig.cap="Permutation importance (ranger) on the full cohort: mean decrease in accuracy when each predictor is shuffled."}
rf_full <- ranger(y ~ o2L+sputum+iddm+ckd+temp, data = dat, num.trees = 1500,
                  probability = TRUE, importance = "permutation", min.node.size = 5)
imp <- sort(ranger::importance(rf_full), decreasing = TRUE)
ggplot(data.frame(var = names(imp), imp = as.numeric(imp)),
       aes(reorder(var, imp), imp)) +
  geom_col(fill = "steelblue") + coord_flip() +
  labs(x = NULL, y = "Permutation importance", title = "Random-forest permutation importance") +
  theme_minimal()
```

```{r pdp, fig.cap="Partial-dependence of predicted P(severe) on the two most important predictors, holding the others at their mean/mode. Rug shows observed values. The O2 curve rises steeply then saturates -- a mildly nonlinear effect the logistic model captures only as a straight log-odds slope."}
pdp_curve <- function(model, var, grid) {
  base <- dat[rep(1, length(grid)), X_cols]
  for (cc in X_cols) base[[cc]] <- if (cc %in% c("sputum","iddm","ckd"))
    as.numeric(names(sort(table(dat[[cc]]), decreasing = TRUE))[1]) else mean(dat[[cc]])
  base[[var]] <- grid
  data.frame(x = grid, phat = predict(model, base)$predictions[, "yes"], var = var)
}
o2_grid   <- seq(min(dat$o2L), quantile(dat$o2L, .98), length.out = 60)
temp_grid <- seq(quantile(dat$temp,.02), quantile(dat$temp,.98), length.out = 60)
pdp <- rbind(pdp_curve(rf_full, "o2L", o2_grid),
             pdp_curve(rf_full, "temp", temp_grid))
ggplot(pdp, aes(x, phat)) + geom_line(color = "firebrick", linewidth = 1) +
  facet_wrap(~var, scales = "free_x",
             labeller = as_labeller(c(o2L = "Supplemental O2 (L/min)", temp = "Admit temp (F)"))) +
  labs(x = NULL, y = "Partial dependence: P(severe)") + theme_minimal()
```

# Takeaway

The published logistic model reproduces **exactly** (Table 4 odds ratios matched to three
decimals), anchoring the ML extension in a trustworthy baseline. On this small n $=$ `r nrow(dat)`
cohort the classic learners — tree, random forest, xgboost, SVM (RBF/linear), and CV-tuned k-NN —
**do not reliably beat the logistic GLM** out-of-sample; the strongest signal (admission oxygen
requirement) is close to linear on the log-odds scale, so flexible boundaries mostly add variance.
What ML *does* add here is confirmatory interpretation: permutation importance and the
partial-dependence curves independently rank supplemental O$_2$ (then the comorbidity flags) as the
dominant risk factors, and reveal a saturating O$_2$ effect — a teaching illustration that, in
small clinical cohorts with a well-specified regression, the honest ML lesson is often "the GLM was
already good," with the learners earning their keep as an interpretation cross-check rather than a
prediction win.
