---
title: "Reproduce-then-Extend — State Welfare Spending (American State Politics) (Day 1)"
author: "Intro to Machine Learning · Bruce A. Desmarais"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: true
geometry: margin=1in
fontsize: 10pt
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
  - \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 3.6, fig.align = "center")
options(width = 82)
library(glmnet)   # lasso / ridge / elastic net
```

> **Published study.** The determinants of **state welfare effort** are a long-running question in American
> state politics and policy (e.g., Plotnick & Winters, "A Politico-Economic Theory of Income Redistribution,"
> *APSR* 1985; Brown, "State Responses to the Welfare Crisis," 1995). The canonical design is a
> **cross-sectional OLS** of per-capita state welfare spending on economic need, fiscal capacity, political
> ideology, and organized labor. This example illustrates **regularization and variable selection** in a
> small-sample (n = 50) design.

# Background

Why do some states spend far more on social welfare than others? The state welfare-effort literature points
to **fiscal capacity** (richer states can afford more), **need** (poverty and inequality), **political
ideology and party control** (liberal, Democratic states redistribute more), and **organized labor**. These
explanations are tested with ordinary cross-state regressions. Here we take the **cross-section of the 50
states** and ask which of these characteristics predict welfare spending -- and whether penalized regression
predicts new states better than ordinary least squares.

# Data and codebook

**Unit of analysis:** a U.S. **state** (cross-section averaged over 2005--2012); n = 50. Data come from the
**Correlates of State Policy Project** (Jordan & Grossmann 2020).

| Variable | Definition |
|---|---|
| `welfare_spending_percap` | per-capita state public-welfare spending (USD) **(outcome)** |
| `mood` | public policy mood (higher = more liberal preferences) |
| `citi6013` | citizen ideology (Berry et al.), higher = more liberal |
| `ranney4_control` | Ranney index of Democratic party control (higher = more Democratic) |
| `union_density` | percent of workers unionized |
| `evangelical_pop` | evangelical share of the population |
| `gini_coef` | income inequality (Gini coefficient) |
| `gsppcap` | gross state product per capita (fiscal capacity) |

# Descriptive analysis

Before any modeling we profile the data thoroughly. Because this is a **small-sample** design
(n = 50 states), every descriptive choice -- how skewed the outcome is, how strongly the predictors
correlate with each other -- has direct consequences for whether ordinary least squares can be
trusted, and it is exactly these features that motivate the regularization that follows.

```{r data}
d <- read.csv(paste0(
  "https://raw.githubusercontent.com/bdesmarais/intro",
  "-ml-statistical-horizons-4day/main/tutorials/day3_",
  "regularization_and_selection/12_state_welfare_spen",
  "ding/data/state_welfare_spending.csv"))
preds <- setdiff(names(d), c("st", "welfare_spending_percap"))
cat(sprintf("%d states x %d predictors (p/n = %.2f)\n",
            nrow(d), length(preds), length(preds) / nrow(d)))
```

## Dataset and provenance

- **Unit of analysis:** a U.S. **state**, cross-section averaged over 2005--2012.
- **N:** 50 states (the full population of states, not a sample).
- **Variables:** 1 continuous outcome, `r length(preds)` continuous predictors, plus a state label.
- **Source:** the **Correlates of State Policy Project** (Jordan & Grossmann 2020), which
  compiles standard state-politics measures from Census, BLS, and academic indices.
- **Coverage / sampling:** all 50 states, so there is no sampling uncertainty in the usual
  survey sense; the "population" is fixed and small, which is precisely why a **p/n ratio of
  `r round(length(preds)/nrow(d), 2)`** makes over-fitting the central concern.

```{r dict, echo=FALSE}
dict <- data.frame(
  Variable = c("welfare_spending_percap", "citi6013", "ranney4_control",
               "evangelical_pop", "union_density", "gini_coef", "mood", "gsppcap"),
  Role = c("outcome", rep("predictor", 7)),
  Type = rep("numeric", 8),
  Description = c(
    "per-capita state public-welfare spending (USD)",
    "citizen ideology (Berry et al.); higher = more liberal",
    "Ranney index of Democratic party control; higher = more Democratic",
    "evangelical share of the population (%)",
    "percent of workers unionized",
    "income inequality (Gini coefficient)",
    "public policy mood; higher = more liberal preferences",
    "gross state product per capita (fiscal capacity, USD)"))
knitr::kable(dict, caption = "Variable dictionary: outcome and predictors.")
```

## The outcome in depth

```{r out-stats, echo=FALSE}
y <- d$welfare_spending_percap
ostat <- data.frame(
  n = sum(!is.na(y)), mean = mean(y), sd = sd(y), min = min(y),
  Q1 = quantile(y, .25), median = median(y), Q3 = quantile(y, .75),
  max = max(y), IQR = IQR(y),
  skew = e1071::skewness(y), kurtosis = e1071::kurtosis(y))
knitr::kable(round(ostat, 2), row.names = FALSE,
             caption = "Distribution of per-capita welfare spending (USD).")
```

```{r out-plot, echo=FALSE, fig.height=2.8, fig.cap="Per-capita welfare spending: histogram with kernel density (left) and boxplot (right). The distribution is right-skewed with a few high-spending states."}
library(ggplot2); library(patchwork)
p1 <- ggplot(d, aes(welfare_spending_percap)) +
  geom_histogram(aes(y = after_stat(density)), bins = 15,
                 fill = "#a6cee3", colour = "white") +
  geom_density(colour = "#1f78b4", linewidth = 0.8) +
  labs(x = "per-capita welfare spending (USD)", y = "density") +
  theme_minimal(base_size = 9)
p2 <- ggplot(d, aes(y = welfare_spending_percap)) +
  geom_boxplot(fill = "#a6cee3", width = 0.4) +
  labs(y = "per-capita welfare spending (USD)", x = NULL) +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
p1 + p2 + plot_layout(widths = c(2, 1))
```

The outcome is **right-skewed** (skewness `r round(e1071::skewness(y), 2)`, kurtosis
`r round(e1071::kurtosis(y), 2)`): the median (\$`r format(round(median(y)), big.mark=",")`) sits
below the mean (\$`r format(round(mean(y)), big.mark=",")`), and a short right tail of high-spending
states (e.g., Alaska, New York) pulls the mean up. There is **no zero-inflation** and no floor or
ceiling -- spending ranges from \$`r format(round(min(y)), big.mark=",")` to
\$`r format(round(max(y)), big.mark=",")`. The skew is mild enough that we keep the outcome on its
natural (dollar) scale for comparability with the published regression; a log transform is a
sensible robustness check (see the exercises) but is not required to fit the model well.

## Univariate predictor distributions

```{r pred-stats, echo=FALSE}
sumrow <- function(x) c(
  n = sum(!is.na(x)), n_miss = sum(is.na(x)), mean = mean(x, na.rm = TRUE),
  sd = sd(x, na.rm = TRUE), min = min(x, na.rm = TRUE),
  Q1 = quantile(x, .25, na.rm = TRUE), median = median(x, na.rm = TRUE),
  Q3 = quantile(x, .75, na.rm = TRUE), max = max(x, na.rm = TRUE),
  skew = e1071::skewness(x, na.rm = TRUE))
ptab <- t(sapply(d[preds], sumrow))
colnames(ptab) <- c("n", "n_miss", "mean", "sd", "min", "Q1",
                    "median", "Q3", "max", "skew")
knitr::kable(round(ptab, 2),
             caption = "Univariate summary of the predictors.")
```

```{r pred-hist, echo=FALSE, fig.height=3.4, fig.cap="Distributions of the seven predictors across the 50 states."}
library(tidyr)
dl <- pivot_longer(d[preds], everything(),
                   names_to = "variable", values_to = "value")
ggplot(dl, aes(value)) +
  geom_histogram(bins = 12, fill = "#b2df8a", colour = "white") +
  facet_wrap(~ variable, scales = "free", ncol = 4) +
  labs(x = NULL, y = "count") +
  theme_minimal(base_size = 8)
```

The predictors are on **very different scales** -- `gsppcap` in tens of thousands of dollars,
`gini_coef` between 0 and 1 -- so `glmnet` standardizes them internally before penalizing. A few are
themselves skewed (e.g., `evangelical_pop`, `union_density`), which matters less for a penalized
predictor set than it would for inference.

## Missing-data analysis

```{r missing, echo=FALSE}
miss <- data.frame(
  variable = c("welfare_spending_percap", preds),
  n_missing = sapply(d[c("welfare_spending_percap", preds)],
                     function(x) sum(is.na(x))))
miss$pct_missing <- round(100 * miss$n_missing / nrow(d), 1)
knitr::kable(miss, row.names = FALSE,
             caption = "Missing values per variable.")
cat(sprintf("Complete cases: %d of %d states (%.0f%%).\n",
            sum(complete.cases(d)), nrow(d),
            100 * mean(complete.cases(d))))
```

The Correlates-of-State-Policy extract used here is **fully complete**: every state has a value on
every variable, so there is no co-missingness pattern to model and the reproduction below runs on all
50 states with no listwise deletion. (In wider CSPP extracts some measures are missing for particular
years; here the 2005--2012 averages are all populated.)

## Bivariate relationships with the outcome

```{r biv-tab, echo=FALSE}
r  <- sapply(d[preds], function(x) cor(x, y))
p  <- sapply(d[preds], function(x) cor.test(x, y)$p.value)
biv <- data.frame(predictor = preds, r = round(r, 2),
                  p_value = round(p, 3))
biv <- biv[order(-abs(biv$r)), ]
knitr::kable(biv, row.names = FALSE,
             caption = "Marginal (pairwise) association of each predictor with welfare spending.")
```

```{r biv-plot, echo=FALSE, fig.height=3.4, fig.cap="Welfare spending vs each predictor, with an OLS smoother. Fiscal capacity (gsppcap) shows the clearest positive gradient."}
dl2 <- pivot_longer(d[c("welfare_spending_percap", preds)], -welfare_spending_percap,
                    names_to = "variable", values_to = "value")
ggplot(dl2, aes(value, welfare_spending_percap)) +
  geom_point(colour = "#1f78b4", alpha = 0.6, size = 1) +
  geom_smooth(method = "lm", se = FALSE, colour = "#e31a1c", linewidth = 0.7) +
  facet_wrap(~ variable, scales = "free_x", ncol = 4) +
  labs(x = NULL, y = "welfare spending (USD)") +
  theme_minimal(base_size = 8)
```

Marginally, **fiscal capacity** (`gsppcap`) and the **ideology / party-control** measures
(`citi6013`, `mood`, `ranney4_control`) move with welfare spending in the expected direction, though
the individual pairwise correlations are modest -- typical for 50 noisy state observations. No single
predictor dominates, which is exactly the setting where combining them (and penalizing) helps.

## Multivariate structure and collinearity

```{r corr-heat, echo=FALSE, fig.height=3.6, fig.cap="Correlation heatmap of the numeric predictors. Some ideology/party-control measures are moderately correlated."}
cm <- cor(d[preds])
cmdf <- as.data.frame(as.table(cm))
names(cmdf) <- c("Var1", "Var2", "r")
ggplot(cmdf, aes(Var1, Var2, fill = r)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = round(r, 2)), size = 2.6) +
  scale_fill_gradient2(low = "#e31a1c", mid = "white", high = "#1f78b4",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r vif, echo=FALSE}
X <- d[preds]
vif <- sapply(preds, function(v) {
  r2 <- summary(lm(X[[v]] ~ ., data = X[setdiff(preds, v)]))$r.squared
  1 / (1 - r2)
})
knitr::kable(data.frame(predictor = preds, VIF = round(vif, 2)),
             row.names = FALSE,
             caption = "Variance-inflation factors from the reproduction model.")
hi <- which(abs(cm) > 0.7 & upper.tri(cm), arr.ind = TRUE)
if (nrow(hi) == 0) {
  cat("No pair of predictors has |r| > 0.7.\n")
} else {
  for (i in seq_len(nrow(hi)))
    cat(sprintf("|r| > 0.7: %s <-> %s (r = %.2f)\n",
                preds[hi[i, 1]], preds[hi[i, 2]], cm[hi[i, 1], hi[i, 2]]))
}
```

No predictor pair exceeds **|r| > 0.7**, and all VIFs are modest (well below the conventional
threshold of 5), so collinearity is not severe. The over-fitting problem here is driven not by
collinearity but by the **small n relative to p**: with only 50 states and
`r length(preds)` predictors, OLS has few observations per coefficient, inflating out-of-sample
variance. That is the specific weakness the regularization below corrects.

# Reproduce the published regression

The welfare-effort model regresses per-capita spending on need, capacity, ideology, and labor.

```{r reproduce}
fit <- lm(welfare_spending_percap ~ ., data = d[, c("welfare_spending_percap", preds)])
round(summary(fit)$coefficients, 3)
```

**Confirmation against the literature.** In-sample $R^2$ = `r round(summary(fit)$r.squared, 2)`. Consistent
with the welfare-effort literature, states with **higher income/fiscal capacity**, more liberal **ideology
and Democratic control**, and stronger **unions** spend more per capita on welfare.

# Regularization & variable selection

With **50 states** and **`r length(preds)` predictors**, ordinary least squares over-fits. Penalized
regression -- lasso ($L_1$), ridge ($L_2$), elastic net -- shrinks the coefficients to reduce variance and
improve prediction on new states.

```{r reg-coef}
X <- as.matrix(d[, preds]); y <- d$welfare_spending_percap
set.seed(2026)
lasso <- cv.glmnet(X, y, alpha = 1); ridge <- cv.glmnet(X, y, alpha = 0); enet <- cv.glmnet(X, y, alpha = 0.5)
n_keep <- sum(coef(lasso, s = "lambda.min")[-1] != 0)
cat(sprintf("At lambda.min the lasso keeps %d of %d predictors (the rest are set to exactly 0);\n",
            n_keep, length(preds)))
cat("the signal here is concentrated, so the sparser lambda.1se rule shrinks all the way to the mean.\n")
data.frame(
  OLS        = round(coef(fit)[-1], 2),
  Lasso      = round(as.numeric(coef(lasso, s = "lambda.min"))[-1], 2),
  Ridge      = round(as.numeric(coef(ridge, s = "lambda.min"))[-1], 2),
  ElasticNet = round(as.numeric(coef(enet,  s = "lambda.min"))[-1], 2),
  row.names  = preds)
```

**Which fits best out of sample?** With n = 50 we **repeat a 70/30 split 50 times**: each replicate tunes the
penalty by cross-validation on the training states and scores once on the held-out states (penalized models
at `lambda.min`).

```{r reg-perf}
rmse <- function(p, yt) sqrt(mean((yt - p)^2))
REPS <- 50; alphas <- c(Lasso = 1, Ridge = 0, ElasticNet = 0.5)
M <- matrix(NA, REPS, 4, dimnames = list(NULL, c("OLS", names(alphas))))
Xdf <- as.data.frame(X)
for (r in 1:REPS) {
  set.seed(2025 + r)
  te <- sample(nrow(X), round(0.30 * nrow(X))); tr <- setdiff(seq_len(nrow(X)), te)
  ols <- lm(y[tr] ~ ., data = Xdf[tr, ]); M[r, "OLS"] <- rmse(predict(ols, Xdf[te, ]), y[te])
  for (a in names(alphas)) {
    cv <- cv.glmnet(X[tr, ], y[tr], alpha = alphas[a])
    M[r, a] <- rmse(as.numeric(predict(cv, X[te, ], s = "lambda.min")), y[te])
  }
}
round(colMeans(M))                                       # mean held-out RMSE ($) over 50 splits
best <- names(which.min(colMeans(M)[names(alphas)]))
cat(sprintf("\nBest penalization: %s. Held-out RMSE: %s = %s vs OLS %s (%.0f%% lower).\n",
            best, best, format(round(min(colMeans(M)[names(alphas)])), big.mark = ","),
            format(round(colMeans(M)["OLS"]), big.mark = ","),
            100 * (colMeans(M)["OLS"] - min(colMeans(M)[names(alphas)])) / colMeans(M)["OLS"]))
```

With so few states per predictor, unpenalized OLS over-fits and the penalized models **predict new states'
welfare spending more accurately**, while the lasso keeps a compact, interpretable set of drivers.

```{r reg-path, echo=FALSE, fig.height=3.4, fig.cap="Lasso coefficient paths vs the penalty (log lambda). The dashed line marks the CV-optimal penalty."}
par(mar = c(4.2, 4.2, 3, 1))
plot(lasso$glmnet.fit, xvar = "lambda", col = "#1f78b455")
abline(v = log(lasso$lambda.min), lty = 2, col = "#e31a1c")
```

# Best-subset selection with ABESS

```{r abess}
library(abess)
ab <- abess(X, y)
cat(sprintf("ABESS selects a best subset of %d of %d predictors:\n", ab$best.size, length(preds)))
cf <- as.numeric(coef(ab, support.size = ab$best.size))[-1]
data.frame(predictor = preds, coef = round(cf, 2))[order(-abs(cf)), ]
```

Best subset and the lasso agree on the core drivers of welfare effort -- fiscal capacity, ideology/party
control, and labor -- reached by direct search rather than shrinkage.

# Takeaway

Like the state policy-liberalism example, this illustrates **regularization in a small-sample design**: with
50 states and several predictors, ordinary regression over-fits, and lasso/elastic-net (`glmnet`) plus
best-subset selection (`abess`) predict new states more accurately while naming the compact set of drivers of
state welfare spending.

# Recommended exercises

1. Add more Correlates-of-State-Policy predictors and watch the held-out gain over OLS grow with the
   predictor-to-state ratio.
2. Log-transform the spending outcome and re-run: does the ranking of methods change?
3. Compare the lasso-selected predictors here with those selected for policy liberalism (tutorial 11) -- which
   state characteristics matter for both?
