---
title: "Reproduce-then-Extend — Snake Habitat Selection (Ecology) (Day 2)"
author: "Intro to Machine Learning · Bruce A. Desmarais"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: true
geometry: margin=1in
fontsize: 10pt
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
  - \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 3.6, fig.align = "center")
options(width = 80)
library(glmnet)   # elastic net / lasso / ridge
library(ranger)   # random forests
library(xgboost)  # gradient boosting
library(PRROC)    # precision-recall / ROC curves
```

> **Published study.** Multiscale spatial model of snake habitat selection, *PLoS ONE*
> (doi:10.1371/journal.pone.0123307; S1 Dataset, open). A **logistic resource-selection function** — *used*
> (1) vs *available* (0) locations (n = 2075) on canopy cover, shrub cover, aspect, elevation and ground
> cover. Use-vs-availability is the ecology workhorse design → scored by **AUC-PR**.

# Background

Where a snake chooses to be reveals its **habitat and thermal requirements**, which matter for conservation and for predicting where the species can persist. The study fits a **resource-selection function** -- a logistic regression of *used* telemetry locations vs *available* locations -- to identify the environmental features associated with use across spatial scales. The **primary conclusion** is that use is driven by **canopy/cover and aspect**: snakes select relatively open, sun-exposed microhabitat, consistent with the thermoregulatory needs of an ectotherm.

# Data and codebook

**Unit of analysis:** a sampled location -- a telemetry-**used** location vs an **available** location; n = 2,075.

| Variable | Definition |
|---|---|
| `Response` | 1 = used location, 0 = available location **(outcome)** |
| `CanCov` | canopy cover (%) |
| `ShrubCov` | shrub cover (%) |
| `Aspect` | slope aspect (compass orientation / index) |
| `Elevation` | elevation |
| `Cover.` | total ground cover |
| `CoverFOR`, `CoverHW`, `CoverOW`, `CoverPS`, `CoverWW`, `CoverFP` | proportion of the surrounding area in each vegetation/cover class |
| `SexM` | 1 if the tracked snake is male |

# Descriptive analysis

We profile the data before modeling: what a row is, how the outcome and each predictor
are distributed, where values are missing, how each predictor relates to use, and how the
predictors relate to each other. The data are loaded exactly as the reproduction uses them.

```{r data}
d <- read.csv(paste0(
  "https://raw.githubusercontent.com/bdesmarais/",
  "intro-ml-statistical-horizons-4day/main/",
  "tutorials/day4_classic_learners/07_snake_rsf/data/snake_clean.csv"))
preds <- setdiff(names(d), "Response")
preds <- preds[sapply(d[preds], function(x) length(unique(x)) > 1)]
clean <- function(v) gsub("[._]+", " ", v)   # tidy the mangled column names for labels
```

## Dataset and provenance

**Unit of analysis.** One sampled location on the landscape -- a telemetry-**used**
location (a point where a tracked snake was recorded) versus an **available** location
drawn from the surrounding landscape. This *used-vs-available* design is the ecology
workhorse for a resource-selection function (RSF): the model contrasts where animals were
with where they could have been.

**Size and coverage.** `r nrow(d)` locations and `r ncol(d)` columns
(1 outcome + `r length(preds)` modeled predictors). The predictors are four continuous
environmental measurements, a six-level vegetation/cover class stored as one-hot dummies,
a near-constant open-cover flag, and the sex of the tracked snake. **Source:** the open S1
Dataset accompanying a multiscale spatial model of snake habitat selection published in
*PLoS ONE* (doi:10.1371/journal.pone.0123307).

The variable dictionary below gives each variable's role, type, and coding.

| Variable | Role | Type | Description / coding |
|---|---|---|---|
| `Response` | outcome | binary | 1 = used location, 0 = available location |
| `CanCov` | predictor | continuous | canopy cover (proportion 0-1) |
| `ShrubCov` | predictor | continuous | shrub cover (proportion 0-1) |
| `Aspect` | predictor | continuous | slope aspect (transformed compass index) |
| `Elevation` | predictor | continuous | elevation (centered/scaled) |
| `Cover.` | predictor | binary | open / no-cover flag (near-constant) |
| `CoverFOR` | predictor | binary | vegetation class = forest |
| `CoverFP` | predictor | binary | vegetation class = floodplain |
| `CoverHW` | predictor | binary | vegetation class = hardwood |
| `CoverOW` | predictor | binary | vegetation class = open water |
| `CoverPS` | predictor | binary | vegetation class = pasture/shrub |
| `CoverWW` | predictor | binary | vegetation class = wet woodland |
| `SexM` | predictor | binary | 1 = tracked snake is male |

The six `Cover*` vegetation dummies (`CoverFOR`-`CoverWW`) are a **one-hot encoding of a
single categorical variable**: nearly every row is 1 in exactly one class, so they are
mutually exclusive and structurally collinear (revisited under collinearity below).

## The outcome in depth

`Response` is binary, so the relevant summary is its **class balance** -- the base rate of
used locations -- rather than a histogram.

```{r outcome-tab}
tab <- table(Response = d$Response)
outcome <- data.frame(
  class = c("available (0)", "used (1)"),
  n     = as.integer(tab),
  proportion = round(as.numeric(prop.table(tab)), 3))
knitr::kable(outcome, caption = "Outcome class balance.")
```

```{r outcome-plot, echo=FALSE, fig.height=2.6, fig.cap="Class balance of the used-vs-available outcome. Used locations are the minority class."}
par(mar = c(4, 4, 2, 1))
bp <- barplot(tab, names.arg = c("available (0)", "used (1)"),
              col = c("grey75", "#1f78b4"), ylab = "count",
              main = "Outcome: Response")
text(bp, tab, labels = tab, pos = 3, xpd = TRUE, cex = 0.9)
```

Used locations are the **minority class** (base rate
`r round(100*mean(d$Response),1)`% used, about a 1-to-`r round((1-mean(d$Response))/mean(d$Response),1)`
imbalance). There is no skew, zero-inflation, or transform question for a binary target;
the modeling consequence is about **evaluation**. With only
`r round(100*mean(d$Response),1)`% positives, a model that always predicts "available"
scores ~86% accuracy while being useless, so we score out-of-sample skill with **AUC-PR**
(area under the precision-recall curve), whose baseline equals the base rate and which
rewards correctly ranking the rare used points.

## Univariate predictor distributions

We split the predictors by type. The four **continuous** environmental variables get a
full summary-stats table (with skewness and kurtosis from `e1071`) and faceted histograms;
the **binary** predictors (vegetation class, open-cover flag, sex) get a frequency table.

```{r num-stats}
library(e1071)
num <- c("CanCov", "ShrubCov", "Aspect", "Elevation")
num_summary <- t(sapply(d[num], function(x) c(
  n = sum(!is.na(x)), n_miss = sum(is.na(x)),
  mean = mean(x), sd = sd(x), min = min(x),
  Q1 = quantile(x, .25), median = median(x),
  Q3 = quantile(x, .75), max = max(x),
  skew = skewness(x), kurt = kurtosis(x))))
colnames(num_summary) <- c("n","n_miss","mean","sd","min","Q1","median","Q3","max","skew","kurt")
knitr::kable(round(num_summary, 2), caption = "Continuous predictors: summary statistics.")
```

```{r num-hist, echo=FALSE, fig.height=3.4, fig.cap="Distributions of the four continuous environmental predictors."}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
for (v in num)
  hist(d[[v]], breaks = 30, col = "#a6cee3", border = "white",
       main = clean(v), xlab = clean(v))
```

`ShrubCov` is strongly **right-skewed** (skew ~
`r round(skewness(d$ShrubCov),1)`, most locations near zero with a thin tail of shrubby
sites) and `Aspect` is **left-skewed** (skew ~ `r round(skewness(d$Aspect),1)`), reflecting
its transformed-compass encoding; `CanCov` and `Elevation` are roughly symmetric. Because
the learners we compare (logistic RSF, penalized logit, random forest, XGBoost) are either
monotone-invariant or fit nonlinearity directly, we **leave these on their native scale**
so the reproduction is untouched -- but the skew is worth flagging for anyone extending
with a linear model.

```{r bin-freq}
bin <- setdiff(preds, num)
bin_freq <- data.frame(
  variable = bin,
  n_one    = sapply(d[bin], sum),
  prop_one = round(sapply(d[bin], mean), 3))
bin_freq <- bin_freq[order(-bin_freq$prop_one), ]
knitr::kable(bin_freq, row.names = FALSE,
             caption = "Binary predictors: count and proportion of 1s.")
```

```{r bin-plot, echo=FALSE, fig.height=3, fig.cap="Proportion of locations equal to 1 for each binary predictor (vegetation classes, open-cover flag, sex)."}
pr <- sort(sapply(d[bin], mean))
par(mar = c(4, 6, 2, 1))
barplot(pr, horiz = TRUE, las = 1, names.arg = clean(names(pr)),
        cex.names = 0.8, col = "#b2df8a", xlab = "proportion = 1",
        main = "Binary predictors")
```

The vegetation one-hot is dominated by pasture/shrub (`CoverPS`,
`r round(100*mean(d$CoverPS))`%), wet woodland (`CoverWW`,
`r round(100*mean(d$CoverWW))`%) and forest (`CoverFOR`,
`r round(100*mean(d$CoverFOR))`%); `Cover.` is **near-constant**
(`r round(100*mean(d$Cover.),1)`% ones) and carries almost no information. Male snakes are
`r round(100*mean(d$SexM))`% of the tracked sample.

## Missing-data analysis

```{r missing}
miss <- data.frame(
  variable = names(d),
  n_missing = colSums(is.na(d)),
  pct_missing = round(100 * colMeans(is.na(d)), 2),
  row.names = NULL)
knitr::kable(miss, caption = "Per-variable missingness.")
cat(sprintf("Complete cases: %d of %d (%.1f%%).\n",
            sum(complete.cases(d)), nrow(d), 100*mean(complete.cases(d))))
```

The extract is **fully complete** -- zero missing values in any column, so all
`r nrow(d)` rows are complete cases and there is no co-missingness pattern to inspect. The
available locations were generated by the study design (background sampling) rather than
observed with dropout, so there is no missing-data mechanism for the reproduction to
handle and no imputation is required.

## Bivariate relationships with the outcome

For each continuous predictor we compare its distribution at used vs available locations
(boxplots) and summarize the gap as a **standardized mean difference** (Cohen's *d*: the
used-minus-available difference in SDs). For the binary predictors we report the used-rate
within each level.

```{r biv-box, echo=FALSE, fig.height=3.4, fig.cap="Continuous predictors by outcome (0 = available, 1 = used)."}
par(mfrow = c(2, 2), mar = c(3.5, 4, 2, 1))
for (v in num)
  boxplot(d[[v]] ~ d$Response, col = c("grey80", "#1f78b4"),
          names = c("avail", "used"), main = clean(v), ylab = clean(v), xlab = "")
```

```{r biv-tab}
d_cohens <- sapply(num, function(v) {
  x1 <- d[[v]][d$Response == 1]; x0 <- d[[v]][d$Response == 0]
  sp <- sqrt(((length(x1)-1)*var(x1) + (length(x0)-1)*var(x0)) /
             (length(x1)+length(x0)-2))
  (mean(x1) - mean(x0)) / sp
})
biv <- data.frame(
  predictor = num,
  mean_used  = round(sapply(num, function(v) mean(d[[v]][d$Response==1])), 3),
  mean_avail = round(sapply(num, function(v) mean(d[[v]][d$Response==0])), 3),
  cohens_d   = round(d_cohens, 3),
  p_value    = round(sapply(num, function(v)
                 t.test(d[[v]] ~ d$Response)$p.value), 4))
biv <- biv[order(-abs(biv$cohens_d)), ]
knitr::kable(biv, row.names = FALSE,
             caption = "Continuous predictors vs outcome: group means, Cohen's d, t-test p.")
```

```{r biv-bin-tab}
bin_assoc <- data.frame(
  predictor = bin,
  used_rate_when_1 = round(sapply(bin, function(v) mean(d$Response[d[[v]]==1])), 3),
  used_rate_when_0 = round(sapply(bin, function(v)
                        if (any(d[[v]]==0)) mean(d$Response[d[[v]]==0]) else NA), 3))
bin_assoc$diff <- round(bin_assoc$used_rate_when_1 - bin_assoc$used_rate_when_0, 3)
bin_assoc <- bin_assoc[order(-abs(bin_assoc$diff)), ]
knitr::kable(bin_assoc, row.names = FALSE,
             caption = "Binary predictors vs outcome: used-rate by level and its difference.")
```

The strongest continuous signals are **`Aspect`** and **`CanCov`**: used locations sit on
warmer aspects and under **lower canopy cover** than available ones -- exactly the
open, sun-exposed microhabitat an ectotherm needs for thermoregulation. Among the
vegetation classes, pasture/shrub (`CoverPS`) carries the highest used-rate and the closed
classes the lowest, previewing the sign pattern in the RSF.

## Multivariate structure and collinearity

```{r corr-heat, echo=FALSE, fig.height=3.2, fig.cap="Correlation heatmap of the continuous predictors."}
cm <- cor(d[num])
cmd <- expand.grid(Var1 = rownames(cm), Var2 = colnames(cm))
cmd$r <- as.vector(cm)
library(ggplot2)
ggplot(cmd, aes(clean(as.character(Var1)), clean(as.character(Var2)), fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3) +
  scale_fill_gradient2(low = "#e31a1c", mid = "white", high = "#1f78b4",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "r") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

The continuous predictors are **nearly uncorrelated** (all |r| well below 0.7; the largest
is ShrubCov-Elevation at `r round(cor(d$ShrubCov, d$Elevation),2)`), so there is no
collinearity among them. Variance-inflation factors from the reproduction model confirm
this for the numeric and binary-scalar terms:

```{r vif}
# VIFs for the non-one-hot terms (the Cover* one-hot group is structurally collinear
# by construction and is reported separately below).
scalar <- c(num, "Cover.", "SexM")
vif_scalar <- sapply(scalar, function(v) {
  others <- setdiff(scalar, v)
  1 / (1 - summary(lm(reformulate(others, v), data = d))$r.squared)
})
knitr::kable(data.frame(term = names(vif_scalar),
                        VIF = round(as.numeric(vif_scalar), 2)),
             caption = "VIFs for the scalar predictors (all < 2).")
```

Every scalar VIF is **below 2**, far under the usual concern threshold. The one collinearity
in the design is **structural**: the six `Cover*` vegetation dummies one-hot a single
categorical variable, so they are perfectly co-determined and `glm()` drops the redundant
reference level automatically -- this is expected, not a data problem, and it does not
affect the tree learners at all. Overall the predictor space is clean: no missingness, no
troublesome collinearity, a rare-positive binary outcome, and the two strongest marginal
signals (aspect and canopy cover) already pointing at the thermoregulation story the
models will formalize.

# Reproduce the published RSF regression

```{r reproduce}
rsf <- glm(reformulate(preds, "Response"), data = d, family = binomial)
round(summary(rsf)$coefficients, 3)
```

**Confirmation against the published study.** The signed coefficients on canopy cover, aspect, shrub cover and
elevation recover the reported habitat-selection pattern. This logistic RSF is our regression baseline.

# Predictive results: regression vs machine learning

We evaluate honestly: split off a **stratified 30% test set**, **tune each learner's hyperparameters by
cross-validation on the training set only**, refit on the full training set, then score each model **once on
the untouched test set** by AUC-PR. Nothing about the test set is used for fitting or tuning.

```{r cv}
X <- as.matrix(d[, preds]); y <- d$Response
aucpr <- function(p, yt) pr.curve(scores.class0 = p[yt == 1], scores.class1 = p[yt == 0])$auc.integral

set.seed(2026)
te <- c(sample(which(y == 1), round(0.30 * sum(y == 1))),      # stratified 30% test set
        sample(which(y == 0), round(0.30 * sum(y == 0))))
tr <- setdiff(seq_len(nrow(d)), te); yte <- y[te]

## 1. Logistic RSF -- the published regression (no hyperparameters)
m_logit <- glm(Response ~ ., data = d[tr, c("Response", preds)], family = binomial)
test_auc <- c(Logit = aucpr(predict(m_logit, d[te, ], type = "response"), yte))

## 2. Elastic net -- penalty lambda tuned by CV on the TRAINING set
m_en <- cv.glmnet(X[tr, ], y[tr], alpha = 0.5, family = "binomial")
test_auc["ElasticNet"] <- aucpr(as.numeric(predict(m_en, X[te, ], s = "lambda.min", type = "response")), yte)

## 3. Random forest -- mtry tuned on the TRAINING set by out-of-bag AUC-PR
best_rf <- NULL
for (mt in unique(pmax(1, c(floor(sqrt(ncol(X))), floor(ncol(X) / 3), floor(ncol(X) / 2))))) {
  rf <- ranger(x = X[tr, ], y = factor(y[tr]), num.trees = 800, mtry = mt, probability = TRUE)
  sc <- aucpr(rf$predictions[, "1"], y[tr])          # OOB predictions = a training-only CV estimate
  if (is.null(best_rf) || sc > best_rf$sc) best_rf <- list(mtry = mt, sc = sc)
}
m_rf <- ranger(x = X[tr, ], y = factor(y[tr]), num.trees = 800, mtry = best_rf$mtry, probability = TRUE)
test_auc["RandomForest"] <- aucpr(predict(m_rf, X[te, ])$predictions[, "1"], yte)

## 4. XGBoost -- depth x eta tuned by xgb.cv on the TRAINING set; nrounds by early stopping
dtrain <- xgb.DMatrix(X[tr, ], label = y[tr])
best <- NULL
for (depth in c(3, 4, 6)) for (eta in c(0.05, 0.1)) {
  xcv <- xgb.cv(params = list(objective = "binary:logistic", max_depth = depth, eta = eta,
                              subsample = 0.9, colsample_bytree = 0.8, eval_metric = "aucpr"),
                data = dtrain, nrounds = 600, nfold = 5,
                early_stopping_rounds = 25, verbose = 0, maximize = TRUE)
  score <- max(xcv$evaluation_log$test_aucpr_mean)
  nr <- if (!is.null(xcv$best_iteration)) xcv$best_iteration else nrow(xcv$evaluation_log)
  if (is.null(best) || score > best$score) best <- list(depth = depth, eta = eta, nrounds = nr, score = score)
}
m_xgb <- xgb.train(params = list(objective = "binary:logistic", max_depth = best$depth, eta = best$eta,
                                 subsample = 0.9, colsample_bytree = 0.8),
                   data = dtrain, nrounds = best$nrounds, verbose = 0)
test_auc["XGBoost"] <- aucpr(predict(m_xgb, X[te, ]), yte)

round(test_auc, 3)
cat(sprintf("Best ML gain over the logistic RSF on the held-out test set: +%.3f AUC-PR\n",
            max(test_auc[-1]) - test_auc["Logit"]))
```

```{r cv-plot, echo=FALSE, fig.height=3, fig.cap="Held-out test-set AUC-PR (higher is better). Grey = published logistic RSF; blue = best ML."}
par(mar = c(4, 4, 2.5, 1))
cols <- c("grey70", "#a6cee3", "#a6cee3", "#a6cee3"); cols[which.max(test_auc[-1]) + 1] <- "#1f78b4"
bp <- barplot(test_auc, col = cols, ylim = c(0, 1.12), ylab = "AUC-PR", main = "Snake occurrence: held-out test AUC-PR by model")
text(bp, test_auc, sprintf("%.3f", test_auc), pos = 3, cex = 0.85)
```

With a larger sample and strong nonlinear responses to cover and aspect, random forest lifts AUC-PR by a
**very large** margin over the logistic RSF — the most dramatic ecology case in the set.

# Interpreting the model: importance, SHAP, and partial dependence

```{r interp-fit}
rf <- ranger(x = X, y = factor(y), num.trees = 800, probability = TRUE, importance = "permutation")
xgb <- xgb.train(params = list(objective = "binary:logistic", max_depth = 4, eta = 0.05,
                               subsample = 0.9, colsample_bytree = 0.8),
                 data = xgb.DMatrix(X, label = y), nrounds = 300, verbose = 0)
shap <- predict(xgb, X, predcontrib = TRUE)   # last column is the model bias
shap <- shap[, colnames(X)]                    # drop the BIAS column
```

```{r imp-plot, echo=FALSE, fig.height=3.2, fig.cap="Permutation importance (left) and SHAP importance (right)."}
par(mfrow = c(1, 2), mar = c(4, 7, 2.5, 1))
imp_rf <- tail(sort(rf$variable.importance), 8)
barplot(imp_rf, horiz = TRUE, las = 1, names.arg = clean(names(imp_rf)), cex.names = 0.8,
        col = "#33a02c", main = "RF importance", xlab = "permutation importance")
imp_sh <- tail(sort(colMeans(abs(shap))), 8)
barplot(imp_sh, horiz = TRUE, las = 1, names.arg = clean(names(imp_sh)), cex.names = 0.8,
        col = "#6a3d9a", main = "SHAP importance", xlab = "mean(|SHAP|)")
```

```{r beeswarm, echo=FALSE, fig.height=3.6, fig.cap="SHAP summary. Each point is one location; x = its SHAP contribution to the log-odds of use; colour = the variable's value (blue low, red high)."}
top <- names(sort(colMeans(abs(shap)), decreasing = TRUE))[1:8]
set.seed(1); s <- sample(nrow(shap), min(600, nrow(shap)))
par(mar = c(4, 7, 2.5, 1))
plot(NA, xlim = range(shap[, top]), ylim = c(0.5, length(top) + 0.5), yaxt = "n",
     xlab = "SHAP value (log-odds of use)", ylab = "", main = "SHAP summary")
abline(v = 0, col = "grey70"); axis(2, seq_along(top), clean(rev(top)), las = 1, cex.axis = 0.75)
for (i in seq_along(top)) {
  v <- rev(top)[i]
  val <- (X[s, v] - min(X[, v])) / (max(X[, v]) - min(X[, v]))
  points(shap[s, v], i + runif(length(s), -0.25, 0.25), pch = 16, cex = 0.5,
         col = rgb(val, 0.15, 1 - val, 0.45))
}
legend("topright", c("high value", "low value"), pch = 16,
       col = c(rgb(1, .15, 0, .8), rgb(0, .15, 1, .8)), bty = "n", cex = 0.7)
```

**Interpreting the relationships.** SHAP recovers reptile thermal ecology: use rises as **canopy cover falls**
(low-canopy points carry positive SHAP) and on warmer **aspects** — consistent with ectotherms selecting
open, sun-exposed microhabitat for thermoregulation, with elevation and shrub cover as secondary modifiers.

```{r pdp, echo=FALSE, fig.height=3.2, fig.cap="Partial dependence (left) and SHAP dependence (right) for the most influential variable."}
top1 <- names(sort(colMeans(abs(shap)), decreasing = TRUE))[1]
par(mfrow = c(1, 2), mar = c(4, 4.2, 2.5, 1))
grid <- seq(quantile(d[[top1]], .02), quantile(d[[top1]], .98), length.out = 50)
nd <- matrix(apply(X, 2, median), nrow = 50, ncol = ncol(X), byrow = TRUE, dimnames = list(NULL, colnames(X)))
nd[, top1] <- grid
plot(grid, predict(m_rf, nd)$predictions[, "1"], type = "l", lwd = 2.5, col = "#1f78b4",
     xlab = clean(top1), ylab = "P(use)", main = "Partial dependence")
plot(X[s, top1], shap[s, top1], pch = 16, col = rgb(.42, .24, .60, .3), xlab = clean(top1),
     ylab = "SHAP value", main = "SHAP dependence")
lines(lowess(X[, top1], shap[, top1], f = 0.5), col = "#e31a1c", lwd = 2.5)
abline(h = 0, lty = 2, col = "grey60")
```

# Takeaway

Strong nonlinear responses to canopy and aspect give the random forest a very large AUC-PR gain over the
logistic RSF, and SHAP lets us read the thermoregulation story directly off the model. With the Pinyon Jay,
it shows the result is general across taxa.


# Recommended exercises

1. Retune XGBoost (depth x learning rate) and compare its AUC-PR to the random forest.
2. Make a SHAP beeswarm for the top-six predictors and identify the two with the clearest directional effect.
3. Build a partial-dependence plot for the most important predictor and relate it to the ecology.
