---
title: "Reproduce-then-Extend — Fatal Militarized Disputes (Gibler & Braithwaite) (Day 2)"
author: "Intro to Machine Learning · Bruce A. Desmarais"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: true
geometry: margin=1in
fontsize: 10pt
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
  - \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 3.6, fig.align = "center")
options(width = 80)
library(glmnet)   # elastic net / lasso / ridge
library(ranger)   # random forests
library(xgboost)  # gradient boosting
library(PRROC)    # precision-recall / ROC curves
library(ggplot2)  # EDA plots
library(dplyr)    # EDA wrangling
library(tidyr)    # reshaping for plots
library(knitr)    # summary tables
library(e1071)    # skewness / kurtosis
library(patchwork)# compose EDA panels
```

> **Published study.** Gibler, D. M. & Braithwaite, A., "Hostile Contiguous States" (Harvard Dataverse
> doi:10.7910/DVN/HQLOOJ) — a binary-response regression for whether a dyadic militarized dispute becomes
> **fatal**, on dyadic predictors. This is *interstate* conflict, distinct from the civil-war cases; fatal
> disputes are rare (~1.3%) → scored by **AUC-PR**.

# Background

Gibler and Braithwaite ask why a small fraction of **militarized interstate disputes escalate to fatalities** while most end without deaths, emphasizing the role of geography. They use a **logistic regression** of whether a dispute turns fatal on dyadic characteristics -- contiguity, whether the dispute is over territory, the balance of military capabilities, alliances, and joint democracy. The **primary conclusion** is that **geography and stakes drive lethal escalation**: contiguous states and territorial disputes are far more likely to produce fatalities, while joint democracy dampens escalation.

# Data and codebook

**Unit of analysis:** a militarized interstate dispute-dyad; n = 20,263 (~1.3% fatal).

| Variable | Definition |
|---|---|
| `fatal` | 1 if the militarized dispute produced at least one battle death **(outcome)** |
| `lowdem` | lower of the two states' Polity democracy scores (weakest-link) |
| `terr500k10yr` | 1 if the dispute is over territory |
| `defense` | 1 if the two states share a defensive alliance |
| `caprat` | national-capability ratio (stronger to weaker state, CINC) |
| `contig` | 1 if the states are geographically contiguous |
| `peaceyears` | years since the dyad's previous dispute |
| `spline1`-`spline3` | cubic-spline bases for peace-years (temporal dependence) |
| `interaction4`, `pnbdem_low` | constructed interaction terms from the published specification |

# Descriptive analysis

```{r data}
d <- read.csv(paste0(
  "https://raw.githubusercontent.com/bdesmarais/intro",
  "-ml-statistical-horizons-4day/main/tutorials/day4_",
  "classic_learners/05_gibler_braithwaite/data/gb20.c",
  "sv"))
preds <- setdiff(names(d), "fatal")
preds <- preds[sapply(d[preds], function(x) length(unique(x)) > 1)]
clean <- function(v) gsub("[._]+", " ", v)   # tidy names for plots
```

## Dataset and provenance

**Unit of analysis.** One row is a **militarized interstate dispute-dyad** — a
pair of states involved in a militarized dispute. The data ship with the
published replication package (Gibler & Braithwaite, Harvard Dataverse
doi:10.7910/DVN/HQLOOJ) and describe the cross-national universe of interstate
disputes rather than a sample, so there is no sampling frame to correct for.

```{r provenance}
c(rows = nrow(d), columns = ncol(d),
  n_predictors = length(preds), n_fatal = sum(d$fatal),
  complete_rows = sum(complete.cases(d)))
```

The frame has `r nrow(d)` disputes and `r ncol(d)` columns — a **binary
outcome** (`fatal`) and `r length(preds)` numeric predictors. Only two
predictors are indicator (0/1) variables (`contig`, `defense`); the rest are
counts or continuous ratios. A variable dictionary:

| Variable | Role | Type | Coding / description |
|---|---|---|---|
| `fatal` | outcome | binary | 1 if the dispute produced >= 1 battle death |
| `contig` | predictor | binary | 1 if the two states are geographically contiguous |
| `terr500k10yr` | predictor | count | intensity of territorial stakes in the dispute |
| `defense` | predictor | binary | 1 if the dyad shares a defensive alliance |
| `caprat` | predictor | ratio | national-capability ratio (CINC), 0-1 |
| `lowdem` | predictor | integer | weaker state's Polity democracy score (-10..10) |
| `pnbdem_low` | predictor | continuous | democracy-based interaction term |
| `interaction4` | predictor | integer | constructed interaction from the published model |
| `peaceyears` | predictor | count | years since the dyad's previous dispute |
| `spline1`-`spline3` | predictor | continuous | cubic-spline bases of `peaceyears` |

## The outcome in depth

`fatal` is a **rare binary event**. The class-balance table and the base rate
below are the single most important facts for how we evaluate models.

```{r outcome-stats}
base_rate <- mean(d$fatal)
data.frame(non_fatal = sum(d$fatal == 0), fatal = sum(d$fatal == 1),
           base_rate = round(base_rate, 4),
           imbalance_ratio = round(sum(d$fatal == 0) / sum(d$fatal == 1), 1))
```

```{r outcome-bar, fig.height=2.6, fig.cap="Class balance of the fatal-dispute outcome. Fatal disputes are the rare positive class (1.3%)."}
ggplot(d, aes(factor(fatal, labels = c("non-fatal", "fatal")))) +
  geom_bar(aes(fill = factor(fatal)), width = 0.6, show.legend = FALSE) +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.3) +
  scale_fill_manual(values = c("grey70", "#1f78b4")) +
  labs(x = NULL, y = "count",
       title = sprintf("Outcome base rate = %.1f%% fatal", 100 * base_rate)) +
  theme_minimal(base_size = 10)
```

Only **`r sum(d$fatal)` of `r nrow(d)`** disputes (**`r sprintf("%.1f%%",
100*base_rate)`**) are fatal — an imbalance of about **76:1**. There is no
distribution shape to transform here; the modeling concern is the *rarity*
itself. With a base rate this low, a model that predicts "never fatal" is
`r sprintf("%.1f%%", 100*(1-base_rate))` accurate yet useless, so plain accuracy
is meaningless. We therefore score every model by **AUC-PR** (area under the
precision-recall curve), whose no-skill baseline is the base rate
(`r round(base_rate,3)`) and which rewards finding the rare fatal cases.

## Univariate predictor distributions

All predictors are numeric, so we summarise them in one table (n, missing,
moments, quartiles, skew) and then show the faceted distributions. Because the
spline bases span hundreds of thousands, we round to keep the table inside the
margin.

```{r pred-summary}
psum <- t(sapply(d[preds], function(x) c(
  n = length(x), n_miss = sum(is.na(x)),
  mean = mean(x), sd = sd(x), min = min(x),
  Q1 = quantile(x, .25), median = median(x),
  Q3 = quantile(x, .75), max = max(x), skew = skewness(x))))
kable_ok <- psum
kable_ok[, 3:9] <- signif(kable_ok[, 3:9], 3)
kable_ok[, "skew"] <- round(kable_ok[, "skew"], 2)
kable(kable_ok, caption = "Univariate summary of the predictors.")
```

```{r pred-hists, fig.height=4.2, fig.cap="Distributions of the eight substantive predictors (spline bases omitted; they are deterministic transforms of peaceyears). Most are right-skewed with heavy zero mass."}
show <- setdiff(preds, c("spline1", "spline2", "spline3"))
d %>% select(all_of(show)) %>%
  pivot_longer(everything(), names_to = "var", values_to = "value") %>%
  mutate(var = clean(var)) %>%
  ggplot(aes(value)) +
  geom_histogram(bins = 30, fill = "#1f78b4", colour = "white", linewidth = 0.1) +
  facet_wrap(~ var, scales = "free", ncol = 4) +
  labs(x = NULL, y = "count") +
  theme_minimal(base_size = 9)
```

The two indicators are heavily lopsided — only **`r round(100*mean(d$contig))`%**
of dyads are contiguous and **`r round(100*mean(d$defense))`%** share a defensive
alliance. `terr500k10yr`, `interaction4` and `peaceyears` have large spikes at
zero and long right tails (skew 2-3). The spline bases are extreme,
near-collinear transforms of `peaceyears` (see the collinearity subsection),
which is why they are excluded from the panels above.

## Missing-data analysis

```{r missing}
miss <- data.frame(
  variable = names(d),
  n_missing = colSums(is.na(d)),
  pct_missing = round(100 * colMeans(is.na(d)), 2),
  row.names = NULL)
miss[order(-miss$n_missing), ]
```

The replication file is **fully complete** — zero missing cells in any of the
`r ncol(d)` columns, so **all `r nrow(d)`** disputes are complete cases. This is
expected: the predictors are constructed from coded dyadic attributes
(contiguity, alliances, capability ratios) that are defined for every dispute.
The reproduction therefore needs no imputation or listwise deletion, and no
missingness mechanism has to be assumed.

## Bivariate relationships with the outcome

For the two indicators we compare fatal rates across groups; for the numeric
predictors we report the point-biserial correlation with `fatal` and the
standardized mean difference (SMD) between fatal and non-fatal disputes.

```{r bivar-binary}
rate_by <- function(g) c(rate0 = mean(d$fatal[g == 0]),
                         rate1 = mean(d$fatal[g == 1]))
biv_bin <- rbind(
  contiguous = rate_by(d$contig),
  defensive_alliance = rate_by(d$defense),
  territorial = c(mean(d$fatal[d$terr500k10yr == 0]),
                  mean(d$fatal[d$terr500k10yr > 0])))
biv_bin <- cbind(round(biv_bin, 4),
                 rel_risk = round(biv_bin[, 2] / biv_bin[, 1], 1))
kable(biv_bin, col.names = c("rate if 0/none", "rate if 1/any", "relative risk"),
      caption = "Fatal rate split by the key categorical / stakes predictors.")
```

```{r bivar-numeric}
biv_num <- data.frame(
  r_with_fatal = sapply(d[preds], function(x) cor(x, d$fatal)),
  std_mean_diff = sapply(d[preds],
    function(x) (mean(x[d$fatal == 1]) - mean(x[d$fatal == 0])) / sd(x)))
biv_num <- round(biv_num[order(-abs(biv_num$std_mean_diff)), ], 3)
kable(biv_num, caption = "Marginal association of each predictor with fatality, ranked by |SMD|.")
```

```{r bivar-plot, fig.height=3, fig.cap="Contiguity is decisive: 26% of contiguous-state disputes are fatal versus 0.4% otherwise (left). Fatal disputes have systematically shorter peace-years (right)."}
p1 <- d %>% group_by(contig) %>% summarise(rate = mean(fatal)) %>%
  ggplot(aes(factor(contig, labels = c("non-contig", "contiguous")), rate)) +
  geom_col(fill = "#1f78b4", width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", 100 * rate)), vjust = -0.3) +
  labs(x = NULL, y = "P(fatal)", title = "Fatal rate by contiguity") +
  theme_minimal(base_size = 9)
p2 <- d %>% mutate(fatal = factor(fatal, labels = c("non-fatal", "fatal"))) %>%
  ggplot(aes(fatal, peaceyears, fill = fatal)) +
  geom_boxplot(show.legend = FALSE, outlier.size = 0.4) +
  scale_fill_manual(values = c("grey70", "#1f78b4")) +
  labs(x = NULL, y = "peace-years", title = "Peace-years by outcome") +
  theme_minimal(base_size = 9)
p1 + p2
```

**Contiguity dominates every other signal.** A dispute between neighbouring
states is fatal **`r round(mean(d$fatal[d$contig==1])/mean(d$fatal[d$contig==0]))`
times** as often (26.1% vs 0.35%), and `contig` is by far the strongest marginal
association (r = 0.43, SMD = 3.8). Territorial disputes roughly triple the fatal
rate, and defensive alliances roughly double it. Fatal disputes also have
**shorter peace-years** (negative SMD): recent conflict predicts renewed lethal
escalation. These marginals already sketch the escalation logic the learners
formalise.

## Multivariate structure and collinearity

```{r corr-heat, fig.height=4, fig.cap="Correlation heatmap of the numeric predictors. The three spline bases are near-perfectly collinear with peaceyears and with one another."}
cm <- cor(d[preds])
cm[lower.tri(cm)] <- NA
as.data.frame(cm) %>% mutate(v1 = rownames(cm)) %>%
  pivot_longer(-v1, names_to = "v2", values_to = "r") %>%
  filter(!is.na(r)) %>%
  mutate(v1 = clean(v1), v2 = clean(v2)) %>%
  ggplot(aes(v2, v1, fill = r)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 2.3) +
  scale_fill_gradient2(low = "#e31a1c", mid = "white", high = "#1f78b4",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r vif}
vif <- sapply(preds, function(p) {
  r2 <- summary(lm(reformulate(setdiff(preds, p), p), data = d))$r.squared
  1 / (1 - r2)
})
round(sort(vif, decreasing = TRUE), 1)
```

The heatmap and the VIFs tell the same story: the **three cubic-spline bases are
near-perfectly collinear** — pairwise |r| > 0.99 among themselves and
0.94-0.96 with `peaceyears`, giving **astronomical VIFs** (tens of thousands to
hundreds of thousands). They are, by construction, deterministic transforms of
`peaceyears` meant to let a *linear* logit bend the peace-years effect. Beyond
that block, the substantive predictors are close to orthogonal (all |r| well
below 0.7, VIFs ~1-2). This matters for the modeling that follows: the logit
must carry the redundant spline block to approximate the nonlinear peace-years
decay, whereas the **tree-based learners model that curvature directly from
`peaceyears` alone** and are unbothered by the collinear splines — a concrete
reason to expect them to win on held-out AUC-PR.

# Reproduce the published regression

The `gb20.csv` frame we model below is a **rare-event subsample** — it keeps all
263 fatal disputes but only a 6% random draw of the non-fatal controls (base
rate inflated from 0.08% to 1.3%) so the machine-learning comparison runs on a
tractable, less-degenerate class balance. Choice-based subsampling like this
leaves the story intact but **shifts the logit intercept and jitters the slopes**,
so it does *not* reproduce the paper's printed coefficients. To confirm the
reproduction against the published table we therefore fit the model on the
**full complete-case sample** (n = 321,041), exactly as Gibler & Braithwaite did.

```{r reproduce}
full <- read.csv(paste0(
  "https://raw.githubusercontent.com/bdesmarais/intro",
  "-ml-statistical-horizons-4day/main/tutorials/day4_",
  "classic_learners/05_gibler_braithwaite/data/gb_ful",
  "l.csv"))
gb_logit <- glm(fatal ~ lowdem + terr500k10yr + interaction4 + defense +
                  caprat + contig + pnbdem_low + peaceyears +
                  spline1 + spline2 + spline3,
                data = full, family = binomial)
round(summary(gb_logit)$coefficients, 4)
```

**Confirmation against the published table.** These coefficients reproduce
**Gibler & Braithwaite (2013), the full fatal-MID logit** (the model in the
replication log, all 11 covariates, robust SEs clustered on dyad): `contig`
= 3.54, `terr500k10yr` = 0.176, `interaction4` = 0.0115, `lowdem` = -0.069,
`peaceyears` = -0.246, intercept = -6.30. Territorial disputes and contiguity
raise the probability of a fatal escalation, joint democracy (`lowdem`) lowers
it, and the joint-democracy × neighborhood-instability interaction is positive —
matching the published estimates term-for-term. (The subsampled `d` used for the
learners below gives the same signs but a shifted intercept, as expected under
case-control sampling.)

# Predictive results: regression vs machine learning

We evaluate honestly: split off a **stratified 30% test set**, **tune each learner's hyperparameters by
cross-validation on the training set only**, refit on the full training set, then score each model **once on
the untouched test set** by AUC-PR. Nothing about the test set is used for fitting or tuning.

```{r cv}
X <- as.matrix(d[, preds]); y <- d$fatal
aucpr <- function(p, yt) pr.curve(scores.class0 = p[yt == 1], scores.class1 = p[yt == 0])$auc.integral

set.seed(2026)
te <- c(sample(which(y == 1), round(0.30 * sum(y == 1))),      # stratified 30% test set
        sample(which(y == 0), round(0.30 * sum(y == 0))))
tr <- setdiff(seq_len(nrow(d)), te); yte <- y[te]

## 1. Logistic RSF -- the published regression (no hyperparameters)
m_logit <- glm(fatal ~ ., data = d[tr, c("fatal", preds)], family = binomial)
test_auc <- c(Logit = aucpr(predict(m_logit, d[te, ], type = "response"), yte))

## 2. Penalized logits (Day 3's toolkit): ridge, lasso, elastic net --
##    penalty lambda tuned by CV on the TRAINING set
##    (ridge needs an extended lambda path: glmnet's default path for alpha = 0
##     stops at a fairly large lambda, which silently over-shrinks when n >> p)
m_ridge <- cv.glmnet(X[tr, ], y[tr], alpha = 0,   family = "binomial",
                     lambda.min.ratio = 1e-6)
m_lasso <- cv.glmnet(X[tr, ], y[tr], alpha = 1,   family = "binomial")
m_en    <- cv.glmnet(X[tr, ], y[tr], alpha = 0.5, family = "binomial")
test_auc["Ridge"]      <- aucpr(as.numeric(predict(m_ridge, X[te, ], s = "lambda.min", type = "response")), yte)
test_auc["Lasso"]      <- aucpr(as.numeric(predict(m_lasso, X[te, ], s = "lambda.min", type = "response")), yte)
test_auc["ElasticNet"] <- aucpr(as.numeric(predict(m_en,    X[te, ], s = "lambda.min", type = "response")), yte)

## 3. Random forest -- mtry tuned on the TRAINING set by out-of-bag AUC-PR
best_rf <- NULL
for (mt in unique(pmax(1, c(floor(sqrt(ncol(X))), floor(ncol(X) / 3), floor(ncol(X) / 2))))) {
  rf <- ranger(x = X[tr, ], y = factor(y[tr]), num.trees = 800, mtry = mt, probability = TRUE)
  sc <- aucpr(rf$predictions[, "1"], y[tr])          # OOB predictions = a training-only CV estimate
  if (is.null(best_rf) || sc > best_rf$sc) best_rf <- list(mtry = mt, sc = sc)
}
m_rf <- ranger(x = X[tr, ], y = factor(y[tr]), num.trees = 800, mtry = best_rf$mtry, probability = TRUE,
               importance = "permutation")
test_auc["RF"] <- aucpr(predict(m_rf, X[te, ])$predictions[, "1"], yte)

## 4. XGBoost -- depth x eta tuned by xgb.cv on the TRAINING set; nrounds by early stopping
dtrain <- xgb.DMatrix(X[tr, ], label = y[tr])
best <- NULL
for (depth in c(3, 4, 6)) for (eta in c(0.05, 0.1)) {
  xcv <- xgb.cv(params = list(objective = "binary:logistic", max_depth = depth, eta = eta,
                              subsample = 0.9, colsample_bytree = 0.8, eval_metric = "aucpr"),
                data = dtrain, nrounds = 600, nfold = 5,
                early_stopping_rounds = 25, verbose = 0, maximize = TRUE)
  score <- max(xcv$evaluation_log$test_aucpr_mean)
  nr <- if (!is.null(xcv$best_iteration)) xcv$best_iteration else nrow(xcv$evaluation_log)
  if (is.null(best) || score > best$score) best <- list(depth = depth, eta = eta, nrounds = nr, score = score)
}
m_xgb <- xgb.train(params = list(objective = "binary:logistic", max_depth = best$depth, eta = best$eta,
                                 subsample = 0.9, colsample_bytree = 0.8),
                   data = dtrain, nrounds = best$nrounds, verbose = 0)
test_auc["XGBoost"] <- aucpr(predict(m_xgb, X[te, ]), yte)

round(test_auc, 3)
cat(sprintf("Best ML gain over the logistic RSF on the held-out test set: +%.3f AUC-PR\n",
            max(test_auc[-1]) - test_auc["Logit"]))
```

```{r cv-plot, fig.height=3, fig.cap="Held-out test-set AUC-PR (higher is better). Grey = published logistic RSF; blue = best ML."}
par(mar = c(4, 4, 2.5, 1))
cols <- c("grey70", rep("#a6cee3", length(test_auc) - 1)); cols[which.max(test_auc[-1]) + 1] <- "#1f78b4"
bp <- barplot(test_auc, col = cols, ylim = c(0, 1.12), ylab = "AUC-PR", cex.names = 0.8,
              main = "Militarized disputes: held-out test AUC-PR by model")
text(bp, test_auc, sprintf("%.3f", test_auc), pos = 3, cex = 0.85)
```

The tree ensembles lift AUC-PR notably over the additive logit by capturing interactions among contiguity,
power asymmetry, democracy and recent peace. The **penalized logits add nothing here** --- with
$n \approx 14{,}000$ training dyads and only 11 predictors, cross-validation drives the penalty essentially
to zero and lasso/elastic net ride the plain logit, exactly Day 3's lesson about *when* regularization pays
run in reverse. (Ridge sits a touch lower still: the near-collinear peace-years spline block resists
ridge's grouped shrinkage. Note the `lambda.min.ratio` argument above --- glmnet's default ridge path
stops at a fairly large $\lambda$, and without extending it ridge is silently over-shrunk.)

# Interpreting the model: importance, SHAP, and partial dependence

We interpret the **same models we tuned and fit on the training set above** (`m_rf`,
`m_xgb`) --- no refitting on the full data, so the held-out test set plays no role in
this section either.

```{r interp-fit}
Xtr  <- X[tr, ]
shap <- predict(m_xgb, Xtr, predcontrib = TRUE)   # last column is the model bias
shap <- shap[, colnames(X)]                        # drop the BIAS column
```

```{r imp-plot, fig.height=3.2, fig.cap="Permutation importance (left) and SHAP importance (right)."}
par(mfrow = c(1, 2), mar = c(4, 7, 2.5, 1))
imp_rf <- tail(sort(m_rf$variable.importance), 8)
barplot(imp_rf, horiz = TRUE, las = 1, names.arg = clean(names(imp_rf)), cex.names = 0.8,
        col = "#33a02c", main = "RF importance", xlab = "permutation importance")
imp_sh <- tail(sort(colMeans(abs(shap))), 8)
barplot(imp_sh, horiz = TRUE, las = 1, names.arg = clean(names(imp_sh)), cex.names = 0.8,
        col = "#6a3d9a", main = "SHAP importance", xlab = "mean(|SHAP|)")
```

```{r beeswarm, fig.height=3.6, fig.cap="SHAP summary. Each point is one dispute; x = its SHAP contribution to the log-odds of a fatal outcome; colour = the variable's value (blue low, red high)."}
top <- names(sort(colMeans(abs(shap)), decreasing = TRUE))[1:8]
set.seed(1); s <- sample(nrow(shap), min(600, nrow(shap)))
par(mar = c(4, 7, 2.5, 1))
plot(NA, xlim = range(shap[, top]), ylim = c(0.5, length(top) + 0.5), yaxt = "n",
     xlab = "SHAP value (log-odds of fatal)", ylab = "", main = "SHAP summary")
abline(v = 0, col = "grey70"); axis(2, seq_along(top), clean(rev(top)), las = 1, cex.axis = 0.75)
for (i in seq_along(top)) {
  v <- rev(top)[i]
  val <- (Xtr[s, v] - min(Xtr[, v])) / (max(Xtr[, v]) - min(Xtr[, v]))
  points(shap[s, v], i + runif(length(s), -0.25, 0.25), pch = 16, cex = 0.5,
         col = rgb(val, 0.15, 1 - val, 0.45))
}
legend("topright", c("high value", "low value"), pch = 16,
       col = c(rgb(1, .15, 0, .8), rgb(0, .15, 1, .8)), bty = "n", cex = 0.7)
```

**Interpreting the relationships.** SHAP makes the escalation logic explicit: **contiguity dominates** —
disputes between neighbouring states carry large positive SHAP, the single strongest driver — while
**peace-years** are protective and larger **capability ratio** (power asymmetry) and lower **joint democracy**
push risk up. Fatal escalation concentrates among contiguous, power-asymmetric, non-democratic dyads with
little recent peace.

```{r pdp, fig.height=3.2, fig.cap="Partial dependence (left) and SHAP dependence (right) for peace-years."}
par(mfrow = c(1, 2), mar = c(4, 4.2, 2.5, 1))
grid <- seq(quantile(Xtr[, "peaceyears"], .02), quantile(Xtr[, "peaceyears"], .98), length.out = 50)
nd <- matrix(apply(Xtr, 2, median), nrow = 50, ncol = ncol(X), byrow = TRUE, dimnames = list(NULL, colnames(X)))
nd[, "peaceyears"] <- grid
plot(grid, predict(m_rf, nd)$predictions[, "1"], type = "l", lwd = 2.5, col = "#1f78b4",
     xlab = "peaceyears", ylab = "P(fatal)", main = "Partial dependence")
plot(Xtr[s, "peaceyears"], shap[s, "peaceyears"], pch = 16, col = rgb(.42, .24, .60, .3),
     xlab = "peaceyears", ylab = "SHAP value", main = "SHAP dependence")
lines(lowess(Xtr[, "peaceyears"], shap[, "peaceyears"], f = 0.5), col = "#e31a1c", lwd = 2.5)
abline(h = 0, lty = 2, col = "grey60")
```

Peace-years act **nonlinearly** — risk falls fastest in the first years after a dispute and then flattens — a
decay the linear logit (with its peace-year splines) only crudely approximates and trees capture directly.

# Takeaway

Escalation to fatality depends on interactions among contiguity, power asymmetry, democracy and recent peace
that a linear logit cannot fully represent. Boosting captures them for a notable precision–recall gain, and
SHAP reads the drivers back out — extending the regression-to-ML result to *interstate* conflict.

# Independent exercises (each about 10 minutes)

1. **Read a second SHAP plot.** Produce a SHAP dependence plot for one
   predictor *other than* the top-ranked one (reuse the plotting code above).
   Is its shape linear, threshold-like, or interactive --- and could the
   published logit have represented it?

2. **Perturb the forest.** Refit the random forest with `mtry = 2` (instead of
   the tuned `best_rf$mtry`) and recompute its test AUC-PR with `aucpr()`.
   Does the ranking of the learners change?

3. **Tree depth and interactions.** The slides explain that boosted trees
   capture interactions through depth. Refit XGBoost with `max_depth = 1`
   (stumps --- no interactions possible), keeping the other tuned settings and
   `nrounds = best$nrounds`, and recompute test AUC-PR with `aucpr()`. How
   much of the boosting gain over the logit disappears when interactions are
   ruled out?
