---
title: "Model Uncertainty in Cross-Country Growth Regressions"
subtitle: "Fernández, Ley & Steel (2001, *J. Applied Econometrics*) --- Reproduction and ML extension (Day 3: Regularization & Variable Selection)"
author: "Intro-ML teaching tutorial"
date: "`r format(Sys.Date())`"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_width: 7
    fig_height: 4.5
header-includes:
  - \setlength{\emergencystretch}{3em}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{fvextra}
  - \AtBeginDocument{\DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}}
  - \AtBeginDocument{\DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}}
  - \fvset{breaklines=true}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.pos = "H")
options(width = 95)
set.seed(2024)
```

# The study and why it anchors a variable-selection lesson

**Citation.** Carmen Fernández, Eduardo Ley & Mark F. J. Steel (2001), "Model
uncertainty in cross-country growth regressions," *Journal of Applied
Econometrics* **16**(5): 563--576. Data archive:
`http://qed.econ.queensu.ca/jae/2001-v16.5/fernandez-ley-steel/` (also shipped,
identically, in the R package **BMS** as `datafls`).

**The economic problem.** Empirical growth economics asks: *which* national
characteristics explain why some countries grew faster than others between 1960
and 1992? Dozens of candidate determinants have been proposed --- initial GDP,
schooling, life expectancy, religion shares, institutions, geography, trade
openness, investment. Fernández, Ley & Steel assemble **41 candidate regressors
for 72 countries** and confront the central difficulty head-on: with so many
correlated predictors and so few observations, *no single regression is
credible*. Their answer is Bayesian Model Averaging (BMA) --- average over all
$2^{41}\approx 2.2\times10^{12}$ possible models, weighting each by its posterior
probability, and report each variable's **posterior inclusion probability
(PIP)**: the probability that its coefficient is non-zero.

This is *exactly* the Day-3 problem --- many correlated predictors, standardization,
and formal variable selection --- dressed in 2001 Bayesian clothing. We first
**reproduce their headline result exactly**, then show what the modern
regularization toolkit (ridge / lasso / elastic net / best-subset) adds and how
its selections line up with their BMA verdict.

```{r load}
library(BMS)      # ships the exact FLS 2001 data as datafls; provides bms()
d  <- read.csv(paste0(
  "https://raw.githubusercontent.com/bdesmarais/intro",
  "-ml-statistical-horizons-4day/main/tutorials/day3_",
  "regularization_and_selection/fernandez_ley_steel_g",
  "rowth/data/fls_growth.csv"))   # a verbatim copy of BMS::datafls
y  <- d$y                          # growth of real GDP per capita, 1960-1992
X  <- as.matrix(d[, -1])           # the 41 candidate determinants
p  <- ncol(X); n <- nrow(X)
c(countries = n, predictors = p)
```

The outcome `y` is average annual growth of real per-capita GDP, 1960--1992
(Penn World Table). Its mean is about 2.1% with s.d. 1.8%.

# Descriptive analysis

Before any modeling we meet the data properly. Good exploratory work is not a
formality here: with **41 candidate predictors over only 72 countries**, the
descriptive picture --- how the outcome is shaped, which predictors carry
signal, and above all how tightly the predictors are entangled with one another
--- *is* the argument for the model-averaging and regularization machinery that
follows. This section is deliberately thorough.

## Dataset, provenance and unit of analysis

The **unit of analysis is a country**. The file `fls_growth.csv` is a verbatim
copy of `BMS::datafls`, itself the exact dataset archived with Fernández, Ley &
Steel (2001) at the *Journal of Applied Econometrics* data archive. There are
**`r nrow(d)` countries** and **`r ncol(d)` columns** --- the growth outcome plus
**`r ncol(d) - 1` candidate determinants**. Each row is a single cross-section:
every variable is either an *initial-year* value (e.g. 1960 income, `GDP60`) or a
*period average / summary* over the 1960--1992 growth window. There is therefore
**no time dimension to exploit** --- this is one long, wide cross-section, not a
panel. The 72 countries are those for which the full battery of 41 regressors was
available; the sample is not a random draw but the intersection of several source
datasets (Penn World Table, Barro--Lee, Sala-i-Martin's compilations), which is
itself a mild selection caveat worth remembering when generalizing.

```{r eda-types, echo=TRUE}
# classify each predictor as 0/1 binary vs continuous
pred      <- d[ , -1]                       # drop the outcome column
is_binary <- sapply(pred, function(x) all(x %in% c(0, 1)))
bin_vars  <- names(is_binary)[is_binary]    # 0/1 dummies
num_vars  <- names(is_binary)[!is_binary]   # continuous / count
data.frame(kind = c("continuous", "binary (0/1)", "total predictors"),
           count = c(length(num_vars), length(bin_vars), ncol(pred)))
```

So of the 41 regressors, **`r length(num_vars)` are continuous** (incomes,
enrolment shares, life expectancy, geography, investment ratios, ...) and
**`r length(bin_vars)` are 0/1 indicators** (colonial heritage, region, war,
trade orientation). A compact **variable dictionary** for the outcome and a
representative slice of the modeling predictors:

```{r eda-dictionary}
dict <- read.csv(text = "
variable,type,description
y,continuous (outcome),Avg annual growth of real GDP per capita 1960-1992
GDP60,continuous,Log real GDP per capita in 1960 (initial income)
LifeExp,continuous,Life expectancy at birth (years) circa 1960
EquipInv,continuous,Equipment investment share of GDP
NequipInv,continuous,Non-equipment investment share of GDP
PrScEnroll,continuous,Primary-school enrolment rate 1960
HighEnroll,continuous,Higher-education enrolment rate
RuleofLaw,continuous,Rule-of-law index (0-1)
YrsOpen,continuous,Fraction of years economy open to trade
Age,continuous,Age of national state / years since independence
Abslat,continuous,Absolute latitude of the country
Area,continuous,Land area
Popg,continuous,Population growth rate
Confucian,continuous,Fraction of population Confucian
Muslim,continuous,Fraction of population Muslim
Protestants,continuous,Fraction of population Protestant
SubSahara,binary,=1 if Sub-Saharan Africa
LatAmerica,binary,=1 if Latin America
Spanish,binary,=1 former Spanish colony
French,binary,=1 former French colony
Brit,binary,=1 former British colony
WarDummy,binary,=1 country took part in a war 1960-1990
OutwarOr,binary,=1 outward-oriented trade regime
", strip.white = TRUE)
knitr::kable(dict, caption = "Variable dictionary (outcome + a representative subset of the 41 predictors). Continuous shares are proportions in [0,1] unless noted.")
```

The full set of 41 predictors is profiled in the summary table further below;
the dictionary lists the outcome plus the highest-salience determinants and one
example of each dummy family.

## The outcome in depth: average GDP-per-capita growth, 1960--1992

```{r eda-outcome, fig.height=3.4, fig.cap="Distribution of average annual growth of real GDP per capita, 1960--1992, across 72 countries. Solid red = kernel density; dashed = mean; dotted = median. The distribution is close to symmetric with a mild right lean and no extreme outliers."}
library(ggplot2)
ggplot(data.frame(y = d$y), aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 18,
                 fill = "steelblue", colour = "white", alpha = 0.85) +
  geom_density(colour = "firebrick", linewidth = 0.8) +
  geom_vline(xintercept = mean(d$y),   linetype = 2) +
  geom_vline(xintercept = median(d$y), linetype = 3) +
  labs(x = "average annual growth, 1960-1992", y = "density") +
  theme_minimal(base_size = 11)
```

A full numeric profile of the outcome, including shape diagnostics from
**e1071** (skewness and excess kurtosis):

```{r eda-outcome-stats}
library(e1071)
q <- quantile(d$y)
out_stats <- data.frame(
  statistic = c("n","mean","sd","min","Q1","median","Q3","max",
                "IQR","skewness","kurtosis (excess)"),
  value = round(c(length(d$y), mean(d$y), sd(d$y), min(d$y),
                  q[2], median(d$y), q[4], max(d$y), IQR(d$y),
                  skewness(d$y), kurtosis(d$y)), 4))
knitr::kable(out_stats, row.names = FALSE,
  caption = "Full summary of the outcome (annual GDP-per-capita growth rate).")
```

Growth ranges from about $-2\%$ (a handful of countries that *shrank* over three
decades) to roughly $+7\%$ per year, centred near $+2\%$. The mean and median
sit almost on top of each other, the skewness is small and positive
(`r round(skewness(d$y), 2)`) and the excess kurtosis is near zero
(`r round(kurtosis(d$y), 2)`) --- i.e. the distribution is close to Gaussian,
with the right tail a touch heavier than the left because a few East-Asian
"miracle" economies grew exceptionally fast. There is **no floor, ceiling, or
zero-inflation**, and no observation is an obvious outlier. A log or Box--Cox
transform of the outcome is therefore **not warranted**; growth enters the models
on its natural scale, which also keeps the reproduction faithful to FLS.

## Univariate distributions of the predictors

For the continuous predictors, a faceted grid of histograms shows the spread of
scales and shapes at a glance --- from tightly bounded [0,1] shares to
heavy-tailed area and income variables. We show a representative dozen (the
highest-salience determinants plus a couple of skewed geographic ones):

```{r eda-num-grid, fig.height=4.6, fig.cap="Univariate distributions of twelve representative continuous predictors. Note the mix of near-symmetric variables (GDP60, LifeExp) and strongly skewed / bounded ones (Area, EquipInv, PrScEnroll), which is why penalized methods standardize before fitting."}
library(dplyr)
show_num <- c("GDP60","LifeExp","EquipInv","NequipInv","PrScEnroll",
              "HighEnroll","RuleofLaw","YrsOpen","Age","Abslat",
              "Popg","Area")
long <- do.call(rbind, lapply(show_num, function(v)
          data.frame(variable = v, value = d[[v]])))
ggplot(long, aes(value)) +
  geom_histogram(bins = 15, fill = "steelblue", colour = "white") +
  facet_wrap(~ variable, scales = "free", ncol = 4) +
  labs(x = NULL, y = "count") +
  theme_minimal(base_size = 9)
```

A **full summary-statistics table for ALL `r length(num_vars)` continuous
predictors** (rounded; `skew` from e1071) --- so nothing is hidden even though we
only plot a subset:

```{r eda-num-table}
num_summary <- data.frame(
  variable = num_vars,
  n        = sapply(num_vars, function(v) sum(!is.na(d[[v]]))),
  n_miss   = sapply(num_vars, function(v) sum(is.na(d[[v]]))),
  mean     = round(sapply(num_vars, function(v) mean(d[[v]])), 3),
  sd       = round(sapply(num_vars, function(v) sd(d[[v]])), 3),
  min      = round(sapply(num_vars, function(v) min(d[[v]])), 3),
  median   = round(sapply(num_vars, function(v) median(d[[v]])), 3),
  max      = round(sapply(num_vars, function(v) max(d[[v]])), 3),
  skew     = round(sapply(num_vars, function(v) e1071::skewness(d[[v]])), 2),
  row.names = NULL)
knitr::kable(num_summary, row.names = FALSE,
  caption = "Summary statistics for all 34 continuous predictors (n, missing, mean, sd, min, median, max, skew). Scales differ by orders of magnitude, foreshadowing the need to standardize.")
```

The table makes the scaling problem concrete: `Area` and `RFEXDist` run into the
thousands while enrolment and religion shares live in [0,1]. Left unstandardized,
a single penalty would punish the large-scale variables far more than the shares
--- which is exactly why every penalized fit below standardizes first.

For the **`r length(bin_vars)` binary indicators**, base rates (the share of
countries coded 1) are far more informative than a mean/sd, and a bar chart makes
the class balance immediate:

```{r eda-binary}
bin_tab <- data.frame(
  dummy      = bin_vars,
  n_ones     = sapply(bin_vars, function(v) sum(d[[v]] == 1)),
  proportion = round(sapply(bin_vars, function(v) mean(d[[v]])), 3))
bin_tab <- bin_tab[order(-bin_tab$proportion), ]
knitr::kable(bin_tab, row.names = FALSE,
  caption = "Binary (0/1) indicators, with the share of countries coded 1.")
```

```{r eda-binary-bar, fig.height=2.8, fig.cap="Base rates of the seven 0/1 indicator predictors. Several are quite rare (e.g. outward-orientation), so their coefficients rest on few countries and are prime candidates for shrinkage."}
ggplot(bin_tab, aes(x = reorder(dummy, proportion), y = proportion)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(x = NULL, y = "share of countries = 1") +
  theme_minimal(base_size = 10)
```

Some dummies are rare --- only a handful of countries carry a 1 --- so any
coefficient estimated for them leans on very little data, another reason the
methods below prefer to shrink rather than trust raw OLS.

## Missing-data analysis

```{r eda-missing}
miss_by_var <- colSums(is.na(d))
data.frame(
  metric = c("total missing cells", "variables with any NA",
             "complete rows", "rows total"),
  value  = c(sum(is.na(d)), sum(miss_by_var > 0),
             sum(complete.cases(d)), nrow(d)))
```

The dataset is **fully complete**: zero missing cells, all
`r sum(complete.cases(d))` countries usable, no co-missingness pattern to
diagnose and no imputation or listwise deletion required. This is by
construction --- the 72 countries are precisely those with a complete record on
all 41 regressors, so *completeness is baked into how the sample was drawn*
(the flip side of the selection caveat noted above). Every method below therefore
sees the identical 72 countries, which makes their out-of-sample comparison a
clean apples-to-apples race.

## Bivariate relationships with growth

We now line up growth against each of the most salient predictors. First the two
headline **continuous** determinants, each with an OLS smoother and its raw
correlation with growth:

```{r eda-bivar, fig.height=3.2, fig.cap="Growth vs the two headline continuous determinants, with least-squares smoother and Pearson r. Left: conditional convergence --- higher initial income is associated with slower subsequent growth. Right: higher equipment investment is associated with faster growth."}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(d$GDP60, d$y, pch = 19, col = "steelblue",
     xlab = "GDP60 (log initial income, 1960)",
     ylab = "annual growth 1960-1992")
abline(lm(y ~ GDP60, data = d), col = "firebrick", lwd = 2)
title(sprintf("Convergence (r = %.2f)", cor(d$GDP60, d$y)),
      line = 0.6, cex.main = 0.95)
plot(d$EquipInv, d$y, pch = 19, col = "steelblue",
     xlab = "EquipInv (equipment investment share)",
     ylab = "annual growth 1960-1992")
abline(lm(y ~ EquipInv, data = d), col = "firebrick", lwd = 2)
title(sprintf("Investment (r = %.2f)", cor(d$EquipInv, d$y)),
      line = 0.6, cex.main = 0.95)
```

Then the strongest **categorical** contrast --- the Sub-Saharan growth deficit
--- as a boxplot with a difference-in-means:

```{r eda-bivar-sub, fig.height=3.0, fig.cap="Growth by region. Sub-Saharan African economies grew markedly slower than the rest of the sample over 1960--1992."}
boxplot(y ~ SubSahara, data = d, col = c("grey85","firebrick"),
        names = c("Rest of world", "Sub-Saharan Africa"),
        ylab = "annual growth 1960-1992", xlab = "")
gap <- mean(d$y[d$SubSahara == 0]) - mean(d$y[d$SubSahara == 1])
title(sprintf("Mean gap = %.1f pct pts / yr", 100 * gap),
      line = 0.5, cex.main = 0.95)
```

To summarize *every* predictor's marginal association compactly, we rank all 41
by the absolute correlation of the outcome with the predictor (for the dummies
this equals a standardized difference in means), and show the strongest links:

```{r eda-marginal}
cors <- sapply(colnames(d)[-1], function(v) cor(d[[v]], d$y))
marg <- data.frame(variable = names(cors),
                   cor_with_y = round(cors, 3),
                   abs_cor = round(abs(cors), 3),
                   row.names = NULL)
marg <- marg[order(-marg$abs_cor), ]
knitr::kable(head(marg, 12), row.names = FALSE,
  caption = "Twelve predictors most strongly correlated (in absolute value) with growth. Sign shown; these are the raw marginal associations, before adjusting for the heavy collinearity documented next.")
```

The leaders --- initial income (negative: convergence), equipment investment,
life expectancy, primary schooling, the Confucian share and the Sub-Saharan
dummy --- are exactly the determinants FLS's BMA later ranks at the top. But
these are *marginal* associations; the next subsection shows why we cannot simply
read them as effects.

## Multivariate structure and collinearity

With many predictors describing overlapping facets of "development", collinearity
is the central statistical hazard: if income, health, schooling and investment
all rise together, no single regression can cleanly attribute growth to one of
them. A correlation heatmap over the salient continuous predictors makes the
entanglement visible (a 41x41 heatmap would be an unreadable postage stamp):

```{r eda-corr, fig.height=4.4, fig.cap="Correlation heatmap among the salient continuous predictors. Initial income, life expectancy, schooling and equipment investment form a strongly positively correlated block --- the collinearity that makes single-regression inference unstable and motivates shrinkage/selection."}
key_num <- c("GDP60","LifeExp","EquipInv","NequipInv","PrScEnroll",
             "HighEnroll","RuleofLaw","YrsOpen","Age","Abslat")
C  <- cor(d[ , key_num])
cm <- as.data.frame(as.table(C)); names(cm) <- c("v1","v2","r")
ggplot(cm, aes(v1, v2, fill = r)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 2.7) +
  scale_fill_gradient2(low = "firebrick", mid = "white",
                       high = "steelblue", midpoint = 0,
                       limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "r") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

To quantify what "strong" means, list every predictor pair (across ALL 41
continuous+binary regressors) with $|r| > 0.7$:

```{r eda-collinear}
Cfull <- cor(d[ , -1])
Cfull[upper.tri(Cfull, diag = TRUE)] <- NA
hi <- which(abs(Cfull) > 0.7, arr.ind = TRUE)
if (nrow(hi) == 0) {
  cat("No predictor pair exceeds |r| = 0.7.\n")
} else {
  hipairs <- data.frame(
    var1 = rownames(Cfull)[hi[, 1]],
    var2 = colnames(Cfull)[hi[, 2]],
    r    = round(Cfull[hi], 3))
  hipairs <- hipairs[order(-abs(hipairs$r)), ]
  knitr::kable(hipairs, row.names = FALSE,
    caption = "All predictor pairs with |r| > 0.7 across the full 41-predictor set.")
}
```

Life expectancy, initial income, schooling and equipment investment move together
(pairwise $r$ up to $\approx 0.8$): richer, healthier, better-schooled economies
cluster. Ordinary least squares cannot cleanly attribute growth to any one of
them --- precisely the identification problem FLS confront with BMA and we revisit
with lasso and best-subset. As a final compact view of the joint structure among
the four headline determinants, a `GGally::ggpairs` panel shows their pairwise
scatters, marginal densities and correlations together:

```{r eda-ggpairs, fig.height=4.6, fig.cap="Pairwise structure of the four headline determinants (GGally::ggpairs): scatterplots below the diagonal, densities on it, correlations above. The uniformly positive off-diagonal correlations are the collinearity that shrinkage and selection are built to handle."}
suppressMessages(library(GGally))
ggpairs(d[ , c("GDP60","LifeExp","PrScEnroll","EquipInv")],
        lower = list(continuous = wrap("points", alpha = 0.5,
                                       size = 0.7, colour = "steelblue"))) +
  theme_minimal(base_size = 8)
```

**What the descriptives foreshadow.** Growth is a well-behaved, near-symmetric
continuous outcome with modest spread ($\approx 1.8$ percentage points), no
missingness, and no transformation needed. Its strongest raw associations are the
classic convergence signal (negative slope on initial income `GDP60`), a positive
equipment-investment gradient, and a large Sub-Saharan growth deficit --- exactly
the determinants FLS's BMA ranks at the top. But those salient predictors are
themselves tightly intercorrelated (several pairs with $|r|>0.7$), and their
scales span orders of magnitude. That combination --- 41 correlated predictors on
wildly different scales over just 72 countries --- is the whole reason the
reproduction below turns to model averaging, why standardization is mandatory, and
why the regularization and best-subset extensions earn their keep out-of-sample.

# Reproduction gate 1 --- the data are the published data

FLS print descriptive statistics for every variable. Matching them to printed
precision proves we hold the exact published dataset.

```{r prov}
pubstats <- data.frame(
  var      = c("y","PrScEnroll","LifeExp","GDP60","Muslim","RuleofLaw","EquipInv"),
  pub_mean = c(0.02073, 0.7953, 56.58, 7.492, 0.1475, 0.5509, 0.04429),
  pub_std  = c(0.01813, 0.2440, 11.37, 0.8785, 0.2925, 0.3329, 0.03426))
pubstats$our_mean <- round(sapply(pubstats$var, function(v) mean(d[[v]])), 5)
# FLS report the population s.d. (divide by n); our sample sd divides by n-1
pubstats$our_std  <- round(sapply(pubstats$var,
                       function(v) sd(d[[v]]) * sqrt((n-1)/n)), 4)
knitr::kable(pubstats, caption = "Data provenance: our data vs FLS (2001) printed descriptive statistics.")
```

Means and standard deviations match the published figures to printed precision.

# Reproduce the published main result: BMA posterior inclusion probabilities

FLS's Table I ranks the 41 regressors by posterior inclusion probability under
their **benchmark $g$-prior** ($g = 1/\max(n,k^2) = 1/k^2$ here) and a uniform
prior over model size. We reproduce it with `BMS::bms()` using an MC$^3$ sampler
(full enumeration of $2^{41}$ models is infeasible).

```{r bma, cache=TRUE, results='hide'}
mfls <- bms(d, burn = 1e5, iter = 1e6, g = "BRIC", mprior = "uniform",
            nmodel = 10000, mcmc = "bd", user.int = FALSE)
```

```{r bma-corr}
cat("Sampler convergence (corr of analytic vs MCMC model probs):",
    round(mfls$info$corr.pmp, 3), "\n")
pip_bma <- coef(mfls, exact = TRUE)[, "PIP"]   # posterior inclusion probs
```

The convergence correlation ($\approx 0.99$) confirms the chain has identified
the high-probability region. Now compare our reproduced PIPs to the numbers
printed in FLS Table I (p. 9):

```{r bma-match}
pub <- data.frame(
  var = c("GDP60","Confucian","LifeExp","EquipInv","SubSahara","Muslim",
          "RuleofLaw","YrsOpen","EcoOrg","Protestants","NequipInv","Mining"),
  published_PIP = c(1.000,0.995,0.946,0.942,0.757,0.656,
                    0.516,0.502,0.471,0.461,0.441,0.431))
pub$reproduced_PIP <- round(pip_bma[pub$var], 3)
pub$abs_diff <- round(abs(pub$published_PIP - pub$reproduced_PIP), 3)
knitr::kable(pub, caption = "Reproduction of FLS (2001) Table I: the 12 highest posterior-inclusion-probability growth determinants. Published vs reproduced.")
cat("Top-12 mean absolute PIP difference:",
    round(mean(pub$abs_diff), 4), "\n")
```

The reproduced posterior inclusion probabilities match the published values to
within Monte-Carlo error (mean absolute difference $\approx 0.01$). **The gate is
passed**: the same variables emerge as important, in the same order, with the
same magnitudes. FLS's substantive story --- initial GDP (conditional
convergence), equipment investment, life expectancy, a Confucian dummy, and
Sub-Saharan Africa dominate; most of the other 33 regressors are dispensable ---
is reproduced.

# The OLS that motivates everything

Before regularizing, look at why you *cannot* just run one big OLS on all 41
predictors with $n=72$.

```{r ols}
ols <- lm(y ~ ., data = d)
s <- summary(ols)
data.frame(R2 = round(s$r.squared,3), adjR2 = round(s$adj.r.squared,3),
           residual_df = ols$df.residual, n = n, params = length(coef(ols)))
```

The full model spends 42 parameters on 72 observations, leaving only 30
residual degrees of freedom. In-sample $R^2 = 0.96$ looks spectacular --- and is
almost entirely overfitting, as the out-of-sample comparison below will show.
This is the canonical setting where **regularization and selection** are not
optional niceties but the difference between a useful model and a useless one.

# Standardization --- why and how

Penalized methods shrink coefficients toward zero, so the *scale* of each
predictor matters: a coefficient on `LifeExp` (range roughly 40--75 years) and one
on a 0/1 religion dummy are penalized on wildly different footings unless we put
all predictors on a common scale. `glmnet` standardizes internally
(`standardize = TRUE`) --- fits on $z$-scored predictors, then reports
coefficients back on the original scale. Here is the effect explicitly:

```{r standardize}
Xs <- scale(X)   # mean 0, sd 1
round(rbind(`orig sd` = apply(X, 2, sd)[1:6],
            `scaled sd` = apply(Xs, 2, sd)[1:6]), 3)
```

# Ridge, lasso and elastic net via `glmnet`

`glmnet` fits the elastic-net family, minimizing
$\frac{1}{2n}\lVert y - X\beta\rVert^2 + \lambda\big[\alpha\lVert\beta\rVert_1 +
\tfrac{1-\alpha}{2}\lVert\beta\rVert_2^2\big]$:
$\alpha=1$ is **lasso** (sparse; sets coefficients exactly to zero),
$\alpha=0$ is **ridge** (shrinks but keeps all), and intermediate $\alpha$ is
**elastic net**.

```{r glmnet-fits}
library(glmnet)
fit_ridge <- glmnet(X, y, alpha = 0,   standardize = TRUE)
fit_lasso <- glmnet(X, y, alpha = 1,   standardize = TRUE)
fit_enet  <- glmnet(X, y, alpha = 0.5, standardize = TRUE)
```

## Coefficient paths

As $\lambda$ falls (right to left) predictors enter the model. Lasso admits them
one at a time and keeps most at exactly zero; ridge moves them all together.

```{r paths, fig.height=3.6, fig.cap="Coefficient paths. Left: lasso ($\\alpha=1$) --- sparse, coefficients enter one at a time. Right: ridge ($\\alpha=0$) --- all coefficients shrink together, none exactly zero."}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot(fit_lasso, xvar = "lambda"); title("Lasso path", line = 2.2)
plot(fit_ridge, xvar = "lambda"); title("Ridge path", line = 2.2)
```

## Cross-validation to choose $\lambda$ (`lambda.min` and `lambda.1se`)

```{r cvlambda, fig.height=3.6, fig.cap="10-fold CV curves. Dotted lines mark lambda.min (loss-minimizing) and lambda.1se (sparsest model within 1 s.e.)."}
set.seed(1)
cv_lasso <- cv.glmnet(X, y, alpha = 1,   nfolds = 10, standardize = TRUE)
cv_ridge <- cv.glmnet(X, y, alpha = 0,   nfolds = 10, standardize = TRUE)
cv_enet  <- cv.glmnet(X, y, alpha = 0.5, nfolds = 10, standardize = TRUE)
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot(cv_lasso); title("Lasso CV", line = 2.4)
plot(cv_enet);  title("Elastic-net CV", line = 2.4)

sel <- function(cvfit, s) {
  b <- coef(cvfit, s = s); nm <- rownames(b)[which(b != 0)]
  setdiff(nm, "(Intercept)")
}
lasso_min <- sel(cv_lasso, "lambda.min")
lasso_1se <- sel(cv_lasso, "lambda.1se")
data.frame(rule = c("lambda.min","lambda.1se"),
           n_kept = c(length(lasso_min), length(lasso_1se)),
           lambda = signif(c(cv_lasso$lambda.min, cv_lasso$lambda.1se), 3))
```

Interpreting shrinkage: at `lambda.min` the lasso keeps `r length(lasso_min)`
predictors, while the more conservative `lambda.1se` rule --- the sparsest model
whose CV error is within one standard error of the minimum --- keeps only
`r length(lasso_1se)`:

```{r lasso1se-list}
lasso_1se
```

## Feature importance from standardized coefficients

With predictors standardized, the magnitude of each lasso coefficient is directly
comparable across variables --- an importance ranking. (glmnet reports on the
original scale, so we multiply by each predictor's s.d. to standardize.)

```{r importance, fig.height=3.8, fig.cap="Standardized lasso (lambda.1se) coefficients: signed importance of the retained growth determinants."}
sdx  <- apply(X, 2, sd)
b1se <- as.numeric(coef(cv_lasso, s = "lambda.1se"))[-1]
imp  <- data.frame(var = colnames(X), std_coef = b1se * sdx)
imp  <- imp[imp$std_coef != 0, ]
imp  <- imp[order(imp$std_coef), ]
par(mar = c(4, 7, 2, 1))
barplot(imp$std_coef, names.arg = imp$var, horiz = TRUE, las = 1,
        col = ifelse(imp$std_coef > 0, "steelblue", "firebrick"),
        xlab = "standardized coefficient (effect on annual growth)")
abline(v = 0)
```

# Best-subset selection with `abess`, contrasted with lasso

Lasso selects by convex $\ell_1$ shrinkage; **best-subset** selection instead
searches directly for the size-$s$ subset of predictors giving the best fit.
Exact best-subset is NP-hard, but the **abess** algorithm (adaptive best-subset)
solves it efficiently and picks the support size by an information criterion.

```{r abess}
library(abess)
keepX   <- apply(X, 2, function(z) stats::sd(z) > 0)   # guard against any constant column
ab      <- abess(X[, keepX, drop = FALSE], y, family = "gaussian")   # support size chosen by GIC
ab_best <- extract(ab)
abess_support <- ab_best$support.vars
data.frame(method = "abess (best subset)",
           support_size = length(abess_support))
abess_support
```

**How the three selectors compare.** Lay each method's kept set beside the BMA
"ground truth" (variables with reproduced PIP $>0.4$):

```{r selection-compare}
bma_important <- names(sort(pip_bma, decreasing = TRUE))[pip_bma[
                   order(pip_bma, decreasing = TRUE)] > 0.4]
allv <- colnames(X)
tab <- data.frame(
  variable   = allv,
  BMA_PIP    = round(pip_bma[allv], 2),
  BMA_top    = ifelse(allv %in% bma_important, "*", ""),
  lasso_1se  = ifelse(allv %in% lasso_1se, "*", ""),
  abess      = ifelse(allv %in% abess_support, "*", ""))
tab <- tab[order(-tab$BMA_PIP), ]
knitr::kable(head(tab, 16), row.names = FALSE,
  caption = "Which predictors each method keeps (top 16 by BMA PIP). A '*' marks inclusion. abess best-subset closely tracks the BMA important set; lasso-1se is sparser and drops some high-PIP variables it cannot separate from correlated rivals.")
```

The contrast is the lesson: **abess best-subset recovers almost exactly the FLS
BMA important set** --- initial GDP, Confucian, life expectancy, equipment
investment, Sub-Saharan Africa, Muslim, rule of law --- whereas **lasso at
`lambda.1se` is sparser** and, because it shrinks correlated predictors together,
substitutes some proxies and drops a few high-PIP variables. Neither is "wrong":
lasso optimizes prediction under shrinkage; best-subset optimizes fit at a fixed
sparsity; BMA hedges over all models.

# Hyperparameter tuning: $\lambda$ and $\alpha$ together

Elastic net has two knobs. We already tune $\lambda$ by CV; now tune $\alpha$
over a grid and read off the CV-optimal combination.

```{r tune-alpha, fig.height=3.8, fig.cap="Tuning the elastic-net mixing parameter alpha. Each point is the minimum 10-fold CV MSE at that alpha; lower is better."}
alphas <- seq(0, 1, by = 0.1)
set.seed(7)
foldid <- sample(rep(1:10, length.out = n))    # common folds => fair alpha comparison
cvm <- sapply(alphas, function(a) {
  min(cv.glmnet(X, y, alpha = a, foldid = foldid, standardize = TRUE)$cvm)
})
best_alpha <- alphas[which.min(cvm)]
plot(alphas, cvm, type = "b", pch = 19, col = "steelblue",
     xlab = expression(alpha~"(0 = ridge, 1 = lasso)"),
     ylab = "min 10-fold CV MSE")
abline(v = best_alpha, lty = 2)
cat("CV-optimal alpha:", best_alpha, "\n")
```

# Out-of-sample comparison: does regularization actually help?

The honest test. We repeat a 70/30 train--test split `r 100` times; on each split
we fit every method on the training rows (choosing $\lambda$ by inner CV) and
score RMSE on the held-out rows. We include the full OLS and a mean-only baseline.

```{r oos, cache=TRUE}
rmse <- function(a, b) sqrt(mean((a - b)^2))
R <- 100
methods <- c("OLS (full)","Ridge","Lasso","Elastic net","abess","Mean baseline")
res <- matrix(NA, R, length(methods), dimnames = list(NULL, methods))
set.seed(2024)
for (r in 1:R) {
  te <- sample(n, size = round(0.30 * n)); tr <- setdiff(1:n, te)
  Xtr <- X[tr,]; Xte <- X[te,]; ytr <- y[tr]; yte <- y[te]
  ols_r <- lm(y ~ ., data = data.frame(y = ytr, Xtr))
  res[r,"OLS (full)"] <- rmse(yte, predict(ols_r, newdata = data.frame(Xte)))
  for (a in c(0, 1, 0.5)) {
    cv <- cv.glmnet(Xtr, ytr, alpha = a, nfolds = 5, standardize = TRUE)
    pr <- as.numeric(predict(cv, newx = Xte, s = "lambda.min"))
    nm <- c("Ridge","Lasso","Elastic net")[match(a, c(0,1,0.5))]
    res[r, nm] <- rmse(yte, pr)
  }
  keep <- apply(Xtr, 2, function(z) stats::sd(z) > 0)  # a rare 0/1 dummy can be constant in a fold
  ab_r <- abess(Xtr[, keep, drop = FALSE], ytr, family = "gaussian")
  res[r,"abess"] <- rmse(yte, as.numeric(predict(ab_r, newx = Xte[, keep, drop = FALSE])))
  res[r,"Mean baseline"] <- rmse(yte, mean(ytr))
}
oos <- data.frame(method = methods,
                  mean_RMSE = round(colMeans(res), 5),
                  se_RMSE   = round(apply(res, 2, sd) / sqrt(R), 5))
oos <- oos[order(oos$mean_RMSE), ]
knitr::kable(oos, row.names = FALSE,
  caption = "Out-of-sample RMSE over 100 repeated 70/30 splits (smaller is better). y has s.d. 0.018.")
```

```{r oos-plot, fig.height=3.6, fig.cap="Distribution of held-out RMSE across 100 splits. The full OLS is off the useful scale --- it overfits so badly it loses to predicting the mean."}
ord <- oos$method
boxplot(res[, ord], las = 2, col = "grey85", ylab = "held-out RMSE",
        main = "Out-of-sample prediction error by method")
abline(h = sd(y), lty = 3)
```

**Reading the result.** The full OLS is *worse than predicting the training mean*
--- its in-sample $R^2=0.96$ was a mirage. Lasso and elastic net predict best,
with ridge and abess close behind; every regularized/selected model beats both
OLS and the naive baseline. With 41 correlated predictors and 72 countries,
**shrinkage and selection are what make the growth regression usable
out-of-sample.**

# Takeaway

Fernández, Ley & Steel (2001) confronted 41 correlated growth determinants and
72 countries with Bayesian model averaging; we reproduced their headline
posterior-inclusion ranking to Monte-Carlo precision (initial GDP, equipment
investment, life expectancy, Confucian share, Sub-Saharan Africa, Muslim, rule of
law). The modern regularization toolkit tells the same substantive story from a
predictive angle: the naive full OLS overfits catastrophically (worse than a mean
baseline out-of-sample), while lasso and elastic net predict best, and
**best-subset (`abess`) selection independently recovers almost exactly the BMA
important set**. The teaching point is that lasso's $\ell_1$ shrinkage, best-subset
search, and Bayesian model averaging are three routes to the same conclusion ---
that only a handful of these 41 determinants carry robust, transferable
signal --- and that with $p$ large relative to $n$, choosing among them is not a
refinement of OLS but a prerequisite for it.
