---
title: "Reproduce-then-Extend — Civil-War Onset (Fearon & Laitin 2003) (Day 1)"
author: "Intro to Machine Learning · Bruce A. Desmarais"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: true
geometry: margin=1in
fontsize: 10pt
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
  - \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines,breakanywhere}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 3.6, fig.align = "center")
options(width = 82)
library(PRROC)   # ROC and precision-recall curves
```

> **Published study.** Fearon, J. D. & Laitin, D. D. (2003). "Ethnicity, Insurgency, and Civil War."
> *American Political Science Review* 97(1):75–90. A **logistic regression** of civil-war onset on 11
> country-year predictors (Sambanis 2006 imputed data; Harvard Dataverse doi:10.7910/DVN/KRKWK8). Onset is a
> **rare event** (~1.6%); this example illustrates **out-of-sample classification** and the right evaluation
> metric for imbalanced data.

# Background

Fearon and Laitin ask **what conditions make civil war likely** worldwide from 1945 to 1999. Their central aim is to adjudicate between two explanations: civil war as a product of ethnic and religious **grievance**, versus civil war as a product of conditions that favor **insurgency** -- weak states, poverty, rough terrain, and populations governments cannot control. They use **logistic regression** of civil-war onset on structural covariates to test which set of factors predicts onset. The **primary conclusion** is that insurgency-favoring conditions (low GDP per capita, large population, mountainous terrain, political instability, new states) strongly predict civil war, while ethnic and religious fractionalization do **not** -- a direct challenge to the ethnic-grievance thesis.

# Data and codebook

**Unit of analysis:** country-year; n = 7,140 (116 civil-war onsets, ~1.6%).

| Variable | Definition |
|---|---|
| `warstds` | civil-war onset: 1 in the first year of a new civil war **(outcome)** |
| `warhist` | prior civil-war history in the country |
| `ln_gdpen` | log GDP per capita (lagged one year) |
| `lpopns` | log national population |
| `lmtnest` | log % of the country that is mountainous terrain |
| `ncontig` | 1 if the state has noncontiguous territory |
| `oil` | 1 if fuel exports exceed one-third of total exports (oil exporter) |
| `nwstate` | 1 in the first two years of a newly independent state |
| `inst3` | political instability: a >=3-point Polity change in the prior 3 years |
| `pol4` | Polity score (-10 autocracy to +10 democracy) |
| `ef` | ethnolinguistic fractionalization |
| `relfrac` | religious fractionalization |

# Descriptive results

```{r data}
d <- read.csv("https://raw.githubusercontent.com/bdesmarais/intro-ml-statistical-horizons-4day/main/tutorials/day1_intro_and_evaluation/02_fearon_laitin/data/SambnisImp.csv")
fl <- c("warhist", "ln_gdpen", "lpopns", "lmtnest", "ncontig", "oil", "nwstate",
        "inst3", "pol4", "ef", "relfrac")
d <- d[complete.cases(d[, c("warstds", fl)]), c("warstds", fl)]

round(t(sapply(d, function(x) c(mean = mean(x), sd = sd(x), min = min(x), max = max(x)))), 2)
```

```{r desc-plot, fig.height=3, fig.cap="Class balance of the rare onset outcome (left) and the covariates that differ most between onset and non-onset years (right)."}
par(mfrow = c(1, 2))
par(mar = c(4, 4, 2.5, 1))
barplot(table(d$warstds), names.arg = c("no onset", "onset"),
        col = c("grey75", "#1f78b4"), main = "Outcome: warstds", ylab = "count")

# standardized mean difference (onset - non-onset) for each predictor
smd <- sapply(d[fl], function(x) (mean(x[d$warstds == 1]) - mean(x[d$warstds == 0])) / sd(x))
smd <- tail(smd[order(abs(smd))], 8)
par(mar = c(4, 6, 2.5, 1))
barplot(smd, horiz = TRUE, las = 1, cex.names = 0.85,
        col = ifelse(smd > 0, "#1f78b4", "#e31a1c"), main = "Onset vs non-onset",
        xlab = "std. mean difference")
abline(v = 0, col = "grey50")
```

# Reproduce the published regression

```{r reproduce}
fl_logit <- glm(warstds ~ warhist + ln_gdpen + lpopns + lmtnest + ncontig + oil +
                  nwstate + inst3 + pol4 + ef + relfrac, data = d, family = binomial)
summary(fl_logit)
```

**Confirmation against the published study.** The coefficients carry the signs and significance reported in
Fearon & Laitin's Table 1: civil war is more likely in **poorer** countries (`ln_gdpen` negative), **more
populous** ones (`lpopns` positive), **more mountainous** terrain (`lmtnest` positive), **new states**
(`nwstate` positive) and amid **political instability** (`inst3` positive), while ethnic/religious
fractionalization are not significant — their central claim that *insurgency conditions, not ethnic grievance,
drive civil war*.

# Out-of-sample prediction with the regression

We fit on a random 70% and score the held-out 30%. With a rare outcome, accuracy is useless (predicting "no
war" everywhere scores 98%); the honest evaluation is the **ROC** and especially the **precision-recall** curve.

```{r oos}
set.seed(2026)
train <- sample(nrow(d), floor(0.7 * nrow(d)))
test  <- setdiff(seq_len(nrow(d)), train)

fit  <- glm(warstds ~ ., data = d[train, ], family = binomial)
pred <- predict(fit, d[test, ], type = "response")   # predicted onset probability
y    <- d$warstds[test]

roc <- roc.curve(scores.class0 = pred[y == 1], scores.class1 = pred[y == 0], curve = TRUE)
pr  <- pr.curve(scores.class0 = pred[y == 1], scores.class1 = pred[y == 0], curve = TRUE)
cat(sprintf("Held-out test: %d country-years, onset base rate %.1f%%\n", length(test), 100 * mean(y)))
cat(sprintf("  Out-of-sample ROC-AUC = %.3f\n  Out-of-sample PR-AUC  = %.3f  (a random model scores %.3f)\n",
            roc$auc, pr$auc.integral, mean(y)))
```

```{r oos-plot, echo=FALSE, fig.height=3.4, fig.cap="Out-of-sample ROC curve (left) and precision-recall curve (right) for the held-out logit predictions."}
par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.5, 1))
plot(roc$curve[, 1], roc$curve[, 2], type = "l", lwd = 2, col = "#1f78b4",
     xlab = "false positive rate", ylab = "true positive rate", main = "ROC (held-out)")
abline(0, 1, lty = 2, col = "grey60")
plot(pr$curve[, 1], pr$curve[, 2], type = "l", lwd = 2, col = "#33a02c",
     xlab = "recall", ylab = "precision", main = "Precision-Recall (held-out)")
abline(h = mean(y), lty = 2, col = "grey60")
```

The ROC-AUC looks strong (~0.8), but the **PR-AUC** is far more modest — the honest picture when only 1.6% of
cases are onsets. This is the key evaluation lesson: *on imbalanced data, ROC-AUC flatters a model and
precision-recall tells the truth.*

# Takeaway

This example illustrates **out-of-sample classification with a published regression** and the
**evaluation-metric choice** that matters most for rare events. The Fearon–Laitin logit predicts onset above
chance out of sample, but reporting precision–recall (not accuracy or ROC alone) is what keeps the assessment
honest. (A later example — Gibler–Braithwaite — extends this rare-event setting with tree-based learners.)


# Recommended exercises

1. Replace the single split with stratified 5-fold CV and report mean AUC-PR (the outcome is ~1.6% positive, so use AUC-PR, not accuracy).
2. Fit an L1-penalized logistic regression and compare its AUC-PR and retained variables to the full logit.
3. Plot the calibration (reliability) curve of the predicted probabilities and comment on over/under-confidence.
