---
title: "Predicting Secondary-School Final Grades: Reproduction and Day-2 Regression-Extensions Lab"
subtitle: "Reproducing El Aissaoui et al. (2020), *A Multiple Linear Regression-Based Approach to Predict Student Performance*"
author: "Intro to Machine Learning --- SH Day 2: Regression Extensions"
date: "`r format(Sys.Date())`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
header-includes:
  - \setlength{\emergencystretch}{3em}
  - \usepackage{fvextra}
  - \AtBeginDocument{\DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}}
  - \AtBeginDocument{\DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines}}
  - \fvset{breaklines=true}
  - \usepackage{booktabs}
  - \usepackage{float}
---

```{r setup, include=FALSE}
options(width = 80)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 6, fig.height = 4,
                      fig.pos = "H")
set.seed(20240501)
options(digits = 5)
```

# The study and the reproduction gate

**Citation.** El Aissaoui, O., El Alami El Madani, Y., Oughdir, L., Dakkak, A., & El Allioui, Y. (2020).
*A Multiple Linear Regression-Based Approach to Predict Student Performance.* In M. Ezziyyani (Ed.),
Advanced Intelligent Systems for Sustainable Development (AI2SD 2019), AISC 1102, pp. 9--23. Springer.
DOI: \texttt{10.1007/978-3-030-36653-7\_2}.

**Data source.** Cortez, P. & Silva, A. (2008), *Student Performance* dataset, UCI Machine Learning
Repository, DOI \texttt{10.24432/C50G6X} --- file \texttt{student-mat.csv} (Mathematics course, 395 students,
33 columns). The dataset is fully public and ships with the course as a local copy.

**The published model.** The paper's task is a plain OLS (`lm`) predicting the final Mathematics grade
`G3` (numeric, 0--20). Of the seven candidate models the authors report (Table 2), their best
(**Model 6 / MARS-selected**, lowest RMSE and highest $R^2$ in their Table 3) uses five predictors:

$$
\widehat{G3} = 17.7935 + 0.8627\,\text{Medu} - 0.5603\,\text{goout}
- 1.2458\,\text{romanticyes} - 0.4423\,\text{age} - 1.6562\,\text{schoolsupyes}.
$$

Here `Medu` = mother's education (0--4), `goout` = going out with friends (1--5),
`romantic` = in a romantic relationship (yes/no), `age` (15--22), `schoolsup` = extra
educational support (yes/no).

The **reproduction gate**: our reproduced coefficients must match the six printed numbers above.
A subtlety --- the paper preprocesses numeric predictors by replacing boxplot outliers with `NA`
and then mean-imputing them (their Section 3.2). We replicate that exactly.

```{r load-data}
raw <- read.csv(paste0(
  "https://raw.githubusercontent.com/bdesmarais/intro",
  "-ml-statistical-horizons-4day/main/tutorials/day2_",
  "regression_extensions/elaissaoui_student_grades/da",
  "ta/student-mat.csv"), sep = ";", stringsAsFactors = FALSE)
cat("Rows x cols:", nrow(raw), "x", ncol(raw), "\n")
cat("Outcome G3: range", paste(range(raw$G3), collapse = "-"),
    "| mean", round(mean(raw$G3), 3), "| sd", round(sd(raw$G3), 3), "\n")

# Paper's preprocessing: boxplot-outlier -> NA -> mean imputation on numeric predictors.
impute_outliers <- function(x) {
  out <- boxplot.stats(x)$out          # points beyond 1.5*IQR
  x[x %in% out] <- NA
  x[is.na(x)]  <- mean(x, na.rm = TRUE) # mean imputation
  x
}
dat <- raw
dat$romantic  <- factor(dat$romantic,  levels = c("no", "yes"))
dat$schoolsup <- factor(dat$schoolsup, levels = c("no", "yes"))
for (v in c("Medu", "goout", "age")) dat[[v]] <- impute_outliers(dat[[v]])
```

# Descriptive analysis

Before fitting anything, we meet the data properly. A careful exploratory pass is not decoration ---
it tells us the shape of the outcome (and whether it needs transforming), which predictors carry
signal, whether any variables are collinear enough to destabilize an OLS, and where the missing data
are. Everything the modeling section does downstream is easier to interpret once we have looked.

## Dataset and provenance

The **unit of analysis is the individual student**. There are `r nrow(dat)` students enrolled in the
Mathematics course, each described by `r ncol(raw)` variables. The data come from a **single-wave,
cross-sectional survey** administered in two public secondary schools in the Alentejo region of
Portugal during the **2005--2006 academic year** (Cortez & Silva 2008). Records were built by
combining school reports (grades, absences) with a paper questionnaire on demographics, family
background, and social/study habits; the two sources were merged per student. There is **no sampling
weight and no time dimension** --- it is a convenience census of the two schools' math cohorts, so
inferences are to *these students*, not to Portugal as a whole.

The 33 columns fall into four blocks: (i) **demographics** (`sex`, `age`, `address`, ...),
(ii) **family background** (`Medu`, `Fedu`, `Mjob`, `Fjob`, `Pstatus`, ...), (iii) **study / social
behavior** (`studytime`, `failures`, `goout`, `absences`, alcohol use, ...), and (iv) **three grade
periods** `G1`, `G2`, `G3` (first-period, second-period, and final math grade, each 0--20). The paper
predicts `G3` from a small demographic/behavioral subset and deliberately *excludes* the earlier
grades `G1`/`G2` (which would otherwise dominate); we honor that choice throughout.

```{r eda-dictionary}
# Variable dictionary for the outcome + the model's predictors (+ the Day-2 smoothing target).
dict <- data.frame(
  variable = c("G3", "Medu", "goout", "romantic", "age",
               "schoolsup", "absences"),
  role = c("outcome", "predictor", "predictor", "predictor",
           "predictor", "predictor", "Day-2 nonlinear target"),
  type = c("numeric (0-20)", "ordinal 0-4", "ordinal 1-5",
           "binary", "numeric (years)", "binary", "count"),
  description = c(
    "Final math grade",
    "Mother's education level",
    "Frequency of going out with friends",
    "In a romantic relationship",
    "Student age",
    "Extra school educational support",
    "Number of school absences"),
  coding = c("0=fail .. 20=top", "0=none,1=primary,2=5-9th,3=sec,4=higher",
             "1=very low .. 5=very high", "no / yes", "15-22 years",
             "no / yes", "0-93 days"))
knitr::kable(dict, row.names = FALSE,
             caption = "Variable dictionary: outcome, the five published predictors, and the continuous predictor (absences) we bend with polynomials/splines/GAM on Day 2.")
```

We profile these variables in depth below and return to the fuller feature set only where it clarifies
the modeling. We also load the plotting/EDA helpers used in this section.

```{r eda-libs}
library(ggplot2)
library(dplyr)
suppressMessages(library(GGally))
suppressMessages(library(patchwork))
theme_set(theme_minimal(base_size = 10))
```

## The outcome `G3`, in depth

```{r eda-outcome, fig.height=3.0, fig.cap="Final math grade G3 (0--20): histogram with overlaid density (left) and boxplot (right). The tall bar at 0 is the ~10% of students recorded with a zero final grade."}
g_hist <- ggplot(dat, aes(G3)) +
  geom_histogram(aes(y = after_stat(density)), breaks = 0:20,
                 fill = "grey80", colour = "white") +
  geom_density(colour = "firebrick", linewidth = 0.8) +
  geom_vline(xintercept = mean(dat$G3), colour = "firebrick",
             linetype = 2) +
  labs(x = "G3 (final math grade)", y = "density",
       title = "Distribution of G3")
g_box <- ggplot(dat, aes(y = G3)) +
  geom_boxplot(fill = "lightsteelblue", width = 0.4) +
  labs(x = NULL, y = "G3", title = "Boxplot") +
  theme(axis.text.x = element_blank())
g_hist + g_box + plot_layout(widths = c(3, 1))
```

```{r eda-outcome-stats}
# Full descriptive stats for the outcome, incl. skewness & kurtosis (e1071).
q <- quantile(dat$G3, c(.25, .5, .75))
os <- data.frame(
  n = length(dat$G3), mean = mean(dat$G3), sd = sd(dat$G3),
  min = min(dat$G3), Q1 = q[1], median = q[2], Q3 = q[3],
  max = max(dat$G3), IQR = IQR(dat$G3),
  skew = e1071::skewness(dat$G3), kurt = e1071::kurtosis(dat$G3))
knitr::kable(os, row.names = FALSE, digits = 2,
             caption = "Outcome G3: full summary statistics (skewness & kurtosis from e1071).")
```

`G3` is centered near `r round(mean(dat$G3),1)` (sd `r round(sd(dat$G3),1)`) and would look roughly
bell-shaped were it not for a conspicuous **floor spike at 0**: `r sum(dat$G3==0)` students
(`r round(100*mean(dat$G3==0))`%) are recorded with a zero final grade --- students who did not sit,
or failed outright, the final. That spike pulls the mean below the median and gives a mild **left
skew** (skewness `r round(e1071::skewness(dat$G3),2)`); the small negative excess kurtosis says the
non-zero mass is a touch flatter than a normal. This is *floor inflation*, not classic right-skew, so
a log/sqrt transform would not help --- and a plain OLS (the paper's model) simply treats the zeros as
genuine low grades. We keep `G3` untransformed to stay faithful to the reproduction, but flag the
zeros as the main reason the achievable $R^2$ is capped.

## Univariate distributions of the predictors

```{r eda-num-univ, fig.height=3.2, fig.cap="Numeric/ordinal predictors used downstream. Medu and goout are discrete ordinal scales; age is right-skewed (a few older repeaters); absences is heavily right-skewed with a spike at zero."}
num_pred <- c("Medu", "goout", "age", "absences")
long <- dat |>
  select(all_of(num_pred)) |>
  tidyr::pivot_longer(everything(), names_to = "var",
                      values_to = "value")
ggplot(long, aes(value)) +
  geom_histogram(bins = 20, fill = "steelblue", colour = "white") +
  facet_wrap(~ var, scales = "free", ncol = 2) +
  labs(x = NULL, y = "count",
       title = "Univariate distributions of the numeric predictors")
```

```{r eda-num-tab}
# Full summary table for the numeric predictors, including skew.
sumrow <- function(v) {
  x <- dat[[v]]; q <- quantile(x, c(.25, .5, .75))
  data.frame(variable = v, n = sum(!is.na(x)),
             n_miss = sum(is.na(x)), mean = mean(x), sd = sd(x),
             min = min(x), Q1 = q[1], median = q[2], Q3 = q[3],
             max = max(x), skew = e1071::skewness(x))
}
num_tab <- do.call(rbind, lapply(num_pred, sumrow))
knitr::kable(num_tab, row.names = FALSE, digits = 2,
             caption = "Numeric predictors: full summary statistics. Medu/goout/age enter the model; absences is the continuous predictor we smooth on Day 2.")
```

`Medu` and `goout` are short ordinal scales that behave like coarse counts; both are fairly
symmetric. `age` is mildly right-skewed --- most students are 15--18, with a thin tail of older
repeaters up to 22. `absences` is the extreme case: strongly **right-skewed** (skew
`r round(e1071::skewness(dat$absences),1)`), a large mass at zero, and a long thin tail (max
`r max(dat$absences)`). That heavy skew is exactly why the paper's outlier-imputation step touches it
and why, later, a high-degree polynomial in `absences` is so prone to chase the few extreme points ---
a caution we return to when we fit polynomials, splines, and a GAM smooth to it on Day 2.

```{r eda-cat-univ, fig.height=2.6, fig.cap="The two binary predictors. Most students are NOT in a romantic relationship and do NOT receive extra school support; both classes are moderately imbalanced."}
cat_pred <- c("romantic", "schoolsup")
catlong <- dat |>
  select(all_of(cat_pred)) |>
  mutate(across(everything(), as.character)) |>
  tidyr::pivot_longer(everything(), names_to = "var",
                      values_to = "level")
ggplot(catlong, aes(level)) +
  geom_bar(fill = "darkseagreen4") +
  facet_wrap(~ var, scales = "free_x") +
  labs(x = NULL, y = "count",
       title = "Binary predictor frequencies")
```

```{r eda-cat-tab}
cat_tab <- rbind(
  romantic  = table(dat$romantic),
  schoolsup = table(dat$schoolsup))
cat_tab <- cbind(cat_tab,
                 prop_yes = round(cat_tab[, "yes"] / rowSums(cat_tab), 3))
knitr::kable(cat_tab,
             caption = "Binary predictors: frequency counts and proportion 'yes'.")
```

About `r round(100*mean(dat$romantic=="yes"))`% of students report being in a romantic relationship
and only `r round(100*mean(dat$schoolsup=="yes"))`% receive extra educational support. Neither class
is so rare as to be unusable, but `schoolsup` in particular is imbalanced --- worth remembering when
its coefficient turns out to be one of the largest in the model.

## Missing-data analysis

```{r eda-missing}
# Per-variable missing counts/percentages across the WHOLE frame.
miss <- sapply(dat, function(x) sum(is.na(x)))
miss_tab <- data.frame(
  variable = names(miss), n_missing = as.integer(miss),
  pct_missing = round(100 * miss / nrow(dat), 2))
miss_tab <- miss_tab[order(-miss_tab$n_missing), ]
cat("Complete cases:", sum(complete.cases(dat)), "of", nrow(dat), "\n")
cat("Total missing cells:", sum(miss), "\n")
knitr::kable(head(miss_tab, 6), row.names = FALSE,
             caption = "Missingness by variable (top rows). The analysis frame is complete.")
```

The distributed UCI dataset is **fully complete** --- every one of the `r nrow(dat)` rows is a
complete case, so there is no missingness pattern to model and no co-missing variables. The only `NA`s
in this pipeline are **transient and self-inflicted**: the paper's preprocessing sets boxplot outliers
in `Medu`, `goout`, and `age` to `NA` and immediately mean-imputes them inside `impute_outliers()`, so
by the time we describe or model the data there are zero missing cells. Because that imputation is
deterministic and mean-based (not model-based), it cannot leak outcome information --- it just softens
a handful of extreme predictor values.

## Bivariate relationships with the outcome

```{r eda-bivar-num, fig.height=3.0, fig.cap="G3 against the two ordinal model predictors, as grouped boxplots. Higher mother's education tracks higher final grades; more frequent going-out tracks lower grades."}
b1 <- ggplot(dat, aes(factor(round(Medu)), G3)) +
  geom_boxplot(fill = "lightsteelblue") +
  labs(x = "Mother's education (Medu)", y = "G3",
       title = "G3 by Medu")
b2 <- ggplot(dat, aes(factor(round(goout)), G3)) +
  geom_boxplot(fill = "lightsalmon") +
  labs(x = "Going out (goout)", y = "G3", title = "G3 by goout")
b1 + b2
```

```{r eda-bivar-abs, fig.height=2.9, fig.cap="G3 vs. absences with linear (solid) and loess (dashed) smoothers. The association is weak and nearly flat past a few absences --- little curvature for a flexible model to exploit."}
ggplot(dat, aes(absences, G3)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, colour = "steelblue") +
  geom_smooth(method = "loess", se = FALSE, colour = "firebrick",
              linetype = 2) +
  labs(x = "Absences (count)", y = "G3",
       title = "G3 vs. absences")
```

```{r eda-bivar-cat, fig.height=2.9, fig.cap="G3 by the two binary predictors. Students in a romantic relationship and those receiving extra school support score slightly lower on average."}
v1 <- ggplot(dat, aes(romantic, G3)) +
  geom_violin(fill = "plum3", alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", colour = "black") +
  labs(x = "romantic", y = "G3", title = "G3 by romantic")
v2 <- ggplot(dat, aes(schoolsup, G3)) +
  geom_violin(fill = "khaki3", alpha = 0.6) +
  stat_summary(fun = mean, geom = "point", colour = "black") +
  labs(x = "schoolsup", y = "G3", title = "G3 by schoolsup")
v1 + v2
```

```{r eda-bivar-tab}
# One tidy table: each predictor's marginal association with G3.
num_assoc <- lapply(c("Medu", "goout", "age", "absences"), function(v) {
  ct <- cor.test(dat[[v]], dat$G3)
  data.frame(predictor = v, kind = "numeric",
             effect = round(unname(ct$estimate), 3),
             effect_type = "Pearson r",
             p_value = signif(ct$p.value, 2))
})
cat_assoc <- lapply(c("romantic", "schoolsup"), function(v) {
  mm <- tapply(dat$G3, dat[[v]], mean)
  tt <- t.test(G3 ~ dat[[v]], data = dat)
  data.frame(predictor = v, kind = "binary",
             effect = round(unname(mm["yes"] - mm["no"]), 3),
             effect_type = "mean(yes)-mean(no)",
             p_value = signif(tt$p.value, 2))
})
assoc <- do.call(rbind, c(num_assoc, cat_assoc))
knitr::kable(assoc, row.names = FALSE,
             caption = "Marginal association of each predictor with G3 (Pearson r for numeric; yes-minus-no grade gap with a two-sample t-test for binary).")
```

The bivariate picture lines up cleanly with the published coefficient signs: **`Medu` is positively**
associated with `G3` (r `r round(cor(dat$Medu, dat$G3),2)`), while `goout`, `age`, being in a
`romantic` relationship, and needing `schoolsup` all pull grades **down**. Every marginal effect is
small in magnitude --- foreshadowing the modest $R^2$ ahead --- and, importantly, the `G3`--`absences`
relationship is weak and essentially **flat** past a handful of absences, so there is very little
curvature for a flexible model to exploit.

## Multivariate structure and collinearity

```{r eda-corr-heat, fig.height=3.2, fig.cap="Correlation heatmap among G3 and the numeric predictors. All off-diagonal correlations are weak (|r| < 0.25); nothing approaches the |r| > 0.7 collinearity threshold."}
ggcorr(dat[, c("G3", num_pred)], label = TRUE, label_round = 2,
       hjust = 0.85, layout.exp = 1) +
  labs(title = "Correlations among G3 and numeric predictors")
```

```{r eda-corr-tab}
cmat <- cor(dat[, c("G3", num_pred)])
knitr::kable(round(cmat, 3),
             caption = "Pearson correlations among G3 and the numeric predictors.")
```

```{r eda-vif}
# Variance-inflation factors from the reproduction model (base-R, no car).
vif_base <- function(fit) {
  X <- model.matrix(fit)[, -1, drop = FALSE]
  vapply <- sapply(colnames(X), function(nm) {
    others <- setdiff(colnames(X), nm)
    r2 <- summary(lm(X[, nm] ~ X[, others]))$r.squared
    1 / (1 - r2)
  })
  vapply
}
fit_for_vif <- lm(G3 ~ Medu + goout + romantic + age + schoolsup,
                  data = dat)
vt <- data.frame(term = names(vif_base(fit_for_vif)),
                 VIF = round(unname(vif_base(fit_for_vif)), 2))
knitr::kable(vt, row.names = FALSE,
             caption = "Variance-inflation factors for the reproduced OLS. All near 1: no collinearity.")
```

Correlations among the predictors are uniformly weak (all $|r| < 0.25$, well under the $|r| > 0.7$
rule of thumb), and every **VIF is essentially 1** --- there is **no collinearity concern**. Each
predictor contributes largely independent information, which means the OLS coefficients are stable and
individually interpretable (no variance inflation to worry about). The correlations *with* `G3` are
likewise small, so we should expect a **modest but real** predictive signal, no collinearity trouble,
and --- a preview of Day 2 --- little genuine nonlinearity, so the flexible learners (polynomials,
splines, GAM) we fit later should land close to the straight line rather than beating it decisively.

## Reproduced regression vs. published

```{r reproduce}
form <- G3 ~ Medu + goout + romantic + age + schoolsup
fit_ols <- lm(form, data = dat)

published <- c(`(Intercept)` = 17.7935, Medu = 0.8627, goout = -0.5603,
               romanticyes = -1.2458, age = -0.4423, schoolsupyes = -1.6562)
repro <- round(coef(fit_ols)[names(published)], 4)
cmp <- data.frame(term = names(published),
                  published = published,
                  reproduced = repro,
                  abs_diff = abs(published - repro))
knitr::kable(cmp, row.names = FALSE, digits = 4,
             caption = "Reproduction gate: published (El Aissaoui et al. Model 6) vs. reproduced OLS coefficients.")
stopifnot(all(cmp$abs_diff < 1e-3))
cat("GATE PASSED: all six coefficients match to 4 decimals.\n")
```

Every coefficient matches the printed table to four decimal places. The gate is passed; we now keep
this model fixed as the **reproduced baseline** and use its outcome (`G3`) and predictor set to
demonstrate the full Day-2 regression-extensions toolkit.

\newpage

# Day 2: Regression extensions

Everything below uses the **same outcome (`G3`) and the same five predictors** as the reproduced OLS,
so that every extension is anchored to the published model. Day 2 keeps the *regression* frame but
progressively relaxes its two strongest assumptions --- **linearity in the predictors** and
**a Gaussian, continuous response** --- and asks each time whether the extra machinery earns its
keep out of sample. We proceed in four moves: (1) read the OLS as a fitted multiple regression and
interrogate its residuals; (2) swap the Gaussian likelihood for a Bernoulli one (logistic regression
and the MLE that powers it); (3) bend a single continuous predictor with polynomials, splines, and a
GAM smooth; and (4) put linear, polynomial, spline, and GAM side by side under one honest
cross-validation.

```{r day2-libs}
library(splines)     # ns(), bs()  -- natural & B-spline bases
library(mgcv)        # gam()       -- penalized smooths, GCV-selected wiggliness
suppressMessages(library(gridExtra))
rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))
r2   <- function(y, yhat) 1 - sum((y - yhat)^2) / sum((y - mean(y))^2)
```

## OLS as multiple regression: interpretation

The reproduced model is an ordinary least-squares fit,
$G3 = \beta_0 + \beta_1\text{Medu} + \beta_2\text{goout} + \beta_3\text{romantic} + \beta_4\text{age}
+ \beta_5\text{schoolsup} + \varepsilon$, with the $\beta$'s chosen to minimize the residual sum of
squares. Because there is no collinearity here (all VIFs $\approx 1$), each slope is a clean
**partial** effect: the expected change in the final grade for a one-unit change in that predictor,
*holding the other four fixed*.

```{r ols-summary}
sm <- summary(fit_ols)
coef_tab <- data.frame(
  term     = rownames(sm$coefficients),
  estimate = round(sm$coefficients[, "Estimate"], 4),
  std_err  = round(sm$coefficients[, "Std. Error"], 4),
  t_value  = round(sm$coefficients[, "t value"], 3),
  p_value  = signif(sm$coefficients[, "Pr(>|t|)"], 3))
knitr::kable(coef_tab, row.names = FALSE,
             caption = "Reproduced OLS: coefficients, standard errors, and significance.")
cat("Residual SE:", round(sm$sigma, 3),
    "on", sm$df[2], "df | Multiple R^2:", round(sm$r.squared, 4),
    "| Adjusted R^2:", round(sm$adj.r.squared, 4), "\n")
cat("F(", sm$fstatistic[2], ",", sm$fstatistic[3], ") =",
    round(sm$fstatistic[1], 2),
    "  p =", signif(pf(sm$fstatistic[1], sm$fstatistic[2],
                       sm$fstatistic[3], lower.tail = FALSE), 3), "\n")
```

Reading the table: each **additional level of mother's education** raises the predicted final grade
by about `r round(coef(fit_ols)["Medu"], 2)` points; each **extra unit of going out** lowers it by
about `r abs(round(coef(fit_ols)["goout"], 2))`; being in a **romantic relationship** costs about
`r abs(round(coef(fit_ols)["romanticyes"], 2))` points and needing **extra school support** about
`r abs(round(coef(fit_ols)["schoolsupyes"], 2))`; each **year of age** subtracts about
`r abs(round(coef(fit_ols)["age"], 2))`. The overall $F$-test is highly significant (the predictors
jointly matter), yet Multiple $R^2$ is only `r round(sm$r.squared, 3)` --- a genuinely useful but
modest fit, exactly the honest ceiling the floor-spike at zero and the weak marginal correlations
foretold.

## Residual diagnostics

OLS inference rests on four assumptions --- **linearity**, **constant error variance
(homoscedasticity)**, **approximate normality of the errors**, and **no single point dominating the
fit**. The classic four-plot panel checks all four at once. We draw it with base R's own `plot.lm`
so the output is exactly what an analyst sees at the console.

```{r resid-diag, fig.width=7, fig.height=6.4, fig.cap="Standard OLS diagnostic panel. Top-left: residuals vs. fitted (linearity + mean-zero errors). Top-right: normal QQ (normality). Bottom-left: scale-location (homoscedasticity). Bottom-right: residuals vs. leverage with Cook's-distance contours (influential points)."}
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2.5, 1.2))
# 1: resid-vs-fitted, 2: QQ, 3: scale-location, 5: resid-vs-leverage (Cook's contours)
plot(fit_ols, which = c(1, 2, 3, 5))
par(op)
```

```{r resid-diag-stats}
# Numeric companions to the four plots.
res  <- resid(fit_ols)
std  <- rstandard(fit_ols)
hatv <- hatvalues(fit_ols)
cook <- cooks.distance(fit_ols)
p    <- length(coef(fit_ols))
n    <- nrow(dat)
diag_tab <- data.frame(
  quantity = c("Shapiro-Wilk W (normality)",
               "Breusch-Pagan-style corr(|resid|, fitted)",
               "Max leverage (avg = p/n = ...)",
               "Max Cook's distance",
               "# points with Cook's D > 4/n"),
  value = c(round(shapiro.test(res)$statistic, 3),
            round(cor(abs(res), fitted(fit_ols)), 3),
            round(max(hatv), 3),
            round(max(cook), 4),
            sum(cook > 4 / n)))
knitr::kable(diag_tab, row.names = FALSE,
             caption = sprintf("Numeric diagnostics. Average leverage p/n = %.3f; Cook's-distance flag threshold 4/n = %.4f.", p / n, 4 / n))
```

The **residuals-vs-fitted** plot shows the fitted values living in a narrow band (the weak model does
not spread predictions far) with a faint downward smear --- a symptom of the floor at `G3 = 0` rather
than of genuine curvature. The **QQ plot** bends away from the line in the lower tail: those are the
zero-grade students, whose large negative residuals are heavier than a normal would produce (the
Shapiro-Wilk $W$ confirms mild non-normality). **Scale-location** is roughly flat, so the constant-
variance assumption is not badly violated. **Residuals-vs-leverage** shows no point crossing a
Cook's-distance contour of concern (max Cook's D $=$ `r round(max(cook), 3)`, well below the classic
$0.5$/$1$ alarm levels); a handful exceed the gentle $4/n$ rule but none is individually decisive.
The verdict: OLS is *usable* here, but the zero-inflated lower tail is precisely the kind of
departure that motivates the extensions below.

## A logistic-regression + MLE aside

OLS assumes a **continuous, Gaussian** response. A great deal of applied work instead has a
**binary** outcome, and the natural regression there is **logistic regression**. To show the
machinery on data we already understand, we binarize `G3` at the Portuguese pass mark of **10/20**:
`pass = 1` if `G3 >= 10`, else `0`. This is a deliberately *constructed* label for teaching the
method --- we are **not** claiming pass/fail is the "right" target, only using it to expose the
likelihood.

```{r logit-fit}
dat$pass <- as.integer(dat$G3 >= 10)   # 1 = passed (G3 >= 10), 0 = failed
cat("Pass rate:", round(mean(dat$pass), 3),
    "  (", sum(dat$pass), "of", nrow(dat), "students )\n")

fit_logit <- glm(pass ~ Medu + goout + romantic + age + schoolsup,
                 data = dat, family = binomial(link = "logit"))
lo <- summary(fit_logit)$coefficients
logit_tab <- data.frame(
  term      = rownames(lo),
  log_odds  = round(lo[, "Estimate"], 4),
  std_err   = round(lo[, "Std. Error"], 4),
  odds_ratio = round(exp(lo[, "Estimate"]), 4),
  p_value   = signif(lo[, "Pr(>|z|)"], 3))
knitr::kable(logit_tab, row.names = FALSE,
             caption = "Logistic regression for pass (G3 >= 10). Coefficients are on the log-odds scale; exp() gives odds ratios.")
```

**Where the coefficients come from --- maximum likelihood.** OLS has a closed-form least-squares
solution; logistic regression does not. Instead we choose the coefficients that maximize the
**Bernoulli likelihood** of the observed 0/1 labels,
$$
\ell(\beta) = \sum_{i=1}^{n} \Big[\, y_i \log p_i + (1 - y_i)\log(1 - p_i) \,\Big],
\qquad
p_i = \frac{1}{1 + e^{-x_i^\top \beta}},
$$
where $p_i$ is the model's predicted probability that student $i$ passes. `glm` finds the $\hat\beta$
maximizing $\ell$ by iteratively reweighted least squares. The coefficients are **log-odds**: a
one-level rise in `Medu` multiplies the *odds* of passing by
`r round(exp(coef(fit_logit)["Medu"]), 2)` (its odds ratio), holding the rest fixed. We can read the
optimized likelihood and the **deviance** ($-2\ell$, the logistic analogue of the residual sum of
squares) straight off the fit:

```{r logit-deviance}
ll   <- as.numeric(logLik(fit_logit))          # maximized log-likelihood
dev  <- fit_logit$deviance                      # residual deviance = -2*logLik
ndev <- fit_logit$null.deviance                 # deviance of intercept-only model
pseudo_r2 <- 1 - dev / ndev                     # McFadden-style pseudo-R^2
cat("Maximized log-likelihood:", round(ll, 3), "\n")
cat("Residual deviance:", round(dev, 2),
    " (= -2 * logLik =", round(-2 * ll, 2), ")\n")
cat("Null deviance:", round(ndev, 2),
    " | McFadden pseudo-R^2:", round(pseudo_r2, 4), "\n")

# In-sample classification at a 0.5 cutoff, vs. always-predict-majority baseline.
phat <- predict(fit_logit, type = "response")
acc  <- mean((phat >= 0.5) == dat$pass)
base_acc <- max(mean(dat$pass), 1 - mean(dat$pass))
cat("Accuracy @0.5:", round(acc, 3),
    " | majority-class baseline:", round(base_acc, 3), "\n")
```

The **deviance** is literally $-2$ times the maximized log-likelihood; comparing the residual
deviance (`r round(dev, 1)`) to the null deviance (`r round(ndev, 1)`) shows how much the five
predictors improve the likelihood over an intercept-only model, summarized by a McFadden pseudo-$R^2$
of `r round(pseudo_r2, 3)`. The same five weak predictors that gave a modest linear $R^2$ give a
modest classifier: it clears the majority-class baseline only slightly. The lesson is the *method*,
not the score --- changing the **likelihood** (Gaussian $\to$ Bernoulli) is what turns a linear
regression into a logistic one, and MLE is the common engine underneath both.

## Polynomial regression on a continuous predictor

Now we relax **linearity**. The most continuous predictor is `absences` (a count with a long right
tail). Polynomial regression stays inside `lm` --- we simply add `poly(absences, d)` terms on top of
the other four predictors --- but lets the `absences` effect curve. We fit degrees 1--4 and plot the
implied `absences` curve at the mean of the other predictors.

```{r poly-fit, fig.height=3.6, fig.cap="Fitted G3-vs-absences curves for polynomial degrees 1-4 (other predictors held at their means/reference). Higher degrees wiggle to chase the sparse right tail; degree 1 (the straight line) is nearly flat, matching the descriptives."}
# Reference row: hold the non-absences predictors at typical values.
ref <- data.frame(
  Medu = mean(dat$Medu), goout = mean(dat$goout),
  romantic = factor("no", levels = c("no", "yes")),
  age = mean(dat$age),
  schoolsup = factor("no", levels = c("no", "yes")))
grid_abs <- data.frame(absences = seq(min(dat$absences), max(dat$absences),
                                      length.out = 200))
newdat <- cbind(grid_abs, ref[rep(1, nrow(grid_abs)), ])

poly_fit <- function(d)
  lm(G3 ~ poly(absences, d) + Medu + goout + romantic + age + schoolsup,
     data = dat)
degs <- 1:4
curves <- do.call(rbind, lapply(degs, function(d) {
  f <- poly_fit(d)
  data.frame(absences = grid_abs$absences,
             fit = predict(f, newdat), degree = factor(d))
}))

ggplot() +
  geom_point(data = dat, aes(absences, G3), alpha = 0.18) +
  geom_line(data = curves, aes(absences, fit, colour = degree),
            linewidth = 0.9) +
  labs(x = "Absences (count)", y = "Predicted G3",
       colour = "Poly degree",
       title = "Polynomial fits of G3 on absences") +
  theme(legend.position = "right")
```

```{r poly-anova}
# Nested F-tests: does each extra polynomial degree buy a significant in-sample gain?
fits <- lapply(degs, poly_fit)
an <- anova(fits[[1]], fits[[2]], fits[[3]], fits[[4]])
poly_gof <- data.frame(
  degree = degs,
  df = sapply(fits, function(f) f$rank),
  RSS = round(sapply(fits, function(f) sum(resid(f)^2)), 1),
  R2  = round(sapply(fits, function(f) summary(f)$r.squared), 4),
  AIC = round(sapply(fits, AIC), 1))
knitr::kable(poly_gof, row.names = FALSE,
             caption = "Polynomial-in-absences fits: in-sample RSS/R2/AIC by degree. AIC rewards fit but penalizes added terms.")
```

The straight line (degree 1) is nearly flat --- exactly what the weak `G3`--`absences` correlation
predicted. Higher degrees add visible wiggle in the sparse right tail, but AIC (which penalizes the
extra terms) barely moves, hinting that the curvature is fitting noise rather than signal. We will
confirm that suspicion out of sample below.

## Splines: a more stable basis for curvature

A high-degree global polynomial is notoriously unstable at the edges (it swings wildly where data
are thin). **Regression splines** fix this by fitting **low-degree piecewise polynomials joined
smoothly at knots**, so flexibility is *local*. We use `splines::ns()` (a **natural cubic spline**,
linear beyond the boundary knots for tame tails) and `splines::bs()` (an unconstrained cubic
**B-spline**), each with a few interior knots, and overlay their `absences` curves.

```{r spline-fit, fig.height=3.6, fig.cap="Natural cubic spline (ns, solid) and B-spline (bs, dashed) fits of G3 on absences, with the OLS straight line for reference. Splines bend locally around the dense low-absence region and stay tame in the sparse tail (especially ns, which is linear beyond the boundary knots)."}
fit_ns <- lm(G3 ~ ns(absences, df = 4) + Medu + goout + romantic +
               age + schoolsup, data = dat)
fit_bs <- lm(G3 ~ bs(absences, df = 5) + Medu + goout + romantic +
               age + schoolsup, data = dat)

spline_curves <- rbind(
  data.frame(absences = grid_abs$absences,
             fit = predict(fit_ns, newdat), model = "natural spline (ns)"),
  data.frame(absences = grid_abs$absences,
             fit = predict(fit_bs, newdat), model = "B-spline (bs)"),
  data.frame(absences = grid_abs$absences,
             fit = predict(poly_fit(1), newdat), model = "OLS (linear)"))

ggplot() +
  geom_point(data = dat, aes(absences, G3), alpha = 0.18) +
  geom_line(data = spline_curves,
            aes(absences, fit, colour = model, linetype = model),
            linewidth = 0.9) +
  scale_linetype_manual(values = c("natural spline (ns)" = 1,
                                   "B-spline (bs)" = 2,
                                   "OLS (linear)" = 1)) +
  labs(x = "Absences (count)", y = "Predicted G3", colour = NULL,
       linetype = NULL, title = "Spline vs. linear fits of G3 on absences") +
  theme(legend.position = "right")
```

Both splines trace a mild hump: predicted grade edges up over the first several absences and then
settles, staying far tamer in the sparse tail than the degree-4 polynomial did. The natural spline in
particular is *linear beyond the boundary knots* by construction, which is why its right tail does not
fly off. Whether this local curvature is real signal --- or just a flexible basis fitting the noisy
low-absence cloud --- is, again, a question only held-out error can settle.

## A GAM smooth (mgcv)

A **generalized additive model** goes one step further: instead of us choosing a degree or knot
count, `mgcv::gam` fits a **penalized smooth** `s(absences)` and selects its wiggliness
automatically by generalized cross-validation (GCV), trading fit against a roughness penalty. The
other four predictors stay linear, so the model is
$G3 = f(\text{absences}) + \beta\text{-terms} + \varepsilon$.

```{r gam-fit, fig.height=3.4, fig.cap="mgcv::gam penalized smooth of G3 on absences (other predictors linear). The GCV-selected effective degrees of freedom quantify how wiggly the data justify; a near-linear fit means edf close to 1."}
fit_gam <- gam(G3 ~ s(absences) + Medu + goout + romantic + age + schoolsup,
               data = dat, method = "GCV.Cp")
edf <- summary(fit_gam)$s.table[, "edf"]
cat("Smooth term s(absences): effective df =", round(edf, 3),
    " (1 = a straight line; larger = wigglier)\n")

gam_curve <- data.frame(absences = grid_abs$absences,
                        fit = predict(fit_gam, newdat))
ggplot() +
  geom_point(data = dat, aes(absences, G3), alpha = 0.18) +
  geom_line(data = gam_curve, aes(absences, fit),
            colour = "darkorange3", linewidth = 1) +
  labs(x = "Absences (count)", y = "Predicted G3",
       title = sprintf("GAM smooth of G3 on absences (edf = %.2f)", edf))
```

The GAM's **effective degrees of freedom** for the smooth is `r round(edf, 2)`. If it sits close to
1, the penalized fit has effectively decided a straight line is all the data support --- the model
*chose* simplicity, rather than us imposing it. That is the appeal of a GAM: it lets the data set the
flexibility, and here it declines to bend `absences` much, echoing every warning from the
descriptives.

## Out-of-sample comparison: linear vs. polynomial vs. spline vs. GAM

The four flexible fits all look plausible in-sample, but added flexibility always lowers *training*
error --- the honest test is **held-out** error. We score linear, polynomial (degree 3), natural
spline, and GAM under one **10-fold cross-validation**, refitting each model inside every fold.

```{r cv-compare, fig.cap="10-fold CV RMSE (lower is better) for the four regression extensions on G3. Bars near-identical: with these weak, near-linear predictors, the flexible models do not beat the straight line out of sample."}
K <- 10
set.seed(20240501)
folds <- sample(rep(1:K, length.out = nrow(dat)))

# Each learner: a function taking (train, test) -> predicted G3 on test.
learners <- list(
  `OLS (linear)` = function(tr, te) {
    f <- lm(G3 ~ absences + Medu + goout + romantic + age + schoolsup, tr)
    predict(f, te)
  },
  `Polynomial (d=3)` = function(tr, te) {
    f <- lm(G3 ~ poly(absences, 3) + Medu + goout + romantic + age +
              schoolsup, tr)
    predict(f, te)
  },
  `Natural spline (ns df=4)` = function(tr, te) {
    f <- lm(G3 ~ ns(absences, df = 4) + Medu + goout + romantic + age +
              schoolsup, tr)
    predict(f, te)
  },
  `GAM (mgcv s())` = function(tr, te) {
    f <- gam(G3 ~ s(absences) + Medu + goout + romantic + age + schoolsup,
             data = tr, method = "GCV.Cp")
    as.numeric(predict(f, te))
  })

cv_one <- function(fn) {
  rm <- numeric(K)
  for (k in 1:K) {
    tr <- dat[folds != k, ]; te <- dat[folds == k, ]
    rm[k] <- rmse(te$G3, fn(tr, te))
  }
  mean(rm)
}
cv_rmse <- sapply(learners, cv_one)

cv_tab <- data.frame(model = names(cv_rmse),
                     cv_rmse = round(unname(cv_rmse), 4))
cv_tab <- cv_tab[order(cv_tab$cv_rmse), ]
knitr::kable(cv_tab, row.names = FALSE,
             caption = "10-fold CV RMSE for the four regression extensions (lower is better).")

ggplot(cv_tab, aes(reorder(model, cv_rmse), cv_rmse)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_text(aes(label = round(cv_rmse, 3)), vjust = -0.4, size = 3) +
  coord_cartesian(ylim = c(min(cv_tab$cv_rmse) - 0.05,
                           max(cv_tab$cv_rmse) + 0.05)) +
  labs(x = NULL, y = "10-fold CV RMSE",
       title = "Held-out RMSE: linear vs. polynomial vs. spline vs. GAM") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

The four held-out RMSEs are **within a hair of one another**, and the plain linear fit is right in
the pack (typically the best or tied-best). This is the central Day-2 lesson made concrete: the
extensions --- polynomials, splines, GAMs --- are the right tools *when the response genuinely curves*,
but here the descriptives told us `absences` relates to `G3` almost linearly, the GAM's own
GCV chose a near-straight smooth, and so the flexible models spend their extra degrees of freedom
fitting noise that does not generalize. **Reach for a spline or GAM when a residual plot or a domain
argument says the relationship bends --- not by default.**

# Takeaway

We exactly reproduced El Aissaoui et al.'s (2020) best OLS for predicting secondary-school final
Mathematics grades (all six coefficients match to four decimals: intercept 17.7935, Medu 0.8627,
goout -0.5603, romanticyes -1.2458, age -0.4423, schoolsupyes -1.6562). Extending it into a Day-2
regression-extensions lab, we (i) read the fit as a multiple regression and ran the full residual-
diagnostic panel (residuals-vs-fitted, QQ, scale-location, leverage/Cook's), which exposed a
zero-inflated lower tail; (ii) swapped the Gaussian likelihood for a Bernoulli one --- logistic
regression on a pass/fail label --- and used it to explain the MLE, log-odds/odds-ratios, and
deviance; (iii) bent `absences` with polynomials, natural/B-splines, and an mgcv GAM smooth, plotting
each fitted curve; and (iv) compared all four under one 10-fold cross-validation. The honest verdict:
these five demographic/behavioral predictors relate to the grade almost linearly, so the flexible
extensions match --- but do not beat --- the simple published regression out of sample. The tools
matter; knowing *when* the data warrant them, from diagnostics and held-out error, matters more.

# Independent exercises (each about 10 minutes)

1. **Bend a second predictor.** Add a fifth entry to the `learners` list that
   wraps `goout` in `ns(goout, df = 4)` (keep `absences` linear, as in the OLS
   learner). Score it with `cv_one()` and compare with the four values in
   `cv_tab`. Does it beat the plain linear model?

2. **The price of extra flexibility.** In the natural-spline learner, change
   `ns(absences, df = 4)` to `df = 8` and re-run `cv_one()`. Does doubling the
   basis size improve or worsen held-out RMSE?

3. **Odds ratios.** Add `goout` to the pass/fail logistic regression and
   report `exp(coef(...))` for it, with a one-sentence interpretation of the
   odds ratio.
